---
title: "UM Atlas - Cancer Metaprogram Analysis (Melanocytic Cells: Healthy + Malignant)"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Figures 3, 4 (cancer metaprogram scoring with UCell on GeneNMF modules; EMT/Interferon analyses)
```

#  {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

This report analyzes **67 cancer metaprograms** (Weizmann 3CA) in **melanocytic cells** from the UM atlas, including both **healthy melanocytes** and **malignant melanoma cells** to enable direct comparison across disease progression.

**Dataset:**
- ~194,476 total melanocytic cells
  - 6,528 healthy melanocytes
  - 187,946 malignant melanoma cells (130,551 primary + 57,395 metastatic)
- 61 patients
- 5 organs (eye, liver, brain, skin, thyroid)

**Metaprogram categories:**
- Cell cycle (MP1-5)
- Stress & hypoxia (MP6-9)
- Protein processing (MP10-14)
- EMT (MP15-20)
- Interferon/MHC (MP22-24)
- Tissue-specific programs (MP37-50)
- And more...

**Comparisons:**
- **Origin (healthy melanocytes → primary UM → metastatic UM)** ← KEY ANALYSIS
- Primary mutation (GNAQ, GNA11, PLCB4, none)
- Secondary mutation (BAP1, SF3B1, EIF1AX, none)
- Organ (eye, liver, brain, skin, thyroid)
- Patient-level heterogeneity

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)

# Load libraries
library(Seurat)
library(qs)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(cowplot)
library(UCell)
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(ggridges)
library(ggpubr)  # For statistical comparisons on violin plots

# Load Victoria's official color palettes
source("colors.R")

# Set output directory
output_dir <- "Outputs/Metaprograms/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Set parallel processing
library(future)
plan("multicore", workers = 16)
options(future.globals.maxSize = 8000 * 1024^2)

# Increased base font size for better readability (per collaborator request)
theme_set(theme_minimal(base_size = 14))
```

## Load Data {.tabset}

### Load UM Atlas

```{r load_um_atlas}
# Load UM atlas (updated dataset from Victoria, Jan 2025)
um_atlas <- qread("../13_final_Dataset_reclustered_28012025.qs", nthreads = 16)

# Fix GANQ → GNAQ typo
um_atlas$primary_mutation[um_atlas$primary_mutation == "GANQ"] <- "GNAQ"

cat("UM Atlas loaded:\n")
cat("  - Cells:", ncol(um_atlas), "\n")
cat("  - Features:", nrow(um_atlas), "\n\n")

# Filter to melanocytic cells: healthy melanocytes + malignant melanoma cells
malignant <- subset(um_atlas, subset = majority_celltype %in% c("Melanocytes", "Melanoma cells"))

cat("Melanocytic cells (Healthy Melanocytes + Malignant Melanoma):\n")
cat("  - Total cells:", ncol(malignant), "\n")
cat("  - Percentage of atlas:", round(100 * ncol(malignant) / ncol(um_atlas), 2), "%\n\n")

# Display distributions by cell type
cat("Cell type distribution:\n")
print(table(malignant$majority_celltype))

# Display distributions by origin
cat("\n\nOrigin distribution (KEY for healthy → primary → metastatic comparison):\n")
origin_table <- table(malignant$origin, malignant$majority_celltype)
print(origin_table)
cat("\nTotal by origin:\n")
print(rowSums(origin_table))

cat("\n\nPrimary mutation distribution:\n")
print(table(malignant$primary_mutation))

cat("\n\nSecondary mutation distribution:\n")
print(table(malignant$secondary_mutation))

cat("\n\nOrgan distribution:\n")
print(table(malignant$organ))

cat("\n\nPrimary tumor location (pUM) distribution:\n")
print(table(malignant$LocationPrimaryTumor))

cat("\n\nSamples (biological replicates merged):", length(unique(malignant$orig.ident)), "\n")
```

### Load Metaprograms

```{r load_metaprograms}
# Load Weizmann 3CA metaprograms
excel_path <- "../Weizman_3CA_meta_programs_2025-01-29.xlsx"

# Read first sheet (Malignant programs)
mp_df <- read_excel(excel_path, sheet = "Malignant")

cat("Metaprograms loaded:\n")
cat("  - Programs:", ncol(mp_df), "\n")
cat("  - Genes per program:", nrow(mp_df), "\n\n")

# Convert to list format for UCell
metaprograms <- list()
for (col in colnames(mp_df)) {
  genes <- mp_df[[col]]
  genes <- genes[!is.na(genes)]
  metaprograms[[col]] <- genes
}

cat("Metaprogram categories:\n")
cat("  - Cell Cycle (MP1-5):", sum(grepl("^MP[1-5] ", names(metaprograms))), "\n")
cat("  - Stress (MP6-9):", sum(grepl("^MP[6-9] ", names(metaprograms))), "\n")
cat("  - Protein Processing (MP10-14):", sum(grepl("^MP1[0-4] ", names(metaprograms))), "\n")
cat("  - EMT (MP15-20):", sum(grepl("^MP1[5-9] |^MP20 ", names(metaprograms))), "\n")
cat("  - Interferon/MHC (MP22-24):", sum(grepl("^MP2[2-4] ", names(metaprograms))), "\n")
cat("  - Total:", length(metaprograms), "\n\n")

cat("Example programs:\n")
print(names(metaprograms)[1:10])
```

## Compute UCell Scores

```{r compute_ucell}
# Compute UCell scores for all metaprograms using 16 cores
cat("Computing UCell scores for", length(metaprograms), "metaprograms...\n")
cat("Using 16 cores for parallel processing\n\n")

start_time <- Sys.time()

malignant <- AddModuleScore_UCell(
  malignant,
  features = metaprograms,
  ncores = 16,
  name = NULL  # Use gene set names directly
)

end_time <- Sys.time()
cat("UCell computation completed in", round(difftime(end_time, start_time, units = "mins"), 2), "minutes\n\n")

# Get UCell score column names
ucell_cols <- grep("^MP[0-9]+", colnames(malignant@meta.data), value = TRUE)
cat("UCell scores added:", length(ucell_cols), "columns\n\n")

cat("Sample UCell scores:\n")
print(summary(malignant@meta.data[, ucell_cols[1:5]]))

# Save melanocytic cells (healthy + malignant) with UCell scores
cat("\nSaving melanocytic cells (healthy + malignant) with UCell scores...\n")
qsave(malignant,
      file.path("..", "melanocytic_metaprograms.qs2"),
      nthreads = 16)
cat("Saved to: melanocytic_metaprograms.qs2\n")
```

## Metaprogram Overview {.tabset}

### Top Metaprograms

```{r top_metaprograms, fig.width = 14, fig.height = 10}
# Calculate mean scores for each metaprogram
mp_means <- sapply(ucell_cols, function(col) mean(malignant@meta.data[[col]], na.rm = TRUE))
mp_means_df <- data.frame(
  Metaprogram = names(mp_means),
  Mean_Score = mp_means
) |>
  arrange(desc(Mean_Score))

# Top 20 metaprograms by mean score
top20 <- head(mp_means_df, 20)

p <- ggplot(top20, aes(x = reorder(Metaprogram, Mean_Score), y = Mean_Score)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  theme_cowplot() +
  labs(title = "Top 20 Metaprograms by Mean UCell Score (All Melanocytic Cells)",
       subtitle = "Includes healthy melanocytes + primary UM + metastatic UM",
       x = "", y = "Mean UCell Score") +
  theme(axis.text.y = element_text(size = 10))

print(p)
ggsave(file.path(output_dir, "top20_metaprograms.png"), p, width = 14, height = 10, dpi = 300, bg = "white")

# Export table
write.csv(mp_means_df, file.path(output_dir, "metaprogram_mean_scores.csv"), row.names = FALSE)
```

### Metaprogram Correlation

```{r mp_correlation, fig.width = 16, fig.height = 15}
# Calculate correlation matrix of top 30 metaprograms
top30_mps <- head(mp_means_df$Metaprogram, 30)
cor_mat <- cor(malignant@meta.data[, top30_mps], use = "complete.obs")

# Shorten names for display
rownames(cor_mat) <- gsub("^(MP[0-9]+) .*", "\\1", rownames(cor_mat))
colnames(cor_mat) <- gsub("^(MP[0-9]+) .*", "\\1", colnames(cor_mat))

# Create heatmap
col_fun <- colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

ht <- Heatmap(
  cor_mat,
  name = "Pearson\nCorrelation",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  column_title = "Top 30 Metaprogram Correlations (All Melanocytic Cells)",
  heatmap_legend_param = list(direction = "vertical")
)

png(file.path(output_dir, "metaprogram_correlation.png"), width = 16, height = 15, units = "in", res = 300)
draw(ht)
dev.off()

# Display in report
draw(ht)
```

## UMAP Visualizations {.tabset}

### Cell Cycle Programs

```{r umap_cell_cycle, fig.width = 18, fig.height = 12}
cell_cycle_mps <- grep("^MP[1-5] ", ucell_cols, value = TRUE)

plots <- lapply(cell_cycle_mps, function(mp) {
  FeaturePlot(malignant, features = mp, reduction = "umap", raster = FALSE) +
    coord_fixed() +
    scale_color_viridis_c(option = "magma") +
    ggtitle(gsub("^(MP[0-9]+) .*", "\\1", mp))
})

combined <- wrap_plots(plots, ncol = 3)
print(combined)
ggsave(file.path(output_dir, "umap_cell_cycle.png"), combined, width = 18, height = 12, dpi = 300, bg = "white")
```

### EMT Programs

```{r umap_emt, fig.width = 18, fig.height = 12}
emt_mps <- grep("^MP1[5-9] |^MP20 ", ucell_cols, value = TRUE)

plots <- lapply(emt_mps, function(mp) {
  FeaturePlot(malignant, features = mp, reduction = "umap", raster = FALSE) +
    coord_fixed() +
    scale_color_viridis_c(option = "magma") +
    ggtitle(gsub("^(MP[0-9]+) .*", "\\1", mp))
})

combined <- wrap_plots(plots, ncol = 3)
print(combined)
ggsave(file.path(output_dir, "umap_emt.png"), combined, width = 18, height = 12, dpi = 300, bg = "white")
```

### Stress & Hypoxia Programs

```{r umap_stress, fig.width = 18, fig.height = 8}
stress_mps <- grep("^MP[6-9] ", ucell_cols, value = TRUE)

plots <- lapply(stress_mps, function(mp) {
  FeaturePlot(malignant, features = mp, reduction = "umap", raster = FALSE) +
    coord_fixed() +
    scale_color_viridis_c(option = "magma") +
    ggtitle(gsub("^(MP[0-9]+) .*", "\\1", mp))
})

combined <- wrap_plots(plots, ncol = 4)
print(combined)
ggsave(file.path(output_dir, "umap_stress.png"), combined, width = 18, height = 8, dpi = 300, bg = "white")
```

### Interferon/MHC Programs

```{r umap_interferon, fig.width = 18, fig.height = 6}
ifn_mps <- grep("^MP2[2-4] ", ucell_cols, value = TRUE)

plots <- lapply(ifn_mps, function(mp) {
  FeaturePlot(malignant, features = mp, reduction = "umap", raster = FALSE) +
    coord_fixed() +
    scale_color_viridis_c(option = "magma") +
    ggtitle(gsub("^(MP[0-9]+) .*", "\\1", mp))
})

combined <- wrap_plots(plots, ncol = 3)
print(combined)
ggsave(file.path(output_dir, "umap_interferon.png"), combined, width = 18, height = 6, dpi = 300, bg = "white")
```

### Top 30 Metaprograms - Comprehensive View

```{r umap_top30_comprehensive, fig.width = 24, fig.height = 40}
# Get top 30 metaprograms
top30_mps <- head(mp_means_df$Metaprogram, 30)

plots <- lapply(top30_mps, function(mp) {
  FeaturePlot(malignant, features = mp, reduction = "umap", raster = FALSE) +
    coord_fixed() +
    scale_color_viridis_c(option = "magma") +
    ggtitle(gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp)) +
    theme(plot.title = element_text(size = 10))
})

combined <- wrap_plots(plots, ncol = 5)
print(combined)
ggsave(file.path(output_dir, "umap_top30_comprehensive.png"), combined, width = 24, height = 40, dpi = 300, bg = "white")
```

## Comparisons by Primary Mutation {.tabset}

### Overview - Primary Mutation

```{r primary_mutation_overview}
# Filter to known mutations
mal_known_prim <- subset(malignant, subset = primary_mutation != "unknown")

cat("Cells by primary mutation (excluding unknown):\n")
print(table(mal_known_prim$primary_mutation))
```

### Violin Plots - Top Metaprograms

```{r violin_primary_top, fig.width = 16, fig.height = 20}
top10_mps <- head(mp_means_df$Metaprogram, 10)

plots <- lapply(top10_mps, function(mp) {
  VlnPlot(mal_known_prim, features = mp, group.by = "primary_mutation", pt.size = 0) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black") +
    ggtitle(gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp)) +
    theme(legend.position = "none", axis.title.x = element_blank())
})

combined <- wrap_plots(plots, ncol = 2)
print(combined)
ggsave(file.path(output_dir, "violin_primary_mutation_top10.png"), combined, width = 16, height = 20, dpi = 300, bg = "white")
```

### Ridge Plots - Cell Cycle

```{r ridge_primary_cellcycle, fig.width = 14, fig.height = 12}
cell_cycle_mps <- grep("^MP[1-5] ", ucell_cols, value = TRUE)

plots <- lapply(cell_cycle_mps, function(mp) {
  ggplot(mal_known_prim@meta.data, aes(x = .data[[mp]], y = primary_mutation, fill = primary_mutation)) +
    geom_density_ridges(alpha = 0.7) +
    theme_ridges() +
    labs(title = gsub("^(MP[0-9]+) .*", "\\1", mp), x = "UCell Score", y = "") +
    theme(legend.position = "none")
})

combined <- wrap_plots(plots, ncol = 3)
print(combined)
ggsave(file.path(output_dir, "ridge_primary_cellcycle.png"), combined, width = 14, height = 12, dpi = 300, bg = "white")
```

### DotPlot - EMT Programs

```{r dotplot_primary_emt, fig.width = 10, fig.height = 8}
emt_mps <- grep("^MP1[5-9] |^MP20 ", ucell_cols, value = TRUE)

p <- DotPlot(mal_known_prim, features = emt_mps, group.by = "primary_mutation") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("EMT Metaprograms by Primary Mutation")

print(p)
ggsave(file.path(output_dir, "dotplot_primary_emt.png"), p, width = 10, height = 8, dpi = 300, bg = "white")
```

## Comparisons by Secondary Mutation {.tabset}

### Overview - Secondary Mutation

```{r secondary_mutation_overview}
# Filter to known mutations
mal_known_sec <- subset(malignant, subset = secondary_mutation != "unknown")

cat("Cells by secondary mutation (excluding unknown):\n")
print(table(mal_known_sec$secondary_mutation))
```

### Violin Plots - Top Metaprograms

```{r violin_secondary_top, fig.width = 16, fig.height = 20}
top10_mps <- head(mp_means_df$Metaprogram, 10)

plots <- lapply(top10_mps, function(mp) {
  VlnPlot(mal_known_sec, features = mp, group.by = "secondary_mutation", pt.size = 0) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black") +
    ggtitle(gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp)) +
    theme(legend.position = "none", axis.title.x = element_blank())
})

combined <- wrap_plots(plots, ncol = 2)
print(combined)
ggsave(file.path(output_dir, "violin_secondary_mutation_top10.png"), combined, width = 16, height = 20, dpi = 300, bg = "white")
```

### Ridge Plots - Stress Programs

```{r ridge_secondary_stress, fig.width = 14, fig.height = 8}
stress_mps <- grep("^MP[6-9] ", ucell_cols, value = TRUE)

plots <- lapply(stress_mps, function(mp) {
  ggplot(mal_known_sec@meta.data, aes(x = .data[[mp]], y = secondary_mutation, fill = secondary_mutation)) +
    geom_density_ridges(alpha = 0.7) +
    theme_ridges() +
    labs(title = gsub("^(MP[0-9]+) .*", "\\1", mp), x = "UCell Score", y = "") +
    theme(legend.position = "none")
})

combined <- wrap_plots(plots, ncol = 2)
print(combined)
ggsave(file.path(output_dir, "ridge_secondary_stress.png"), combined, width = 14, height = 8, dpi = 300, bg = "white")
```

### Heatmap - Top 20 by Secondary Mutation

```{r heatmap_secondary_top20, fig.width = 12, fig.height = 10}
top20_mps <- head(mp_means_df$Metaprogram, 20)

# Calculate mean scores by secondary mutation
mean_scores <- sapply(top20_mps, function(mp) {
  tapply(mal_known_sec@meta.data[[mp]], mal_known_sec$secondary_mutation, mean, na.rm = TRUE)
})

# Shorten MP names
colnames(mean_scores) <- gsub("^(MP[0-9]+) .*", "\\1", colnames(mean_scores))

# Create heatmap
col_fun <- colorRamp2(
  quantile(mean_scores, c(0, 0.5, 1)),
  c("blue", "white", "red")
)

ht <- Heatmap(
  mean_scores,
  name = "Mean\nUCell Score",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 9),
  column_title = "Top 20 Metaprograms by Secondary Mutation",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", mean_scores[i, j]), x, y, gp = gpar(fontsize = 7))
  }
)

png(file.path(output_dir, "heatmap_secondary_top20.png"), width = 12, height = 10, units = "in", res = 300)
draw(ht)
dev.off()

draw(ht)
```

## ★ KEY ANALYSIS: Disease Progression (Healthy → Primary → Metastatic) {.tabset}

### Overview - Disease Progression

```{r origin_overview}
cat("Cells by disease stage (origin):\n")
origin_counts <- table(malignant$origin, malignant$majority_celltype)
print(origin_counts)
cat("\nTotal cells per stage:\n")
print(rowSums(origin_counts))
cat("\n\nThis is the KEY comparison requested: Healthy melanocytes → Primary UM → Metastatic UM\n")
```

### Violin Plots - Top Metaprograms

```{r violin_origin_top, fig.width = 16, fig.height = 20}
top10_mps <- head(mp_means_df$Metaprogram, 10)

plots <- lapply(top10_mps, function(mp) {
  VlnPlot(malignant, features = mp, group.by = "origin", pt.size = 0) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black") +
    ggtitle(gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp)) +
    scale_fill_manual(values = origin_colors) +  # Victoria's official colors
    labs(subtitle = "Healthy melanocytes vs Primary UM vs Metastatic UM") +
    theme(legend.position = "none", axis.title.x = element_blank(),
          text = element_text(size = 14))
})

combined <- wrap_plots(plots, ncol = 2)
print(combined)
ggsave(file.path(output_dir, "violin_origin_top10.png"), combined, width = 16, height = 20, dpi = 300, bg = "white")
```

### Statistical Analysis - MP6, MP9, MP17, MP45 (Collaborator Request)

**Specific comparison requested:** Healthy vs Primary vs Metastasis for stress/hypoxia (MP6, MP9), EMT-III (MP17), and tissue-specific (MP45) metaprograms.

```{r violin_mp6_mp9_mp17_mp45_stats, fig.width = 16, fig.height = 12}
# Find the specific metaprogram columns
stress_mps <- grep("^MP[6|9] ", ucell_cols, value = TRUE)
mp17_col <- grep("^MP17 ", ucell_cols, value = TRUE)
mp45_col <- grep("^MP45 ", ucell_cols, value = TRUE)
selected_mps <- c(stress_mps, mp17_col, mp45_col)

cat("Selected metaprograms for statistical analysis:\n")
print(selected_mps)

# Define pairwise comparisons
comparisons <- list(
  c("healthy", "primary"),
  c("primary", "metastasis"),
  c("healthy", "metastasis")
)

# Create violin plots with statistics for each metaprogram
plots_with_stats <- lapply(selected_mps, function(mp) {
  # Prepare data
  plot_data <- malignant@meta.data %>%
    dplyr::select(origin, score = all_of(mp)) %>%
    filter(!is.na(origin), !is.na(score))

  # Set factor levels for proper ordering
  plot_data$origin <- factor(plot_data$origin, levels = c("healthy", "primary", "metastasis"))

  # Create violin plot with Wilcoxon tests
  p <- ggviolin(plot_data, x = "origin", y = "score", fill = "origin",
                palette = origin_colors,  # Victoria's colors
                add = "boxplot", add.params = list(fill = "white", width = 0.1)) +
    stat_compare_means(comparisons = comparisons, method = "wilcox.test",
                       label = "p.signif", hide.ns = FALSE,
                       tip.length = 0.02, bracket.size = 0.5) +
    stat_compare_means(method = "kruskal.test", label.y.npc = 0.95,
                       label = "p.format") +
    scale_x_discrete(limits = c("healthy", "primary", "metastasis")) +
    labs(title = gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp),
         subtitle = "Wilcoxon pairwise tests + Kruskal-Wallis overall",
         x = "Origin", y = "UCell Score") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", size = 14),
          axis.title = element_text(size = 12),
          axis.text = element_text(size = 11))

  return(p)
})

# Combine plots (2 columns for 4 metaprograms)
combined_stats <- wrap_plots(plots_with_stats, ncol = 2)
print(combined_stats)

ggsave(file.path(output_dir, "violin_mp6_mp9_mp17_mp45_with_statistics.png"),
       combined_stats, width = 16, height = 14, dpi = 300, bg = "white")

# Print detailed statistics
cat("\n\n=== Detailed Statistical Results ===\n\n")
for (mp in selected_mps) {
  cat(paste0("\n--- ", mp, " ---\n"))

  plot_data <- malignant@meta.data %>%
    dplyr::select(origin, score = all_of(mp)) %>%
    filter(!is.na(origin), !is.na(score))

  # Kruskal-Wallis overall test
  kw <- kruskal.test(score ~ origin, data = plot_data)
  cat("Kruskal-Wallis test (overall):\n")
  cat("  Chi-squared:", round(kw$statistic, 4), "\n")
  cat("  p-value:", format.pval(kw$p.value, digits = 3), "\n\n")

  # Pairwise Wilcoxon tests
  pw <- pairwise.wilcox.test(plot_data$score, plot_data$origin,
                              p.adjust.method = "BH")
  cat("Pairwise Wilcoxon tests (BH-adjusted):\n")
  print(pw$p.value)
  cat("\n")

  # Summary statistics by group
  cat("Group means:\n")
  group_stats <- plot_data %>%
    group_by(origin) %>%
    summarise(
      n = n(),
      mean = round(mean(score), 4),
      median = round(median(score), 4),
      sd = round(sd(score), 4)
    )
  print(as.data.frame(group_stats))
}
```

### Ridge Plots - EMT Programs

```{r ridge_origin_emt, fig.width = 14, fig.height = 12}
emt_mps <- grep("^MP1[5-9] |^MP20 ", ucell_cols, value = TRUE)

plots <- lapply(emt_mps, function(mp) {
  ggplot(malignant@meta.data, aes(x = .data[[mp]], y = origin, fill = origin)) +
    geom_density_ridges(alpha = 0.7) +
    scale_fill_manual(values = origin_colors) +
    theme_ridges() +
    labs(title = gsub("^(MP[0-9]+) .*", "\\1", mp), x = "UCell Score", y = "") +
    theme(legend.position = "none")
})

combined <- wrap_plots(plots, ncol = 3)
print(combined)
ggsave(file.path(output_dir, "ridge_origin_emt.png"), combined, width = 14, height = 12, dpi = 300, bg = "white")
```

### DotPlot - Interferon Programs

```{r dotplot_origin_interferon, fig.width = 8, fig.height = 6}
ifn_mps <- grep("^MP2[2-4] ", ucell_cols, value = TRUE)

# Set origin factor levels for proper ordering: healthy → primary → metastasis
malignant$origin <- factor(malignant$origin, levels = c("healthy", "primary", "metastasis"))

p <- DotPlot(malignant, features = ifn_mps, group.by = "origin") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Interferon/MHC Metaprograms by Origin")

print(p)
ggsave(file.path(output_dir, "dotplot_origin_interferon.png"), p, width = 8, height = 6, dpi = 300, bg = "white")
```

### DotPlot - Interferon Programs by Treatment (Metastases)

```{r dotplot_ifn_treatment_metastasis, fig.width = 10, fig.height = 6}
# Focus on metastatic samples and split by treatment
met_cells <- subset(malignant, origin == "metastasis")

cat("Metastatic cells:", ncol(met_cells), "\n\n")
cat("Treatment distribution in metastases:\n")
print(table(met_cells$TreatmentStage_ofProcessedSample, useNA = "always"))

# Create DotPlot by treatment stage
ifn_mps <- grep("^MP2[2-4] ", ucell_cols, value = TRUE)

p <- DotPlot(met_cells, features = ifn_mps, group.by = "TreatmentStage_ofProcessedSample") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Interferon/MHC Metaprograms by Treatment (Metastases Only)")

print(p)
ggsave(file.path(output_dir, "dotplot_ifn_treatment_metastasis.png"), p, width = 10, height = 6, dpi = 300, bg = "white")
```

## Comparisons by Organ {.tabset}

### Overview - Organ

```{r organ_overview}
cat("Cells by organ:\n")
print(table(malignant$organ))
```

### Violin Plots - Top Metaprograms

```{r violin_organ_top, fig.width = 18, fig.height = 20}
top10_mps <- head(mp_means_df$Metaprogram, 10)

plots <- lapply(top10_mps, function(mp) {
  VlnPlot(malignant, features = mp, group.by = "organ", pt.size = 0) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black") +
    ggtitle(gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp)) +
    theme(legend.position = "none", axis.title.x = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1))
})

combined <- wrap_plots(plots, ncol = 2)
print(combined)
ggsave(file.path(output_dir, "violin_organ_top10.png"), combined, width = 18, height = 20, dpi = 300, bg = "white")
```

### Heatmap - Top 30 by Organ

```{r heatmap_organ_top30, fig.width = 14, fig.height = 12}
top30_mps <- head(mp_means_df$Metaprogram, 30)

# Calculate mean scores by organ
mean_scores <- sapply(top30_mps, function(mp) {
  tapply(malignant@meta.data[[mp]], malignant$organ, mean, na.rm = TRUE)
})

# Shorten MP names
colnames(mean_scores) <- gsub("^(MP[0-9]+) .*", "\\1", colnames(mean_scores))

# Create heatmap
col_fun <- colorRamp2(
  quantile(mean_scores, c(0, 0.5, 1)),
  c("blue", "white", "red")
)

ht <- Heatmap(
  mean_scores,
  name = "Mean\nUCell Score",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 8),
  column_title = "Top 30 Metaprograms by Organ"
)

png(file.path(output_dir, "heatmap_organ_top30.png"), width = 14, height = 12, units = "in", res = 300)
draw(ht)
dev.off()

draw(ht)
```

## Combined Analysis: Mutation × Origin {.tabset}

### Cell Cycle Programs

```{r combined_mut_origin_cellcycle, fig.width = 18, fig.height = 12}
# Filter to known primary mutations
mal_known <- subset(malignant, subset = primary_mutation != "unknown")

cell_cycle_mps <- grep("^MP[1-5] ", ucell_cols, value = TRUE)

plots <- lapply(cell_cycle_mps, function(mp) {
  ggplot(mal_known@meta.data, aes(x = primary_mutation, y = .data[[mp]], fill = origin)) +
    geom_violin(alpha = 0.7, scale = "width", position = position_dodge(0.9)) +
    geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.size = 0.3) +
    scale_fill_manual(values = origin_colors) +
    theme_cowplot() +
    ggtitle(gsub("^(MP[0-9]+) .*", "\\1", mp)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title.x = element_blank(),
          legend.position = "bottom")
})

combined <- wrap_plots(plots, ncol = 3, guides = "collect") & theme(legend.position = "bottom")
print(combined)
ggsave(file.path(output_dir, "combined_mutation_origin_cellcycle.png"), combined, width = 18, height = 12, dpi = 300, bg = "white")
```

### EMT Programs

```{r combined_mut_origin_emt, fig.width = 18, fig.height = 12}
emt_mps <- grep("^MP1[5-9] |^MP20 ", ucell_cols, value = TRUE)

plots <- lapply(emt_mps, function(mp) {
  ggplot(mal_known@meta.data, aes(x = primary_mutation, y = .data[[mp]], fill = origin)) +
    geom_violin(alpha = 0.7, scale = "width", position = position_dodge(0.9)) +
    geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.size = 0.3) +
    scale_fill_manual(values = origin_colors) +
    theme_cowplot() +
    ggtitle(gsub("^(MP[0-9]+) .*", "\\1", mp)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title.x = element_blank(),
          legend.position = "bottom")
})

combined <- wrap_plots(plots, ncol = 3, guides = "collect") & theme(legend.position = "bottom")
print(combined)
ggsave(file.path(output_dir, "combined_mutation_origin_emt.png"), combined, width = 18, height = 12, dpi = 300, bg = "white")
```

## Patient-Level Heterogeneity {.tabset}

### Boxplot - Top Metaprograms

```{r patient_boxplot_top, fig.width = 16, fig.height = 20}
# Calculate per-sample mean scores for top 10 metaprograms (replicates merged)
top10_mps <- head(mp_means_df$Metaprogram, 10)

# Filter samples with >= 50 malignant cells (using orig.ident for merged replicates)
sample_counts <- table(malignant$orig.ident)
samples_filt <- names(sample_counts[sample_counts >= 50])
mal_filt <- subset(malignant, subset = orig.ident %in% samples_filt)

cat("Samples with >=50 malignant cells:", length(samples_filt), "\n")

plots <- lapply(top10_mps, function(mp) {
  sample_stats <- mal_filt@meta.data |>
    group_by(orig.ident, origin, primary_mutation) |>
    summarise(mean_score = mean(.data[[mp]], na.rm = TRUE), .groups = "drop")

  ggplot(sample_stats, aes(x = primary_mutation, y = mean_score, fill = origin)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, size = 0.5, alpha = 0.5) +
    scale_fill_manual(values = origin_colors) +
    theme_cowplot() +
    ggtitle(gsub("^(MP[0-9]+) (.*)", "\\1: \\2", mp)) +
    ylab("Mean UCell Score per Sample") +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
})

combined <- wrap_plots(plots, ncol = 2)
print(combined)
ggsave(file.path(output_dir, "sample_boxplot_top10.png"), combined, width = 16, height = 20, dpi = 300, bg = "white")
```

### Heatmap - Sample Heterogeneity

```{r patient_heatmap, fig.width = 16, fig.height = 12}
top20_mps <- head(mp_means_df$Metaprogram, 20)

# Calculate mean scores per sample (using orig.ident for merged replicates)
sample_scores <- sapply(top20_mps, function(mp) {
  tapply(mal_filt@meta.data[[mp]], mal_filt$orig.ident, mean, na.rm = TRUE)
})

# Shorten MP names
colnames(sample_scores) <- gsub("^(MP[0-9]+) .*", "\\1", colnames(sample_scores))

# Get sample metadata
sample_meta <- mal_filt@meta.data |>
  group_by(orig.ident) |>
  summarise(
    origin = unique(origin)[1],
    primary_mutation = unique(primary_mutation)[1],
    .groups = "drop"
  )

# Create annotation
ha <- rowAnnotation(
  Origin = sample_meta$origin[match(rownames(sample_scores), sample_meta$orig.ident)],
  Mutation = sample_meta$primary_mutation[match(rownames(sample_scores), sample_meta$orig.ident)],
  col = list(
    Origin = origin_colors,
    Mutation = primary_mutations_col
  )
)

# Create heatmap
col_fun <- colorRamp2(
  quantile(sample_scores, c(0, 0.5, 1), na.rm = TRUE),
  c("blue", "white", "red")
)

ht <- Heatmap(
  sample_scores,
  name = "Mean\nUCell Score",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 7),
  column_names_gp = gpar(fontsize = 9),
  column_title = "Sample Heterogeneity - Top 20 Metaprograms",
  right_annotation = ha
)

png(file.path(output_dir, "sample_heatmap_top20.png"), width = 16, height = 12, units = "in", res = 300)
draw(ht)
dev.off()

draw(ht)
```

## Export Results

```{r export_results}
# Export UCell scores for all cells (using orig.ident as Sample for merged replicates)
ucell_scores <- malignant@meta.data[, c("orig.ident", "majority_celltype", "primary_mutation",
                                         "secondary_mutation", "origin", "organ", "LocationPrimaryTumor", ucell_cols)]
colnames(ucell_scores)[1] <- "Sample"  # Rename orig.ident to Sample

write.csv(ucell_scores,
          file.path(output_dir, "malignant_ucell_scores_all.csv"),
          row.names = FALSE)

# Export summary statistics by primary mutation
prim_mut_stats <- mal_known_prim@meta.data |>
  group_by(primary_mutation) |>
  summarise(across(all_of(ucell_cols), ~mean(.x, na.rm = TRUE)), .groups = "drop")

write.csv(prim_mut_stats,
          file.path(output_dir, "ucell_stats_primary_mutation.csv"),
          row.names = FALSE)

# Export summary statistics by secondary mutation
sec_mut_stats <- mal_known_sec@meta.data |>
  group_by(secondary_mutation) |>
  summarise(across(all_of(ucell_cols), ~mean(.x, na.rm = TRUE)), .groups = "drop")

write.csv(sec_mut_stats,
          file.path(output_dir, "ucell_stats_secondary_mutation.csv"),
          row.names = FALSE)

# Export summary statistics by origin
origin_stats <- malignant@meta.data |>
  group_by(origin) |>
  summarise(across(all_of(ucell_cols), ~mean(.x, na.rm = TRUE)), .groups = "drop")

write.csv(origin_stats,
          file.path(output_dir, "ucell_stats_origin.csv"),
          row.names = FALSE)

cat("Exported results:\n")
cat("  - malignant_ucell_scores_all.csv\n")
cat("  - ucell_stats_primary_mutation.csv\n")
cat("  - ucell_stats_secondary_mutation.csv\n")
cat("  - ucell_stats_origin.csv\n")
```

## Comparisons by Treatment Stage {.tabset}

Stratify metaprogram activity by treatment stage at the time of sample processing.

### Treatment Overview

```{r treatment_overview}
# Check TreatmentStage_ofProcessedSample in melanocytic cells
cat("=== Treatment Stage Distribution (Melanocytic Cells) ===\n\n")

if ("TreatmentStage_ofProcessedSample" %in% colnames(malignant@meta.data)) {
  treatment_table <- table(malignant$TreatmentStage_ofProcessedSample, useNA = "always")
  print(treatment_table)

  # Cross-tabulation with origin
  cat("\n=== Treatment Stage by Origin ===\n")
  print(table(malignant$TreatmentStage_ofProcessedSample, malignant$origin, useNA = "always"))

  # Simplify treatment categories
  malignant$treatment_simplified <- case_when(
    malignant$TreatmentStage_ofProcessedSample %in% c("no treatment", NA) ~ "Untreated",
    malignant$TreatmentStage_ofProcessedSample %in% c("Enucleation", "Iridocyclectomy", "Plaque", "Radiotherapy") ~ "Primary Treatment",
    grepl("Recurrence", malignant$TreatmentStage_ofProcessedSample) ~ "Post-Recurrence",
    malignant$TreatmentStage_ofProcessedSample %in% c("Tebentafusp", "Tebentafusp/Mekinist", "pip_nivo") ~ "Systemic Therapy",
    grepl("NCT", malignant$TreatmentStage_ofProcessedSample) ~ "Clinical Trial",
    TRUE ~ "Other"
  )

  cat("\n=== Simplified Treatment Categories ===\n")
  print(table(malignant$treatment_simplified))
} else {
  cat("TreatmentStage_ofProcessedSample column not found.\n")
}
```

### Metaprograms by Treatment (Primary Samples)

```{r treatment_primary_samples, fig.width=16, fig.height=24}
if ("TreatmentStage_ofProcessedSample" %in% colnames(malignant@meta.data)) {
  # Focus on primary tumors with different treatments
  primary_cells <- subset(malignant, origin == "primary")
  cat("Primary tumor cells:", ncol(primary_cells), "\n")

  # Get top 10 metaprograms
  top_mps <- ucell_cols[1:min(10, length(ucell_cols))]

  # Violin plots by treatment
  plots <- lapply(top_mps, function(mp) {
    VlnPlot(primary_cells, features = mp, group.by = "TreatmentStage_ofProcessedSample", pt.size = 0) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
            legend.position = "none") +
      ggtitle(gsub("_UCell", "", mp))
  })

  combined <- wrap_plots(plots, ncol = 2)
  print(combined)

  ggsave(
    file.path(output_dir, "violin_treatment_primary_top10.png"),
    combined, width = 16, height = 20, dpi = 300, bg = "white"
  )
}
```

### Metaprograms by Treatment (Metastatic Samples)

```{r treatment_metastatic_samples, fig.width=16, fig.height=24}
if ("TreatmentStage_ofProcessedSample" %in% colnames(malignant@meta.data)) {
  # Focus on metastatic samples with different treatments
  met_cells <- subset(malignant, origin == "metastasis")
  cat("Metastatic cells:", ncol(met_cells), "\n")

  # Get treatment distribution in metastatic samples
  cat("\nTreatment distribution in metastatic samples:\n")
  print(table(met_cells$TreatmentStage_ofProcessedSample))

  # Violin plots by treatment for metastatic samples
  plots <- lapply(top_mps, function(mp) {
    VlnPlot(met_cells, features = mp, group.by = "TreatmentStage_ofProcessedSample", pt.size = 0) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
            legend.position = "none") +
      ggtitle(gsub("_UCell", "", mp))
  })

  combined <- wrap_plots(plots, ncol = 2)
  print(combined)

  ggsave(
    file.path(output_dir, "violin_treatment_metastatic_top10.png"),
    combined, width = 16, height = 20, dpi = 300, bg = "white"
  )
}
```

### Pre-Treatment vs Post-Treatment Comparison

```{r pretreatment_comparison, fig.width=14, fig.height=10}
if ("treatment_simplified" %in% colnames(malignant@meta.data)) {
  # Compare untreated vs treated (primary therapy) in primary samples
  primary_cells <- subset(malignant, origin == "primary")

  # Filter to primary treatment vs post-recurrence
  treatment_compare <- subset(
    primary_cells,
    treatment_simplified %in% c("Primary Treatment", "Post-Recurrence")
  )

  cat("Cells for comparison:\n")
  print(table(treatment_compare$treatment_simplified))

  if (ncol(treatment_compare) > 100) {
    # Statistical comparison
    cat("\n=== Statistical Comparison: Primary Treatment vs Post-Recurrence ===\n\n")

    results <- data.frame(
      metaprogram = character(),
      mean_primary = numeric(),
      mean_recurrence = numeric(),
      p_value = numeric(),
      stringsAsFactors = FALSE
    )

    for (mp in top_mps) {
      primary_vals <- treatment_compare@meta.data[[mp]][treatment_compare$treatment_simplified == "Primary Treatment"]
      recurrence_vals <- treatment_compare@meta.data[[mp]][treatment_compare$treatment_simplified == "Post-Recurrence"]

      if (length(primary_vals) > 10 && length(recurrence_vals) > 10) {
        test <- wilcox.test(primary_vals, recurrence_vals)
        results <- rbind(results, data.frame(
          metaprogram = gsub("_UCell", "", mp),
          mean_primary = mean(primary_vals, na.rm = TRUE),
          mean_recurrence = mean(recurrence_vals, na.rm = TRUE),
          p_value = test$p.value
        ))
      }
    }

    results$p_adj <- p.adjust(results$p_value, method = "BH")
    results$significant <- results$p_adj < 0.05
    results <- results[order(results$p_value), ]

    print(results)

    write.csv(
      results,
      file.path(output_dir, "treatment_comparison_stats.csv"),
      row.names = FALSE
    )

    # Visualization
    results_long <- results |>
      pivot_longer(cols = c(mean_primary, mean_recurrence),
                   names_to = "group", values_to = "mean_score") |>
      mutate(group = gsub("mean_", "", group))

    p_compare <- ggplot(results_long, aes(x = reorder(metaprogram, -mean_score), y = mean_score, fill = group)) +
      geom_bar(stat = "identity", position = "dodge") +
      scale_fill_manual(values = c("primary" = "#E41A1C", "recurrence" = "#377EB8"),
                        labels = c("Primary Treatment", "Post-Recurrence")) +
      labs(
        title = "Metaprogram Activity: Primary Treatment vs Post-Recurrence",
        x = "Metaprogram",
        y = "Mean UCell Score",
        fill = "Treatment Stage"
      ) +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    print(p_compare)

    ggsave(
      file.path(output_dir, "barplot_primary_vs_recurrence.png"),
      p_compare, width = 12, height = 6, dpi = 300, bg = "white"
    )
  }
}
```

### Systemic Therapy Response (Metastases)

```{r systemic_therapy, fig.width=14, fig.height=14}
if ("treatment_simplified" %in% colnames(malignant@meta.data)) {
  # Focus on metastatic samples: untreated vs systemic therapy
  met_cells <- subset(malignant, origin == "metastasis")

  cat("=== Metastatic Samples by Treatment ===\n")
  print(table(met_cells$treatment_simplified))

  # Compare untreated vs systemic therapy in metastatic samples
  met_compare <- subset(
    met_cells,
    treatment_simplified %in% c("Untreated", "Systemic Therapy")
  )

  if (ncol(met_compare) > 100) {
    cat("\nCells for comparison (Untreated vs Systemic Therapy):\n")
    print(table(met_compare$treatment_simplified))

    # Violin plots
    plots <- lapply(top_mps[1:min(6, length(top_mps))], function(mp) {
      VlnPlot(met_compare, features = mp, group.by = "treatment_simplified", pt.size = 0) +
        theme(legend.position = "none") +
        ggtitle(gsub("_UCell", "", mp))
    })

    combined <- wrap_plots(plots, ncol = 2)
    print(combined)

    ggsave(
      file.path(output_dir, "violin_systemic_therapy_effect.png"),
      combined, width = 12, height = 12, dpi = 300, bg = "white"
    )
  }
}
```

### Export Treatment Statistics

```{r export_treatment_stats}
if ("TreatmentStage_ofProcessedSample" %in% colnames(malignant@meta.data)) {
  # Export statistics by treatment
  treatment_stats <- malignant@meta.data |>
    filter(!is.na(TreatmentStage_ofProcessedSample)) |>
    group_by(TreatmentStage_ofProcessedSample) |>
    summarise(
      n_cells = n(),
      across(all_of(ucell_cols), ~mean(.x, na.rm = TRUE)),
      .groups = "drop"
    )

  write.csv(
    treatment_stats,
    file.path(output_dir, "ucell_stats_treatment.csv"),
    row.names = FALSE
  )

  # Export simplified treatment stats
  if ("treatment_simplified" %in% colnames(malignant@meta.data)) {
    simplified_stats <- malignant@meta.data |>
      group_by(treatment_simplified) |>
      summarise(
        n_cells = n(),
        across(all_of(ucell_cols), ~mean(.x, na.rm = TRUE)),
        .groups = "drop"
      )

    write.csv(
      simplified_stats,
      file.path(output_dir, "ucell_stats_treatment_simplified.csv"),
      row.names = FALSE
    )
  }

  cat("Exported treatment statistics:\n")
  cat("  - ucell_stats_treatment.csv\n")
  cat("  - ucell_stats_treatment_simplified.csv\n")
}
```

## Session Info

```{r session_info}
sessionInfo()
```
