---
title: "CNV Arm Analysis"
author: "Prachi Shukla"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# By Arm CNV Extraction

```{r libraries, message=FALSE}

library(data.table)
library(tidyverse)
library(qs2)
library(GWASTools)
library(GenomicRanges)
```

```{r loading_files}

# infercnv_res_path = "../Data/Derived/inferCNV/"
# files = dir(infercnv_res_path, pattern = "HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.3.pred_cnv_genes.dat", recursive = TRUE)
# 
# genes = lapply(paste0(infercnv_res_path, files), function(x) fread(x) %>% mutate(Sample = gsub(".*/(.*)/.*", "\\1", x)))
# 
# names(genes) = sapply(genes, function(x) unique(x$Sample))
# qs_save(genes, "../Data/Derived/inferCNV_Output/20251024_HMM_gene_CNV_all.qs2")

genes = qs_read("../20251024_HMM_gene_CNV_all.qs2")

data("centromeres.hg38", package = "GWASTools")
centromeres <- centromeres.hg38 %>%
  transmute(chr = paste0("chr", chrom), start = left.base, end = right.base) 


# chromosomes 13, 14, 15, 21, 22 are acrocentric, so p arms (13p/14p/15p/21p/22p) have essentially no annotated genes in GRCh38
```



```{r}

centro_gr <- GRanges(seqnames = centromeres$chr,
                     ranges = IRanges(start = centromeres$start, end = centromeres$end))

for(i in names(genes)){

genes_gr <- GRanges(seqnames = genes[[i]]$chr,
                    ranges = IRanges(start = genes[[i]]$start, end = genes[[i]]$end),
                    gene = genes[[i]]$gene)

# Find overlaps with the centromere region (label "cen")
hits <- findOverlaps(genes_gr, centro_gr)
genes[[i]]$arm <- "unassigned"
genes[[i]]$arm[unique(queryHits(hits))] <- "cen"

# For non-centromeric genes[[i]], use the centromere midpoint as the p/q boundary
centro_mid <- centromeres %>%
  mutate(mid = floor((start + end) / 2)) %>%
  select(chr, mid)


 genes[[i]] <- genes[[i]] %>%
   dplyr::rename(infercnv_subcluster = cell_group_name) %>% 
  left_join(centro_mid, by = c("chr")) %>%
  mutate(arm = ifelse(arm == "cen", "cen",
               ifelse(end <= mid, "p",
               ifelse(start >= mid, "q", "cen"))),
         state_cat = case_when(
           state <= 2 ~ "loss",
           state >= 4 ~ "gain",
           state == 3 ~ "neutral",
         )) %>%
  select(infercnv_subcluster, gene, chr, arm, state, state_cat)
}


```


```{r eval=FALSE}

summary_cnv = list()
for(i in names(genes)){

  ## ARM summary (p/q only; ignore 'cen')
  arm_sum <- genes[[i]][arm %in% c("p","q"), .(
    n_genes    = .N,
    n_gain     = sum(state_cat == "gain"),
    n_loss     = sum(state_cat == "loss"),
    frac_gain  = mean(state_cat == "gain"),
    frac_loss  = mean(state_cat == "loss"),
    majority   = c("gain","loss","neutral")[
                   which.max(c(sum(state_cat == "gain"),
                               sum(state_cat == "loss"),
                               sum(state_cat == "neutral")))
                 ]
  ), by = .(infercnv_subcluster, chr, arm)][
    , chrom_arm := paste0(chr, arm)][
    order(infercnv_subcluster, chr, arm)]

  arm_sum[, strict_call := fifelse(frac_gain >= 0.6, "gain",
                            fifelse(frac_loss >= 0.6, "loss", "neutral"))]


  # wide table of strict calls for p and q
  calls_wide <- dcast(
    arm_sum[arm %in% c("p","q")],
    infercnv_subcluster + chr ~ arm,
    value.var = "strict_call"
  )

# counts of genes
    counts_wide <- dcast(
    arm_sum[arm %in% c("p","q")],
    infercnv_subcluster + chr ~ arm,
    value.var = "n_genes"
  )
  setnames(counts_wide, c("p","q"), c("n_genes_p","n_genes_q"))

  both_wide <- merge(calls_wide, counts_wide, by = c("infercnv_subcluster","chr"), all = TRUE)

  # AND rule:
  # - if p missing (e.g., acrocentric), inherit q; if q missing, inherit p
  # - gain only if p==gain & q==gain; loss only if p==loss & q==loss; else different
  both_wide[, both_call := fifelse(is.na(p), q,
                            fifelse(is.na(q), p,
                            fifelse(p=="gain" & q=="gain", "gain",
                            fifelse(p=="loss" & q=="loss", "loss", 
                            fifelse(p=="neutral" & q=="neutral", "neutral", 
                                    "different")))))]

  # build "both" rows; fractions aren’t meaningful for AND, set to NA
  both_sum <- both_wide[, .(
    infercnv_subcluster,
    chr,
    arm        = "both",
    chrom_arm  = paste0(chr, "both"),
    n_genes    = rowSums(cbind(n_genes_p, n_genes_q), na.rm = TRUE),
    n_gain     = NA_integer_,
    n_loss     = NA_integer_,
    frac_gain  = NA_real_,
    frac_loss  = NA_real_,
    majority   = both_call,
    strict_call= both_call
  )]

  ## combine arm + both
  summary_cnv[[i]] <- rbindlist(list(arm_sum, both_sum), use.names = TRUE, fill = TRUE)
}

combined = list(
  Genes = rbindlist(genes, idcol = "Sample"),
  CNV_Summary = rbindlist(summary_cnv, idcol = "Sample") %>%
    mutate(infercnv_subcluster = gsub(".*\\.(.*)", "\\1", infercnv_subcluster))
)

qs_save(combined, "../20251031_by_arm_chr_CNV.qs2")

```

# Analysis

```{r libraries, message=FALSE}

library(ggplot2)
library(data.table)
library(tidyverse)
library(patchwork)
library(RColorBrewer)
library(viridisLite)
library(scales)
library(qs)
library(qs2)
library(Seurat)
library(SCpubr)
library(forcats)
library(grid)
library(scales)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

theme_set(theme_classic())

cnv_type = c("gain" = "#9e002a", "loss" = "#00428c")

bar_label = c("metastasis" = "#ef476f", "primary" = "#26547c", "healthy" = "#06d6a0")

chrom_arm_thresh = 100
```


```{r loading_files}

coi = c("gain" =  "chr6p", "gain" =  "chr8q", "gain" =  "chr1q", "gain" = "chr8both",
"loss" = "chr1p", "loss" = "chr3both", "loss" = "chr6q", "loss" = "chr8p", "loss" = "chr16q", "loss" = "chr9p")

so = qs::qread("../01_integrated_melanoma_melanocyte_subset.qs")

cnv_arms = qs_read("../20251031_by_arm_chr_CNV.qs2")$CNV_Summary


```


# Introduction
The inferCNV results were extracted for each sample and gene locations split by centromere. Chromosomes 13, 14, 15, 21, 22 are acrocentric and have no genes in their p arms. CNV for an arm are called if ≥ 60% of genes are have a certain CNV state. A 'both' suffix means the p and q arm calls concur (both gain or both loss). Chromosome arms with less than `r chrom_arm_thresh` cells are removed.

```{r chromosome_info}

chr_call = cnv_arms %>% select(Sample, infercnv_subcluster, chrom_arm, strict_call) %>% distinct()

# Adding chromosome mutation information
chr_call = so@meta.data %>% select(Sample, sample_barcode, infercnv_subcluster, seurat_clusters_CNV, origin, Malignant) %>% 
  merge(., chr_call, by = c("Sample", "infercnv_subcluster"))  %>% arrange(Sample, infercnv_subcluster) 

# Removing healthy cells
# chr_call_filt = chr_call %>% filter(Malignant == "Malignant")

```


```{r removing_lowly_occuring_calls}

cells_by_arm <- chr_call %>% group_by(Sample, origin, infercnv_subcluster, chrom_arm) %>% count(name = "n_cells") %>% filter(n_cells > chrom_arm_thresh) %>% 
  unite(col = "good_combs", c("Sample", "infercnv_subcluster", "chrom_arm"))

```



```{r counting}
# All chromosomes
# If a sample has a chromosome arm with a loss or a gain as counted above it is counted. If both arms have the same CNV the sample is counted under 'both'. 
arm_by_sample <- chr_call %>% 
  unite(col = "good_combs", c("Sample", "infercnv_subcluster", "chrom_arm"), remove = FALSE) %>% 
  filter(good_combs %in% cells_by_arm$good_combs) %>% 
  select(Sample, infercnv_subcluster, chrom_arm, strict_call, origin) 

# summarize to one row per Sample x chrom_arm
arm_by_sample <- arm_by_sample %>%
  
  mutate(is_gain = strict_call == "gain",
         is_loss = strict_call == "loss") %>%
  group_by(Sample, chrom_arm, origin) %>%
  summarise(
    has_gain = any(is_gain, na.rm = TRUE),
    has_loss = any(is_loss, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    CNV = case_when(
      has_gain & !has_loss ~ "gain",
      has_loss & !has_gain ~ "loss",
      has_gain & has_loss & str_ends(chrom_arm, "both") ~ "different",
      TRUE                 ~ NA_character_
    )
  ) %>% 
  select(Sample, chrom_arm, CNV, origin) %>% filter(!is.na(CNV)) %>% mutate(chr_id = str_remove(chrom_arm, "(p|q|both)$")) %>%
  group_by(Sample, chr_id, origin) %>%
  filter(
    if (any(str_ends(chrom_arm, "both") & CNV != "different")) {
      str_ends(chrom_arm, "both")      # keep only 'both'
    } else {
      !str_ends(chrom_arm, "both")     # keep p/q (no 'both' or 'both' is "different")
    }
  ) %>% ungroup()

```

# Selected Chromosomes

```{r, fig.width=6, fig.height=10}

chr_counts <- arm_by_sample %>%
  filter(chrom_arm %in% coi) %>%
  count(name = "Num_Samps", chrom_arm, CNV, origin) %>%
  mutate(
    order_ = as.numeric(gsub("^chr(\\d+).*$", "\\1", chrom_arm))
  )

arms_add <- coi

chr_counts2 <- chr_counts %>%
  bind_rows(
    expand_grid(
      chrom_arm = arms_add,
      CNV       = unique(chr_counts$CNV),
      origin    = "healthy"
    ) %>% mutate(Num_Samps = 0L)
  ) %>%
  distinct(chrom_arm, CNV, origin, .keep_all = TRUE) %>%
  mutate(
    order_ = as.numeric(gsub("^chr(\\d+).*$", "\\1", chrom_arm)),
    Num_Samps = if_else(CNV == "loss", -Num_Samps, Num_Samps)
  )

## 2) Same y-axis scale for all three plots (symmetric around 0 is usually nicest)
m <- max(abs(chr_counts2$Num_Samps), na.rm = TRUE)
ylims <- c(-m, m)

base_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1),
  panel.spacing.x = unit(0.15, "lines")
)

p_healthy <- ggplot(chr_counts2 %>% filter(origin == "healthy"),
  aes(x = fct_reorder(chrom_arm, order_), y = Num_Samps, fill = CNV)
) +
  geom_col() +
  guides(fill = "none") +
  labs(x = "Chromosome Arm", y = "Instances of CNV", title = "Healthy") +
  scale_fill_manual(values = cnv_type) +
  facet_grid(~order_, scales = "free_x", space = "free_x") +
  coord_cartesian(ylim = ylims) +
  geom_hline(yintercept = 0, colour = "lightgrey", linewidth = 0.01) +
  base_theme

p_primary <- ggplot(chr_counts2 %>% filter(origin == "primary"),
  aes(x = fct_reorder(chrom_arm, order_), y = Num_Samps, fill = CNV)
) +
  geom_col() +
  labs(x = "Chromosome Arm", fill = "CNV Call", y = "Instances of CNV", title = "Primary") +
  scale_fill_manual(values = cnv_type) +
  facet_grid(~order_, scales = "free_x", space = "free_x") +
  coord_cartesian(ylim = ylims) +
  geom_hline(yintercept = 0, colour = "lightgrey", linewidth = 0.01) +
  base_theme

p_mets <- ggplot(chr_counts2 %>% filter(origin == "metastasis"),
  aes(x = fct_reorder(chrom_arm, order_), y = Num_Samps, fill = CNV)
) +
  geom_col() +
  guides(fill = "none") +
  labs(x = "Chromosome Arm", y = "Instances of CNV", title = "Metastasis") +
  scale_fill_manual(values = cnv_type) +
  facet_grid(~order_, scales = "free_x", space = "free_x") +
  coord_cartesian(ylim = ylims) +
  geom_hline(yintercept = 0, colour = "lightgrey", linewidth = 0.01) +
  base_theme

# pdf("UM_chromosome_arm_CNV_analysis.pdf", width = 5, height = 7)
(p_healthy / p_primary / p_mets) + plot_layout(guides = "collect")
# dev.off()
```


