---
title: "p31662 - CytoTRACE2 Analysis of Uveal Melanoma Atlas"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Figures 2, 3C, 5E (CytoTRACE2 cellular potency / stemness analysis on the UM atlas)
```

#  {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

**Purpose**: Analyze cellular differentiation potential and stemness in uveal melanoma atlas using CytoTRACE 2.

**CytoTRACE 2** predicts cellular potency from scRNA-seq data, outputting:
- Continuous scores (0 = differentiated, 1 = totipotent)
- Discrete potency categories (Differentiated, Unipotent, Oligopotent, Multipotent, Pluripotent, Totipotent)
- Relative ordering within the dataset

**Reference**: https://github.com/digitalcytometry/cytotrace2

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)

library(Seurat)
library(CytoTRACE2)
library(qs)
library(tidyverse)
library(ggplot2)
library(dittoSeq)
library(cowplot)
library(patchwork)
library(dplyr)
library(ggpubr)

# Load Victoria's official color palettes
source("colors.R")

# CRITICAL: Reorder origin_colors to match disease progression (healthy → primary → metastasis)
# This ensures consistent ordering across all plots
origin_order <- c("healthy", "primary", "metastasis")
origin_colors <- origin_colors[origin_order]

output_dir <- "Outputs/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Color palettes
# Potency category colors (CytoTRACE2 standard)
labels <- c("Differentiated", "Unipotent", "Oligopotent",
            "Multipotent", "Pluripotent", "Totipotent")
colors.cyto <- c("#9E0142", "#F46D43", "#FEE08B",
                 "#E6F598", "#66C2A5", "#5E4FA2")
names(colors.cyto) <- labels

# Use Victoria's cell type colors
celltype_colors <- majority_celltype_col

# Extended colors for additional cell types (fallback)
base_colors <- unname(pals::polychrome())
mycols <- colorRampPalette(base_colors)(100)

# Increased base font size for better readability (per collaborator request)
theme_set(theme_minimal(base_size = 14))
```

## Load Data

```{r load_data}
# Load Victoria's updated dataset (Dec 2025) - has correct cells and UMAP
cat("Loading Victoria's updated UM atlas (Dec 2025)...\n")
um_atlas <- qs::qread(
  "../13_final_Dataset_reclustered_28012025.qs",
  nthreads = 32
)

cat("Loaded new UM atlas:\n")
cat("  - Cells:", ncol(um_atlas), "\n")
cat("  - Features:", nrow(um_atlas), "\n")
cat("  - Assays:", names(um_atlas@assays), "\n\n")

# Load pre-computed CytoTRACE2 results from old object and transfer by barcode matching
cat("Loading pre-computed CytoTRACE2 results...\n")
old_atlas <- qs::qread(
  "../UM_atlas_cytotrace2.qs2",
  nthreads = 32
)

cat("Old atlas with CytoTRACE2:\n")
cat("  - Cells:", ncol(old_atlas), "\n\n")

# Transfer CytoTRACE2 scores by matching cell barcodes
cat("Transferring CytoTRACE2 scores to new object...\n")
common_cells <- intersect(colnames(um_atlas), colnames(old_atlas))
cat("  - Common cells:", length(common_cells), "\n")
cat("  - New cells (no CytoTRACE2):", sum(!colnames(um_atlas) %in% common_cells), "\n")
cat("  - Removed cells (had CytoTRACE2):", sum(!colnames(old_atlas) %in% common_cells), "\n\n")

# Transfer CytoTRACE2 columns
cytotrace_cols <- c("CytoTRACE2_Score", "CytoTRACE2_Relative", "CytoTRACE2_Potency",
                    "preKNN_CytoTRACE2_Score", "preKNN_CytoTRACE2_Potency")

for (col in cytotrace_cols) {
  if (col %in% colnames(old_atlas@meta.data)) {
    um_atlas@meta.data[[col]] <- NA
    um_atlas@meta.data[common_cells, col] <- old_atlas@meta.data[common_cells, col]
    cat("  Transferred:", col, "-", sum(!is.na(um_atlas@meta.data[[col]])), "cells\n")
  }
}

# Convert integer potency to factor with proper labels
potency_labels <- c("Differentiated", "Unipotent", "Oligopotent",
                    "Multipotent", "Pluripotent", "Totipotent")
um_atlas$CytoTRACE2_Potency <- factor(
  potency_labels[um_atlas$CytoTRACE2_Potency],
  levels = potency_labels
)
um_atlas$preKNN_CytoTRACE2_Potency <- factor(
  potency_labels[um_atlas$preKNN_CytoTRACE2_Potency],
  levels = potency_labels
)
cat("  Converted potency columns to factor with labels\n")

# Clean up old object to free memory
rm(old_atlas)
gc()

cat("\nCytoTRACE2 transfer complete!\n")
cat("Coverage:", round(100 * sum(!is.na(um_atlas@meta.data$CytoTRACE2_Score)) / ncol(um_atlas), 1), "%\n\n")

# Display metadata columns
cat("Metadata columns:\n")
print(colnames(um_atlas@meta.data))

# Check for cell type annotations
celltype_cols <- grep("cell.*type|cluster|annotation", colnames(um_atlas@meta.data),
                      value = TRUE, ignore.case = TRUE)
cat("\nPotential cell type columns:\n")
print(celltype_cols)
```

## Run CytoTRACE2 {.tabset}

### Compute Scores

```{r run_cytotrace2, eval = FALSE}
# Seurat v5 compatibility: Join layers before extracting counts
cat("Joining RNA assay layers for CytoTRACE2 compatibility...\n")
um_atlas[["RNA"]] <- JoinLayers(um_atlas[["RNA"]])
cat("Layers joined successfully.\n\n")

# Extract count matrix
cat("Extracting count matrix...\n")
count_matrix <- GetAssayData(um_atlas, assay = "RNA", layer = "counts")
count_matrix <- as.matrix(count_matrix)
cat("Count matrix extracted: ", nrow(count_matrix), "genes x", ncol(count_matrix), "cells\n\n")

# Run CytoTRACE2 on the full dataset
cat("Running CytoTRACE2 on full dataset...\n")
cytotrace2_result <- cytotrace2(
  count_matrix,
  is_seurat = FALSE,
  ncores = 32,
  species = "human",
  batch_size = 10000
)

cat("CytoTRACE2 computation complete!\n\n")

# Extract results from the CytoTRACE2 output
cat("Extracting results...\n")
cytotrace2_df <- cytotrace2_result$CytoTRACE2_results

# Add results to Seurat object metadata
cat("Adding results to Seurat object...\n")
um_atlas$CytoTRACE2_Score <- cytotrace2_df$CytoTRACE2_Score[match(colnames(um_atlas), rownames(cytotrace2_df))]
um_atlas$CytoTRACE2_Relative <- cytotrace2_df$CytoTRACE2_Relative[match(colnames(um_atlas), rownames(cytotrace2_df))]
um_atlas$CytoTRACE2_Potency <- cytotrace2_df$CytoTRACE2_Potency[match(colnames(um_atlas), rownames(cytotrace2_df))]
um_atlas$preKNN_CytoTRACE2_Score <- cytotrace2_df$preKNN_CytoTRACE2_Score[match(colnames(um_atlas), rownames(cytotrace2_df))]
um_atlas$preKNN_CytoTRACE2_Potency <- cytotrace2_df$preKNN_CytoTRACE2_Potency[match(colnames(um_atlas), rownames(cytotrace2_df))]

cat("CytoTRACE2 analysis complete!\n")
cat("  - CytoTRACE2_Score: ", sum(!is.na(um_atlas$CytoTRACE2_Score)), "cells\n")
cat("  - CytoTRACE2_Potency categories: ", table(um_atlas$CytoTRACE2_Potency), "\n\n")

# Save results
output_path <- file.path(
  "..",
  "UM_atlas_cytotrace2.qs2"
)
cat("Saving to:", output_path, "\n")
qs::qsave(um_atlas, output_path, nthreads = 32)

cat("\nResults saved!\n")
```

### Check CytoTRACE2 Results

```{r load_results}
# Check if CytoTRACE2 results exist in current object (computed in previous chunk)
# NOTE: We no longer load from cached file to ensure we use the correct NEW dataset

# Flag to track if we have CytoTRACE2 results
has_cytotrace2 <- "CytoTRACE2_Score" %in% colnames(um_atlas@meta.data) &&
                  sum(!is.na(um_atlas$CytoTRACE2_Score)) > 0

if (has_cytotrace2) {
  cat("CytoTRACE2 results found in current object.\n")
  cat("Cells with CytoTRACE2 scores:", sum(!is.na(um_atlas$CytoTRACE2_Score)), "\n")
} else {
  cat("CytoTRACE2 results not found.\n")
  cat("\nTo compute CytoTRACE2 scores:\n")
  cat("1. Set eval=TRUE in the 'Compute Scores' chunk above\n")
  cat("2. Re-render this document\n")
  cat("\nSkipping visualization sections...\n")
}

# Initialize has_malignant for later use
has_malignant <- FALSE

# Check CytoTRACE2 columns if we have results
if (has_cytotrace2) {
  cytotrace_cols <- grep("CytoTRACE2", colnames(um_atlas@meta.data), value = TRUE)
  cat("\nCytoTRACE2 columns:\n")
  print(cytotrace_cols)

  if ("CytoTRACE2_Score" %in% colnames(um_atlas@meta.data)) {
    cat("\nCytoTRACE2 Score summary:\n")
    print(summary(um_atlas$CytoTRACE2_Score))

    cat("\nCytoTRACE2 Potency distribution:\n")
    print(table(um_atlas$CytoTRACE2_Potency))
  } else {
    has_cytotrace2 <- FALSE
    cat("\nERROR: CytoTRACE2 columns not found in loaded object!\n")
  }
}
```

## Data Cleaning {.tabset}

### Fix Primary Mutation Typo

```{r fix_primary_mutation}
# Fix typo: GANQ should be GNAQ (applied after loading data)
cat("Fixing primary_mutation typo: GANQ → GNAQ\n")
cat("\nBefore correction:\n")
print(table(um_atlas$primary_mutation))

# Merge GANQ into GNAQ
um_atlas$primary_mutation[um_atlas$primary_mutation == "GANQ"] <- "GNAQ"

cat("\nAfter correction:\n")
print(table(um_atlas$primary_mutation))
cat("\n✅ GANQ merged into GNAQ\n")
```

## UMAP Visualizations {.tabset}

### Cell Type Reference

```{r umap_celltype, fig.width = 10, fig.height = 9, eval = has_cytotrace2}
# Use majority_celltype for cell type annotations
celltype_col <- "majority_celltype"

if (celltype_col %in% colnames(um_atlas@meta.data)) {
  cat("Using cell type column:", celltype_col, "\n")
  cat("\nCell type distribution:\n")
  print(table(um_atlas@meta.data[[celltype_col]]))
} else {
  cat("ERROR: majority_celltype column not found in metadata!\n")
  cat("Available columns:", paste(colnames(um_atlas@meta.data), collapse = ", "), "\n")
}

# Set up colors based on number of groups
n_groups <- length(unique(um_atlas@meta.data[[celltype_col]]))
if (n_groups <= length(mycols)) {
  group_colors <- mycols[1:n_groups]
} else {
  group_colors <- mycols
}

p_celltype <- DimPlot(
  um_atlas,
  reduction = "umap",
  group.by = celltype_col,
  cols = celltype_colors,
  label = TRUE,
  repel = TRUE,
  order = TRUE,
  raster = FALSE
) +
  coord_fixed() +
  ggtitle(paste("Cell Type Annotations -", celltype_col))

print(p_celltype)
ggsave(file.path(output_dir, "umap_celltypes.png"), p_celltype,
       width = 10, height = 9, dpi = 300, bg = "white")
```

### CytoTRACE2 Relative Score

```{r umap_relative, fig.width = 9, fig.height = 8, eval = has_cytotrace2}
# Setup UMAP limits for consistent aspect ratio
x_limits <- range(um_atlas@reductions$umap@cell.embeddings[, 1], na.rm = TRUE)
y_limits <- range(um_atlas@reductions$umap@cell.embeddings[, 2], na.rm = TRUE)

p_relative <- FeaturePlot(
  um_atlas,
  reduction = "umap",
  features = "CytoTRACE2_Relative",
  raster = FALSE
) +
  scale_colour_gradientn(
    colours = c("#000004FF", "#3B0F70FF", "#8C2981FF",
                "#DE4968FF", "#FE9F6DFF", "#FCFDBFFF"),
    na.value = "transparent",
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    labels = c("0.0 (More diff.)", "0.2", "0.4", "0.6", "0.8", "1.0 (Less diff.)"),
    name = "Relative\norder",
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  ) +
  coord_fixed() +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  ggtitle("CytoTRACE 2 Relative Score") +
  xlab("UMAP1") + ylab("UMAP2") +
  theme(
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20))
  )

print(p_relative)
ggsave(file.path(output_dir, "umap_cytotrace2_relative.png"), p_relative,
       width = 9, height = 8, dpi = 300, bg = "white")
```

### CytoTRACE2 Absolute Score

```{r umap_absolute, fig.width = 9, fig.height = 8, eval = has_cytotrace2}
p_absolute <- FeaturePlot(
  um_atlas,
  reduction = "umap",
  features = "CytoTRACE2_Score",
  raster = FALSE
) +
  scale_colour_gradientn(
    colours = c("#000004FF", "#3B0F70FF", "#8C2981FF",
                "#DE4968FF", "#FE9F6DFF", "#FCFDBFFF"),
    na.value = "transparent",
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    labels = c("0.0 (More diff.)", "0.2", "0.4", "0.6", "0.8", "1.0 (Less diff.)"),
    name = "CytoTRACE2\nScore",
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  ) +
  coord_fixed() +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  ggtitle("CytoTRACE 2 Absolute Score") +
  xlab("UMAP1") + ylab("UMAP2") +
  theme(
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20))
  )

print(p_absolute)
ggsave(file.path(output_dir, "umap_cytotrace2_absolute.png"), p_absolute,
       width = 9, height = 8, dpi = 300, bg = "white")
```

### CytoTRACE2 Potency Categories

```{r umap_potency, fig.width = 10, fig.height = 8, eval = has_cytotrace2}
p_potency <- DimPlot(
  um_atlas,
  reduction = "umap",
  group.by = "CytoTRACE2_Potency",
  cols = colors.cyto,
  order = TRUE,
  raster = FALSE
) +
  coord_fixed() +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  ggtitle("CytoTRACE 2 Potency Categories") +
  xlab("UMAP1") + ylab("UMAP2")

print(p_potency)
ggsave(file.path(output_dir, "umap_cytotrace2_potency.png"), p_potency,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Pre-KNN Score (Raw)

```{r umap_preknn, fig.width = 9, fig.height = 8, eval = has_cytotrace2}
p_preknn <- FeaturePlot(
  um_atlas,
  reduction = "umap",
  features = "preKNN_CytoTRACE2_Score",
  raster = FALSE
) +
  scale_colour_gradientn(
    colours = c("#000004FF", "#3B0F70FF", "#8C2981FF",
                "#DE4968FF", "#FE9F6DFF", "#FCFDBFFF"),
    na.value = "transparent",
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    labels = c("0.0 (More diff.)", "0.2", "0.4", "0.6", "0.8", "1.0 (Less diff.)"),
    name = "Pre-KNN\nScore",
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  ) +
  coord_fixed() +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  ggtitle("CytoTRACE 2 Pre-KNN Score (before smoothing)") +
  xlab("UMAP1") + ylab("UMAP2") +
  theme(
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20))
  )

print(p_preknn)
ggsave(file.path(output_dir, "umap_cytotrace2_preknn.png"), p_preknn,
       width = 9, height = 8, dpi = 300, bg = "white")
```

## Violin Plots {.tabset}

### By Cell Type

```{r violin_celltype, fig.width = 14, fig.height = 9, eval = has_cytotrace2}
p_vln_celltype <- VlnPlot(
  um_atlas,
  features = "CytoTRACE2_Score",
  group.by = celltype_col,
  cols = celltype_colors,
  sort = TRUE,
  pt.size = 0
) +
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.2),
    limits = c(0, 1),
    sec.axis = sec_axis(
      trans = ~.,
      breaks = seq(0, 1, by = 1/12),
      labels = c("", "Differentiated", "", "Unipotent", "",
                 "Oligopotent", "", "Multipotent", "",
                 "Pluripotent", "", "Totipotent", "")
    )
  ) +
  labs(x = "Cell Type", y = "Potency Score") +
  ggtitle("Developmental Potential by Cell Type") +
  theme(
    legend.position = "None",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.ticks.y.right = element_line(
      color = c("black", NA, "black", NA, "black", NA,
                "black", NA, "black", NA, "black", NA, "black")
    ),
    aspect.ratio = 0.8,
    axis.ticks.length.y.right = unit(0.3, "cm")
  )

print(p_vln_celltype)
ggsave(file.path(output_dir, "violin_cytotrace2_celltype.png"), p_vln_celltype,
       width = 14, height = 9, dpi = 300, bg = "white")
```

### By Sample (if available)

```{r violin_sample, fig.width = 12, fig.height = 8, eval = has_cytotrace2}
# Use orig.ident to merge biological replicates from same tumor
if ("orig.ident" %in% colnames(um_atlas@meta.data)) {

  n_samples <- length(unique(um_atlas$orig.ident))
  # Use Victoria's sample colors if available
  available_samples <- unique(um_atlas$orig.ident)
  sample_colors_use <- sample_col[names(sample_col) %in% available_samples]
  if (length(sample_colors_use) < n_samples) {
    sample_colors_use <- mycols[1:min(n_samples, length(mycols))]
  }

  p_vln_sample <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    group.by = "orig.ident",
    cols = sample_colors_use,
    sort = TRUE,
    pt.size = 0
  ) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    labs(x = "Sample", y = "Potency Score") +
    ggtitle("Developmental Potential by Sample") +
    theme(
      legend.position = "None",
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      aspect.ratio = 0.8,
      axis.ticks.length.y.right = unit(0.3, "cm")
    )

  print(p_vln_sample)
  ggsave(file.path(output_dir, "violin_cytotrace2_sample.png"), p_vln_sample,
         width = 12, height = 8, dpi = 300, bg = "white")

} else {
  cat("No 'Sample' column found in metadata. Skipping sample-level violin plot.\n")
}
```

### Split by Sample and Cell Type

```{r violin_split, fig.width = 16, fig.height = 9, eval = has_cytotrace2}
if ("orig.ident" %in% colnames(um_atlas@meta.data)) {

  # Median statistic function
  median.stat <- function(x) {
    out <- quantile(x, probs = c(0.5))
    names(out) <- c("ymed")
    return(out)
  }

  p_vln_split <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    split.by = "orig.ident",  # Use orig.ident to merge biological replicates
    group.by = celltype_col,
    cols = celltype_colors,
    sort = TRUE,
    pt.size = 0
  ) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    labs(x = "Cell Type", y = "Potency Score") +
    ggtitle("Developmental Potential by Cell Type and Sample") +
    theme(
      axis.text = element_text(size = 8),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      aspect.ratio = 0.8,
      axis.ticks.length.y.right = unit(0.3, "cm")
    ) +
    RestoreLegend() +
    stat_summary(
      fun.y = median.stat,
      geom = "point",
      size = 3,
      colour = "white",
      position = position_dodge(0.9),
      key_glyph = "blank"
    )

  print(p_vln_split)
  ggsave(file.path(output_dir, "violin_cytotrace2_split.png"), p_vln_split,
         width = 16, height = 9, dpi = 300, bg = "white")

} else {
  cat("No 'Sample' column found in metadata. Skipping split violin plot.\n")
}
```

## Metadata Comparisons {.tabset}

### By Origin (Healthy, Primary, Metastasis)

```{r violin_origin, fig.width = 12, fig.height = 8, eval = has_cytotrace2}
if ("origin" %in% colnames(um_atlas@meta.data)) {

  cat("Origin distribution:\n")
  print(table(um_atlas$origin))

  # Enforce disease progression order: healthy → primary → metastasis
  um_atlas$origin <- factor(um_atlas$origin, levels = c("healthy", "primary", "metastasis"))

  # Violin plot with medians (using Victoria's origin_colors)
  p_origin <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    group.by = "origin",
    pt.size = 0,
    cols = origin_colors
  ) +
    scale_x_discrete(limits = origin_order) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", linewidth = 0.8) +
    labs(title = "CytoTRACE2 Score by Sample Origin (Healthy → Primary → Metastasis)",
         x = "Origin", y = "Potency Score") +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      axis.ticks.length.y.right = unit(0.3, "cm")
    )

  print(p_origin)
  ggsave(file.path(output_dir, "violin_cytotrace2_origin.png"), p_origin,
         width = 12, height = 8, dpi = 300, bg = "white")

  # Statistical summary
  origin_stats <- um_atlas@meta.data |>
    dplyr::group_by(origin) |>
    dplyr::summarise(
      n_cells = n(),
      mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
      sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    )

  cat("\nCytoTRACE2 statistics by origin:\n")
  print(origin_stats)

} else {
  cat("No 'origin' column found in metadata.\n")
}
```

### By Primary Mutation

```{r violin_mutation, fig.width = 14, fig.height = 8, eval = has_cytotrace2}
if ("primary_mutation" %in% colnames(um_atlas@meta.data)) {

  cat("Primary mutation distribution:\n")
  print(table(um_atlas$primary_mutation))

  # Violin plot
  p_mutation <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    group.by = "primary_mutation",
    pt.size = 0
  ) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", linewidth = 0.8) +
    labs(title = "CytoTRACE2 Score by Primary Mutation",
         x = "Primary Mutation", y = "Potency Score") +
    theme(
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      axis.ticks.length.y.right = unit(0.3, "cm")
    )

  print(p_mutation)
  ggsave(file.path(output_dir, "violin_cytotrace2_mutation.png"), p_mutation,
         width = 14, height = 8, dpi = 300, bg = "white")

  # Statistical summary
  mutation_stats <- um_atlas@meta.data |>
    dplyr::group_by(primary_mutation) |>
    dplyr::summarise(
      n_cells = n(),
      mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
      sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(n_cells))

  cat("\nCytoTRACE2 statistics by primary mutation:\n")
  print(mutation_stats)

} else {
  cat("No 'primary_mutation' column found in metadata.\n")
}
```

### By Organ

```{r violin_organ, fig.width = 12, fig.height = 8, eval = has_cytotrace2}
if ("organ" %in% colnames(um_atlas@meta.data)) {

  cat("Organ distribution:\n")
  print(table(um_atlas$organ))

  # Violin plot
  p_organ <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    group.by = "organ",
    pt.size = 0,
    cols = c("eye" = "#377EB8", "liver" = "#E41A1C", "brain" = "#984EA3",
             "skin" = "#FF7F00", "thyroid" = "#4DAF4A")
  ) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", linewidth = 0.8) +
    labs(title = "CytoTRACE2 Score by Organ",
         x = "Organ", y = "Potency Score") +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      axis.ticks.length.y.right = unit(0.3, "cm")
    )

  print(p_organ)
  ggsave(file.path(output_dir, "violin_cytotrace2_organ.png"), p_organ,
         width = 12, height = 8, dpi = 300, bg = "white")

  # Statistical summary
  organ_stats <- um_atlas@meta.data |>
    dplyr::group_by(organ) |>
    dplyr::summarise(
      n_cells = n(),
      mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
      sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(n_cells))

  cat("\nCytoTRACE2 statistics by organ:\n")
  print(organ_stats)

} else {
  cat("No 'organ' column found in metadata.\n")
}
```

### By Location (pUM Focus)

```{r violin_location, fig.width = 16, fig.height = 8, eval = has_cytotrace2}
if ("location" %in% colnames(um_atlas@meta.data)) {

  cat("Location distribution:\n")
  print(table(um_atlas$location))

  # Violin plot for all locations
  p_location <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    group.by = "location",
    pt.size = 0
  ) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", linewidth = 0.8) +
    labs(title = "CytoTRACE2 Score by Anatomical Location",
         x = "Location", y = "Potency Score") +
    theme(
      axis.text = element_text(size = 9),
      axis.title = element_text(size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      axis.ticks.length.y.right = unit(0.3, "cm")
    )

  print(p_location)
  ggsave(file.path(output_dir, "violin_cytotrace2_location.png"), p_location,
         width = 16, height = 8, dpi = 300, bg = "white")

  # Statistical summary
  location_stats <- um_atlas@meta.data |>
    dplyr::group_by(location) |>
    dplyr::summarise(
      n_cells = n(),
      mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
      sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(n_cells))

  cat("\nCytoTRACE2 statistics by location:\n")
  print(location_stats)

  # Focus on pUM locations (Choroid, Iris, Ciliary Body)
  pum_locations <- c("choroid", "iris", "ciliary body")
  if (any(tolower(um_atlas$location) %in% pum_locations)) {

    um_atlas_pum <- subset(um_atlas, subset = tolower(location) %in% pum_locations)

    cat("\n\nFocus on pUM locations (Choroid, Iris, Ciliary Body):\n")
    cat("Number of cells:", ncol(um_atlas_pum), "\n")
    print(table(um_atlas_pum$location))

    p_pum_location <- VlnPlot(
      um_atlas_pum,
      features = "CytoTRACE2_Score",
      group.by = "location",
      pt.size = 0,
      cols = c("choroid" = "#E41A1C", "iris" = "#377EB8", "ciliary body" = "#4DAF4A")
    ) +
      scale_y_continuous(
        breaks = seq(0, 1, by = 0.2),
        limits = c(0, 1),
        sec.axis = sec_axis(
          trans = ~.,
          breaks = seq(0, 1, by = 1/12),
          labels = c("", "Differentiated", "", "Unipotent", "",
                     "Oligopotent", "", "Multipotent", "",
                     "Pluripotent", "", "Totipotent", "")
        )
      ) +
      stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", linewidth = 0.8) +
      labs(title = "CytoTRACE2 Score in pUM Locations (Choroid, Iris, Ciliary Body)",
           x = "pUM Location", y = "Potency Score") +
      theme(
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5, margin = margin(b = 20)),
        axis.ticks.y.right = element_line(
          color = c("black", NA, "black", NA, "black", NA,
                    "black", NA, "black", NA, "black", NA, "black")
        ),
        axis.ticks.length.y.right = unit(0.3, "cm")
      )

    print(p_pum_location)
    ggsave(file.path(output_dir, "violin_cytotrace2_pum_locations.png"), p_pum_location,
           width = 10, height = 8, dpi = 300, bg = "white")

    pum_location_stats <- um_atlas_pum@meta.data |>
      dplyr::group_by(location) |>
      dplyr::summarise(
        n_cells = n(),
        mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
        median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
        sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
        .groups = "drop"
      )

    cat("\nCytoTRACE2 statistics for pUM locations:\n")
    print(pum_location_stats)
  }

} else {
  cat("No 'location' column found in metadata.\n")
}
```

### Sample-Level Dispersion

```{r boxplot_patients, fig.width = 18, fig.height = 10, eval = has_cytotrace2}
# Use orig.ident to merge biological replicates from same tumor
if ("orig.ident" %in% colnames(um_atlas@meta.data)) {

  cat("Sample distribution (biological replicates merged):\n")
  sample_counts <- table(um_atlas$orig.ident)
  print(sample_counts)
  cat("\nTotal samples:", length(sample_counts), "\n")

  # Compute sample-level summary statistics
  sample_stats <- um_atlas@meta.data |>
    dplyr::group_by(orig.ident) |>
    dplyr::summarise(
      n_cells = n(),
      mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
      sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
      origin = first(origin),
      organ = first(organ),
      primary_mutation = first(primary_mutation),
      .groups = "drop"
    ) |>
    dplyr::rename(Sample = orig.ident) |>
    dplyr::arrange(origin, organ)

  # Boxplot by sample (ordered by origin)
  p_sample_box <- ggplot(sample_stats, aes(x = reorder(Sample, mean_potency), y = mean_potency, fill = origin)) +
    geom_boxplot(aes(group = origin), alpha = 0.3, outlier.shape = NA) +
    geom_point(aes(color = origin, size = n_cells), alpha = 0.8) +
    scale_fill_manual(values = origin_colors) +
    scale_color_manual(values = origin_colors) +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    labs(
      title = "Sample-Level CytoTRACE2 Mean Potency Score Distribution",
      subtitle = "Points colored by origin, sized by cell count",
      x = "Sample",
      y = "Mean Potency Score"
    ) +
    theme_cowplot() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5),
      legend.position = "right"
    )

  print(p_sample_box)
  ggsave(file.path(output_dir, "boxplot_cytotrace2_samples.png"), p_sample_box,
         width = 18, height = 10, dpi = 300, bg = "white")

  # Save sample-level statistics
  write.csv(sample_stats, file.path(output_dir, "cytotrace2_stats_by_sample.csv"), row.names = FALSE)
  cat("\nSample-level statistics saved to:", file.path(output_dir, "cytotrace2_stats_by_sample.csv"), "\n")

  # Summary by origin
  origin_sample_summary <- sample_stats |>
    dplyr::group_by(origin) |>
    dplyr::summarise(
      n_samples = n(),
      mean_of_means = mean(mean_potency, na.rm = TRUE),
      sd_of_means = sd(mean_potency, na.rm = TRUE),
      median_of_medians = median(median_potency, na.rm = TRUE),
      .groups = "drop"
    )

  cat("\nSample-level summary by origin:\n")
  print(origin_sample_summary)

  # Violin plot by sample (showing dispersion within each sample)
  p_sample_violin <- VlnPlot(
    um_atlas,
    features = "CytoTRACE2_Score",
    group.by = "orig.ident",
    cols = sample_col,
    pt.size = 0
  ) +
    labs(title = "CytoTRACE2 Score Distribution by Sample (Within-Sample Dispersion)",
         x = "Sample", y = "Potency Score") +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )

  print(p_sample_violin)
  ggsave(file.path(output_dir, "violin_cytotrace2_samples_dispersion.png"), p_sample_violin,
         width = 20, height = 8, dpi = 300, bg = "white")

} else {
  cat("No 'orig.ident' column found in metadata.\n")
}
```

### Cell Type Statistics

```{r celltype_stats, fig.width = 12, fig.height = 6, eval = has_cytotrace2}
# Using majority_celltype as the official cell type annotation
cat("Cell type distribution (majority_celltype):\n")
print(table(um_atlas$majority_celltype))

# Statistical summary by cell type
majority_stats <- um_atlas@meta.data |>
  dplyr::group_by(majority_celltype) |>
  dplyr::summarise(
    n_cells = n(),
    mean_potency = mean(CytoTRACE2_Score, na.rm = TRUE),
    median_potency = median(CytoTRACE2_Score, na.rm = TRUE),
    sd_potency = sd(CytoTRACE2_Score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(desc(median_potency))

cat("\nCytoTRACE2 statistics by cell type (sorted by median potency):\n")
print(majority_stats)

# Save stats
write.csv(majority_stats, file.path(output_dir, "cytotrace2_stats_by_celltype.csv"), row.names = FALSE)
```

## Compositional Analysis {.tabset}

### Potency by Cell Type

```{r barplot_celltype, fig.width = 12, fig.height = 7, eval = has_cytotrace2}
p_bar_celltype <- dittoBarPlot(
  um_atlas,
  var = "CytoTRACE2_Potency",
  group.by = celltype_col,
  color.panel = colors.cyto
) +
  ggtitle("CytoTRACE2 Potency Composition by Cell Type") +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p_bar_celltype)
ggsave(file.path(output_dir, "barplot_potency_celltype.png"), p_bar_celltype,
       width = 12, height = 7, dpi = 300, bg = "white")
```

### Potency by Cell Type (Counts)

```{r barplot_celltype_counts, fig.width = 12, fig.height = 7, eval = has_cytotrace2}
p_bar_counts <- dittoBarPlot(
  um_atlas,
  var = "CytoTRACE2_Potency",
  group.by = celltype_col,
  color.panel = colors.cyto,
  scale = "count"
) +
  ggtitle("CytoTRACE2 Potency Counts by Cell Type") +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p_bar_counts)
ggsave(file.path(output_dir, "barplot_potency_celltype_counts.png"), p_bar_counts,
       width = 12, height = 7, dpi = 300, bg = "white")
```

### Potency by Sample (if available)

```{r barplot_sample, fig.width = 12, fig.height = 7, eval = has_cytotrace2}
# Use orig.ident to merge biological replicates from same tumor
if ("orig.ident" %in% colnames(um_atlas@meta.data)) {

  p_bar_sample <- dittoBarPlot(
    um_atlas,
    var = "CytoTRACE2_Potency",
    group.by = celltype_col,
    split.by = "orig.ident",
    color.panel = colors.cyto
  ) +
    ggtitle("CytoTRACE2 Potency by Cell Type and Sample") +
    theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

  print(p_bar_sample)
  ggsave(file.path(output_dir, "barplot_potency_sample.png"), p_bar_sample,
         width = 12, height = 7, dpi = 300, bg = "white")

} else {
  cat("No 'Sample' column found in metadata. Skipping sample-level bar plot.\n")
}
```

## Gene Correlation Analysis {.tabset}

### Top Stemness Markers

Genes positively correlated with high CytoTRACE2 scores (stemness/undifferentiated state).

```{r correlation_stemness, fig.width = 11, fig.height = 8, eval = has_cytotrace2}
# Extract CytoTRACE2 scores
cytotrace2_scores <- as.numeric(um_atlas@meta.data$CytoTRACE2_Score)

# Extract RNA assay data (Seurat v5 syntax)
expr_matrix <- GetAssayData(um_atlas, assay = "RNA", layer = "data")
expr_matrix <- as.matrix(expr_matrix)

# Calculate gene-wise correlations
cat("Calculating correlations for", nrow(expr_matrix), "genes...\n")
correlations <- apply(expr_matrix, 1, function(x) {
  cor(cytotrace2_scores, x, use = "complete.obs")
})

# Create correlation dataframe
cor_df <- data.frame(
  gene = names(correlations),
  correlation = correlations,
  stringsAsFactors = FALSE
) |>
  dplyr::filter(!is.na(correlation)) |>
  dplyr::arrange(desc(abs(correlation)))

# Plot top positively correlated genes (stemness markers)
p_stemness <- cor_df |>
  dplyr::top_n(25, wt = correlation) |>
  dplyr::arrange(desc(correlation)) |>
  ggplot(aes(x = reorder(gene, correlation), y = correlation)) +
  geom_col(fill = "deepskyblue") +
  theme_cowplot() +
  ggtitle("Top 25 Genes Correlated with Stemness (High CytoTRACE2)") +
  ylab("Correlation with CytoTRACE2 Score") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

print(p_stemness)
ggsave(file.path(output_dir, "genes_correlated_stemness.png"), p_stemness,
       width = 11, height = 8, dpi = 300, bg = "white")
```

### Top Differentiation Markers

Genes negatively correlated with CytoTRACE2 scores (differentiation markers).

```{r correlation_differentiation, fig.width = 11, fig.height = 8, eval = has_cytotrace2}
# Plot top negatively correlated genes (differentiation markers)
p_diff <- cor_df |>
  dplyr::top_n(25, wt = -correlation) |>
  dplyr::arrange(correlation) |>
  ggplot(aes(x = reorder(gene, correlation), y = correlation)) +
  geom_col(fill = "salmon") +
  theme_cowplot() +
  ggtitle("Top 25 Genes Correlated with Differentiation (Low CytoTRACE2)") +
  ylab("Correlation with CytoTRACE2 Score") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

print(p_diff)
ggsave(file.path(output_dir, "genes_correlated_differentiation.png"), p_diff,
       width = 11, height = 8, dpi = 300, bg = "white")
```

### Combined Plot

```{r correlation_combined, fig.width = 11, fig.height = 15, eval = has_cytotrace2}
p_combined <- p_stemness / p_diff
print(p_combined)
ggsave(file.path(output_dir, "genes_correlated_combined.png"), p_combined,
       width = 11, height = 15, dpi = 300, bg = "white")
```

### Export Correlation Table

```{r export_correlations, eval = has_cytotrace2}
# Export full correlation table
cor_output <- file.path(output_dir, "cytotrace2_gene_correlations.csv")
write.csv(cor_df, cor_output, row.names = FALSE)
cat("Saved gene correlation table to:", cor_output, "\n")

# Display top 10 in each direction
cat("\nTop 10 stemness markers:\n")
print(cor_df |> dplyr::top_n(10, wt = correlation) |> dplyr::arrange(desc(correlation)))

cat("\nTop 10 differentiation markers:\n")
print(cor_df |> dplyr::top_n(10, wt = -correlation) |> dplyr::arrange(correlation))
```

## Malignant Cell Analysis (if applicable) {.tabset}

### Identify Malignant Population

```{r identify_malignant, eval = has_cytotrace2}
# Identify melanocytic cells: healthy melanocytes + malignant melanoma cells
has_malignant <- FALSE
malignant_subset <- NULL

if ("majority_celltype" %in% colnames(um_atlas@meta.data)) {
  cat("Checking for Melanocytes and Melanoma cells in majority_celltype...\n")
  cat("\nmajority_celltype distribution:\n")
  print(table(um_atlas@meta.data$majority_celltype))

  # Check if both "Melanocytes" and "Melanoma cells" categories exist
  melanocytic_celltypes <- c("Melanocytes", "Melanoma cells")
  available_types <- intersect(melanocytic_celltypes, unique(um_atlas@meta.data$majority_celltype))

  if (length(available_types) > 0) {
    malignant_subset <- subset(um_atlas, subset = majority_celltype %in% available_types)

    # Remove 2 erroneous melanocytes from mUM15 metastasis sample
    # These cells lack CNV data and were misclassified (should not be in metastasis)
    erroneous_cells <- c("mUM15_TAAGAGACAGCTCCGA-1_48", "mUM15_GTCACGGGTGTGTGCC-1_48")
    cells_before <- ncol(malignant_subset)
    malignant_subset <- malignant_subset[, !colnames(malignant_subset) %in% erroneous_cells]
    cells_after <- ncol(malignant_subset)

    has_malignant <- TRUE
    cat("\n✅ Identified melanocytic cells:\n")
    print(table(malignant_subset$majority_celltype))
    cat("Total cells:", cells_after, "(removed", cells_before - cells_after, "erroneous cells from mUM15)\n")

    cat("\nBy origin:\n")
    origin_table <- table(malignant_subset$origin, malignant_subset$majority_celltype)
    print(origin_table)
    cat("\nTotal by origin:\n")
    print(rowSums(origin_table))
  } else {
    cat("\n⚠️ Neither 'Melanocytes' nor 'Melanoma cells' categories found in majority_celltype.\n")
    cat("Available categories:", paste(unique(um_atlas@meta.data$majority_celltype), collapse = ", "), "\n")
  }
} else {
  cat("\n❌ majority_celltype column not found in metadata.\n")
}

if (!has_malignant) {
  cat("\nNo melanocytic cell population identified. Skipping melanocytic-specific analysis.\n")
}
```

### Malignant Cell UMAP

```{r malignant_umap, fig.width = 10, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  p_mal_umap <- FeaturePlot(
    malignant_subset,
    reduction = "umap",
    features = "CytoTRACE2_Score",
    raster = FALSE
  ) +
    scale_colour_gradientn(
      colours = c("#000004FF", "#3B0F70FF", "#8C2981FF",
                  "#DE4968FF", "#FE9F6DFF", "#FCFDBFFF"),
      na.value = "transparent",
      limits = c(0, 1),
      breaks = seq(0, 1, by = 0.2),
      labels = c("0.0 (More diff.)", "0.2", "0.4", "0.6", "0.8", "1.0 (Less diff.)"),
      name = "CytoTRACE2\nScore",
      guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
    ) +
    coord_fixed() +
    ggtitle("CytoTRACE 2 Score - Malignant Cells") +
    xlab("UMAP1") + ylab("UMAP2")

  print(p_mal_umap)
  ggsave(file.path(output_dir, "umap_cytotrace2_malignant.png"), p_mal_umap,
         width = 10, height = 8, dpi = 300, bg = "white")
}
```

### Malignant Gene Correlations

```{r malignant_correlations, fig.width = 11, fig.height = 15, eval = has_malignant}
if (has_malignant) {

  # Function for correlation analysis
  correlation_plot <- function(data, title_suffix) {
    expr_matrix <- GetAssayData(data, assay = "RNA", layer = "data")
    expr_matrix <- as.matrix(expr_matrix)

    score <- as.numeric(data@meta.data$CytoTRACE2_Score)

    correlations <- apply(expr_matrix, 1, function(x) {
      cor(score, x, use = "complete.obs")
    })

    cor_df <- data.frame(
      gene = names(correlations),
      correlation = correlations,
      stringsAsFactors = FALSE
    ) |>
      dplyr::filter(!is.na(correlation))

    p1 <- cor_df |>
      dplyr::top_n(20, wt = correlation) |>
      dplyr::arrange(desc(correlation)) |>
      ggplot(aes(x = reorder(gene, correlation), y = correlation)) +
      geom_col(fill = "deepskyblue") +
      theme_cowplot() +
      ggtitle(paste("Genes Correlated with Stemness in", title_suffix)) +
      ylab("Correlation") +
      xlab("") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    p2 <- cor_df |>
      dplyr::top_n(20, wt = -correlation) |>
      dplyr::arrange(correlation) |>
      ggplot(aes(x = reorder(gene, correlation), y = correlation)) +
      geom_col(fill = "salmon") +
      theme_cowplot() +
      ggtitle(paste("Genes Anticorrelated with Stemness in", title_suffix)) +
      ylab("Correlation") +
      xlab("") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))

    list(combined = p1 / p2, cor_df = cor_df)
  }

  # Run analysis
  mal_results <- correlation_plot(malignant_subset, "Malignant Cells")

  print(mal_results$combined)
  ggsave(file.path(output_dir, "genes_correlated_malignant.png"),
         mal_results$combined,
         width = 11, height = 15, dpi = 300, bg = "white")

  # Export malignant correlation table
  write.csv(
    mal_results$cor_df |> dplyr::arrange(desc(abs(correlation))),
    file.path(output_dir, "cytotrace2_malignant_gene_correlations.csv"),
    row.names = FALSE
  )

  cat("\nTop malignant stemness markers:\n")
  print(mal_results$cor_df |> dplyr::top_n(10, wt = correlation) |>
          dplyr::arrange(desc(correlation)))
}
```

### Healthy Melanocytes Only - Gene Correlations

```{r healthy_melanocytes_correlation, fig.width = 11, fig.height = 15, eval = has_malignant}
if (has_malignant) {
  # Subset to healthy melanocytes only (origin == "healthy")
  healthy_melanocytes <- subset(malignant_subset, subset = origin == "healthy")

  cat("\n=== Healthy Melanocytes Gene Correlation Analysis ===\n")
  cat("Total healthy melanocytes:", ncol(healthy_melanocytes), "\n")
  cat("This analysis reveals normal melanocyte differentiation programs\n")
  cat("without malignant transformation.\n\n")

  # Run correlation analysis using the existing function
  healthy_results <- correlation_plot(healthy_melanocytes, "Healthy Melanocytes")

  print(healthy_results$combined)
  ggsave(file.path(output_dir, "genes_correlated_healthy_melanocytes.png"),
         healthy_results$combined,
         width = 11, height = 15, dpi = 300, bg = "white")

  # Export healthy melanocytes correlation table
  write.csv(
    healthy_results$cor_df |> dplyr::arrange(desc(abs(correlation))),
    file.path(output_dir, "cytotrace2_healthy_melanocytes_gene_correlations.csv"),
    row.names = FALSE
  )

  cat("\nTop healthy melanocyte stemness markers (positive correlations):\n")
  print(healthy_results$cor_df |> dplyr::top_n(10, wt = correlation) |>
          dplyr::arrange(desc(correlation)))

  cat("\nTop healthy melanocyte differentiation markers (negative correlations):\n")
  print(healthy_results$cor_df |> dplyr::top_n(10, wt = -correlation) |>
          dplyr::arrange(correlation))
}
```

## Melanocytic Cell Focus - UM Atlas Key Comparisons {.tabset}

### Overview

This section focuses on **melanocytic cells** (healthy melanocytes + malignant melanoma cells) to address the key biological questions of the UM atlas:

1. **Disease progression** across healthy melanocytes → primary UM → metastatic UM (KEY COMPARISON)
2. **Secondary mutations** in pUM tumors (BAP1, SF3B1, EIF1AX)
3. **Anatomical locations** in pUM cohort (Choroid, Iris, Ciliary Body)

```{r malignant_focus_overview, eval = has_malignant}
if (has_malignant) {
  cat("Melanocytic Cell Population Statistics:\n")
  cat("========================================\n\n")
  cat("Total melanocytic cells:", ncol(malignant_subset), "\n\n")

  # Cell type distribution
  cat("Cell type distribution:\n")
  print(table(malignant_subset$majority_celltype))

  # Secondary mutation distribution
  cat("\n\nSecondary mutation distribution:\n")
  print(table(malignant_subset$secondary_mutation))

  cat("\n\nLocation distribution:\n")
  print(table(malignant_subset$location))

  cat("\n\nOrigin distribution:\n")
  origin_celltype_table <- table(malignant_subset$origin, malignant_subset$majority_celltype)
  print(origin_celltype_table)
  cat("\nTotal by origin:\n")
  print(rowSums(origin_celltype_table))
}
```

### Secondary Mutations (BAP1, SF3B1, EIF1AX)

```{r malignant_secondary_mutations, fig.width = 12, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter to EIF1AX, SF3B1, BAP1 only (remove unknown and none)
  mal_known_mut <- malignant_subset@meta.data |>
    dplyr::filter(secondary_mutation %in% c("EIF1AX", "SF3B1", "BAP1")) |>
    dplyr::mutate(secondary_mutation = factor(secondary_mutation,
                                               levels = c("EIF1AX", "SF3B1", "BAP1")))

  # Boxplot with statistical comparisons
  p_mut <- ggplot(mal_known_mut,
                   aes(x = secondary_mutation, y = CytoTRACE2_Score, fill = secondary_mutation)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
    scale_fill_manual(values = second_mutations_col) +
    stat_compare_means(comparisons = list(c("EIF1AX", "SF3B1"), c("SF3B1", "BAP1"), c("EIF1AX", "BAP1")),
                       method = "wilcox.test", label = "p.signif") +
    theme_cowplot() +
    ggtitle("CytoTRACE2 Scores by Secondary Mutation in Malignant Cells") +
    xlab("Secondary Mutation") +
    ylab("CytoTRACE2 Score") +
    theme(legend.position = "none",
          axis.text.x = element_text(size = 12, angle = 0))

  print(p_mut)
  ggsave(file.path(output_dir, "malignant_cytotrace2_secondary_mutations.png"), p_mut,
         width = 12, height = 8, dpi = 300, bg = "white")

  # Summary statistics
  mut_stats <- mal_known_mut |>
    dplyr::group_by(secondary_mutation) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(mean_score))

  cat("\nCytoTRACE2 Statistics by Secondary Mutation:\n")
  print(mut_stats)

  write.csv(mut_stats,
            file.path(output_dir, "malignant_cytotrace2_stats_secondary_mutations.csv"),
            row.names = FALSE)
}
```

### pUM Anatomical Locations

```{r malignant_pum_locations, fig.width = 12, fig.height = 10, eval = has_malignant}
if (has_malignant) {
  library(ggpubr)

  # Focus on primary UM locations: choroid, iris, ciliary body
  pum_locations <- c("choroid", "ciliary body", "iris")
  mal_pum <- subset(malignant_subset, subset = location %in% pum_locations)

  if (ncol(mal_pum) > 0) {
    # Statistical tests
    cat("=== Statistical Analysis: CytoTRACE2 Scores by Intraocular Location ===\n\n")

    # Kruskal-Wallis test (non-parametric, 3+ groups)
    kw_test <- kruskal.test(CytoTRACE2_Score ~ location, data = mal_pum@meta.data)
    cat("Kruskal-Wallis test:\n")
    cat("  Chi-squared =", round(kw_test$statistic, 2), "\n")
    cat("  df =", kw_test$parameter, "\n")
    cat("  p-value =", format(kw_test$p.value, scientific = TRUE, digits = 3), "\n\n")

    # Pairwise Wilcoxon with Bonferroni correction
    pairwise_tests <- pairwise.wilcox.test(
      mal_pum@meta.data$CytoTRACE2_Score,
      mal_pum@meta.data$location,
      p.adjust.method = "bonferroni"
    )
    cat("Pairwise Wilcoxon tests (Bonferroni corrected):\n")
    print(pairwise_tests)
    cat("\n")

    # Violin plot with statistical annotations (Victoria's location colors)
    comparisons_list <- list(c("choroid", "iris"), c("ciliary body", "iris"), c("choroid", "ciliary body"))

    p_loc <- ggplot(mal_pum@meta.data,
                     aes(x = location, y = CytoTRACE2_Score, fill = location)) +
      geom_violin(alpha = 0.7, scale = "width") +
      geom_boxplot(width = 0.2, outlier.size = 0.5, alpha = 0.5) +
      scale_fill_manual(values = primary_location_colors) +
      stat_compare_means(comparisons = comparisons_list,
                         method = "wilcox.test",
                         label = "p.signif",
                         hide.ns = FALSE) +
      stat_compare_means(method = "kruskal.test",
                         label.y = 1.15,
                         label = "p.format") +
      theme_cowplot() +
      ggtitle("CytoTRACE2 Scores by Anatomical Location in pUM Malignant Cells",
              subtitle = "Statistical comparison: Kruskal-Wallis + pairwise Wilcoxon (Bonferroni)") +
      xlab("Anatomical Location") +
      ylab("CytoTRACE2 Score") +
      theme(legend.position = "none",
            axis.text.x = element_text(size = 12),
            plot.subtitle = element_text(size = 10))

    print(p_loc)
    ggsave(file.path(output_dir, "malignant_cytotrace2_pum_locations.png"), p_loc,
           width = 12, height = 10, dpi = 300, bg = "white")

    # Summary statistics
    loc_stats <- mal_pum@meta.data |>
      dplyr::group_by(location) |>
      dplyr::summarise(
        n_cells = n(),
        mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
        sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
        median_score = median(CytoTRACE2_Score, na.rm = TRUE),
        .groups = "drop"
      ) |>
      dplyr::arrange(desc(mean_score))

    cat("\nCytoTRACE2 Statistics by pUM Location:\n")
    cat("Cells analyzed:", ncol(mal_pum), "\n\n")
    print(loc_stats)

    write.csv(loc_stats,
              file.path(output_dir, "malignant_cytotrace2_stats_pum_locations.csv"),
              row.names = FALSE)
  } else {
    cat("No malignant cells found in pUM locations (choroid, iris, ciliary body)\n")
  }
}
```

### ★ KEY: Disease Progression (Healthy → Primary → Metastatic)

```{r malignant_origin, fig.width = 10, fig.height = 12, eval = has_malignant}
if (has_malignant) {
  library(ggbeeswarm)

  # Enforce disease progression order
  malignant_subset$origin <- factor(malignant_subset$origin,
                                     levels = c("healthy", "primary", "metastasis"))

  # 1. Violin plot (all cells merged) - standard approach (Victoria's colors)
  p_violin_all <- ggplot(malignant_subset@meta.data,
                          aes(x = origin, y = CytoTRACE2_Score, fill = origin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.2, outlier.size = 0.5, alpha = 0.5) +
    scale_x_discrete(limits = origin_order) +
    scale_fill_manual(values = origin_colors) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1),
      sec.axis = sec_axis(
        trans = ~.,
        breaks = seq(0, 1, by = 1/12),
        labels = c("", "Differentiated", "", "Unipotent", "",
                   "Oligopotent", "", "Multipotent", "",
                   "Pluripotent", "", "Totipotent", "")
      )
    ) +
    stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "black", linewidth = 1) +
    theme_cowplot() +
    ggtitle("CytoTRACE2 Scores by Origin (All Cells Merged)",
            subtitle = "Healthy melanocytes → Primary UM → Metastatic UM") +
    xlab("Disease Stage") +
    ylab("CytoTRACE2 Score") +
    theme(
      legend.position = "none",
      axis.text.x = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5),
      axis.ticks.y.right = element_line(
        color = c("black", NA, "black", NA, "black", NA,
                  "black", NA, "black", NA, "black", NA, "black")
      ),
      axis.ticks.length.y.right = unit(0.3, "cm")
    )

  # 2. Split violin plot by cell type (Melanocytes vs Melanoma cells)
  p_violin_split <- ggplot(malignant_subset@meta.data,
                            aes(x = origin, y = CytoTRACE2_Score, fill = majority_celltype)) +
    geom_violin(alpha = 0.7, scale = "width", position = position_dodge(0.9)) +
    geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.size = 0.3, alpha = 0.5) +
    scale_x_discrete(limits = origin_order) +
    scale_fill_manual(
      values = c("Melanocytes" = "#E6F5E6", "Melanoma cells" = "#8DA0CB"),
      name = "Cell Type"
    ) +
    scale_y_continuous(
      breaks = seq(0, 1, by = 0.2),
      limits = c(0, 1)
    ) +
    theme_cowplot() +
    ggtitle("CytoTRACE2 Scores by Origin (Split by Cell Type)",
            subtitle = "Comparing healthy melanocytes vs malignant melanoma cells") +
    xlab("Disease Stage") +
    ylab("CytoTRACE2 Score") +
    theme(
      legend.position = "right",
      axis.text.x = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5)
    )

  # 3. Comprehensive sample-level statistics with all metadata (replicates merged)
  sample_stats_full <- malignant_subset@meta.data |>
    dplyr::group_by(orig.ident, origin, majority_celltype, organ, location,
                    primary_mutation, secondary_mutation, Age_atDiagnosisY,
                    Gender, LocationPrimaryTumor, Patient_nr) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::filter(n_cells >= 10)  # Filter patients with at least 10 melanocytic cells

  # Create age bins for visualization (convert to numeric first)
  sample_stats_full <- sample_stats_full |>
    dplyr::mutate(
      Age_atDiagnosisY_num = as.numeric(as.character(Age_atDiagnosisY)),
      age_bin = cut(Age_atDiagnosisY_num,
                    breaks = c(0, 50, 60, 70, 100),
                    labels = c("<50", "50-60", "60-70", ">70"),
                    include.lowest = TRUE)
    )

  # Get polychrome palette for multi-category variables
  polychrome_colors <- unname(pals::polychrome())

  # Enforce disease progression order in sample_stats_full
  sample_stats_full$origin <- factor(sample_stats_full$origin,
                                      levels = c("healthy", "primary", "metastasis"))

  # Boxplot by Origin (using Victoria's colors)
  p_boxplot_origin <- ggplot(sample_stats_full,
                               aes(x = origin, y = mean_score, fill = origin)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_quasirandom(aes(color = origin, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_x_discrete(limits = origin_order) +
    scale_fill_manual(values = origin_colors, name = "Uveal Location") +
    scale_color_manual(values = origin_colors, name = "Uveal Location") +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
            subtitle = "Each point = one patient (sized by cell count, ≥10 cells)") +
    xlab("Disease Progression") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  # Additional boxplot variants colored by metadata variables

  # 1. Boxplot colored by Organ (using Victoria's organ_colors from config)
  p_boxplot_organ <- ggplot(sample_stats_full,
                             aes(x = origin, y = mean_score)) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
    geom_quasirandom(aes(color = organ, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_x_discrete(limits = origin_order) +
    scale_color_manual(values = organ_colors, name = "Organ") +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    stat_compare_means(aes(group = organ), method = "kruskal.test",
                       label = "p.format", label.y = max(sample_stats_full$mean_score, na.rm = TRUE) + 0.05) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
            subtitle = "Colored by Organ | Each point = one patient (sized by cell count, ≥10 cells)") +
    xlab("Disease Stage") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_organ)
  ggsave(file.path(output_dir, "patient_boxplot_by_organ.png"),
         p_boxplot_organ, width = 12, height = 8, dpi = 300, bg = "white")

  # 2. Boxplot colored by Primary Mutation
  n_mut <- length(unique(sample_stats_full$primary_mutation))
  mutation_colors <- setNames(polychrome_colors[1:n_mut], sort(unique(sample_stats_full$primary_mutation)))

  p_boxplot_mutation <- ggplot(sample_stats_full,
                                 aes(x = origin, y = mean_score)) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
    geom_quasirandom(aes(color = primary_mutation, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_x_discrete(limits = origin_order) +
    scale_color_manual(values = mutation_colors, name = "Primary Mutation") +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
            subtitle = "Colored by Primary Mutation | Each point = one patient (sized by cell count, ≥10 cells)") +
    xlab("Disease Stage") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_mutation)
  ggsave(file.path(output_dir, "patient_boxplot_by_primary_mutation.png"),
         p_boxplot_mutation, width = 12, height = 8, dpi = 300, bg = "white")

  # 3. Boxplot colored by Secondary Mutation
  sample_stats_full_2nd <- sample_stats_full |>
    dplyr::filter(!is.na(secondary_mutation) & secondary_mutation != "")

  if (nrow(sample_stats_full_2nd) > 0) {
    n_mut2 <- length(unique(sample_stats_full_2nd$secondary_mutation))
    mutation2_colors <- setNames(polychrome_colors[1:n_mut2], sort(unique(sample_stats_full_2nd$secondary_mutation)))

    p_boxplot_mutation2 <- ggplot(sample_stats_full_2nd,
                                   aes(x = origin, y = mean_score)) +
      geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
      geom_quasirandom(aes(color = secondary_mutation, size = n_cells),
                       alpha = 0.7, dodge.width = 0.8) +
      scale_x_discrete(limits = origin_order) +
      scale_color_manual(values = mutation2_colors, name = "Secondary Mutation") +
      scale_size_continuous(name = "Cell count", range = c(2, 8)) +
      theme_cowplot() +
      ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
              subtitle = "Colored by Secondary Mutation | Each point = one patient (sized by cell count, ≥10 cells)") +
      xlab("Disease Stage") +
      ylab("Mean CytoTRACE2 Score per Patient") +
      theme(legend.position = "right", axis.text.x = element_text(size = 12),
            plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
            plot.subtitle = element_text(size = 11, hjust = 0.5))

    print(p_boxplot_mutation2)
    ggsave(file.path(output_dir, "patient_boxplot_by_secondary_mutation.png"),
           p_boxplot_mutation2, width = 12, height = 8, dpi = 300, bg = "white")
  }

  # 4. Boxplot colored by Uveal Location
  patient_stats_uveal <- sample_stats_full |>
    dplyr::filter(!is.na(location) & location != "" & location != "NA")

  if (nrow(patient_stats_uveal) > 0) {
    n_loc <- length(unique(patient_stats_uveal$location))
    location_colors <- setNames(polychrome_colors[1:n_loc], sort(unique(patient_stats_uveal$location)))

    p_boxplot_location <- ggplot(patient_stats_uveal,
                                  aes(x = origin, y = mean_score)) +
      geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
      geom_quasirandom(aes(color = location, size = n_cells),
                       alpha = 0.7, dodge.width = 0.8) +
      scale_x_discrete(limits = origin_order) +
      scale_color_manual(values = location_colors, name = "Uveal Location") +
      scale_size_continuous(name = "Cell count", range = c(2, 8)) +
      theme_cowplot() +
      ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
              subtitle = "Colored by Uveal Location | Each point = one patient (sized by cell count, ≥10 cells)") +
      xlab("Disease Stage") +
      ylab("Mean CytoTRACE2 Score per Patient") +
      theme(legend.position = "right", axis.text.x = element_text(size = 12),
            plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
            plot.subtitle = element_text(size = 11, hjust = 0.5))

    print(p_boxplot_location)
    ggsave(file.path(output_dir, "patient_boxplot_by_uveal_location.png"),
           p_boxplot_location, width = 12, height = 8, dpi = 300, bg = "white")
  }

  # 5. Boxplot colored by Age Class (using Victoria's ageGroup_col from config)
  # Map age_bin categories to ageGroup_col names
  age_to_agegroup <- c("<50" = "Young", "50-60" = "Middle", "60-70" = "Middle", ">70" = "Old")
  sample_stats_full <- sample_stats_full |>
    dplyr::mutate(age_group = factor(age_to_agegroup[as.character(age_bin)],
                                      levels = c("Young", "Middle", "Old")))

  p_boxplot_age <- ggplot(sample_stats_full,
                          aes(x = origin, y = mean_score)) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
    geom_quasirandom(aes(color = age_group, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_x_discrete(limits = origin_order) +
    scale_color_manual(values = ageGroup_col, name = "Age Class", na.translate = FALSE) +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
            subtitle = "Colored by Age Class | Each point = one patient (sized by cell count, ≥10 cells)") +
    xlab("Disease Progression") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_age)
  ggsave(file.path(output_dir, "patient_boxplot_by_age.png"),
         p_boxplot_age, width = 12, height = 8, dpi = 300, bg = "white")

  # 6. Boxplot colored by Gender (using Victoria's gender_col from config)
  p_boxplot_gender <- ggplot(sample_stats_full,
                              aes(x = origin, y = mean_score)) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
    geom_quasirandom(aes(color = Gender, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_x_discrete(limits = origin_order) +
    scale_color_manual(values = gender_col, name = "Gender") +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
            subtitle = "Colored by Gender | Each point = one patient (sized by cell count, ≥10 cells)") +
    xlab("Disease Progression") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_gender)
  ggsave(file.path(output_dir, "patient_boxplot_by_gender.png"),
         p_boxplot_gender, width = 12, height = 8, dpi = 300, bg = "white")

  # 7. Boxplot colored by Patient
  n_patients <- length(unique(sample_stats_full$Patient_nr))
  patient_colors <- setNames(polychrome_colors[1:n_patients], sort(unique(sample_stats_full$Patient_nr)))

  p_boxplot_patient <- ggplot(sample_stats_full,
                               aes(x = origin, y = mean_score)) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, fill = "gray80") +
    geom_quasirandom(aes(color = Patient_nr, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_x_discrete(limits = origin_order) +
    scale_color_manual(values = patient_colors, name = "Patient") +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Disease Stage",
            subtitle = "Colored by Patient | Each point = one patient (sized by cell count, ≥10 cells)") +
    xlab("Disease Stage") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_patient)
  ggsave(file.path(output_dir, "patient_boxplot_by_patient.png"),
         p_boxplot_patient, width = 14, height = 8, dpi = 300, bg = "white")

  # Combine plots (violins + boxplot)
  # Use relative heights: violins get more space, boxplot is compact
  p_combined_origin <- (p_violin_all / p_violin_split / p_boxplot_origin) +
    plot_layout(heights = c(2, 2, 1))

  print(p_combined_origin)
  ggsave(file.path(output_dir, "melanocytic_cytotrace2_origin_comprehensive.png"),
         p_combined_origin,
         width = 10, height = 12, dpi = 300, bg = "white")

  # Cell-level summary statistics
  origin_stats <- malignant_subset@meta.data |>
    dplyr::group_by(origin) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(origin)

  cat("\n=== CELL-LEVEL Statistics by Origin ===\n")
  print(origin_stats)

  # Patient-level summary statistics
  patient_summary_origin <- sample_stats_full |>
    dplyr::group_by(origin) |>
    dplyr::summarise(
      n_patients = n(),
      mean_of_patient_means = mean(mean_score, na.rm = TRUE),
      sd_of_patient_means = sd(mean_score, na.rm = TRUE),
      median_of_patient_means = median(mean_score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(origin)

  cat("\n=== PATIENT-LEVEL Statistics by Origin ===\n")
  cat("(Patients with ≥10 melanocytic cells)\n\n")
  print(patient_summary_origin)

  # Cell type breakdown by origin
  celltype_origin_stats <- malignant_subset@meta.data |>
    dplyr::group_by(origin, majority_celltype) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(origin, majority_celltype)

  cat("\n=== By Cell Type and Origin ===\n")
  print(celltype_origin_stats)

  # Export statistics
  write.csv(origin_stats,
            file.path(output_dir, "melanocytic_cytotrace2_stats_origin_cell_level.csv"),
            row.names = FALSE)
  write.csv(patient_summary_origin,
            file.path(output_dir, "melanocytic_cytotrace2_stats_origin_patient_level.csv"),
            row.names = FALSE)
  write.csv(celltype_origin_stats,
            file.path(output_dir, "melanocytic_cytotrace2_stats_origin_by_celltype.csv"),
            row.names = FALSE)
  write.csv(sample_stats_full,
            file.path(output_dir, "melanocytic_cytotrace2_patient_data_full.csv"),
            row.names = FALSE)
}
```

### Patient-Level Boxplots - Multiple Perspectives {.tabset}

Explore patient-level CytoTRACE2 scores colored by different clinical and biological variables.

#### By Organ (Metastases)

```{r malignant_boxplot_organ, fig.width = 12, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter to metastases only and remove NA organs
  patient_stats_metastasis <- sample_stats_full |>
    dplyr::filter(origin == "metastasis" & !is.na(organ) & organ != "")

  # Get unique organs and assign colors
  unique_organs <- unique(patient_stats_metastasis$organ)
  n_organs <- length(unique_organs)
  organ_colors_fill <- polychrome_colors[1:n_organs]
  organ_colors_line <- adjustcolor(organ_colors_fill, red.f = 0.7, green.f = 0.7, blue.f = 0.7)
  names(organ_colors_fill) <- unique_organs
  names(organ_colors_line) <- unique_organs

  p_boxplot_organ <- ggplot(patient_stats_metastasis,
                             aes(x = organ, y = mean_score, fill = organ)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_quasirandom(aes(color = organ, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_fill_manual(values = organ_colors_fill) +
    scale_color_manual(values = organ_colors_line) +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Metastasis Organ",
            subtitle = "Metastasis samples only (≥10 cells per patient)") +
    xlab("Metastasis Organ") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_organ)
  ggsave(file.path(output_dir, "patient_boxplot_by_organ.png"),
         p_boxplot_organ, width = 12, height = 8, dpi = 300, bg = "white")
}
```

#### By Primary Mutation

```{r malignant_boxplot_mutation, fig.width = 12, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter out unknown mutations
  patient_stats_mutation <- sample_stats_full |>
    dplyr::filter(!is.na(primary_mutation) & primary_mutation != "" & primary_mutation != "unknown")

  # Get unique mutations and assign colors
  unique_mutations <- unique(patient_stats_mutation$primary_mutation)
  n_mutations <- length(unique_mutations)
  mutation_colors_fill <- polychrome_colors[1:n_mutations]
  mutation_colors_line <- adjustcolor(mutation_colors_fill, red.f = 0.7, green.f = 0.7, blue.f = 0.7)
  names(mutation_colors_fill) <- unique_mutations
  names(mutation_colors_line) <- unique_mutations

  p_boxplot_mutation <- ggplot(patient_stats_mutation,
                                aes(x = primary_mutation, y = mean_score, fill = primary_mutation)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_quasirandom(aes(color = primary_mutation, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_fill_manual(values = mutation_colors_fill) +
    scale_color_manual(values = mutation_colors_line) +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Primary Mutation",
            subtitle = "All disease stages (≥10 cells per patient)") +
    xlab("Primary Mutation") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_mutation)
  ggsave(file.path(output_dir, "patient_boxplot_by_mutation.png"),
         p_boxplot_mutation, width = 12, height = 8, dpi = 300, bg = "white")
}
```

#### By Uveal Location

```{r malignant_boxplot_uveal, fig.width = 12, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter to primary + healthy with known location
  patient_stats_uveal <- sample_stats_full |>
    dplyr::filter(origin %in% c("healthy", "primary") &
                  !is.na(LocationPrimaryTumor) &
                  LocationPrimaryTumor != "" &
                  LocationPrimaryTumor != "unknown")

  # Assign colors for iris, choroid, ciliary body
  uveal_colors <- c("iris" = "#E41A1C", "choroid" = "#377EB8", "ciliary body" = "#4DAF4A")
  uveal_colors_line <- c("iris" = "#B3151A", "choroid" = "#2D6594", "ciliary body" = "#3E8C3B")

  p_boxplot_uveal <- ggplot(patient_stats_uveal,
                             aes(x = LocationPrimaryTumor, y = mean_score, fill = LocationPrimaryTumor)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_quasirandom(aes(color = LocationPrimaryTumor, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_fill_manual(values = uveal_colors, na.value = "grey50") +
    scale_color_manual(values = uveal_colors_line, na.value = "grey30") +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Uveal Location",
            subtitle = "Healthy and primary tumors only (≥10 cells per patient)") +
    xlab("Primary Tumor Location") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 11),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_uveal)
  ggsave(file.path(output_dir, "patient_boxplot_by_uveal_location.png"),
         p_boxplot_uveal, width = 12, height = 8, dpi = 300, bg = "white")
}
```

#### By Patient

```{r malignant_boxplot_patient, fig.width = 16, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Sort patients by mean score for better visualization
  patient_order <- sample_stats_full |>
    dplyr::group_by(Patient_nr) |>
    dplyr::summarise(mean_score_avg = mean(mean_score, na.rm = TRUE), .groups = "drop") |>
    dplyr::arrange(mean_score_avg) |>
    dplyr::pull(Patient_nr)

  sample_stats_full$Patient_nr <- factor(sample_stats_full$Patient_nr, levels = patient_order)

  # Assign unique colors to each patient
  n_patients <- length(unique(sample_stats_full$Patient_nr))
  patient_colors <- rep(polychrome_colors, length.out = n_patients)
  names(patient_colors) <- levels(sample_stats_full$Patient_nr)

  p_boxplot_patient <- ggplot(sample_stats_full,
                               aes(x = Patient_nr, y = mean_score, fill = Patient_nr)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA, show.legend = FALSE) +
    geom_quasirandom(aes(size = n_cells), alpha = 0.7, dodge.width = 0.8, show.legend = TRUE) +
    scale_fill_manual(values = patient_colors) +
    scale_size_continuous(name = "Cell count", range = c(2, 6)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Patient",
            subtitle = "All patients sorted by mean score (≥10 cells per patient)") +
    xlab("Patient ID") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5),
          legend.position = "right")

  print(p_boxplot_patient)
  ggsave(file.path(output_dir, "patient_boxplot_by_patient.png"),
         p_boxplot_patient, width = 16, height = 8, dpi = 300, bg = "white")
}
```

#### By Age

```{r malignant_boxplot_age, fig.width = 12, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter out patients with missing age
  patient_stats_age <- sample_stats_full |>
    dplyr::filter(!is.na(age_bin))

  # Define age colors
  age_colors <- c("<50" = "#FEE5D9", "50-60" = "#FCAE91", "60-70" = "#FB6A4A", ">70" = "#CB181D")
  age_colors_line <- c("<50" = "#CCBAB4", "50-60" = "#CA8B74", "60-70" = "#C9553B", ">70" = "#A21317")

  p_boxplot_age <- ggplot(patient_stats_age,
                           aes(x = age_bin, y = mean_score, fill = age_bin)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_quasirandom(aes(color = age_bin, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_fill_manual(values = age_colors) +
    scale_color_manual(values = age_colors_line) +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Age at Diagnosis",
            subtitle = "All disease stages (≥10 cells per patient)") +
    xlab("Age Group (Years)") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 11),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_age)
  ggsave(file.path(output_dir, "patient_boxplot_by_age.png"),
         p_boxplot_age, width = 12, height = 8, dpi = 300, bg = "white")
}
```

#### By Gender

```{r malignant_boxplot_gender, fig.width = 10, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter out patients with missing gender
  patient_stats_gender <- sample_stats_full |>
    dplyr::filter(!is.na(Gender) & Gender != "")

  # Define gender colors
  gender_colors <- c("Female" = "#F781BF", "Male" = "#377EB8", "female" = "#F781BF", "male" = "#377EB8")
  gender_colors_line <- c("Female" = "#C6679F", "Male" = "#2D6594", "female" = "#C6679F", "male" = "#2D6594")

  p_boxplot_gender <- ggplot(patient_stats_gender,
                              aes(x = Gender, y = mean_score, fill = Gender)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_quasirandom(aes(color = Gender, size = n_cells),
                     alpha = 0.7, dodge.width = 0.8) +
    scale_fill_manual(values = gender_colors) +
    scale_color_manual(values = gender_colors_line) +
    scale_size_continuous(name = "Cell count", range = c(2, 8)) +
    theme_cowplot() +
    ggtitle("Patient-Level CytoTRACE2 Scores by Gender",
            subtitle = "All disease stages (≥10 cells per patient)") +
    xlab("Gender") +
    ylab("Mean CytoTRACE2 Score per Patient") +
    theme(legend.position = "right", axis.text.x = element_text(size = 12),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 11, hjust = 0.5))

  print(p_boxplot_gender)
  ggsave(file.path(output_dir, "patient_boxplot_by_gender.png"),
         p_boxplot_gender, width = 10, height = 8, dpi = 300, bg = "white")
}
```

### Combined Analysis: Mutations × Origin

```{r malignant_mutation_origin, fig.width = 14, fig.height = 8, eval = has_malignant}
if (has_malignant) {
  # Filter out unknown mutations
  mal_known <- subset(malignant_subset, subset = secondary_mutation != "unknown")

  # Ensure disease progression order
  mal_known$origin <- factor(mal_known$origin, levels = c("healthy", "primary", "metastasis"))

  # Faceted violin plot (Victoria's colors)
  p_comb <- ggplot(mal_known@meta.data,
                    aes(x = origin, y = CytoTRACE2_Score, fill = origin)) +
    geom_violin(alpha = 0.7, scale = "width") +
    geom_boxplot(width = 0.3, outlier.size = 0.3, alpha = 0.5) +
    facet_wrap(~ secondary_mutation, nrow = 1) +
    scale_x_discrete(limits = origin_order) +
    scale_fill_manual(values = origin_colors) +
    theme_cowplot() +
    ggtitle("CytoTRACE2 Scores: Secondary Mutations across Origins (Malignant Cells)") +
    xlab("Origin") +
    ylab("CytoTRACE2 Score") +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_comb)
  ggsave(file.path(output_dir, "malignant_cytotrace2_mutation_origin.png"), p_comb,
         width = 14, height = 8, dpi = 300, bg = "white")

  # Summary statistics
  comb_stats <- mal_known@meta.data |>
    dplyr::group_by(secondary_mutation, origin) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(secondary_mutation, desc(mean_score))

  cat("\nCytoTRACE2 Statistics by Secondary Mutation and Origin:\n")
  print(comb_stats)

  write.csv(comb_stats,
            file.path(output_dir, "malignant_cytotrace2_stats_mutation_origin.csv"),
            row.names = FALSE)
}
```

### Sample-Level Dispersion

```{r malignant_patient_dispersion, fig.width = 14, fig.height = 10, eval = has_malignant}
# Use orig.ident to merge biological replicates from same tumor
if (has_malignant && "orig.ident" %in% colnames(malignant_subset@meta.data)) {
  # Calculate per-sample statistics (replicates merged)
  sample_stats_mal <- malignant_subset@meta.data |>
    dplyr::group_by(orig.ident, origin, secondary_mutation) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::rename(Sample = orig.ident) |>
    dplyr::filter(n_cells >= 10)  # Filter samples with at least 10 malignant cells

  # Boxplot showing sample-level dispersion
  p_sample <- ggplot(sample_stats_mal,
                   aes(x = secondary_mutation, y = mean_score, fill = origin)) +
    geom_boxplot(alpha = 0.7, outlier.size = 1) +
    geom_jitter(width = 0.2, size = 1.5, alpha = 0.4) +
    scale_fill_manual(values = origin_colors) +
    theme_cowplot() +
    ggtitle("Sample-Level CytoTRACE2 Dispersion by Secondary Mutation and Origin\n(Malignant Cells, n≥10)") +
    xlab("Secondary Mutation") +
    ylab("Mean CytoTRACE2 Score per Sample") +
    theme(legend.position = "right",
          axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_sample)
  ggsave(file.path(output_dir, "malignant_sample_dispersion.png"), p_sample,
         width = 14, height = 10, dpi = 300, bg = "white")

  cat("\nSample-level statistics (biological replicates merged):\n")
  cat("Number of samples with ≥10 malignant cells:", nrow(sample_stats_mal), "\n\n")

  write.csv(sample_stats_mal,
            file.path(output_dir, "malignant_cytotrace2_sample_stats.csv"),
            row.names = FALSE)
}
```

## Cell Type-Specific CytoTRACE2 Analyses {.tabset}

This section analyzes CytoTRACE2 potency scores and gene correlations for key non-melanocytic cell populations.

### Fibroblasts

```{r fibroblasts_correlation, fig.width = 11, fig.height = 15, eval = has_cytotrace2}
if (has_cytotrace2 && "Fibroblasts" %in% unique(um_atlas@meta.data$majority_celltype)) {
  # Subset to Fibroblasts
  fibroblasts <- subset(um_atlas, subset = majority_celltype == "Fibroblasts")

  cat("\n=== Fibroblasts Gene Correlation Analysis ===\n")
  cat("Total fibroblast cells:", ncol(fibroblasts), "\n\n")

  # Run correlation analysis
  fibro_results <- correlation_plot(fibroblasts, "Fibroblasts")

  print(fibro_results$combined)
  ggsave(file.path(output_dir, "genes_correlated_fibroblasts.png"),
         fibro_results$combined,
         width = 11, height = 15, dpi = 300, bg = "white")

  # Export correlation table
  write.csv(
    fibro_results$cor_df |> dplyr::arrange(desc(abs(correlation))),
    file.path(output_dir, "cytotrace2_fibroblasts_gene_correlations.csv"),
    row.names = FALSE
  )

  cat("\nTop fibroblast stemness markers:\n")
  print(fibro_results$cor_df |> dplyr::top_n(10, wt = correlation) |>
          dplyr::arrange(desc(correlation)))

  cat("\nTop fibroblast differentiation markers:\n")
  print(fibro_results$cor_df |> dplyr::top_n(10, wt = -correlation) |>
          dplyr::arrange(correlation))
}
```

### T/NK Cells

```{r tnk_correlation, fig.width = 11, fig.height = 15, eval = has_cytotrace2}
if (has_cytotrace2 && "T/NK cells" %in% unique(um_atlas@meta.data$majority_celltype)) {
  # Subset to T/NK cells
  tnk_cells <- subset(um_atlas, subset = majority_celltype == "T/NK cells")

  cat("\n=== T/NK Cells Gene Correlation Analysis ===\n")
  cat("Total T/NK cells:", ncol(tnk_cells), "\n\n")

  # Run correlation analysis
  tnk_results <- correlation_plot(tnk_cells, "T/NK Cells")

  print(tnk_results$combined)
  ggsave(file.path(output_dir, "genes_correlated_tnk.png"),
         tnk_results$combined,
         width = 11, height = 15, dpi = 300, bg = "white")

  # Export correlation table
  write.csv(
    tnk_results$cor_df |> dplyr::arrange(desc(abs(correlation))),
    file.path(output_dir, "cytotrace2_tnk_gene_correlations.csv"),
    row.names = FALSE
  )

  cat("\nTop T/NK stemness markers:\n")
  print(tnk_results$cor_df |> dplyr::top_n(10, wt = correlation) |>
          dplyr::arrange(desc(correlation)))

  cat("\nTop T/NK differentiation markers:\n")
  print(tnk_results$cor_df |> dplyr::top_n(10, wt = -correlation) |>
          dplyr::arrange(correlation))
}
```

### Myeloid Cells

```{r myeloid_correlation, fig.width = 11, fig.height = 15, eval = has_cytotrace2}
if (has_cytotrace2 && "Myeloid cells" %in% unique(um_atlas@meta.data$majority_celltype)) {
  # Subset to Myeloid cells
  myeloid_cells <- subset(um_atlas, subset = majority_celltype == "Myeloid cells")

  cat("\n=== Myeloid Cells Gene Correlation Analysis ===\n")
  cat("Total myeloid cells:", ncol(myeloid_cells), "\n\n")

  # Run correlation analysis
  myeloid_results <- correlation_plot(myeloid_cells, "Myeloid Cells")

  print(myeloid_results$combined)
  ggsave(file.path(output_dir, "genes_correlated_myeloid.png"),
         myeloid_results$combined,
         width = 11, height = 15, dpi = 300, bg = "white")

  # Export correlation table
  write.csv(
    myeloid_results$cor_df |> dplyr::arrange(desc(abs(correlation))),
    file.path(output_dir, "cytotrace2_myeloid_gene_correlations.csv"),
    row.names = FALSE
  )

  cat("\nTop myeloid stemness markers:\n")
  print(myeloid_results$cor_df |> dplyr::top_n(10, wt = correlation) |>
          dplyr::arrange(desc(correlation)))

  cat("\nTop myeloid differentiation markers:\n")
  print(myeloid_results$cor_df |> dplyr::top_n(10, wt = -correlation) |>
          dplyr::arrange(correlation))
}
```

## Summary Statistics {.tabset}

### Overall Distribution

```{r summary_stats, eval = has_cytotrace2}
cat("CytoTRACE2 Score Statistics:\n")
cat("===========================\n\n")

cat("Overall:\n")
print(summary(um_atlas$CytoTRACE2_Score))

cat("\n\nBy Potency Category:\n")
print(table(um_atlas$CytoTRACE2_Potency))

cat("\n\nPotency Proportions:\n")
print(prop.table(table(um_atlas$CytoTRACE2_Potency)))
```

### By Cell Type

```{r summary_by_celltype, eval = has_cytotrace2}
cat("CytoTRACE2 Score by Cell Type:\n")
cat("==============================\n\n")

# Mean and SD by cell type
celltype_stats <- um_atlas@meta.data |>
  dplyr::group_by(.data[[celltype_col]]) |>
  dplyr::summarise(
    n_cells = n(),
    mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
    sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
    median_score = median(CytoTRACE2_Score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(desc(mean_score))

print(celltype_stats)

# Export
write.csv(
  celltype_stats,
  file.path(output_dir, "cytotrace2_stats_by_celltype.csv"),
  row.names = FALSE
)
```

### By Sample

```{r summary_by_sample, eval = has_cytotrace2}
# Use orig.ident to merge biological replicates from same tumor
if ("orig.ident" %in% colnames(um_atlas@meta.data)) {
  cat("CytoTRACE2 Score by Sample (biological replicates merged):\n")
  cat("==========================================================\n\n")

  sample_stats <- um_atlas@meta.data |>
    dplyr::group_by(orig.ident) |>
    dplyr::summarise(
      n_cells = n(),
      mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
      sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
      median_score = median(CytoTRACE2_Score, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::rename(Sample = orig.ident) |>
    dplyr::arrange(desc(mean_score))

  print(sample_stats)

  # Export
  write.csv(
    sample_stats,
    file.path(output_dir, "cytotrace2_stats_by_sample.csv"),
    row.names = FALSE
  )
} else {
  cat("No 'Sample' column found. Skipping sample-level statistics.\n")
}
```

## Healthy vs Primary Melanocytes: Stemness Comparison {.tabset}

This section compares CytoTRACE2 stemness-correlated genes between healthy melanocytes and primary melanoma cells.

### Venn Diagram: Stemness Gene Overlap

```{r venn_healthy_primary, fig.width=10, fig.height=8, eval = has_cytotrace2}
library(VennDiagram)
library(ggvenn)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)

# Load gene correlations
healthy_genes <- read.csv(file.path(output_dir, "cytotrace2_healthy_melanocytes_gene_correlations.csv"))
malignant_genes <- read.csv(file.path(output_dir, "cytotrace2_malignant_gene_correlations.csv"))

# Get top 100 stemness genes (positive correlation with CytoTRACE2 score = stemness)
n_top <- 100
healthy_stemness <- healthy_genes |>
  dplyr::filter(correlation > 0) |>
  dplyr::arrange(desc(correlation)) |>
  head(n_top) |>
  dplyr::pull(gene)

malignant_stemness <- malignant_genes |>
  dplyr::filter(correlation > 0) |>
  dplyr::arrange(desc(correlation)) |>
  head(n_top) |>
  dplyr::pull(gene)

# Create Venn diagram data
venn_data <- list(
  "Healthy Melanocytes" = healthy_stemness,
  "Primary Melanoma" = malignant_stemness
)

# Plot with ggvenn
p_venn <- ggvenn(
  venn_data,
  fill_color = c("#4DAF4A", "#E41A1C"),
  stroke_size = 0.5,
  set_name_size = 5,
  text_size = 4
) +
  ggtitle(paste0("Top ", n_top, " Stemness Genes: Healthy vs Primary Melanoma")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
print(p_venn)

ggsave(
  file.path(output_dir, "venn_stemness_healthy_vs_primary.png"),
  p_venn, width = 10, height = 8, dpi = 300, bg = "white"
)

# Calculate overlaps
overlap <- intersect(healthy_stemness, malignant_stemness)
healthy_unique <- setdiff(healthy_stemness, malignant_stemness)
malignant_unique <- setdiff(malignant_stemness, healthy_stemness)

cat("\n=== Stemness Gene Overlap Summary ===\n")
cat("Shared genes:", length(overlap), "\n")
cat("Healthy-specific genes:", length(healthy_unique), "\n")
cat("Malignant-specific genes:", length(malignant_unique), "\n")
```

### Shared Stemness Genes

```{r shared_stemness_genes, eval = has_cytotrace2}
cat("=== Genes associated with stemness in BOTH healthy and malignant melanocytes ===\n\n")

if (length(overlap) > 0) {
  # Get correlations for shared genes
  shared_df <- data.frame(
    gene = overlap,
    correlation_healthy = healthy_genes$correlation[match(overlap, healthy_genes$gene)],
    correlation_malignant = malignant_genes$correlation[match(overlap, malignant_genes$gene)]
  ) |>
    dplyr::mutate(
      mean_correlation = (correlation_healthy + correlation_malignant) / 2
    ) |>
    dplyr::arrange(desc(mean_correlation))

  print(shared_df)

  write.csv(
    shared_df,
    file.path(output_dir, "stemness_genes_shared_healthy_malignant.csv"),
    row.names = FALSE
  )
}
```

### Unique Stemness Genes

```{r unique_stemness_genes, eval = has_cytotrace2}
cat("=== Healthy-Specific Stemness Genes ===\n")
cat("(Genes correlated with stemness only in healthy melanocytes)\n\n")

if (length(healthy_unique) > 0) {
  healthy_unique_df <- healthy_genes |>
    dplyr::filter(gene %in% healthy_unique) |>
    dplyr::arrange(desc(correlation))
  print(healthy_unique_df)

  write.csv(
    healthy_unique_df,
    file.path(output_dir, "stemness_genes_healthy_unique.csv"),
    row.names = FALSE
  )
}

cat("\n=== Malignant-Specific Stemness Genes ===\n")
cat("(Genes correlated with stemness only in malignant melanoma cells)\n\n")

if (length(malignant_unique) > 0) {
  malignant_unique_df <- malignant_genes |>
    dplyr::filter(gene %in% malignant_unique) |>
    dplyr::arrange(desc(correlation))
  print(malignant_unique_df)

  write.csv(
    malignant_unique_df,
    file.path(output_dir, "stemness_genes_malignant_unique.csv"),
    row.names = FALSE
  )
}
```

### GO Enrichment: Stemness Genes

```{r enrichment_stemness, fig.width=10, fig.height=8, eval = has_cytotrace2}
# Reusable function for GO BP enrichment
run_enrichment <- function(genes, category_name, output_prefix) {
  if (length(genes) < 5) {
    cat("Too few genes for enrichment:", category_name, "\n")
    return(NULL)
  }

  # GO Biological Process
  ego <- enrichGO(
    gene = genes,
    OrgDb = org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )

  if (!is.null(ego) && nrow(ego) > 0) {
    p_go <- dotplot(ego, showCategory = 15, title = paste(category_name, "- GO BP"))
    print(p_go)
    ggsave(file.path(output_dir, paste0(output_prefix, "_GO_BP_dotplot.png")),
           p_go, width = 10, height = 8, dpi = 300, bg = "white")
  } else {
    cat("No significant GO BP terms for:", category_name, "\n")
  }

  return(ego)
}

cat("=== GO Enrichment Analysis: Stemness Genes ===\n\n")

# Run enrichment for 3 stemness categories
cat("--- Shared Stemness Genes ---\n")
enr_shared_stemness <- run_enrichment(overlap, "Shared Stemness Genes", "enrichment_stemness_shared")

cat("\n--- Healthy-Specific Stemness Genes ---\n")
enr_healthy_stemness <- run_enrichment(healthy_unique, "Healthy-Specific Stemness", "enrichment_stemness_healthy_unique")

cat("\n--- Malignant-Specific Stemness Genes ---\n")
enr_malignant_stemness <- run_enrichment(malignant_unique, "Malignant-Specific Stemness", "enrichment_stemness_malignant_unique")
```

### Venn Diagram: Differentiation Gene Overlap

```{r venn_differentiation, fig.width=10, fig.height=8, eval = has_cytotrace2}
# Get top 100 differentiation genes (NEGATIVE correlation = differentiation)
healthy_diff <- healthy_genes |>
  dplyr::filter(correlation < 0) |>
  dplyr::arrange(correlation) |>
  head(n_top) |>
  dplyr::pull(gene)

malignant_diff <- malignant_genes |>
  dplyr::filter(correlation < 0) |>
  dplyr::arrange(correlation) |>
  head(n_top) |>
  dplyr::pull(gene)

# Create Venn diagram data
venn_data_diff <- list(
  "Healthy Melanocytes" = healthy_diff,
  "Primary Melanoma" = malignant_diff
)

# Plot with ggvenn
p_venn_diff <- ggvenn(
  venn_data_diff,
  fill_color = c("#4DAF4A", "#E41A1C"),
  stroke_size = 0.5,
  set_name_size = 5,
  text_size = 4
) +
  ggtitle(paste0("Top ", n_top, " Differentiation Genes: Healthy vs Primary Melanoma")) +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
print(p_venn_diff)

ggsave(
  file.path(output_dir, "venn_differentiation_healthy_vs_primary.png"),
  p_venn_diff, width = 10, height = 8, dpi = 300, bg = "white"
)

# Calculate overlaps
overlap_diff <- intersect(healthy_diff, malignant_diff)
healthy_unique_diff <- setdiff(healthy_diff, malignant_diff)
malignant_unique_diff <- setdiff(malignant_diff, healthy_diff)

cat("\n=== Differentiation Gene Overlap Summary ===\n")
cat("Shared genes:", length(overlap_diff), "\n")
cat("Healthy-specific genes:", length(healthy_unique_diff), "\n")
cat("Malignant-specific genes:", length(malignant_unique_diff), "\n")

# Save differentiation gene lists
write.csv(
  data.frame(
    gene = overlap_diff,
    correlation_healthy = healthy_genes$correlation[match(overlap_diff, healthy_genes$gene)],
    correlation_malignant = malignant_genes$correlation[match(overlap_diff, malignant_genes$gene)]
  ),
  file.path(output_dir, "differentiation_genes_shared_healthy_malignant.csv"),
  row.names = FALSE
)

write.csv(
  healthy_genes |> dplyr::filter(gene %in% healthy_unique_diff) |> dplyr::arrange(correlation),
  file.path(output_dir, "differentiation_genes_healthy_unique.csv"),
  row.names = FALSE
)

write.csv(
  malignant_genes |> dplyr::filter(gene %in% malignant_unique_diff) |> dplyr::arrange(correlation),
  file.path(output_dir, "differentiation_genes_malignant_unique.csv"),
  row.names = FALSE
)
```

### GO Enrichment: Differentiation Genes

```{r enrichment_differentiation, fig.width=10, fig.height=8, eval = has_cytotrace2}
cat("=== GO Enrichment Analysis: Differentiation Genes ===\n\n")

# Run enrichment for 3 differentiation categories
cat("--- Shared Differentiation Genes ---\n")
enr_shared_diff <- run_enrichment(overlap_diff, "Shared Differentiation Genes", "enrichment_diff_shared")

cat("\n--- Healthy-Specific Differentiation Genes ---\n")
enr_healthy_diff <- run_enrichment(healthy_unique_diff, "Healthy-Specific Differentiation", "enrichment_diff_healthy_unique")

cat("\n--- Malignant-Specific Differentiation Genes ---\n")
enr_malignant_diff <- run_enrichment(malignant_unique_diff, "Malignant-Specific Differentiation", "enrichment_diff_malignant_unique")
```

### Statistical Comparison

```{r healthy_vs_primary_stats, fig.width=10, fig.height=6, eval = has_cytotrace2}
# Compare CytoTRACE2 scores between healthy and primary melanocytes
melanocytic_cells <- um_atlas@meta.data |>
  dplyr::filter(majority_celltype %in% c("Melanocytes", "Melanoma cells")) |>
  dplyr::filter(origin %in% c("healthy", "primary"))

cat("=== CytoTRACE2 Score Comparison: Healthy vs Primary ===\n\n")

# Statistical test
wilcox_result <- wilcox.test(
  CytoTRACE2_Score ~ origin,
  data = melanocytic_cells,
  alternative = "two.sided"
)

cat("Wilcoxon rank-sum test:\n")
cat("W =", wilcox_result$statistic, "\n")
cat("p-value =", format(wilcox_result$p.value, scientific = TRUE), "\n\n")

# Summary stats
origin_stats <- melanocytic_cells |>
  dplyr::group_by(origin) |>
  dplyr::summarise(
    n_cells = n(),
    mean_score = mean(CytoTRACE2_Score, na.rm = TRUE),
    sd_score = sd(CytoTRACE2_Score, na.rm = TRUE),
    median_score = median(CytoTRACE2_Score, na.rm = TRUE),
    .groups = "drop"
  )
print(origin_stats)

# Violin plot comparison (using Victoria's origin_colors)
melanocytic_cells$origin <- factor(melanocytic_cells$origin, levels = c("healthy", "primary", "metastasis"))
p_compare <- ggplot(melanocytic_cells, aes(x = origin, y = CytoTRACE2_Score, fill = origin)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_x_discrete(limits = origin_order) +
  scale_fill_manual(values = origin_colors) +
  labs(
    title = "CytoTRACE2 Stemness: Healthy Melanocytes vs Primary Melanoma",
    subtitle = paste0("Wilcoxon p = ", format(wilcox_result$p.value, scientific = TRUE, digits = 3)),
    x = "Origin",
    y = "CytoTRACE2 Score (Stemness)"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")
print(p_compare)

ggsave(
  file.path(output_dir, "violin_healthy_vs_primary_comparison.png"),
  p_compare, width = 8, height = 6, dpi = 300, bg = "white"
)
```

## Stemness Signature for Xenium Validation {.tabset}

Extract top stemness genes from primary melanoma for spatial validation on Xenium data.

### Top Melanoma Stemness Genes

```{r extract_xenium_signature, eval = has_cytotrace2}
# Get top 50 melanoma stemness genes for Xenium validation
n_signature <- 50

melanoma_stemness_signature <- malignant_genes |>
  dplyr::filter(correlation > 0) |>
  dplyr::arrange(desc(correlation)) |>
  head(n_signature)

cat("=== Top", n_signature, "Melanoma Stemness Genes for Xenium Validation ===\n\n")
print(melanoma_stemness_signature)

# Save signature
write.csv(
  melanoma_stemness_signature,
  file.path(output_dir, "melanoma_stemness_signature_xenium.csv"),
  row.names = FALSE
)

cat("\nSignature saved to: Outputs/melanoma_stemness_signature_xenium.csv\n")
cat("\nTop 20 genes:\n")
cat(paste(melanoma_stemness_signature$gene[1:20], collapse = ", "), "\n")
```

### Check Xenium Panel Availability

```{r check_xenium_panel, eval = has_cytotrace2}
# Load Xenium data to check which genes are available
xenium_path <- "../Xenium_3cohorts_merged"

if (dir.exists(xenium_path)) {
  xenium_files <- list.files(xenium_path, pattern = "\\.qs2?$|\\.rds$", full.names = TRUE)
  cat("Found Xenium files:\n")
  print(basename(xenium_files))
} else {
  cat("Xenium analysis directory not found at expected location.\n")
  cat("Expected: ", xenium_path, "\n")
}

# Export gene list as simple text file for easy lookup
writeLines(
  melanoma_stemness_signature$gene,
  file.path(output_dir, "melanoma_stemness_genes_list.txt")
)
cat("\nGene list saved as plain text: Outputs/melanoma_stemness_genes_list.txt\n")
```

## Myeloid Cell Investigation {.tabset}

Investigating why myeloid cells show high correlation with melanocytic genes in CytoTRACE2 analysis.

### Myeloid Gene Correlations Issue

```{r myeloid_investigation, fig.width=12, fig.height=8, eval = has_cytotrace2}
# Load myeloid gene correlations
myeloid_genes <- read.csv(file.path(output_dir, "cytotrace2_myeloid_gene_correlations.csv"))

cat("=== Top 30 CytoTRACE2-Correlated Genes in Myeloid Cells ===\n\n")
print(head(myeloid_genes, 30))

# Define melanocyte markers
melanocyte_markers <- c("PMEL", "MLANA", "TYR", "TYRP1", "DCT", "MITF", "SOX10",
                        "PAX3", "CDH19", "RAB38", "OCA2", "EDNRB", "TRPM1")

# Check how many melanocyte markers are in top correlated genes
top_100_myeloid <- head(myeloid_genes, 100)$gene
melanocyte_overlap <- intersect(top_100_myeloid, melanocyte_markers)

cat("\n=== Melanocyte Markers in Top 100 Myeloid CytoTRACE2-Correlated Genes ===\n")
cat("Found:", length(melanocyte_overlap), "of", length(melanocyte_markers), "markers\n")
cat("Markers present:", paste(melanocyte_overlap, collapse = ", "), "\n")
```

### Expression Analysis

```{r myeloid_expression_check, fig.width=14, fig.height=10, eval = has_cytotrace2}
# Subset to myeloid cells
myeloid_cells <- subset(um_atlas, majority_celltype == "Myeloid cells")
cat("Number of myeloid cells:", ncol(myeloid_cells), "\n")

# Check expression of melanocyte markers vs myeloid markers
myeloid_markers <- c("CD68", "CD14", "CD163", "MARCO", "CSF1R", "LYZ", "C1QA", "C1QB", "C1QC")
all_markers <- c(melanocyte_markers, myeloid_markers)

# Get available markers
available_markers <- intersect(all_markers, rownames(myeloid_cells))
cat("\nAvailable markers for analysis:", length(available_markers), "/", length(all_markers), "\n")

# Check doublet scores in myeloid cells
if ("doubletScore" %in% colnames(myeloid_cells@meta.data)) {
  cat("\n=== Doublet Scores in Myeloid Cells ===\n")
  cat("Mean doublet score:", mean(myeloid_cells$doubletScore, na.rm = TRUE), "\n")
  cat("Median doublet score:", median(myeloid_cells$doubletScore, na.rm = TRUE), "\n")
  cat("% doublets (score > 0.5):",
      round(100 * mean(myeloid_cells$doubletScore > 0.5, na.rm = TRUE), 2), "%\n")
}

# Expression dotplot
if (length(available_markers) > 0) {
  p_dot <- DotPlot(
    myeloid_cells,
    features = available_markers,
    group.by = "origin"
  ) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle("Expression of Melanocyte vs Myeloid Markers in 'Myeloid' Cells")
  print(p_dot)

  ggsave(
    file.path(output_dir, "dotplot_myeloid_marker_expression.png"),
    p_dot, width = 10, height = 10, dpi = 300, bg = "white"
  )
}
```

### Myeloid Cells by Origin and CytoTRACE2

```{r myeloid_by_origin, fig.width=12, fig.height=6, eval = has_cytotrace2}
# Check CytoTRACE2 distribution in myeloid cells by origin
# Ensure disease progression order
myeloid_cells$origin <- factor(myeloid_cells$origin, levels = c("healthy", "primary", "metastasis"))

p_myeloid_origin <- ggplot(
  myeloid_cells@meta.data,
  aes(x = origin, y = CytoTRACE2_Score, fill = origin)
) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_x_discrete(limits = origin_order) +
  scale_fill_manual(values = origin_colors) +
  labs(
    title = "CytoTRACE2 Score in Myeloid Cells by Origin",
    x = "Origin",
    y = "CytoTRACE2 Score"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")
print(p_myeloid_origin)

# Check if high-stemness myeloid cells express melanocyte markers
myeloid_meta <- myeloid_cells@meta.data
myeloid_meta$stemness_group <- cut(
  myeloid_meta$CytoTRACE2_Score,
  breaks = c(0, 0.2, 0.4, 1),
  labels = c("Low", "Medium", "High")
)

cat("\n=== Myeloid Cells by Stemness Group ===\n")
print(table(myeloid_meta$stemness_group, useNA = "always"))
```

### Hypothesis: Doublets or Phagocytosis

```{r myeloid_hypothesis, fig.width=12, fig.height=8, eval = has_cytotrace2}
cat("=== Potential Explanations for Melanocyte Gene Expression in Myeloid Cells ===\n\n")

cat("1. DOUBLETS: Myeloid-melanoma doublets could explain co-expression\n")
cat("   - Check: Doublet scores should be elevated\n")
cat("   - Check: Co-expression of myeloid + melanocyte markers in same cells\n\n")

cat("2. PHAGOCYTOSIS: Macrophages may have engulfed melanoma cells\n")
cat("   - Check: Primarily in macrophage subsets (CD68+, CD163+)\n")
cat("   - Check: Tumor-associated macrophages in malignant samples\n\n")

cat("3. MISANNOTATION: Some cells may be incorrectly labeled\n")
cat("   - Check: Reference-based annotation confidence scores\n")
cat("   - Check: Manual marker inspection\n\n")

# Check correlation between melanocyte and myeloid marker expression
if (all(c("PMEL", "CD68") %in% rownames(myeloid_cells))) {
  myeloid_cells <- JoinLayers(myeloid_cells)
  expr_mat <- GetAssayData(myeloid_cells, layer = "data")

  if ("PMEL" %in% rownames(expr_mat) && "CD68" %in% rownames(expr_mat)) {
    pmel_expr <- expr_mat["PMEL", ]
    cd68_expr <- expr_mat["CD68", ]

    cor_test <- cor.test(as.numeric(pmel_expr), as.numeric(cd68_expr))

    cat("=== Co-expression Analysis ===\n")
    cat("Correlation between PMEL and CD68 in myeloid cells:\n")
    cat("Pearson r =", round(cor_test$estimate, 3), "\n")
    cat("p-value =", format(cor_test$p.value, scientific = TRUE), "\n")

    # Scatter plot
    plot_df <- data.frame(
      PMEL = as.numeric(pmel_expr),
      CD68 = as.numeric(cd68_expr),
      CytoTRACE2 = myeloid_cells$CytoTRACE2_Score
    )

    p_scatter <- ggplot(plot_df, aes(x = CD68, y = PMEL, color = CytoTRACE2)) +
      geom_point(alpha = 0.5, size = 1) +
      scale_color_viridis_c() +
      labs(
        title = "CD68 vs PMEL Expression in Myeloid Cells",
        subtitle = paste0("r = ", round(cor_test$estimate, 3)),
        x = "CD68 (Myeloid marker)",
        y = "PMEL (Melanocyte marker)",
        color = "CytoTRACE2\nScore"
      ) +
      theme_classic(base_size = 12)
    print(p_scatter)

    ggsave(
      file.path(output_dir, "scatter_myeloid_cd68_vs_pmel.png"),
      p_scatter, width = 8, height = 6, dpi = 300, bg = "white"
    )
  }
}
```

## Session Info

```{r session_info}
sessionInfo()
```

## Interpretation Guide

### CytoTRACE2 Score Ranges

- **0.0 - 0.167**: Differentiated (terminally differentiated cells)
- **0.167 - 0.333**: Unipotent (single lineage potential)
- **0.333 - 0.5**: Oligopotent (few lineages)
- **0.5 - 0.667**: Multipotent (multiple lineages)
- **0.667 - 0.833**: Pluripotent (most lineages)
- **0.833 - 1.0**: Totipotent (all lineages)

### Expected Patterns in Uveal Melanoma

- **Malignant cells with high stemness** may represent cancer stem cells with higher tumorigenic potential
- **Differentiated tumor cells** may be more terminally differentiated and less aggressive
- **Immune cells** typically show lower potency scores (differentiated)
- **Endothelial/stromal cells** may show intermediate scores

### Key Outputs

All results saved to: `../Outputs/`

- Seurat object with CytoTRACE2: `../UM_atlas_cytotrace2.qs2`
- UMAP plots showing potency scores and categories
- Violin plots by cell type and sample
- Compositional bar plots
- Gene correlation tables (stemness and differentiation markers)
- Summary statistics tables
