---
title: "p31662 UM Atlas - CytoTRACE2 vs Metaprogram Correlation Analysis"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Methods (correlation between CytoTRACE2 stemness scores and the 67 GeneNMF cancer metaprograms)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Overview

This report analyzes the correlation between CytoTRACE2 stemness scores and specific cancer metaprogram expression in malignant cells from the UM Atlas dataset.

**Requested metaprograms:**

- **MP6 (Hypoxia)**: Stress/hypoxia response
- **MP9**: Stress-related metaprogram
- **MP45**: Tissue-specific metaprogram

## Data Loading

```{r load_libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(cowplot)
library(qs)
library(readr)

# Load Victoria's color palettes
source("colors.R")

# CRITICAL: Reorder origin_colors to match disease progression (healthy → primary → metastasis)
origin_order <- c("healthy", "primary", "metastasis")
origin_colors <- origin_colors[origin_order]

output_dir <- "Outputs/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r load_data}
# Load melanocytic cells with metaprogram scores
cat("Loading melanocytic cells with metaprogram scores...\n")
melanocytic <- qread(
  "../melanocytic_metaprograms.qs2",
  nthreads = 16
)
cat("  - Cells:", ncol(melanocytic), "\n")

# Load UM atlas with CytoTRACE2 scores
cat("\nLoading CytoTRACE2 data from UM atlas...\n")
um_atlas <- qread(
  "../UM_atlas_cytotrace2.qs2",
  nthreads = 16
)
cat("  - Cells:", ncol(um_atlas), "\n")

# Find common cells
common_cells <- intersect(colnames(melanocytic), colnames(um_atlas))
cat("\nCommon cells:", length(common_cells), "\n")

# Transfer CytoTRACE2 scores to melanocytic object
melanocytic$CytoTRACE2_Score <- NA
melanocytic@meta.data[common_cells, "CytoTRACE2_Score"] <- um_atlas@meta.data[common_cells, "CytoTRACE2_Score"]

# Clean up
rm(um_atlas)
gc()

# Filter to cells with both CytoTRACE2 and MP scores
has_cyto <- !is.na(melanocytic$CytoTRACE2_Score)
cat("Cells with CytoTRACE2 scores:", sum(has_cyto), "\n")
```

```{r identify_mp_columns}
# Find metaprogram columns
ucell_cols <- grep("^MP[0-9]+", colnames(melanocytic@meta.data), value = TRUE)
cat("Available metaprogram columns:", length(ucell_cols), "\n\n")

# Find specific MPs requested
mp6_col <- grep("^MP6 ", ucell_cols, value = TRUE)
mp9_col <- grep("^MP9 ", ucell_cols, value = TRUE)
mp45_col <- grep("^MP45 ", ucell_cols, value = TRUE)

cat("Selected columns:\n")
cat("  MP6:", mp6_col, "\n")
cat("  MP9:", mp9_col, "\n")
cat("  MP45:", mp45_col, "\n")

# Create analysis dataframe (malignant cells with CytoTRACE2)
analysis_df <- melanocytic@meta.data %>%
  filter(!is.na(CytoTRACE2_Score)) %>%
  select(CytoTRACE2_Score, origin, all_of(c(mp6_col, mp9_col, mp45_col)))

# Rename for clarity
colnames(analysis_df) <- c("CytoTRACE2_Score", "origin", "MP6", "MP9", "MP45")

cat("\nAnalysis dataframe:\n")
cat("  - Cells:", nrow(analysis_df), "\n")
cat("  - Columns:", paste(colnames(analysis_df), collapse = ", "), "\n")

# Enforce disease progression order
analysis_df$origin <- factor(analysis_df$origin, levels = c("healthy", "primary", "metastasis"))
```

# {.tabset}

## MP6 (Hypoxia) Correlation

```{r mp6_correlation, fig.width = 10, fig.height = 8}
# Spearman correlation
cor_mp6 <- cor.test(analysis_df$CytoTRACE2_Score, analysis_df$MP6,
                    method = "spearman", exact = FALSE)

cat("=== MP6 (Hypoxia) vs CytoTRACE2 Score ===\n")
cat("Spearman rho:", round(cor_mp6$estimate, 4), "\n")
cat("p-value:", format(cor_mp6$p.value, scientific = TRUE, digits = 3), "\n\n")

# Scatterplot
p_mp6 <- ggplot(analysis_df, aes(x = CytoTRACE2_Score, y = MP6)) +
  geom_point(aes(color = origin), alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  stat_cor(method = "spearman", label.x = 0.05, label.y = max(analysis_df$MP6) * 0.95,
           size = 5, cor.coef.name = "rho") +
  scale_color_manual(values = origin_colors) +
  labs(
    title = paste0("CytoTRACE2 Score vs MP6 (", mp6_col, ")"),
    subtitle = paste0("n = ", nrow(analysis_df), " malignant/melanocytic cells"),
    x = "CytoTRACE2 Score (Stemness)",
    y = "MP6 Score (UCell)",
    color = "Origin"
  ) +
  theme_cowplot(font_size = 14) +
  theme(legend.position = "right")

print(p_mp6)
ggsave(file.path(output_dir, "correlation_cytotrace2_mp6.png"), p_mp6,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Faceted by Origin

```{r mp6_faceted, fig.width = 14, fig.height = 6}
p_mp6_facet <- ggplot(analysis_df, aes(x = CytoTRACE2_Score, y = MP6)) +
  geom_point(aes(color = origin), alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  stat_cor(method = "spearman", size = 4, cor.coef.name = "rho") +
  scale_color_manual(values = origin_colors) +
  facet_wrap(~ origin, nrow = 1) +
  labs(
    title = "CytoTRACE2 vs MP6 by Disease Stage",
    x = "CytoTRACE2 Score (Stemness)",
    y = "MP6 Score (UCell)"
  ) +
  theme_cowplot(font_size = 12) +
  theme(legend.position = "none")

print(p_mp6_facet)
ggsave(file.path(output_dir, "correlation_cytotrace2_mp6_faceted.png"), p_mp6_facet,
       width = 14, height = 6, dpi = 300, bg = "white")
```

## MP9 Correlation

```{r mp9_correlation, fig.width = 10, fig.height = 8}
# Spearman correlation
cor_mp9 <- cor.test(analysis_df$CytoTRACE2_Score, analysis_df$MP9,
                    method = "spearman", exact = FALSE)

cat("=== MP9 vs CytoTRACE2 Score ===\n")
cat("Spearman rho:", round(cor_mp9$estimate, 4), "\n")
cat("p-value:", format(cor_mp9$p.value, scientific = TRUE, digits = 3), "\n\n")

# Scatterplot
p_mp9 <- ggplot(analysis_df, aes(x = CytoTRACE2_Score, y = MP9)) +
  geom_point(aes(color = origin), alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  stat_cor(method = "spearman", label.x = 0.05, label.y = max(analysis_df$MP9) * 0.95,
           size = 5, cor.coef.name = "rho") +
  scale_color_manual(values = origin_colors) +
  labs(
    title = paste0("CytoTRACE2 Score vs MP9 (", mp9_col, ")"),
    subtitle = paste0("n = ", nrow(analysis_df), " malignant/melanocytic cells"),
    x = "CytoTRACE2 Score (Stemness)",
    y = "MP9 Score (UCell)",
    color = "Origin"
  ) +
  theme_cowplot(font_size = 14) +
  theme(legend.position = "right")

print(p_mp9)
ggsave(file.path(output_dir, "correlation_cytotrace2_mp9.png"), p_mp9,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Faceted by Origin

```{r mp9_faceted, fig.width = 14, fig.height = 6}
p_mp9_facet <- ggplot(analysis_df, aes(x = CytoTRACE2_Score, y = MP9)) +
  geom_point(aes(color = origin), alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  stat_cor(method = "spearman", size = 4, cor.coef.name = "rho") +
  scale_color_manual(values = origin_colors) +
  facet_wrap(~ origin, nrow = 1) +
  labs(
    title = "CytoTRACE2 vs MP9 by Disease Stage",
    x = "CytoTRACE2 Score (Stemness)",
    y = "MP9 Score (UCell)"
  ) +
  theme_cowplot(font_size = 12) +
  theme(legend.position = "none")

print(p_mp9_facet)
ggsave(file.path(output_dir, "correlation_cytotrace2_mp9_faceted.png"), p_mp9_facet,
       width = 14, height = 6, dpi = 300, bg = "white")
```

## MP45 Correlation

```{r mp45_correlation, fig.width = 10, fig.height = 8}
# Spearman correlation
cor_mp45 <- cor.test(analysis_df$CytoTRACE2_Score, analysis_df$MP45,
                     method = "spearman", exact = FALSE)

cat("=== MP45 vs CytoTRACE2 Score ===\n")
cat("Spearman rho:", round(cor_mp45$estimate, 4), "\n")
cat("p-value:", format(cor_mp45$p.value, scientific = TRUE, digits = 3), "\n\n")

# Scatterplot
p_mp45 <- ggplot(analysis_df, aes(x = CytoTRACE2_Score, y = MP45)) +
  geom_point(aes(color = origin), alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  stat_cor(method = "spearman", label.x = 0.05, label.y = max(analysis_df$MP45) * 0.95,
           size = 5, cor.coef.name = "rho") +
  scale_color_manual(values = origin_colors) +
  labs(
    title = paste0("CytoTRACE2 Score vs MP45 (", mp45_col, ")"),
    subtitle = paste0("n = ", nrow(analysis_df), " malignant/melanocytic cells"),
    x = "CytoTRACE2 Score (Stemness)",
    y = "MP45 Score (UCell)",
    color = "Origin"
  ) +
  theme_cowplot(font_size = 14) +
  theme(legend.position = "right")

print(p_mp45)
ggsave(file.path(output_dir, "correlation_cytotrace2_mp45.png"), p_mp45,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Faceted by Origin

```{r mp45_faceted, fig.width = 14, fig.height = 6}
p_mp45_facet <- ggplot(analysis_df, aes(x = CytoTRACE2_Score, y = MP45)) +
  geom_point(aes(color = origin), alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black", linewidth = 1) +
  stat_cor(method = "spearman", size = 4, cor.coef.name = "rho") +
  scale_color_manual(values = origin_colors) +
  facet_wrap(~ origin, nrow = 1) +
  labs(
    title = "CytoTRACE2 vs MP45 by Disease Stage",
    x = "CytoTRACE2 Score (Stemness)",
    y = "MP45 Score (UCell)"
  ) +
  theme_cowplot(font_size = 12) +
  theme(legend.position = "none")

print(p_mp45_facet)
ggsave(file.path(output_dir, "correlation_cytotrace2_mp45_faceted.png"), p_mp45_facet,
       width = 14, height = 6, dpi = 300, bg = "white")
```

## Summary Statistics

```{r summary_table}
# Create correlation summary
correlation_summary <- data.frame(
  Metaprogram = c("MP6 (Hypoxia)", "MP9", "MP45"),
  Full_Name = c(mp6_col, mp9_col, mp45_col),
  Spearman_rho = c(cor_mp6$estimate, cor_mp9$estimate, cor_mp45$estimate),
  P_value = c(cor_mp6$p.value, cor_mp9$p.value, cor_mp45$p.value),
  N_cells = nrow(analysis_df)
)

correlation_summary$Significance <- ifelse(correlation_summary$P_value < 0.001, "***",
                                    ifelse(correlation_summary$P_value < 0.01, "**",
                                    ifelse(correlation_summary$P_value < 0.05, "*", "ns")))

cat("=== Correlation Summary ===\n\n")
print(knitr::kable(correlation_summary, digits = 4,
                   caption = "Spearman correlations between CytoTRACE2 Score and Metaprogram Expression"))

# Save summary
write_csv(correlation_summary, file.path(output_dir, "cytotrace2_mp_correlation_summary.csv"))
cat("\nSaved to:", file.path(output_dir, "cytotrace2_mp_correlation_summary.csv"), "\n")
```

### Combined Correlation Matrix

```{r correlation_matrix, fig.width = 8, fig.height = 6}
# Correlation matrix for all 3 MPs + CytoTRACE2
cor_matrix <- cor(analysis_df[, c("CytoTRACE2_Score", "MP6", "MP9", "MP45")],
                  method = "spearman", use = "pairwise.complete.obs")

# Create heatmap
library(pheatmap)
pheatmap(cor_matrix,
         display_numbers = TRUE,
         number_format = "%.3f",
         main = "Spearman Correlation Matrix:\nCytoTRACE2 vs Metaprograms",
         fontsize = 12,
         fontsize_number = 10)
```

## Session Information

```{r session_info}
sessionInfo()
```
