---
title: "UM Atlas - Cancer Gene Module Analysis with GeneNMF"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Methods (GeneNMF module discovery on malignant cells; 67 cancer metaprograms)
```

#  {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

This report applies **GeneNMF** (Non-Negative Matrix Factorization) to discover **recurrent cancer gene programs** (meta-programs) in malignant melanoma cells from the UM atlas.

**Background:**

GeneNMF is a computational method for extracting recurrent gene programs in cancer cells across multiple studies. Inspired by previous work on NMF for scRNA-seq data (Kotliar et al. 2019, Barkely et al. 2022, Gavish et al. 2023), GeneNMF decomposes the expression matrix into interpretable dimensions and calculates "meta-programs" (MPs) defined as consensus gene sets found across multiple samples and initializations.

**Reference:** [Characterizing intratumor heterogeneity with GeneNMF](https://carmonalab.github.io/GeneNMF.demo/NMF_demo_cancercells.html)

**Analysis approach:**
- Apply NMF to malignant melanoma cells from each sample/patient
- Split data by sample and run multiNMF with k=4:9 (testing different ranks)
- Extract consensus meta-programs across all samples
- **Why k=5 meta-programs?** Based on the GeneNMF metrics, 5 meta-programs were selected because:
  1. They capture the major sources of biological variation (sample coverage > 0.7)
  2. Silhouette scores indicate good cluster separation (> 0.1 is acceptable)
  3. Mean similarity values show coherence within each meta-program
  4. This number balances interpretability with comprehensiveness - too few would miss important biology, too many would fragment related programs
- Analyze program activity by mutation status, organ, origin

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)

# Load libraries
library(Seurat)
library(qs2)
library(qs)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggridges)
library(ggpubr)
library(ggsignif)
library(viridis)
library(RColorBrewer)

# bbknn for batch-corrected UMAP
library(bbknnR)

# UCell for scoring gene signatures
library(UCell)

# Excel export
library(writexl)

# GeneNMF package
library(GeneNMF)

# Load Victoria's official color palettes
source("colors.R")

# Set output directory
output_dir <- "Outputs/GeneNMF/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Set parallel processing
library(future)
plan("multicore", workers = 16)
options(future.globals.maxSize = 50000 * 1024^2)  # Increased for bbknn (was 8000, now 50000 MiB)

# Increased base font size
theme_set(theme_minimal(base_size = 14))
```

## Load Data {.tabset}

### Load UM Atlas

```{r load_um_atlas}
# Load UM atlas
um_atlas <- qs::qread("../13_final_Dataset_reclustered.qs", nthreads = 16)

# Fix GANQ -> GNAQ typo
um_atlas$primary_mutation[um_atlas$primary_mutation == "GANQ"] <- "GNAQ"

cat("UM Atlas loaded:\n")
cat("  - Cells:", ncol(um_atlas), "\n")
cat("  - Features:", nrow(um_atlas), "\n\n")

# Filter to malignant melanoma cells only
malignant <- subset(um_atlas, subset = majority_celltype == "Melanoma cells")

cat("Malignant melanoma cells:\n")
cat("  - Total cells:", ncol(malignant), "\n")
cat("  - Percentage of atlas:", round(100 * ncol(malignant) / ncol(um_atlas), 2), "%\n\n")

# Display distributions
cat("Origin distribution:\n")
print(table(malignant$origin))

cat("\n\nPrimary mutation distribution:\n")
print(table(malignant$primary_mutation))

cat("\n\nSecondary mutation distribution:\n")
print(table(malignant$secondary_mutation))

cat("\n\nOrgan distribution:\n")
print(table(malignant$organ))

cat("\n\nSamples:", length(unique(malignant$orig.ident)), "\n")
```

## Data Preparation for GeneNMF {.tabset}

### Subset and Normalize

GeneNMF works best with normalized count data. We'll use the SCTransform normalized data from our atlas.

```{r data_prep}
# Extract only malignant cells and work with a manageable subset
# For NMF, we'll use cells from key samples

# Select samples with sufficient cells
sample_counts <- table(malignant$orig.ident)
selected_samples <- names(sample_counts[sample_counts > 500])

cat("Selected samples with > 500 cells:", length(selected_samples), "\n")
print(head(selected_samples))

# Subset to selected samples
malignant_sub <- subset(malignant, subset = orig.ident %in% selected_samples)

cat("\nSubset size:", ncol(malignant_sub), "cells\n")

# Get the normalized data matrix
# GeneNMF expects a genes x cells matrix
DefaultAssay(malignant_sub) <- "RNA"

# Use variable features for NMF (most informative genes)
variable_features <- VariableFeatures(malignant_sub)
if (length(variable_features) > 3000) {
  variable_features <- variable_features[1:3000]
}

cat("Using", length(variable_features), "variable features for NMF\n")

# Extract normalized counts - Seurat v5 compatible
# Use log-normalized data from the "data" layer
data_matrix <- LayerData(malignant_sub, assay = "RNA", layer = "data")
data_matrix <- data_matrix[variable_features, , drop = FALSE]

# Convert to dense matrix if sparse
if (inherits(data_matrix, "dgCMatrix") ||
    inherits(data_matrix, "Matrix")) {
  data_matrix <- as.matrix(data_matrix)
}

cat("Data matrix dimensions:", nrow(data_matrix), "x", ncol(data_matrix), "\n")

# Note: We'll use the original UMAP from the malignant object for comparison
# The original UMAP coordinates will be extracted during visualization
```

### Prepare Sample Metadata

```{r prepare_metadata}
# Create sample metadata for GeneNMF
sample_meta <- data.frame(
  sample_id = malignant_sub$orig.ident,
  patient = malignant_sub$Patient_nr,
  origin = malignant_sub$origin,
  primary_mutation = malignant_sub$primary_mutation,
  secondary_mutation = malignant_sub$secondary_mutation,
  organ = malignant_sub$organ,
  stringsAsFactors = FALSE
)

# Remove duplicates
sample_meta <- unique(sample_meta)

cat("Number of samples for NMF:", nrow(sample_meta), "\n")
print(head(sample_meta))
```

### Batch-Corrected UMAP with bbknn

```{r bbknn_umap, fig.width = 12, fig.height = 10}
# Run bbknn for batch-corrected neighbor graph and UMAP
# This corrects for sample/batch effects while preserving biological variation

# Cache file to avoid re-computing bbknn on every render
bbknn_cache_file <- "../Outputs/GeneNMF/malignant_sub_bbknn.qs2"

if (file.exists(bbknn_cache_file)) {
  cat("Loading cached bbknn results...\n")
  malignant_sub <- qs2::qs_read(bbknn_cache_file, nthreads = 16)
  cat("Loaded from cache:", bbknn_cache_file, "\n")
} else {
  cat("Running bbknn for batch-corrected UMAP...\n")
  cat("  - Batch key: orig.ident\n")
  cat("  - Cells:", ncol(malignant_sub), "\n")

  # Run bbknn - this creates a batch-corrected UMAP
  malignant_sub <- RunBBKNN(malignant_sub, batch_key = "orig.ident")

  # Save to cache for faster re-renders
  cat("Saving bbknn results to cache...\n")
  qs2::qs_save(malignant_sub, bbknn_cache_file, nthreads = 16)
  cat("Saved to:", bbknn_cache_file, "\n")
}

cat("bbknn UMAP computed!\n")
cat("Available reductions:\n")
print(names(malignant_sub@reductions))

# Check which reduction was created (may be "bbknn" or "bbknn.UMAP")
bbknn_reduction <- if ("bbknn.UMAP" %in% names(malignant_sub@reductions)) {
  "bbknn.UMAP"
} else if ("bbknn" %in% names(malignant_sub@reductions)) {
  "bbknn"
} else {
  "umap"  # Fallback to default umap
}

cat("Using reduction:", bbknn_reduction, "\n")

# Visualize bbknn UMAP with sample annotation
p_bbknn_samples <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                           group.by = "orig.ident", pt.size = 0.5,
                           cols = NULL, raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Batch-Corrected UMAP - Colored by Sample",
       subtitle = "Batch correction applied across samples")

p_bbknn_origin <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                          group.by = "origin", pt.size = 0.5,
                          cols = origin_colors, raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Batch-Corrected UMAP - Colored by Origin")

# Combine plots
p_bbknn_combined <- p_bbknn_samples | p_bbknn_origin
print(p_bbknn_combined)

# Save individual plots
ggsave(file.path(output_dir, "bbknn_umap_by_sample.png"),
       p_bbknn_samples, width = 12, height = 10, dpi = 300, bg = "white")
ggsave(file.path(output_dir, "bbknn_umap_by_origin.png"),
       p_bbknn_origin, width = 12, height = 10, dpi = 300, bg = "white")
```

## Data Exploration {.tabset}

### Sample Sizes

```{r sample_sizes, fig.width = 12, fig.height = 6}
# Count cells per sample
sample_sizes <- table(malignant_sub$orig.ident)
sample_sizes_df <- data.frame(
  Sample = names(sample_sizes),
  nCells = as.numeric(sample_sizes),
  stringsAsFactors = FALSE
)

# Add metadata
sample_sizes_df <- sample_sizes_df %>%
  left_join(distinct(sample_meta, sample_id, origin, primary_mutation,
                     secondary_mutation, organ),
            by = c("Sample" = "sample_id"))

# Order by cell count
sample_sizes_df <- sample_sizes_df[order(sample_sizes_df$nCells, decreasing = TRUE), ]

# Plot
p_samples <- ggplot(sample_sizes_df, aes(x = reorder(Sample, nCells), y = nCells, fill = origin)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = origin_colors) +
  coord_flip() +
  labs(title = "Cell Counts per Sample (Selected for GeneNMF)",
       subtitle = paste("Total:", nrow(sample_sizes_df), "samples |",
                        sum(sample_sizes_df$nCells), "cells"),
       x = "Sample",
       y = "Number of Cells",
       fill = "Origin") +
  theme_minimal() +
  theme(legend.position = "top")

print(p_samples)

ggsave(file.path(output_dir, "sample_sizes.png"),
       p_samples, width = 12, height = 6, dpi = 300, bg = "white")
```

### Mutation Distribution

```{r mutation_dist, fig.width = 14, fig.height = 6}
# Prepare mutation data
mut_data <- malignant_sub@meta.data %>%
  select(orig.ident, primary_mutation, secondary_mutation, origin) %>%
  distinct() %>%
  filter(!primary_mutation %in% c("unknown", "none")) %>%
  filter(!secondary_mutation %in% c("unknown", "none"))

# Primary mutation by origin
p_primary <- ggplot(mut_data, aes(x = primary_mutation, fill = origin)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = origin_colors) +
  labs(title = "Primary Mutation Distribution by Origin",
       x = "Primary Mutation",
       y = "Number of Samples",
       fill = "Origin") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Secondary mutation by origin
p_secondary <- ggplot(mut_data, aes(x = secondary_mutation, fill = origin)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = origin_colors) +
  labs(title = "Secondary Mutation Distribution by Origin",
       x = "Secondary Mutation",
       y = "Number of Samples",
       fill = "Origin") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Combine
p_mut <- p_primary | p_secondary
print(p_mut)

ggsave(file.path(output_dir, "mutation_distribution.png"),
       p_mut, width = 14, height = 6, dpi = 300, bg = "white")
```

### Feature Expression Distribution

```{r feature_dist, fig.width = 12, fig.height = 8}
# Sample some genes for visualization
set.seed(42)
sample_genes <- sample(variable_features, 9)

# Get expression data
expr_data <- LayerData(malignant_sub, assay = "RNA", layer = "data")[sample_genes, ]
expr_long <- as.data.frame(t(expr_data)) %>%
  pivot_longer(cols = all_of(sample_genes), names_to = "Gene", values_to = "Expression") %>%
  mutate(Expression = log1p(Expression))

# Plot
p_expr <- ggplot(expr_long, aes(x = Gene, y = Expression, fill = Gene)) +
  geom_violin(scale = "width", alpha = 0.8) +
  scale_fill_viridis(discrete = TRUE, option = "plasma") +
  coord_flip() +
  labs(title = "Expression Distribution of Sample Variable Features",
       subtitle = "Log1p normalized counts",
       x = "Gene",
       y = "Log1p(Expression)") +
  theme_minimal() +
  theme(legend.position = "none")

print(p_expr)

ggsave(file.path(output_dir, "feature_distribution.png"),
       p_expr, width = 12, height = 8, dpi = 300, bg = "white")
```

### Data Summary Table

```{r data_summary}
# Create summary table
summary_table <- data.frame(
  Category = c("Total Cells", "Total Genes", "Variable Features",
               "Selected Samples", "Primary Tumors", "Metastatic Tumors",
               "Patients", "Organs"),
  Value = c(ncol(malignant_sub),
            nrow(malignant_sub),
            length(variable_features),
            nrow(sample_meta),
            sum(sample_meta$origin == "primary"),
            sum(sample_meta$origin == "metastasis"),
            length(unique(sample_meta$patient)),
            length(unique(sample_meta$organ)))
)

knitr::kable(summary_table,
             caption = "Data Summary for GeneNMF Analysis (Malignant Cells Only)",
             col.names = c("Category", "Value"),
             format = "html")
```

### Verify Malignant-Only Filtering

```{r verify_malignant, fig.width = 10, fig.height = 6}
# Verify we only have malignant cells
cell_type_check <- table(malignant_sub$majority_celltype)
cat("Cell type verification (should be only Melanoma cells):\n")
print(cell_type_check)

# Visual confirmation
p_verify <- data.frame(CellType = names(cell_type_check),
                       nCells = as.numeric(cell_type_check)) %>%
  ggplot(aes(x = CellType, y = nCells, fill = CellType)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Melanoma cells" = "#B00068")) +
  labs(title = "Verification: Malignant Cells Only",
       subtitle = "GeneNMF analysis uses ONLY malignant melanoma cells (no healthy melanocytes)",
       x = "Cell Type",
       y = "Number of Cells") +
  theme_minimal() +
  theme(legend.position = "none")

print(p_verify)
```

## Comparison with Weizmann 3CA Metaprograms {.tabset}

### Load 3CA Metaprogram Scores

The 3CA (Weizmann) metaprograms were previously computed using UCell. This section loads those scores for comparison with future GeneNMF results.

```{r load_3ca}
# Load existing 3CA metaprogram scores from malignant cells
ca3_path <- "../malignant_metaprograms.qs2"

if (file.exists(ca3_path)) {
  cat("Loading 3CA metaprogram scores...\n")
  ca3_obj <- qs::qread(ca3_path, nthreads = 4)

  # Get 3CA UCell score columns
  ca3_mp_cols <- grep("^MP[0-9]+ ", colnames(ca3_obj@meta.data), value = TRUE)

  cat("  - Available 3CA metaprograms:", length(ca3_mp_cols), "\n")
  cat("  - Cells in 3CA object:", ncol(ca3_obj), "\n")

  # Check cell types in 3CA object
  cat("\nCell type distribution in 3CA object:\n")
  print(table(ca3_obj$majority_celltype))

} else {
  cat("3CA metaprogram file not found.\n")
  ca3_obj <- NULL
  ca3_mp_cols <- NULL
}
```

### 3CA Metaprogram Activity by Origin

```{r ca3_by_origin, fig.width = 14, fig.height = 10, eval = !is.null(ca3_obj)}
if (!is.null(ca3_obj) && !is.null(ca3_mp_cols)) {

  # Subset 3CA object to match our samples
  common_cells <- intersect(colnames(malignant_sub), colnames(ca3_obj))
  cat("Common cells between datasets:", length(common_cells), "\n")

  # Get top 3CA metaprograms by variance
  ca3_scores <- ca3_obj@meta.data[common_cells, ca3_mp_cols]
  mp_var <- apply(ca3_scores, 2, var, na.rm = TRUE)
  top_mps <- names(sort(mp_var, decreasing = TRUE))[1:12]

  # Prepare data for plotting
  ca3_meta <- ca3_obj@meta.data[common_cells,
                                c("orig.ident", "origin", "primary_mutation",
                                  "secondary_mutation", top_mps)]

  ca3_long <- ca3_meta %>%
    pivot_longer(cols = all_of(top_mps),
                 names_to = "MetaProgram",
                 values_to = "Score")

  # Plot by origin
  p_ca3_origin <- ggplot(ca3_long, aes(x = origin, y = Score, fill = origin)) +
    geom_boxplot(outlier.alpha = 0.3) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    facet_wrap(~ MetaProgram, ncol = 4, scales = "free_y") +
    scale_fill_manual(values = origin_colors) +
    labs(title = "3CA Weizmann Metaprogram Activity by Origin",
         subtitle = "Top 12 most variable metaprograms in malignant UM cells",
         x = "Origin",
         y = "UCell Score") +
    theme_minimal() +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_ca3_origin)

  ggsave(file.path(output_dir, "ca3_metaprograms_by_origin.png"),
         p_ca3_origin, width = 14, height = 10, dpi = 300, bg = "white")
}
```

### 3CA Metaprogram Heatmap

```{r ca3_heatmap_simple, fig.width = 12, fig.height = 10, eval = !is.null(ca3_obj)}
if (!is.null(ca3_obj) && !is.null(ca3_mp_cols)) {

  # Create summary by sample
  ca3_meta <- ca3_obj@meta.data[common_cells,
                                c("orig.ident", "origin", top_mps)]

  sample_means <- ca3_meta %>%
    group_by(orig.ident, origin) %>%
    summarise(across(all_of(top_mps), mean, na.rm = TRUE), .groups = "drop")

  # Scale for heatmap
  mat_for_heat <- sample_means %>%
    column_to_rownames(var = "orig.ident") %>%
    select(-origin) %>%
    as.matrix()

  mat_scaled <- t(scale(t(mat_for_heat)))

  # Annotation
  annot <- data.frame(row.names = colnames(mat_scaled),
                      Origin = sample_means$origin[match(colnames(mat_scaled),
                                                         sample_means$orig.ident)])

  # Heatmap
  p_heat <- Heatmap(
    mat_scaled,
    name = "Z-score",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 8),
    col = viridis(100),
    top_annotation = HeatmapAnnotation(
      Origin = annot$Origin,
      col = list(Origin = origin_colors)
    ),
    column_title = "Sample",
    row_title = "3CA Metaprogram",
    heatmap_legend_param = list(title = "Z-score")
  )

  print(p_heat)

  # Save
  png(file.path(output_dir, "ca3_metaprogram_heatmap.png"),
      width = 12, height = 10, units = "in", res = 300)
  print(p_heat)
  dev.off()
}
```

## NMF-Based Gene Module Discovery {.tabset}

### Prepare Data for GeneNMF

GeneNMF requires running NMF separately per sample (patient/donor) and then finding consensus meta-programs across all samples. This approach captures recurrent gene programs that are shared across patients.

```{r prepare_genenmf_data}
# Prepare data for GeneNMF workflow
# We need to split by sample and run NMF per sample

set.seed(42)

# Select samples with sufficient cells for NMF
min_cells_per_sample <- 300
sample_counts <- table(malignant$orig.ident)
samples_for_nmf <- names(sample_counts[sample_counts >= min_cells_per_sample])

cat("Samples with sufficient cells (>", min_cells_per_sample, "):", length(samples_for_nmf), "\n")
print(sample_counts[samples_for_nmf])

# Subset to selected samples
malignant_for_nmf <- subset(malignant, subset = orig.ident %in% samples_for_nmf)

cat("\nSubset size:", ncol(malignant_for_nmf), "cells from", length(samples_for_nmf), "samples\n")

# Use HVGs for NMF (GeneNMF works best with highly variable genes)
hvgs <- VariableFeatures(malignant_for_nmf)
if (length(hvgs) > 2000) {
  hvgs <- hvgs[1:2000]
}

cat("Using", length(hvgs), "highly variable genes for NMF\n")
```

### Split by Sample and Run multiNMF

This is the core GeneNMF workflow:
1. Split the Seurat object by sample (donor)
2. Run multiNMF across all samples with multiple k values
3. Extract consensus meta-programs using getMetaPrograms

```{r run_multiNMF}
# Cache file for GeneNMF results
genenmf_cache_file <- file.path(output_dir, "geneNMF_metaprograms.qs2")

if (file.exists(genenmf_cache_file)) {
  cat("Loading cached GeneNMF results...\n")
  geneNMP_list <- qs2::qs_read(genenmf_cache_file, nthreads = 16)
  geneNMF_metaprograms <- geneNMP_list$metaprograms
  geneNMF_programs <- geneNMP_list$programs
  cat("Loaded from cache\n")
} else {
  cat("Running GeneNMF workflow...\n")
  cat("Step 1: Splitting by sample (donor)\n")

  # Split by sample (orig.ident represents patient samples)
  sampleObj.list <- Seurat::SplitObject(malignant_for_nmf, split.by = "orig.ident")

  cat("  - Split into", length(sampleObj.list), "samples\n")

  # Subsample cells per sample for faster computation
  max_cells_per_sample <- 500
  sampleObj.list <- lapply(sampleObj.list, function(obj) {
    if (ncol(obj) > max_cells_per_sample) {
      cells_to_keep <- sample(colnames(obj), max_cells_per_sample)
      return(subset(obj, cells = cells_to_keep))
    } else {
      return(obj)
    }
  })

  cat("  - After subsampling (max", max_cells_per_sample, "cells/sample):\n")
  cell_counts <- sapply(sampleObj.list, ncol)
  print(summary(cell_counts))

  cat("\nStep 2: Running multiNMF with k=4:9\n")
  cat("  - This runs NMF for each k value and finds consensus programs\n")

  # Run multiNMF - this runs NMF across all samples with multiple k values
  # Note: multiNMF accepts assay, k, and min.exp parameters
  geneNMF_programs <- GeneNMF::multiNMF(
    sampleObj.list,
    assay = "RNA",
    k = 4:9,
    min.exp = 0.05
  )

  cat("\nStep 3: Extracting meta-programs with getMetaPrograms\n")
  cat("  - nMP = 5 meta-programs\n")

  # Extract consensus meta-programs
  geneNMF_metaprograms <- GeneNMF::getMetaPrograms(
    geneNMF_programs,
    nMP = 5
  )

  cat("\nGeneNMF completed!\n")
  cat("  - Meta-programs discovered:", length(geneNMF_metaprograms$metaprograms.genes), "\n")

  # Save results to cache
  geneNMP_list <- list(
    metaprograms = geneNMF_metaprograms,
    programs = geneNMF_programs
  )
  qs2::qs_save(geneNMP_list, genenmf_cache_file, nthreads = 16)
  cat("Saved to cache:", genenmf_cache_file, "\n")
}
```

### Meta-Program Metrics and Statistics

This section displays the key quality metrics for each meta-program (from GeneNMF output):
- **Sample Coverage**: Fraction of samples where this MP is detected
- **Silhouette Score**: Coherence/separation of the MP (higher = better defined)
- **Mean Similarity**: Average pairwise similarity between programs contributing to this MP
- **Number Genes**: Size of the meta-program gene signature
- **Number Programs**: Number of individual NMF programs that make up this MP

```{r mp_metrics, fig.width = 14, fig.height = 10}
# Get meta-program names and count
n_mp <- length(geneNMF_metaprograms$metaprograms.genes)
mp_names <- names(geneNMF_metaprograms$metaprograms.genes)

# Extract metrics directly from GeneNMF output
mp_metrics <- geneNMF_metaprograms$metaprograms.metrics

cat("Meta-Programs Summary\n")
cat("====================\n\n")
print(mp_metrics)

cat("\n\n")
cat("Metric Interpretations:\n")
cat("- sampleCoverage: Fraction of samples where this MP was detected\n")
cat("- silhouette: Internal coherence (higher = better, negative = poor)\n")
cat("- meanSimilarity: Average similarity between contributing programs\n")
cat("- numberGenes: Size of the MP gene signature\n")
cat("- numberPrograms: How many individual NMF programs form this MP\n\n")

# Display metrics table as HTML
knitr::kable(mp_metrics,
             caption = "Meta-Program Quality Metrics (from GeneNMF)",
             digits = 3,
             format = "html")

# Plot metrics
metrics_long <- mp_metrics %>%
  rownames_to_column(var = "MetaProgram") %>%
  select(MetaProgram, sampleCoverage, silhouette, meanSimilarity) %>%
  pivot_longer(cols = c(sampleCoverage, silhouette, meanSimilarity),
               names_to = "Metric", values_to = "Value")

p_metrics <- ggplot(metrics_long, aes(x = MetaProgram, y = Value, fill = MetaProgram)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Metric, scales = "free_y", ncol = 3) +
  scale_fill_viridis(discrete = TRUE) +
  labs(title = "Meta-Program Quality Metrics (from GeneNMF)",
       subtitle = "Sample Coverage, Silhouette Score, and Mean Similarity",
       x = "Meta-Program",
       y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

print(p_metrics)

ggsave(file.path(output_dir, "genenmf_metrics.png"),
       p_metrics, width = 14, height = 10, dpi = 300, bg = "white")
```

### Gene Contribution Weight Barplots

Each meta-program is defined by a set of genes with contribution weights. These barplots show the top 10 genes with highest weights for each meta-program (using actual weights from GeneNMF).

```{r gene_weight_barplots, fig.width = 16, fig.height = 12}
# Create barplots of gene weights for each MP using actual weights from GeneNMF
plot_list <- list()

# Check if weights are available
if (!is.null(geneNMF_metaprograms$metaprograms.genes.weights)) {
  for (mp_name in names(geneNMF_metaprograms$metaprograms.genes)) {
    # Get genes and their weights for this MP
    mp_weights <- geneNMF_metaprograms$metaprograms.genes.weights[[mp_name]]

    # Sort by weight (descending) and take top 10
    top_weights <- head(sort(mp_weights, decreasing = TRUE), 10)

    # Create data frame for plotting
    gene_df <- data.frame(
      Gene = factor(names(top_weights), levels = rev(names(top_weights))),
      Weight = as.numeric(top_weights)
    )

    p <- ggplot(gene_df, aes(x = Gene, y = Weight, fill = Weight)) +
      geom_bar(stat = "identity") +
      scale_fill_viridis() +
      coord_flip() +
      labs(title = paste0(mp_name, " - Top 10 Gene Contributions"),
           x = "Gene",
           y = "Contribution Weight") +
      theme_minimal() +
      theme(legend.position = "none")

    plot_list[[mp_name]] <- p
  }

  # Combine all barplots
  p_weights <- wrap_plots(plot_list, ncol = 3)
  print(p_weights)

  ggsave(file.path(output_dir, "genenmf_gene_weights.png"),
         p_weights, width = 16, height = 12, dpi = 300, bg = "white")
} else {
  cat("Gene weights not available in GeneNMF output\n")
  cat("Showing top genes by rank instead:\n")

  for (mp_name in names(geneNMF_metaprograms$metaprograms.genes)) {
    genes <- geneNMF_metaprograms$metaprograms.genes[[mp_name]]
    top_10 <- head(genes, 10)

    cat(paste0("\n", mp_name, ":\n"))
    print(top_10)
  }
}
```

### Module Gene Sets

Display top 10 genes for all meta-programs discovered by GeneNMF:

```{r module_genes}
# GeneNMF metaprogram genes are already in the correct format
# Display top 10 genes for ALL meta-programs in the report
cat("\n\n## GeneNMF Meta-Program Gene Signatures\n\n")

for (i in 1:n_mp) {
  mp_name <- mp_names[i]
  genes <- geneNMF_metaprograms$metaprograms.genes[[i]]
  top_10 <- head(genes, 10)

  cat(paste0("### **", mp_name, "**\n\n"))
  for (j in 1:length(top_10)) {
    cat(paste0(j, ". ", top_10[j], "\n"))
  }
  cat("\n")
}

# Export gene lists to Excel
excel_path <- file.path(output_dir, "GeneNMF_MetaProgram_GeneLists.xlsx")

# Create a list of data frames for export
module_export_list <- list()
for (i in 1:n_mp) {
  mp_name <- mp_names[i]
  genes <- geneNMF_metaprograms$metaprograms.genes[[i]]

  module_export_list[[mp_name]] <- data.frame(
    Gene = genes,
    Rank = seq_along(genes)
  )
}

# Write to Excel
writexl::write_xlsx(module_export_list, excel_path)

cat("\nGene lists exported to:", excel_path, "\n")
```

### Download Gene Lists

**Download:** [GeneNMF Meta-Program Gene Lists (Excel)](../../Outputs/GeneNMF/GeneNMF_MetaProgram_GeneLists.xlsx)

*Note: This link will work after the HTML is uploaded to gstore.*

### Biological Interpretation of the 5 Meta-Programs

Based on the top contributing genes for each NMF factor, here is the biological annotation for the 5 Meta-Programs (MPs) discovered in this Uveal Melanoma (UM) single-cell dataset.

In single-cell tumor analysis, these programs typically split into **cell-intrinsic identity states** (differentiation levels) and **generic cellular states** (stress, cycling, metabolism).

#### Summary Table

| MP | Label | Key Biological Process |
|----|-------|------------------------|
| **MP1** | AP-1 / Immediate Early Genes | Transcriptional activation, MAPK signaling response, dissociation stress |
| **MP2** | Differentiation | Neural crest identity, cell adhesion, tumor suppressor expression (WWOX) |
| **MP3** | Heat Shock Response | Proteotoxic stress response, protein folding (HSP70/40) |
| **MP4** | OxPhos | Mitochondrial respiration, energy production (ATP synthase, Complex III) |
| **MP5** | Translation | Ribosome assembly, protein synthesis, cell growth |

---

#### MP1: AP-1 / Immediate Early Genes (Transcriptional Activation)

**Top Genes:** ZFP36, DUSP1, EGR1, JUNB, FOSB, JUN, GADD45B

**Annotation:** This program represents the AP-1 transcription factor network and Immediate Early Genes (IEGs).

**Biological Context:**
- These genes are rapidly induced in response to a wide variety of cellular stimuli, including growth factors and stress.
- In single-cell data, this signature can sometimes reflect tissue dissociation artifacts ("dissociation stress"), but in Uveal Melanoma, the AP-1 pathway is also **biologically relevant** as a downstream effector of the GNAQ/GNA11 driver mutations (which activate MAPK/ERK signaling).
- DUSP1 is a negative regulator of ERK, often upregulated to control this pathway.

---

#### MP2: Neural Crest / Melanocytic Differentiation (Tumor Identity)

**Top Genes:** MDGA2, WWOX, AGAP1, NPIPB13, RAD51B, LDLRAD4, AFF3

**Annotation:** This appears to be the core Uveal Melanoma Identity / Differentiation program.

**Biological Context:**
- Unlike the other MPs, these genes are not generic metabolic or stress markers.
- **WWOX** (WW Domain Containing Oxidoreductase) is a well-known tumor suppressor located on Chromosome 16q, involved in neural crest development.
- **MDGA2** and **AGAP1** are highly expressed in neural tissues. Since uveal melanocytes are derived from the neural crest, this likely represents a more "differentiated" or "neural-like" state.
- In UM, loss of differentiation (dedifferentiation) is associated with higher risk (Class 2/Monosomy 3). If this MP correlates with "Class 1" tumors, it represents the lower-risk, differentiated state.

---

#### MP3: Heat Shock Response / Proteotoxic Stress

**Top Genes:** JUN, HSPA1B, HSPA1A, JUNB, DNAJB1, IER2, PPP1R15A

**Annotation:** Cellular Stress / Unfolded Protein Response (UPR).

**Biological Context:**
- While it shares JUN with MP1, this program is dominated by Heat Shock Proteins (HSPs) like **HSPA1A/B** (Hsp70) and **DNAJB1** (Hsp40).
- This indicates cells undergoing **proteotoxic stress** (problems with protein folding).
- This is distinct from the transcriptional activation in MP1; these cells are actively trying to prevent protein aggregation, which can be a survival mechanism in the dense tumor microenvironment.
- **PPP1R15A** (GADD34) is a key regulator of the integrated stress response.

---

#### MP4: Oxidative Phosphorylation (OxPhos)

**Top Genes:** UQCR11, RPL35A, RPS5, UQCRQ, TOMM7, ATP5F1E, POLR2L

**Annotation:** Mitochondrial / Oxidative Phosphorylation (OxPhos).

**Biological Context:**
- The presence of **UQCR** genes (Complex III of the electron transport chain) and **ATP5F1E** (ATP Synthase) clearly points to mitochondrial respiration.
- Uveal Melanoma is known to rely heavily on **OxPhos** (unlike some other cancers that switch to glycolysis/Warburg effect).
- High OxPhos activity can sometimes be linked to metastatic potential or drug resistance in melanoma.
- **TOMM7** is part of the mitochondrial protein import machinery, further supporting mitochondrial activity.

---

#### MP5: Translation / Ribosome Biogenesis

**Top Genes:** RPS5, RPS10, RPL35A, RPL21, RPS14, EEF1B2

**Annotation:** Protein Synthesis / Ribosome Biogenesis.

**Biological Context:**
- This list is almost exclusively composed of **Ribosomal Proteins** (RPS/RPL) and translation elongation factors (EEF1B2).
- This program reflects **high biosynthetic activity**.
- It often correlates with the cell cycle (specifically the G1 phase preparing for division) or cells that are rapidly growing and requiring high protein turnover.
- High translation activity is a hallmark of proliferating cancer cells.

---

### Add Meta-Program Scores to Seurat

```{r add_module_scores}
# Add GeneNMF meta-program scores to the full Seurat object (malignant_sub)
# Using UCell's ScoreSignatures_UCell function for proper continuous scores
# This creates continuous 0-1 scores instead of binary values
# Cache the result since this step is time-consuming

# Define meta-program column names from GeneNMF results
mp_cols <- names(geneNMF_metaprograms$metaprograms.genes)

# Cache file for object with UCell scores
scores_cache_file <- file.path(output_dir, "malignant_sub_with_mp_scores.qs2")

if (file.exists(scores_cache_file)) {
  cat("Loading cached Seurat object with meta-program scores...\n")
  malignant_sub <- qs2::qs_read(scores_cache_file, nthreads = 16)
  cat("Loaded from cache\n")
} else {
  mp.genes <- geneNMF_metaprograms$metaprograms.genes

  # Calculate UCell scores for each meta-program using ScoreSignatures_UCell
  # This returns a matrix with cells x metaprograms, continuous 0-1 scale
  cat("Calculating UCell scores for", length(mp.genes), "meta-programs...\n")

  # Use ScoreSignatures_UCell for better control
  mp_scores <- UCell::ScoreSignatures_UCell(
    malignant_sub,
    features = mp.genes,
    ncores = 4
  )

  # Add scores to metadata
  cat("Adding scores to Seurat object metadata...\n")
  for (mp_name in colnames(mp_scores)) {
    malignant_sub@meta.data[[mp_name]] <- mp_scores[, mp_name]
  }

  cat("Saving object with meta-program scores to cache...\n")
  qs2::qs_save(malignant_sub, scores_cache_file, nthreads = 16)
  cat("Saved to:", scores_cache_file, "\n")
}

cat("Meta-program scores ready\n")
cat("Score columns created:", paste(mp_cols, collapse = ", "), "\n")

# Verify scores are continuous (not binary)
for (mp in mp_cols[1:min(3, length(mp_cols))]) {
  scores <- malignant_sub@meta.data[[mp]]
  cat(paste0(mp, ": range=[", round(min(scores, na.rm=TRUE), 3), ", ",
             round(max(scores, na.rm=TRUE), 3), "], ",
             "unique values=", length(unique(scores)), "\n"))
}
```

### Meta-Program Activity by Origin

```{r nmf_by_origin, fig.width = 14, fig.height = 8}
# Get meta-program columns from ScoreSignatures_UCell
# The columns are named after the metaprograms.genes names
mp_cols <- names(geneNMF_metaprograms$metaprograms.genes)

# Prepare data
module_data <- malignant_sub@meta.data[, c("orig.ident", "origin", "Patient_nr",
                                           "primary_mutation", "secondary_mutation",
                                           "organ", "Gender", mp_cols), drop = FALSE]

module_long <- module_data %>%
  select(-primary_mutation, -secondary_mutation, -organ, -Patient_nr, -Gender) %>%
  pivot_longer(cols = all_of(mp_cols),
               names_to = "MetaProgram",
               values_to = "Score") %>%
  mutate(MetaProgram = as.character(MetaProgram),
         MetaProgram = factor(MetaProgram, levels = mp_cols))

# Explicitly set origin as factor
module_long$origin <- as.factor(module_long$origin)

# Plot by origin with ggsignif statistics
p_nmf_origin <- ggplot(module_long, aes(x = origin, y = Score, fill = origin)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  stat_signif(geom = "text", step_increase = 0.1, tip_length = 0.01,
              test = "wilcox.test", comparisons = list(c("primary", "metastasis")),
              map_signif_level = TRUE) +
  facet_wrap(~ MetaProgram, ncol = 5, scales = "free_y") +
  scale_fill_manual(values = origin_colors) +
  labs(title = "GeneNMF Meta-Program Activity by Origin",
       subtitle = "Meta-programs derived from multi-sample NMF (k=4:9). Wilcoxon test comparing primary vs metastasis.",
       x = "Origin",
       y = "UCell Score (0-1)") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_nmf_origin)

ggsave(file.path(output_dir, "genenmf_metaprograms_by_origin.png"),
       p_nmf_origin, width = 14, height = 8, dpi = 300, bg = "white")
```

### Meta-Program Activity by Primary Mutation

```{r nmf_by_primary_mutation, fig.width = 14, fig.height = 8}
# Prepare data for primary mutation
module_long_mutation <- module_data %>%
  select(-origin, -secondary_mutation, -organ, -Patient_nr, -Gender) %>%
  filter(!primary_mutation %in% c("unknown", "none", "NA")) %>%
  pivot_longer(cols = all_of(mp_cols),
               names_to = "MetaProgram",
               values_to = "Score") %>%
  mutate(MetaProgram = as.character(MetaProgram),
         MetaProgram = factor(MetaProgram, levels = mp_cols),
         primary_mutation = as.factor(primary_mutation))

# Plot by primary mutation
p_nmf_mutation <- ggplot(module_long_mutation, aes(x = primary_mutation, y = Score, fill = primary_mutation)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  stat_signif(geom = "text", step_increase = 0.08, tip_length = 0.01,
              test = "wilcox.test", size = 2.5,
              comparisons = combn(unique(module_long_mutation$primary_mutation), 2, simplify = FALSE)) +
  facet_wrap(~ MetaProgram, ncol = 5, scales = "free_y") +
  scale_fill_manual(values = primary_mutations_col) +
  labs(title = "GeneNMF Meta-Program Activity by Primary Mutation",
       subtitle = "Wilcoxon test comparing mutations. Significance shown for all pairwise comparisons.",
       x = "Primary Mutation",
       y = "UCell Score (0-1)") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_nmf_mutation)

ggsave(file.path(output_dir, "genenmf_metaprograms_by_primary_mutation.png"),
       p_nmf_mutation, width = 14, height = 8, dpi = 300, bg = "white")
```

### Meta-Program Activity by Secondary Mutation

```{r nmf_by_secondary_mutation, fig.width = 14, fig.height = 8}
# Prepare data for secondary mutation
module_long_secondary <- module_data %>%
  select(-origin, -primary_mutation, -organ, -Patient_nr, -Gender) %>%
  filter(!secondary_mutation %in% c("unknown", "none", "NA", "")) %>%
  pivot_longer(cols = all_of(mp_cols),
               names_to = "MetaProgram",
               values_to = "Score") %>%
  mutate(MetaProgram = as.character(MetaProgram),
         MetaProgram = factor(MetaProgram, levels = mp_cols),
         secondary_mutation = as.factor(secondary_mutation))

# Plot by secondary mutation
p_nmf_secondary <- ggplot(module_long_secondary, aes(x = secondary_mutation, y = Score, fill = secondary_mutation)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  stat_signif(geom = "text", step_increase = 0.08, tip_length = 0.01,
              test = "wilcox.test", size = 2.5,
              comparisons = combn(unique(module_long_secondary$secondary_mutation), 2, simplify = FALSE)) +
  facet_wrap(~ MetaProgram, ncol = 5, scales = "free_y") +
  scale_fill_manual(values = second_mutations_col) +
  labs(title = "GeneNMF Meta-Program Activity by Secondary Mutation",
       subtitle = "Wilcoxon test comparing mutations. Significance shown for all pairwise comparisons.",
       x = "Secondary Mutation",
       y = "UCell Score (0-1)") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_nmf_secondary)

ggsave(file.path(output_dir, "genenmf_metaprograms_by_secondary_mutation.png"),
       p_nmf_secondary, width = 14, height = 8, dpi = 300, bg = "white")
```

### Meta-Program Activity by Organ

```{r nmf_by_organ, fig.width = 14, fig.height = 8}
# Prepare data for organ
module_long_organ <- module_data %>%
  select(-origin, -primary_mutation, -secondary_mutation, -Patient_nr, -Gender) %>%
  filter(!organ %in% c("unknown", "none", "NA")) %>%
  pivot_longer(cols = all_of(mp_cols),
               names_to = "MetaProgram",
               values_to = "Score") %>%
  mutate(MetaProgram = as.character(MetaProgram),
         MetaProgram = factor(MetaProgram, levels = mp_cols),
         organ = as.factor(organ))

# Plot by organ
p_nmf_organ <- ggplot(module_long_organ, aes(x = organ, y = Score, fill = organ)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  stat_signif(geom = "text", step_increase = 0.08, tip_length = 0.01,
              test = "wilcox.test", size = 2.5,
              comparisons = combn(unique(module_long_organ$organ), 2, simplify = FALSE)) +
  facet_wrap(~ MetaProgram, ncol = 5, scales = "free_y") +
  scale_fill_manual(values = organ_colors) +
  labs(title = "GeneNMF Meta-Program Activity by Organ",
       subtitle = "Wilcoxon test comparing organs. Significance shown for all pairwise comparisons.",
       x = "Organ",
       y = "UCell Score (0-1)") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_nmf_organ)

ggsave(file.path(output_dir, "genenmf_metaprograms_by_organ.png"),
       p_nmf_organ, width = 14, height = 8, dpi = 300, bg = "white")
```

### Meta-Program Activity by Gender

```{r nmf_by_gender, fig.width = 14, fig.height = 8}
# Prepare data for gender
module_long_gender <- module_data %>%
  select(-origin, -primary_mutation, -secondary_mutation, -organ, -Patient_nr) %>%
  filter(!Gender %in% c("unknown", "none", "NA", "")) %>%
  pivot_longer(cols = all_of(mp_cols),
               names_to = "MetaProgram",
               values_to = "Score") %>%
  mutate(MetaProgram = as.character(MetaProgram),
         MetaProgram = factor(MetaProgram, levels = mp_cols),
         Gender = as.factor(Gender))

# Plot by gender
p_nmf_gender <- ggplot(module_long_gender, aes(x = Gender, y = Score, fill = Gender)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  stat_signif(geom = "text", step_increase = 0.1, tip_length = 0.01,
              test = "wilcox.test",
              comparisons = list(c("M", "F")),
              map_signif_level = TRUE) +
  facet_wrap(~ MetaProgram, ncol = 5, scales = "free_y") +
  scale_fill_manual(values = c("M" = "#1f77b4", "F" = "#d62728")) +
  labs(title = "GeneNMF Meta-Program Activity by Gender",
       subtitle = "Wilcoxon test comparing M vs F.",
       x = "Gender",
       y = "UCell Score (0-1)") +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))

print(p_nmf_gender)

ggsave(file.path(output_dir, "genenmf_metaprograms_by_gender.png"),
       p_nmf_gender, width = 14, height = 8, dpi = 300, bg = "white")
```

### Module Heatmap with Clinical Metadata

```{r nmf_heatmap, fig.width = 14, fig.height = 12}
# Create meta-program heatmap by sample with clinical metadata
# Get the first occurrence of each metadata column per sample
get_first_value <- function(meta_col, sample_ids) {
  sapply(sample_ids, function(sid) {
    idx <- which(malignant_sub$orig.ident == sid)[1]
    vals <- malignant_sub@meta.data[idx, meta_col]
    if (is.factor(vals)) as.character(vals) else vals
  }, USE.NAMES = TRUE)
}

# Calculate mean scores per sample (include all metadata columns in the group_by)
module_means <- module_data %>%
  group_by(orig.ident) %>%
  summarise(across(all_of(mp_cols), mean, na.rm = TRUE), .groups = "drop")

# Extract clinical metadata for each sample (first cell per sample)
module_means$origin <- sapply(module_means$orig.ident, function(sid) {
  idx <- which(malignant_sub$orig.ident == sid)[1]
  as.character(malignant_sub$origin[idx])
})
module_means$primary_mutation <- sapply(module_means$orig.ident, function(sid) {
  idx <- which(malignant_sub$orig.ident == sid)[1]
  as.character(malignant_sub$primary_mutation[idx])
})
module_means$secondary_mutation <- sapply(module_means$orig.ident, function(sid) {
  idx <- which(malignant_sub$orig.ident == sid)[1]
  as.character(malignant_sub$secondary_mutation[idx])
})
module_means$organ <- sapply(module_means$orig.ident, function(sid) {
  idx <- which(malignant_sub$orig.ident == sid)[1]
  as.character(malignant_sub$organ[idx])
})
module_means$Gender <- sapply(module_means$orig.ident, function(sid) {
  idx <- which(malignant_sub$orig.ident == sid)[1]
  as.character(malignant_sub$Gender[idx])
})

# Scale for heatmap
mat_for_heat <- module_means %>%
  column_to_rownames(var = "orig.ident") %>%
  select(-origin, -primary_mutation, -secondary_mutation, -organ, -Gender) %>%
  as.matrix()

mat_scaled <- t(scale(t(mat_for_heat)))

# Clinical annotations for heatmap - use left_annotation for row annotations (samples)
# Order: samples are rows, so we annotate rows
row_anno <- rowAnnotation(
  Origin = module_means$origin,
  `Primary Mutation` = module_means$primary_mutation,
  `Secondary Mutation` = module_means$secondary_mutation,
  Organ = module_means$organ,
  Gender = module_means$Gender,
  col = list(
    Origin = origin_colors,
    `Primary Mutation` = primary_mutations_col,
    `Secondary Mutation` = second_mutations_col,
    Organ = organ_colors,
    Gender = c("M" = "#1f77b4", "F" = "#d62728", "unknown" = "gray50")
  ),
  annotation_legend_param = list(
    Origin = list(title_gp = gpar(fontsize = 10), labels_gp = gpar(fontsize = 9)),
    `Primary Mutation` = list(title_gp = gpar(fontsize = 10), labels_gp = gpar(fontsize = 9)),
    `Secondary Mutation` = list(title_gp = gpar(fontsize = 10), labels_gp = gpar(fontsize = 9)),
    Organ = list(title_gp = gpar(fontsize = 10), labels_gp = gpar(fontsize = 9)),
    Gender = list(title_gp = gpar(fontsize = 10), labels_gp = gpar(fontsize = 9))
  )
)

# Heatmap with clinical row annotations
p_nmf_heat <- Heatmap(
  mat_scaled,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 9),
  col = viridis(100),
  left_annotation = row_anno,
  column_title = "GeneNMF Meta-Program",
  row_title = "Sample",
  heatmap_legend_param = list(title = "Z-score", legend_height = unit(3, "cm"))
)

print(p_nmf_heat)

# Save
png(file.path(output_dir, "genenmf_metaprogram_heatmap_with_clinical.png"),
    width = 14, height = 12, units = "in", res = 300)
print(p_nmf_heat)
dev.off()
```

### Meta-Program Correlation Heatmap

This heatmap shows the pairwise Pearson correlation between the 5 discovered meta-programs based on their UCell scores across all malignant cells. Positive correlations suggest co-active programs that tend to be expressed together, while negative correlations indicate mutually exclusive programs.

```{r mp_correlation_heatmap, fig.width = 8, fig.height = 7}
# Calculate pairwise correlation between MP UCell scores
mp_scores_matrix <- malignant_sub@meta.data[, mp_cols]
mp_correlation <- cor(mp_scores_matrix, use = "pairwise.complete.obs")

# Create correlation heatmap with correlation values displayed
p_mp_cor <- Heatmap(
  mp_correlation,
  name = "Correlation",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10),
  col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.2f", mp_correlation[i, j]), x, y, gp = gpar(fontsize = 9))
  },
  column_title = "Meta-Program Correlation Heatmap",
  row_title = "Meta-Program",
  heatmap_legend_param = list(title = "Pearson\nCorrelation")
)

print(p_mp_cor)

# Save
png(file.path(output_dir, "genenmf_mp_correlation_heatmap.png"),
    width = 8, height = 7, units = "in", res = 300)
print(p_mp_cor)
dev.off()

cat("\nMeta-Program correlation summary:\n")
cat("  - Positive correlations suggest co-active programs\n")
cat("  - Negative correlations suggest mutually exclusive programs\n")
```

### Meta-Program UMAP Projections (bbknn)

```{r nmf_umap, fig.width = 20, fig.height = 16}
# Create UMAP plots for GeneNMF meta-programs using bbknn batch-corrected UMAP
# Using order = TRUE to order cells by expression level
n_plots <- min(length(mp_cols), 10)
plot_list <- list()

# Use the bbknn reduction if available, otherwise use default umap
umap_to_use <- if (exists("bbknn_reduction")) bbknn_reduction else "umap"

for (i in 1:n_plots) {
  p <- FeaturePlot(malignant_sub,
                   features = mp_cols[i],
                   reduction = umap_to_use,
                   pt.size = 0.5,
                   order = TRUE,
                   raster = FALSE) +
    scale_color_viridis_c(option = "D") +
    coord_fixed() +
    theme(legend.position = "right") +
    ggtitle(mp_cols[i])

  plot_list[[i]] <- p
}

p_combined <- wrap_plots(plot_list, ncol = 4)
print(p_combined)

ggsave(file.path(output_dir, "genenmf_metaprograms_bbknn_umap.png"),
       p_combined, width = 20, height = 16, dpi = 300, bg = "white")
```

### Meta-Program UMAP Projections (Original/Unintegrated)

In addition to the batch-corrected bbknn UMAP, it is valuable to visualize the GeneNMF meta-programs on the original, unintegrated UMAP space. The original UMAP preserves the intrinsic structure of each sample without any batch correction, allowing us to observe:

1. **Natural sample separation**: How cells from different patients or samples naturally cluster based on their biological differences, without any algorithmic correction for technical batch effects.

2. **Batch effect assessment**: By comparing the original UMAP with the bbknn-corrected UMAP, we can visually assess the degree of batch correction that has been applied. Strong batch effects are visible when cells cluster primarily by sample/patient on the original UMAP but show more intermingling on the bbknn UMAP.

3. **Meta-program robustness**: Whether the GeneNMF meta-programs maintain coherent spatial patterns across different dimensional reduction spaces. Robust meta-programs should show consistent patterns regardless of whether batch correction has been applied.

4. **Technical vs biological variation**: The original UMAP reflects both biological variation and technical artifacts (batch effects), while the bbknn UMAP primarily emphasizes biological signal after minimizing technical variation.

The comparison between these two UMAP representations provides insight into the balance between preserving true biological heterogeneity and removing unwanted technical variation in our single-cell dataset.

```{r nmf_umap_original, fig.width = 20, fig.height = 16}
# Show GeneNMF meta-programs on the original (unintegrated) UMAP for comparison
# This preserves the intrinsic structure without batch correction
# Use the original malignant object which has the umap reduction
if ("umap" %in% names(malignant@reductions)) {
  # Get common cells between malignant_sub and malignant
  common_cells <- intersect(colnames(malignant_sub), colnames(malignant))

  # Create a temporary object with original UMAP for visualization
  # We need to subset malignant to the same cells as malignant_sub
  malignant_orig_vis <- subset(malignant, cells = common_cells)

  # Copy MP scores from malignant_sub to malignant_orig_vis
  # Use match to ensure correct order
  for (mp_col in mp_cols) {
    malignant_orig_vis@meta.data[[mp_col]] <- malignant_sub@meta.data[[mp_col]][match(colnames(malignant_orig_vis), colnames(malignant_sub))]
  }

  # Remove any NA values from MP scores
  for (mp_col in mp_cols) {
    malignant_orig_vis@meta.data[[mp_col]][is.na(malignant_orig_vis@meta.data[[mp_col]])] <- 0
  }

  n_plots_orig <- min(length(mp_cols), 10)
  plot_list_orig <- list()

  for (i in 1:n_plots_orig) {
    p <- FeaturePlot(malignant_orig_vis,
                     features = mp_cols[i],
                     reduction = "umap",
                     pt.size = 0.5,
                     order = TRUE,
                     raster = FALSE) +
      scale_color_viridis_c(option = "D") +
      coord_fixed() +
      theme(legend.position = "right") +
      ggtitle(paste0(mp_cols[i], " (Original UMAP)"))

    plot_list_orig[[i]] <- p
  }

  p_combined_orig <- wrap_plots(plot_list_orig, ncol = 4)
  print(p_combined_orig)

  ggsave(file.path(output_dir, "genenmf_metaprograms_original_umap.png"),
         p_combined_orig, width = 20, height = 16, dpi = 300, bg = "white")
} else {
  cat("Original UMAP not available in malignant object\n")
}
```

### bbknn Clustering Visualization

```{r bbknn_clustering, fig.width = 16, fig.height = 20}
# Visualize clustering on bbknn embedding
# This shows how cells cluster by different metadata on the batch-corrected space

# First, perform unsupervised clustering on the bbknn embedding
# Check how many dimensions are available
n_dims_bbknn <- ncol(Embeddings(malignant_sub, bbknn_reduction))
dims_to_use <- min(30, n_dims_bbknn)
cat("Using", dims_to_use, "dimensions from", bbknn_reduction, "reduction\n")

malignant_sub <- FindNeighbors(malignant_sub, reduction = bbknn_reduction,
                                dims = 1:dims_to_use, graph.name = "bbknn_nn")
malignant_sub <- FindClusters(malignant_sub, graph.name = "bbknn_nn",
                               resolution = 0.2, algorithm = 4)
# The clustering result is stored as 'bbknn_nn_res.0.2'
bbknn_cluster_col <- paste0("bbknn_nn_res.", 0.2)
cat("Unsupervised clustering completed. Number of clusters:",
    length(unique(malignant_sub@meta.data[[bbknn_cluster_col]])), "\n")

# Clustering by origin
p_cluster_origin <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                            group.by = "origin", pt.size = 0.5,
                            cols = origin_colors, raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Clustering by Origin")

# Clustering by primary mutation
p_cluster_mutation <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                             group.by = "primary_mutation", pt.size = 0.5,
                             cols = primary_mutations_col, raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Clustering by Primary Mutation")

# Clustering by secondary mutation
p_cluster_secondary <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                               group.by = "secondary_mutation", pt.size = 0.5,
                               cols = second_mutations_col, raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Clustering by Secondary Mutation")

# Clustering by organ
p_cluster_organ <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                          group.by = "organ", pt.size = 0.5,
                          cols = organ_colors, raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Clustering by Organ")

# Clustering by gender
p_cluster_gender <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                            group.by = "Gender", pt.size = 0.5,
                            cols = c("M" = "#1f77b4", "F" = "#d62728",
                                     "unknown" = "gray50"), raster = FALSE) +
  coord_fixed() +
  labs(title = "bbknn Clustering by Gender")

# Unsupervised clustering (graph-based on bbknn)
n_clusters <- length(unique(malignant_sub@meta.data[[bbknn_cluster_col]]))
cluster_colors <- scales::hue_pal()(n_clusters)
p_cluster_unsup <- DimPlot(malignant_sub, reduction = bbknn_reduction,
                           group.by = bbknn_cluster_col, pt.size = 0.5,
                           cols = cluster_colors, raster = FALSE) +
  coord_fixed() +
  labs(title = "Unsupervised Clustering (bbknn-based)",
       subtitle = paste("Resolution = 0.2, n clusters =", n_clusters))

# Combine plots in a 3x2 grid
p_cluster_combined <- (p_cluster_origin | p_cluster_mutation) /
                      (p_cluster_secondary | p_cluster_organ) /
                      (p_cluster_gender | p_cluster_unsup)

print(p_cluster_combined)

# Save
ggsave(file.path(output_dir, "bbknn_clustering_by_metadata.png"),
       p_cluster_combined, width = 16, height = 20, dpi = 300, bg = "white")
```

### GeneNMF MetaProgram Similarity Heatmap

This section visualizes the pairwise similarity between individual NMF programs with clear cluster (meta-program) assignments using GeneNMF's native `plotMetaPrograms()` function. The dendrogram shows how programs cluster into the 5 meta-programs, and the annotation track indicates which MP each program belongs to.

```{r genenmf_style_heatmap, fig.width = 12, fig.height = 10}
# Use GeneNMF's plotMetaPrograms function for proper visualization
# This shows the 5 MP clusters clearly with hierarchical clustering tree
p_sim <- plotMetaPrograms(
  geneNMF_metaprograms,
  similarity.cutoff = c(0.1, 1),
  showtree = TRUE,
  main = "GeneNMF Program Similarity with Meta-Program Clusters (k=5)",
  palette = viridis(100, option = "H", direction = 1)
)

print(p_sim)

# Save the plot
png(file.path(output_dir, "genenmf_metaprograms_similarity_heatmap.png"),
    width = 12, height = 10, units = "in", res = 300)
print(p_sim)
dev.off()

cat("\nGeneNMF program similarity heatmap with", n_mp, "meta-program clusters\n")
cat("  - Individual NMF programs:", nrow(geneNMF_metaprograms$programs.similarity), "\n")
cat("  - similarity.cutoff = c(0.1, 1) to emphasize clustering structure\n")
```

## Comparison: NMF vs 3CA Metaprograms {.tabset}

### 3CA Metaprogram Heatmap

```{r ca3_heatmap, fig.width = 12, fig.height = 10, eval = !is.null(ca3_obj)}
if (!is.null(ca3_obj) && !is.null(ca3_mp_cols)) {

  # Create summary by sample with proper metadata extraction
  # Get common cells between datasets
  common_cells_3ca <- intersect(colnames(malignant_sub), colnames(ca3_obj))

  # Extract 3CA scores and metadata for common cells
  ca3_meta <- ca3_obj@meta.data[common_cells_3ca,
                                c("orig.ident", "origin", top_mps)]

  # Calculate mean scores per sample
  sample_ca3_means <- ca3_meta %>%
    group_by(orig.ident) %>%
    summarise(across(all_of(top_mps), mean, na.rm = TRUE), .groups = "drop")

  # Extract origin for each sample (first cell per sample)
  get_first_value_ca3 <- function(meta_col, sample_ids) {
    sapply(sample_ids, function(sid) {
      cell_idx <- which(ca3_obj$orig.ident == sid)[1]
      vals <- ca3_obj@meta.data[cell_idx, meta_col]
      if (is.factor(vals)) as.character(vals) else vals
    }, USE.NAMES = TRUE)
  }

  sample_ca3_means$origin <- get_first_value_ca3("origin", sample_ca3_means$orig.ident)

  # Scale for heatmap
  mat_for_heat_ca3 <- sample_ca3_means %>%
    column_to_rownames(var = "orig.ident") %>%
    select(-origin) %>%
    as.matrix()

  mat_scaled_ca3 <- t(scale(t(mat_for_heat_ca3)))

  # Row annotation for origin
  row_anno_ca3 <- rowAnnotation(
    Origin = sample_ca3_means$origin,
    col = list(Origin = origin_colors)
  )

  # Heatmap with origin annotation
  p_ca3_heat <- Heatmap(
    mat_scaled_ca3,
    name = "Z-score",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 8),
    col = viridis(100),
    left_annotation = row_anno_ca3,
    column_title = "3CA Metaprogram",
    row_title = "Sample",
    heatmap_legend_param = list(title = "Z-score", legend_height = unit(3, "cm"))
  )

  print(p_ca3_heat)

  # Save
  png(file.path(output_dir, "ca3_metaprogram_heatmap.png"),
      width = 12, height = 10, units = "in", res = 300)
  print(p_ca3_heat)
  dev.off()

}
```

### Correlation Between NMF Modules and 3CA Metaprograms

```{r nmf_3ca_correlation, fig.width = 12, fig.height = 10, eval = !is.null(ca3_obj)}
if (!is.null(ca3_obj)) {

  # Get common cells
  common_cells <- intersect(colnames(malignant_sub), colnames(ca3_obj))

  # Extract NMF module scores for common cells
  nmf_scores_common <- malignant_sub@meta.data[common_cells, mp_cols]

  # Extract 3CA metaprogram scores for common cells
  ca3_scores_common <- ca3_obj@meta.data[common_cells, ca3_mp_cols]

  # Calculate correlation matrix
  cor_matrix <- cor(nmf_scores_common, ca3_scores_common, use = "pairwise.complete.obs")

  # Plot heatmap
  p_cor <- Heatmap(
    cor_matrix,
    name = "Correlation",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    row_names_gp = gpar(fontsize = 8),
    column_names_gp = gpar(fontsize = 7),
    col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
    row_title = "GeneNMF Meta-Program",
    column_title = "3CA Metaprogram",
    heatmap_legend_param = list(title = "Correlation"),
    column_names_rot = 45
  )

  print(p_cor)

  # Save
  png(file.path(output_dir, "nmf_3ca_correlation_heatmap.png"),
      width = 12, height = 10, units = "in", res = 300)
  print(p_cor)
  dev.off()

  # Find top correlations
  cor_df <- as.data.frame(as.table(cor_matrix))
  colnames(cor_df) <- c("GeneNMF_MP", "CA3_MP", "Correlation")
  cor_df <- cor_df %>%
    arrange(desc(abs(Correlation)))

  cat("\nTop 10 NMF-3CA correlations:\n")
  print(head(cor_df, 10))

}
```

## Session Information

```{r session}
sessionInfo()
```
