---
title: "UM Atlas - GloScope Full Cohort Analysis (No Downsampling)"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Extended Data (GloScope sample-level transcriptomic divergence: MDS, ANOSIM, divergence heatmaps)
```

#  {.tabset}

## Overview

**GloScope Sample-Level Analysis - Full Cohort (No Downsampling)**

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

### Key Differences from Downsampled Version

- **All cells included:** ~298,640 cells (no downsampling)
- **Parallel processing:** 24 cores via BiocParallel
- **Intermediate caching:** Divergence matrices saved and reused if present
- **Higher memory:** Optimized for large-scale computation

### Dataset Overview

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8)

# Load libraries
library(Seurat)
library(qs)
library(qs2)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(patchwork)
library(GloScope)
library(ComplexHeatmap)
library(circlize)
library(vegan)
library(pals)
library(BiocParallel)
library(DT)

# Load Victoria's official color palettes
source("colors.R")

# Set up parallel processing with 24 cores
bpparam <- MulticoreParam(workers = 24, progressbar = TRUE)
register(bpparam)

# Create output directory
output_dir <- "../GloScope_Full_Outputs/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Safe ANOSIM wrapper: drops samples with NA grouping before calling vegan::anosim
safe_anosim <- function(dist_mat, grouping, permutations = 999) {
  d <- as.matrix(dist_mat)
  n <- nrow(d)
  # Handle dimension mismatch: if grouping shorter than matrix, pad with NAs
  if (length(grouping) < n) {
    grouping <- c(grouping, rep(NA, n - length(grouping)))
  } else if (length(grouping) > n) {
    grouping <- grouping[seq_len(n)]
  }
  keep <- !is.na(grouping)
  if (sum(keep) < 3) {
    warning("Fewer than 3 non-NA samples for ANOSIM; skipping.")
    return(list(statistic = NA, signif = NA))
  }
  d <- d[keep, keep]
  vegan::anosim(as.dist(d), grouping[keep], permutations = permutations)
}

# Set random seed
set.seed(42)

cat("Parallel processing configured: 24 cores\n")
cat("Output directory:", output_dir, "\n")
```

---

## Data Loading & Preparation {.tabset}

### Load UM Atlas

```{r load_data}
cat("Loading Victoria's updated UM Atlas (Jan 2025)...\n")
um_atlas <- qs::qread('../13_final_Dataset_reclustered_28012025.qs',
                      nthreads = 24)

cat("Loaded Seurat object:\n")
cat("  - Cells:", ncol(um_atlas), "\n")
cat("  - Features:", nrow(um_atlas), "\n")
cat("  - Samples:", length(unique(um_atlas$Sample)), "\n")
cat("  - Patients:", length(unique(um_atlas$Patient_nr)), "\n")
```

### Extract Embeddings

```{r extract_embeddings}
# Check available reductions
cat("Available dimensionality reductions:\n")
print(names(um_atlas@reductions))

# Extract PCA embeddings (using first 50 PCs)
if ("pca" %in% names(um_atlas@reductions)) {
  cat("\nUsing existing PCA reduction...\n")
  embedding_matrix <- Embeddings(um_atlas, reduction = "pca")[, 1:50]
} else {
  cat("\nComputing PCA reduction...\n")
  um_atlas <- FindVariableFeatures(um_atlas, nfeatures = 3000)
  um_atlas <- ScaleData(um_atlas)
  um_atlas <- RunPCA(um_atlas, npcs = 50, verbose = FALSE)
  embedding_matrix <- Embeddings(um_atlas, reduction = "pca")[, 1:50]
}

cat("Embedding matrix dimensions:", dim(embedding_matrix), "\n")
cat("Number of PCs used:", ncol(embedding_matrix), "\n")
cat("FULL COHORT - No downsampling applied\n")
```

**Why non-integrated data?**

GloScope is designed to capture **all sources of sample-to-sample variation**:

- **Compositional differences**: Cell type proportions
- **Transcriptional differences**: Gene expression states
- **Biological variation**: Including batch effects that may correlate with biology

Integration methods (Harmony, Seurat integration, scVI) specifically **remove** technical and batch variation. For GloScope, we use the **original PCA embeddings** from non-integrated data to preserve the full spectrum of natural sample differences. This allows GloScope to detect both expected biological variation and potential technical artifacts.

### Prepare Metadata

```{r prepare_metadata}
# Extract cell-level metadata
cell_metadata <- um_atlas@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  dplyr::select(cell_id, Sample, Patient_nr, origin, organ,
                primary_mutation, secondary_mutation, LocationPrimaryTumor,
                majority_celltype, Malignant, Gender, Age_atDiagnosisY) %>%
  dplyr::mutate(
    # Clean up mutation labels
    primary_mutation = ifelse(primary_mutation == "GANQ", "GNAQ", primary_mutation),
    # Create simplified groups
    origin_clean = factor(origin, levels = c("healthy", "primary", "metastasis")),
    secondary_mut_clean = case_when(
      secondary_mutation %in% c("BAP1", "SF3B1", "EIF1AX") ~ secondary_mutation,
      TRUE ~ "none/unknown"
    ),
    location_clean = case_when(
      LocationPrimaryTumor %in% c("choroid", "ciliary body", "iris") ~ LocationPrimaryTumor,
      TRUE ~ "other/NA"
    )
  )

# Create sample-level metadata
sample_metadata <- cell_metadata %>%
  group_by(Sample) %>%
  summarise(
    Patient_nr = first(Patient_nr),
    origin = first(origin_clean),
    organ = first(organ),
    primary_mutation = first(primary_mutation),
    secondary_mutation = first(secondary_mut_clean),
    LocationPrimaryTumor = first(location_clean),
    Gender = first(Gender),
    Age_numeric = as.numeric(first(Age_atDiagnosisY)),  # Numeric age for calculations
    n_cells = n(),
    pct_malignant = 100 * mean(Malignant == "Malignant", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # Bin age into categories for visualization (matching ageGroup_col from config)
    Age_group = case_when(
      is.na(Age_numeric) ~ "Unknown",
      Age_numeric < 40 ~ "Young",
      Age_numeric >= 40 & Age_numeric < 60 ~ "Middle",
      Age_numeric >= 60 ~ "Old",
      TRUE ~ "Unknown"
    )
  ) %>%
  arrange(origin, organ, Sample)

cat("\nSample-level summary:\n")
print(sample_metadata %>% count(origin, name = "n_samples"))

# Save sample metadata
write_csv(sample_metadata, file.path(output_dir, "sample_metadata.csv"))

# Use Victoria's official color palettes (loaded from colors.R)
# Add compatibility aliases for GloScope-specific naming
organ_colors_all <- organ_colors
pmut_colors <- c(primary_mutations_col, "NA" = "#CCCCCC")
smut_colors <- c(second_mutations_col, "none/unknown" = "#CCCCCC", "NA" = "#CCCCCC")
gender_colors <- gender_col

# Location colors for simplified primary tumor location analysis
location_colors_simplified <- c(
  "choroid" = location_colors["choroid"],
  "iris" = location_colors["iris"],
  "ciliary body" = location_colors["ciliary body"],
  "other/NA" = "#CCCCCC"
)

# Age group colors - use ageGroup_col from config, add Unknown
age_colors <- c(ageGroup_col, "Unknown" = "#E1E1E1FF")
```

---

## Helper Function: Cached GloScope

**Intelligent caching:** Checks if divergence matrix exists; if yes, loads it; if no, computes and saves it.

```{r gloscope_cached_function}
run_gloscope_cached <- function(embedding_subset, sample_labels, cache_file, analysis_name) {

  # Prefer .qs2 cache, fall back to legacy .rds
  cache_qs2 <- file.path(output_dir, sub("\\.rds$", ".qs2", cache_file))
  cache_rds <- file.path(output_dir, cache_file)

  if (file.exists(cache_qs2)) {
    cat("  [CACHED] Loading existing divergence matrix:", basename(cache_qs2), "\n")
    knn_div <- qs2::qs_read(cache_qs2, nthreads = parallel::detectCores())
  } else if (file.exists(cache_rds)) {
    cat("  [CACHED] Loading legacy .rds divergence matrix:", cache_file, "\n")
    knn_div <- readRDS(cache_rds)
  } else {
    cat("  [COMPUTING] Running GloScope for:", analysis_name, "\n")
    cat("    - Cells:", nrow(embedding_subset), "\n")
    cat("    - Samples:", length(unique(sample_labels)), "\n")

    start_time <- Sys.time()

    knn_div <- gloscope(
      embedding_matrix = embedding_subset,
      cell_sample_ids = sample_labels,
      dens = "KNN",
      dist_mat = "KL",
      k = 50,
      BPPARAM = bpparam  # Use 24 cores
    )

    end_time <- Sys.time()
    cat("    - Computation time:", round(difftime(end_time, start_time, units = "mins"), 2), "minutes\n")

    # Save as qs2
    qs2::qs_save(knn_div, cache_qs2, nthreads = parallel::detectCores())
    cat("    - Saved to:", cache_qs2, "\n")
  }

  return(knn_div)
}
```

---

## Helper Function: MDS Feature Contributions

**Extract genes driving MDS coordinates:** For each MDS plot, we identify which genes (PCA features) contribute most to the MDS1 and MDS2 axes.

```{r mds_feature_contributions_function}
compute_mds_feature_contributions <- function(mds_coords, embedding_matrix,
                                              sample_labels, seurat_obj,
                                              n_top_genes = 30) {
  # Get sample-averaged embeddings in PCA space
  samples <- unique(sample_labels)
  avg_pca <- matrix(0, nrow = length(samples), ncol = ncol(embedding_matrix),
                    dimnames = list(samples, colnames(embedding_matrix)))

  for (samp in samples) {
    cells_in_sample <- which(sample_labels == samp)
    avg_pca[samp, ] <- colMeans(embedding_matrix[cells_in_sample, , drop = FALSE])
  }

  # Align sample order (intersect to handle samples missing from either)
  common_samples <- intersect(rownames(avg_pca), rownames(mds_coords))
  avg_pca <- avg_pca[common_samples, ]
  mds_coords <- mds_coords[common_samples, ]

  # Correlate each PC with MDS1 and MDS2
  cor_mds1 <- cor(avg_pca, mds_coords[, 1])
  cor_mds2 <- cor(avg_pca, mds_coords[, 2])

  # Get PCA loadings (genes × PCs)
  pca_loadings <- seurat_obj@reductions$pca@feature.loadings

  # Compute gene contributions to MDS axes
  gene_contrib_mds1 <- abs(pca_loadings %*% cor_mds1)
  gene_contrib_mds2 <- abs(pca_loadings %*% cor_mds2)

  # Get top genes for each axis
  top_genes_mds1 <- names(sort(gene_contrib_mds1[, 1], decreasing = TRUE)[1:n_top_genes])
  top_genes_mds2 <- names(sort(gene_contrib_mds2[, 1], decreasing = TRUE)[1:n_top_genes])

  # Combine and remove duplicates
  top_genes <- unique(c(top_genes_mds1, top_genes_mds2))

  # Get expression data and filter to genes that exist (Seurat v5 syntax)
  expr_data <- GetAssayData(seurat_obj, assay = "RNA", layer = "data")
  available_genes <- rownames(expr_data)
  top_genes <- intersect(top_genes, available_genes)

  if (length(top_genes) == 0) {
    stop("No top genes found in expression matrix")
  }

  cat("  - Found", length(top_genes), "genes in expression matrix\n")

  # Create heatmap matrix: samples × genes
  # Get cell barcodes from embedding_matrix rownames
  cell_barcodes <- rownames(embedding_matrix)

  expr_matrix <- matrix(0, nrow = length(samples), ncol = length(top_genes),
                       dimnames = list(samples, top_genes))

  for (samp in samples) {
    cells_in_sample_idx <- which(sample_labels == samp)
    cells_in_sample_barcodes <- cell_barcodes[cells_in_sample_idx]

    # Filter to cells that exist in the Seurat object
    cells_in_sample_barcodes <- intersect(cells_in_sample_barcodes, colnames(expr_data))

    if (length(cells_in_sample_barcodes) > 0) {
      expr_matrix[samp, ] <- colMeans(expr_data[top_genes, cells_in_sample_barcodes, drop = FALSE])
    } else {
      warning("No cells found for sample: ", samp)
    }
  }

  # Align with MDS sample order (intersect to handle missing samples)
  common_expr <- intersect(rownames(expr_matrix), rownames(mds_coords))
  expr_matrix <- expr_matrix[common_expr, ]

  return(list(
    expr_matrix = expr_matrix,
    top_genes_mds1 = top_genes_mds1,
    top_genes_mds2 = top_genes_mds2,
    gene_contrib_mds1 = gene_contrib_mds1,
    gene_contrib_mds2 = gene_contrib_mds2
  ))
}

plot_mds_feature_heatmap <- function(feature_data, sample_metadata_subset,
                                    title = "MDS Feature Contributions",
                                    output_file = NULL) {

  # Scale expression matrix for visualization
  expr_scaled <- t(scale(t(feature_data$expr_matrix)))

  # Mark which genes drive which MDS axis
  gene_anno <- data.frame(
    Gene = colnames(expr_scaled),
    MDS_Axis = ifelse(colnames(expr_scaled) %in% feature_data$top_genes_mds1[1:15], "MDS1",
                     ifelse(colnames(expr_scaled) %in% feature_data$top_genes_mds2[1:15], "MDS2", "Both"))
  )
  rownames(gene_anno) <- gene_anno$Gene

  axis_colors <- c("MDS1" = "#E15759", "MDS2" = "#4E79A7", "Both" = "#59A14F")

  col_anno <- HeatmapAnnotation(
    `MDS Axis` = gene_anno$MDS_Axis,
    col = list(`MDS Axis` = axis_colors),
    annotation_name_gp = gpar(fontsize = 9),
    show_legend = TRUE
  )

  # Row annotation with metadata (safe intersection)
  common_row <- intersect(rownames(expr_scaled), rownames(sample_metadata_subset))
  expr_scaled <- expr_scaled[common_row, , drop = FALSE]
  row_anno_data <- sample_metadata_subset[common_row, ]

  # Replace NA values with "Unknown" for annotations
  row_anno_data_clean <- row_anno_data %>%
    dplyr::mutate(
      Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
      Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
    )

  # Add colors for Unknown categories
  gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
  age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

  row_anno <- rowAnnotation(
    Origin = row_anno_data_clean$origin,
    Gender = row_anno_data_clean$Gender,
    Age_group = row_anno_data_clean$Age_group,
    col = list(
      Origin = origin_colors,
      Gender = gender_colors_full,
      Age_group = age_colors_full
    ),
    annotation_name_gp = gpar(fontsize = 9),
    show_legend = c(Origin = TRUE, Gender = TRUE, Age_group = TRUE)
  )

  # Create heatmap
  ht <- Heatmap(
    expr_scaled,
    name = "Scaled\nExpression",
    col = colorRamp2(c(-2, 0, 2), c("#2E91E5", "white", "#E15759")),
    top_annotation = col_anno,
    left_annotation = row_anno,
    show_row_names = FALSE,
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 7),
    column_title = title,
    column_title_gp = gpar(fontsize = 12, fontface = "bold"),
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    heatmap_legend_param = list(
      title = "Scaled\nExpression",
      legend_height = unit(3, "cm")
    )
  )

  if (!is.null(output_file)) {
    png(output_file, width = 4000, height = 2400, res = 300)
    draw(ht)
    dev.off()
  }

  return(ht)
}
```

---

## All Cells Analysis {.tabset}

**How GloScope differs from classical MDS:**

Classical MDS workflows typically represent each sample as a **single point** (e.g., centroid in PCA space) and compute Euclidean distances between centroids. This approach loses information about within-sample variation and heterogeneity.

**GloScope's innovation**:

1. **Samples as distributions**: Each sample is represented as a probability distribution in PCA space using KNN density estimation (k=50)
2. **Divergence, not distance**: Computes KL (Kullback-Leibler) divergence between sample distributions instead of simple Euclidean distances
3. **Captures heterogeneity**: Two samples with the same cell type proportions but different transcriptional heterogeneity will show different divergences
4. **MDS for visualization**: The divergence matrix is then visualized using standard MDS (same as classical approach)

**Example**: Sample A with tightly clustered cells vs. Sample B with diffuse cells (both having same proportions) will appear similar in classical MDS but different in GloScope.

**What we're computing**: `gloscope()` creates the sample-to-sample divergence matrix → `plotMDS()` visualizes it in 2D

### Disease Progression (Origin)

**Comparison:** healthy vs. primary vs. metastasis

```{r gloscope_origin}
cat("Running GloScope analysis for disease origin (ALL CELLS)...\n")

# Prepare sample labels
sample_labels <- cell_metadata$Sample

# Calculate or load divergence
knn_div_origin <- run_gloscope_cached(
  embedding_subset = embedding_matrix,
  sample_labels = sample_labels,
  cache_file = "divergence_origin_full_cohort.rds",
  analysis_name = "Disease Origin (All Cells)"
)

cat("Divergence matrix:", dim(knn_div_origin), "\n")

# MDS visualization
samples_in_div <- rownames(knn_div_origin)
sample_metadata_for_plot <- data.frame(Sample = samples_in_div) %>%
  left_join(sample_metadata, by = "Sample")

mds_origin <- plotMDS(
  dist_mat = knn_div_origin,
  metadata_df = sample_metadata_for_plot,
  sample_id = "Sample",
  group_id = "origin",
  k = 2
)

# Run ANOSIM test first to get p-value for plot
anosim_origin <- safe_anosim(
  as.dist(knn_div_origin),
  grouping = sample_metadata_for_plot$origin,
  permutations = 999
)

cat("\nANOSIM test for disease origin:\n")
cat("  R statistic:", round(anosim_origin$statistic, 4), "\n")
cat("  p-value:", anosim_origin$signif, "\n")

# Format p-value for display
p_display <- ifelse(anosim_origin$signif < 0.001, "p < 0.001",
                    paste0("p = ", format(anosim_origin$signif, digits = 3)))

# Enhance plot with ANOSIM p-value in subtitle
p_origin_mds <- mds_origin$plot +
  geom_point(size = 4) +
  scale_color_manual(values = origin_colors) +
  coord_fixed() +
  labs(
    title = "GloScope MDS: Disease Progression (Full Cohort)",
    subtitle = paste0("n = ", nrow(sample_metadata_for_plot), " samples | ANOSIM: R = ",
                     round(anosim_origin$statistic, 3), ", ", p_display),
    color = "Origin"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

print(p_origin_mds)

ggsave(
  file.path(output_dir, "MDS_origin_full_cohort.png"),
  p_origin_mds,
  width = 10, height = 8, dpi = 300, bg = "white"
)

write_csv(
  data.frame(
    comparison = "disease_origin_full",
    statistic = anosim_origin$statistic,
    p_value = anosim_origin$signif,
    permutations = 999
  ),
  file.path(output_dir, "anosim_origin_full.csv")
)
```

### Divergence Heatmap - Origin

```{r heatmap_origin}
# Prepare annotation
sample_order <- sample_metadata %>% arrange(origin, organ) %>% pull(Sample)
knn_div_ordered <- knn_div_origin[sample_order, sample_order]

# Get annotation data
anno_data <- sample_metadata[match(sample_order, sample_metadata$Sample), ]

# Replace NA values with "Unknown" for annotations
anno_data_clean <- anno_data %>%
  dplyr::mutate(
    Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
    Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
  )

# Add colors for Unknown categories
gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

# Column annotation - using Victoria's patient colors to fix blank rectangles
# Generate patient colors for any patients not in the predefined palette
all_patients <- unique(anno_data_clean$Patient_nr)
patient_colors_full <- patientID_col[all_patients]
missing_patients <- all_patients[is.na(patient_colors_full)]
if (length(missing_patients) > 0) {
  extra_colors <- setNames(
    scales::hue_pal()(length(missing_patients)),
    missing_patients
  )
  patient_colors_full <- c(patientID_col, extra_colors)
}
patient_colors_full <- patient_colors_full[!is.na(names(patient_colors_full))]

col_anno <- HeatmapAnnotation(
  Patient = anno_data_clean$Patient_nr,
  Origin = anno_data_clean$origin,
  Organ = anno_data_clean$organ,
  Gender = anno_data_clean$Gender,
  Age_group = anno_data_clean$Age_group,
  `Primary Mutation` = anno_data_clean$primary_mutation,
  `Secondary Mutation` = anno_data_clean$secondary_mutation,
  col = list(
    Patient = patient_colors_full,
    Origin = origin_colors,
    Organ = organ_colors_all,
    Gender = gender_colors_full,
    Age_group = age_colors_full,
    `Primary Mutation` = pmut_colors,
    `Secondary Mutation` = smut_colors
  ),
  annotation_name_gp = gpar(fontsize = 11),
  show_legend = c(Patient = FALSE, Origin = TRUE, Organ = TRUE,
                  Gender = TRUE, Age_group = TRUE,
                  `Primary Mutation` = TRUE, `Secondary Mutation` = TRUE)
)

# Create heatmap
ht_origin <- Heatmap(
  knn_div_ordered,
  name = "KL Divergence",
  col = colorRamp2(c(0, median(knn_div_origin), max(knn_div_origin)),
                   c("white", "orange", "red")),
  top_annotation = col_anno,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "GloScope Divergence: Disease Origin (Full Cohort - All Cells)",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  heatmap_legend_param = list(
    title = "KL Divergence",
    legend_height = unit(4, "cm")
  )
)

# Plot
png(file.path(output_dir, "heatmap_origin_full_cohort.png"),
    width = 3000, height = 2800, res = 300)
draw(ht_origin)
dev.off()

draw(ht_origin)

# Save heatmap object and data for recreation
qs2::qs_save(list(ht = ht_origin, div_matrix = knn_div_ordered, anno_data = anno_data_clean),
             file.path(output_dir, "gloscope_heatmap_all_cells_objects.qs2"), nthreads = parallel::detectCores())
```

### MDS Feature Contributions - Origin

**NOTE:** MDS feature contribution analysis is temporarily disabled due to gene naming compatibility issues. This will be addressed in a future update.

<!--
**Genes driving MDS1/MDS2 separation:** Identify which genes contribute most to sample positioning in the MDS plot.

```{r mds_features_origin, eval=FALSE}
cat("Computing MDS feature contributions for disease origin...\n")

# Get MDS coordinates
mds_coords_origin <- mds_origin$df[, c("dim1", "dim2")]
rownames(mds_coords_origin) <- mds_origin$df$Sample

# Compute feature contributions
feature_data_origin <- compute_mds_feature_contributions(
  mds_coords = mds_coords_origin,
  embedding_matrix = embedding_matrix,
  sample_labels = sample_labels,
  seurat_obj = um_atlas,
  n_top_genes = 30
)

cat("Top genes driving MDS1:\n")
print(head(feature_data_origin$top_genes_mds1, 15))

cat("\nTop genes driving MDS2:\n")
print(head(feature_data_origin$top_genes_mds2, 15))

# Plot heatmap
ht_features_origin <- plot_mds_feature_heatmap(
  feature_data = feature_data_origin,
  sample_metadata_subset = sample_metadata_for_plot,
  title = "MDS Feature Contributions: Disease Origin (Full Cohort)",
  output_file = file.path(output_dir, "heatmap_mds_features_origin_full.png")
)

draw(ht_features_origin)

# Save top genes list
write_csv(
  data.frame(
    Gene = feature_data_origin$top_genes_mds1[1:30],
    Contribution_MDS1 = feature_data_origin$gene_contrib_mds1[feature_data_origin$top_genes_mds1[1:30], 1]
  ),
  file.path(output_dir, "top_genes_MDS1_origin_full.csv")
)

write_csv(
  data.frame(
    Gene = feature_data_origin$top_genes_mds2[1:30],
    Contribution_MDS2 = feature_data_origin$gene_contrib_mds2[feature_data_origin$top_genes_mds2[1:30], 1]
  ),
  file.path(output_dir, "top_genes_MDS2_origin_full.csv")
)
```
-->

### Primary Mutations

**Comparison:** GNAQ vs. GNA11 vs. PLCB4

```{r gloscope_primary_mutation}
cat("Running GloScope analysis for primary mutations...\n")

# Filter to samples with defined primary mutations
samples_with_pmut <- sample_metadata %>%
  dplyr::filter(primary_mutation %in% c("GNAQ", "GNA11", "PLCB4")) %>%
  pull(Sample)

cells_pmut <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_pmut) %>% pull(cell_id)
embedding_pmut <- embedding_matrix[cells_pmut, ]
sample_labels_pmut <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_pmut) %>% pull(Sample)

# Calculate or load divergence
knn_div_pmut <- run_gloscope_cached(
  embedding_subset = embedding_pmut,
  sample_labels = sample_labels_pmut,
  cache_file = "divergence_primary_mutation_full.rds",
  analysis_name = "Primary Mutations"
)

# MDS visualization
samples_in_div_pmut <- rownames(knn_div_pmut)
sample_metadata_pmut <- data.frame(Sample = samples_in_div_pmut) %>%
  left_join(sample_metadata %>% dplyr::filter(Sample %in% samples_with_pmut), by = "Sample")

mds_pmut <- plotMDS(
  dist_mat = knn_div_pmut,
  metadata_df = sample_metadata_pmut,
  sample_id = "Sample",
  group_id = "primary_mutation",
  k = 2
)

# ANOSIM test first to get p-value for plot
anosim_pmut <- safe_anosim(
  as.dist(knn_div_pmut),
  grouping = sample_metadata_pmut$primary_mutation,
  permutations = 999
)

cat("\nANOSIM test for primary mutations:\n")
cat("  R statistic:", round(anosim_pmut$statistic, 4), "\n")
cat("  p-value:", anosim_pmut$signif, "\n")

p_display_pmut <- ifelse(anosim_pmut$signif < 0.001, "p < 0.001",
                         paste0("p = ", format(anosim_pmut$signif, digits = 3)))

p_pmut_mds <- mds_pmut$plot +
  geom_point(size = 4) +
  scale_color_manual(values = pmut_colors) +
  coord_fixed() +
  labs(
    title = "GloScope MDS: Primary Mutations (Full Cohort)",
    subtitle = paste0("n = ", nrow(sample_metadata_pmut), " samples | ANOSIM: R = ",
                     round(anosim_pmut$statistic, 3), ", ", p_display_pmut),
    color = "Primary Mutation"
  ) +
  theme_minimal(base_size = 12)

print(p_pmut_mds)

ggsave(
  file.path(output_dir, "MDS_primary_mutation_full.png"),
  p_pmut_mds,
  width = 10, height = 8, dpi = 300, bg = "white"
)

write_csv(
  tibble(
    comparison = "primary_mutation_full",
    statistic = anosim_pmut$statistic,
    p_value = anosim_pmut$signif
  ),
  file.path(output_dir, "anosim_primary_mutation_full.csv")
)
```

### Primary Mutation - Malignant Cells Only

```{r gloscope_primary_mutation_malignant}
cat("Running GloScope analysis for Primary Mutation (Malignant Cells Only)...\n")

# Filter to malignant cells from samples with primary mutation data
cells_pmut_malignant <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_with_pmut, Malignant == "Malignant") %>%
  pull(cell_id)

sample_labels_pmut_malignant <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_with_pmut, Malignant == "Malignant") %>%
  pull(Sample)

sample_cell_counts_pmut_mal <- table(sample_labels_pmut_malignant)
valid_samples_pmut_mal <- names(sample_cell_counts_pmut_mal)[sample_cell_counts_pmut_mal >= 50]

cat("  - Total malignant cells:", length(cells_pmut_malignant), "\n")
cat("  - Samples with >=50 malignant cells:", length(valid_samples_pmut_mal), "\n")

if (length(valid_samples_pmut_mal) >= 4) {
  cells_pmut_mal_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_pmut_mal) %>%
    pull(cell_id)

  embedding_pmut_mal <- embedding_matrix[cells_pmut_mal_filtered, ]
  sample_labels_pmut_mal_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_pmut_mal) %>%
    pull(Sample)

  div_matrix_pmut_mal <- run_gloscope_cached(
    embedding_subset = embedding_pmut_mal,
    sample_labels = sample_labels_pmut_mal_filtered,
    cache_file = "divergence_primary_mutation_malignant.rds",
    analysis_name = "Primary Mutation (Malignant Cells)"
  )

  # ANOSIM test
  pmut_vec_mal <- sample_metadata$primary_mutation[match(rownames(div_matrix_pmut_mal), sample_metadata$Sample)]

  anosim_pmut_mal <- safe_anosim(div_matrix_pmut_mal, pmut_vec_mal, permutations = 999)

  cat("\nANOSIM test for Primary Mutation (malignant cells):\n")
  cat("  R statistic:", round(anosim_pmut_mal$statistic, 4), "\n")
  cat("  p-value:", anosim_pmut_mal$signif, "\n")

  write_csv(
    tibble(
      comparison = "primary_mutation_malignant_full",
      statistic = anosim_pmut_mal$statistic,
      p_value = anosim_pmut_mal$signif
    ),
    file.path(output_dir, "anosim_primary_mutation_malignant_full.csv")
  )
} else {
  cat("Insufficient samples with adequate malignant cell representation\n")
}
```

### Primary Mutation - Primary Tumors Only

```{r gloscope_primary_mutation_pUM}
cat("Running GloScope analysis for Primary Mutation (Primary Tumors Only)...\n")

# Filter to primary tumor samples with mutation data
samples_pUM_pmut <- sample_metadata %>%
  dplyr::filter(origin == "primary", primary_mutation %in% c("GNAQ", "GNA11", "PLCB4")) %>%
  pull(Sample)

cells_pmut_pUM <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_pUM_pmut) %>%
  pull(cell_id)

sample_labels_pmut_pUM <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_pUM_pmut) %>%
  pull(Sample)

cat("  - Total cells from pUM:", length(cells_pmut_pUM), "\n")
cat("  - pUM samples with mutation:", length(unique(sample_labels_pmut_pUM)), "\n")

if (length(unique(sample_labels_pmut_pUM)) >= 4) {
  embedding_pmut_pUM <- embedding_matrix[cells_pmut_pUM, ]

  div_matrix_pmut_pUM <- run_gloscope_cached(
    embedding_subset = embedding_pmut_pUM,
    sample_labels = sample_labels_pmut_pUM,
    cache_file = "divergence_primary_mutation_pUM.rds",
    analysis_name = "Primary Mutation (Primary Tumors)"
  )

  # ANOSIM test
  pmut_vec_pUM <- sample_metadata$primary_mutation[match(rownames(div_matrix_pmut_pUM), sample_metadata$Sample)]

  anosim_pmut_pUM <- safe_anosim(div_matrix_pmut_pUM, pmut_vec_pUM, permutations = 999)

  cat("\nANOSIM test for Primary Mutation (primary tumors):\n")
  cat("  R statistic:", round(anosim_pmut_pUM$statistic, 4), "\n")
  cat("  p-value:", anosim_pmut_pUM$signif, "\n")

  write_csv(
    tibble(
      comparison = "primary_mutation_pUM_full",
      statistic = anosim_pmut_pUM$statistic,
      p_value = anosim_pmut_pUM$signif
    ),
    file.path(output_dir, "anosim_primary_mutation_pUM_full.csv")
  )
} else {
  cat("Insufficient pUM samples with mutation data\n")
}
```

### Gender

**Comparison:** Male vs Female

```{r gloscope_gender}
cat("Running GloScope analysis for Gender...\n")

# Filter to samples with defined gender
samples_with_gender <- sample_metadata %>%
  dplyr::filter(!is.na(Gender) & Gender %in% c("M", "F")) %>%
  pull(Sample)

cells_gender <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_gender) %>% pull(cell_id)
embedding_gender <- embedding_matrix[cells_gender, ]
sample_labels_gender <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_gender) %>% pull(Sample)

# Run GloScope
cat(sprintf("  - Cells: %d\n  - Samples: %d\n", nrow(embedding_gender), length(unique(sample_labels_gender))))

div_matrix_gender <- run_gloscope_cached(
  embedding_subset = embedding_gender,
  sample_labels = sample_labels_gender,
  cache_file = "divergence_gender_full_cohort.rds",
  analysis_name = "Gender (All Cells)"
)

cat("Divergence matrix:", nrow(div_matrix_gender), ncol(div_matrix_gender), "\n")
```

```{r mds_gender, fig.width=10, fig.height=8}
# MDS visualization
mds_gender <- limma::plotMDS(div_matrix_gender, plot = FALSE)
mds_coords_gender <- data.frame(
  Sample = rownames(div_matrix_gender),
  MDS1 = mds_gender$x,
  MDS2 = mds_gender$y
) %>%
  left_join(sample_metadata %>% dplyr::select(Sample, Gender), by = "Sample")

p_gender <- ggplot(mds_coords_gender, aes(x = MDS1, y = MDS2, color = Gender, label = Sample)) +
  geom_point(size = 4.5, alpha = 0.7) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
  scale_color_manual(values = c("M" = "#3498db", "F" = "#e74c3c")) +
  coord_fixed() +
  theme_minimal(base_size = 14) +
  labs(title = "MDS - Gender (All Cells)", x = "MDS1", y = "MDS2")

print(p_gender)
ggsave(file.path(output_dir, "mds_gender_full.png"), p_gender, width = 10, height = 8, dpi = 300, bg = "white")

# ANOSIM test
library(vegan)
gender_vec <- sample_metadata$Gender[match(rownames(div_matrix_gender), sample_metadata$Sample)]

anosim_gender <- safe_anosim(div_matrix_gender, gender_vec, permutations = 999)

cat("\nANOSIM test for Gender:\n")
cat("  R statistic:", round(anosim_gender$statistic, 4), "\n")
cat("  p-value:", anosim_gender$signif, "\n")

write_csv(
  tibble(
    comparison = "gender_full",
    statistic = anosim_gender$statistic,
    p_value = anosim_gender$signif
  ),
  file.path(output_dir, "anosim_gender_full.csv")
)
```

### Divergence Heatmap - Gender

```{r heatmap_gender, fig.width=12, fig.height=11}
library(ComplexHeatmap)
library(circlize)

# Prepare annotation — align by matrix rownames (not filter, to preserve dimension)
anno_data_clean <- data.frame(
  Sample = rownames(div_matrix_gender),
  stringsAsFactors = FALSE
) %>%
  left_join(sample_metadata, by = "Sample") %>%
  dplyr::mutate(
    Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown")
  )

# Define color palettes
gender_colors <- c(gender_col, "Unknown" = "#E1E1E1FF")

# Create column annotation
col_anno <- HeatmapAnnotation(
  Gender = anno_data_clean$Gender,
  col = list(Gender = gender_colors),
  show_legend = c(Gender = TRUE),
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(0.4, "cm")
)

# Create heatmap
ht_gender <- Heatmap(
  div_matrix_gender,
  name = "KL Divergence",
  col = colorRamp2(
    c(min(div_matrix_gender), median(div_matrix_gender), max(div_matrix_gender)),
    c("blue", "white", "red")
  ),
  top_annotation = col_anno,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  heatmap_legend_param = list(
    title = "KL Divergence",
    legend_height = unit(4, "cm")
  )
)

png(file.path(output_dir, "heatmap_gender_full.png"),
    width = 2400, height = 2200, res = 300)
draw(ht_gender)
dev.off()

draw(ht_gender)
```

### MDS Feature Contributions - Gender

```{r mds_top_genes_gender}
cat("Identifying top genes driving MDS coordinates for Gender...\n")

# Get MDS coordinates
mds_coords_gender_mat <- data.frame(
  dim1 = mds_coords_gender$MDS1,
  dim2 = mds_coords_gender$MDS2,
  row.names = mds_coords_gender$Sample
)

# Get sample-averaged embeddings
samples_gender <- unique(sample_labels_gender)
avg_pca_gender <- matrix(0, nrow = length(samples_gender), ncol = ncol(embedding_gender),
                         dimnames = list(samples_gender, colnames(embedding_gender)))

for (samp in samples_gender) {
  cells_in_sample <- which(sample_labels_gender == samp)
  avg_pca_gender[samp, ] <- colMeans(embedding_gender[cells_in_sample, , drop = FALSE])
}

# Align sample order (intersect to handle missing samples)
common_gender <- intersect(rownames(avg_pca_gender), rownames(mds_coords_gender_mat))
avg_pca_gender <- avg_pca_gender[common_gender, ]
mds_coords_gender_mat <- mds_coords_gender_mat[common_gender, ]

# Correlate each PC with MDS1 and MDS2
cor_mds1_gender <- cor(avg_pca_gender, mds_coords_gender_mat[, 1])
cor_mds2_gender <- cor(avg_pca_gender, mds_coords_gender_mat[, 2])

# Get PCA loadings
pca_loadings <- um_atlas@reductions$pca@feature.loadings

# Compute gene contributions
gene_contrib_mds1_gender <- abs(pca_loadings %*% cor_mds1_gender)
gene_contrib_mds2_gender <- abs(pca_loadings %*% cor_mds2_gender)

# Get top 30 genes for each axis
# Gene names are already in the rownames of gene_contrib matrices
contrib_vec_mds1_gender <- as.vector(gene_contrib_mds1_gender)
names(contrib_vec_mds1_gender) <- rownames(gene_contrib_mds1_gender)
top_genes_mds1_gender <- names(sort(contrib_vec_mds1_gender, decreasing = TRUE)[1:30])

contrib_vec_mds2_gender <- as.vector(gene_contrib_mds2_gender)
names(contrib_vec_mds2_gender) <- rownames(gene_contrib_mds2_gender)
top_genes_mds2_gender <- names(sort(contrib_vec_mds2_gender, decreasing = TRUE)[1:30])

cat("\nTop 30 genes driving MDS1 (Gender):\n")
print(top_genes_mds1_gender)

cat("\nTop 30 genes driving MDS2 (Gender):\n")
print(top_genes_mds2_gender)

# Save to CSV
write_csv(
  data.frame(
    Gene = top_genes_mds1_gender,
    Contribution = contrib_vec_mds1_gender[top_genes_mds1_gender]
  ),
  file.path(output_dir, "top_genes_MDS1_gender.csv")
)

write_csv(
  data.frame(
    Gene = top_genes_mds2_gender,
    Contribution = contrib_vec_mds2_gender[top_genes_mds2_gender]
  ),
  file.path(output_dir, "top_genes_MDS2_gender.csv")
)
```

### Organ (Metastasis Sites)

**Comparison:** Metastasis sites (liver, brain, skin, thyroid) vs primary eye tumors

```{r gloscope_organ}
cat("Running GloScope analysis for Organ/Metastasis Sites...\n")

# Check organ distribution first
cat("\nOrgan distribution:\n")
organ_counts <- sample_metadata %>% count(organ)
print(organ_counts)

# Use all samples - compare different organs (need at least 2 per group)
samples_by_organ <- sample_metadata %>%
  group_by(organ) %>%
  filter(n() >= 2) %>%  # Need at least 2 samples per group
  ungroup() %>%
  pull(Sample)

cat(sprintf("\nSamples with >=2 per organ group: %d\n", length(samples_by_organ)))

cells_organ <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_by_organ) %>%
  pull(cell_id)

embedding_organ <- embedding_matrix[cells_organ, ]
sample_labels_organ <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_by_organ) %>%
  pull(Sample)

cat(sprintf("  - Cells: %d\n  - Samples: %d\n", nrow(embedding_organ), length(unique(sample_labels_organ))))

# Run GloScope with error handling
div_matrix_organ <- tryCatch({
  run_gloscope_cached(
    embedding_subset = embedding_organ,
    sample_labels = sample_labels_organ,
    cache_file = "divergence_organ_full_cohort.rds",
    analysis_name = "Organ (All Cells)"
  )
}, error = function(e) {
  cat("GloScope error:", conditionMessage(e), "\n")
  cat("Trying with sequential processing...\n")

  # Try without parallel processing — check qs2 first, then legacy rds
  cache_qs2 <- file.path(output_dir, "divergence_organ_full_cohort.qs2")
  cache_rds <- file.path(output_dir, "divergence_organ_full_cohort.rds")
  if (file.exists(cache_qs2)) {
    return(qs2::qs_read(cache_qs2, nthreads = parallel::detectCores()))
  } else if (file.exists(cache_rds)) {
    return(readRDS(cache_rds))
  }

  # Run with BPPARAM = SerialParam()
  GloScope::gloscope(
    embedding_organ,
    sample_labels_organ,
    dens = "GMM",
    dist_mat = "KL",
    BPPARAM = BiocParallel::SerialParam()
  )
})

cat("Divergence matrix:", nrow(div_matrix_organ), "x", ncol(div_matrix_organ), "\n")
```

```{r mds_organ, fig.width=12, fig.height=9}
# MDS plot with dual encoding: Color = Secondary Mutation, Shape = Disease Progression
mds_organ <- limma::plotMDS(div_matrix_organ, plot = FALSE)
mds_coords_organ <- data.frame(
  Sample = rownames(div_matrix_organ),
  MDS1 = mds_organ$x,
  MDS2 = mds_organ$y
) %>%
  left_join(sample_metadata %>% dplyr::select(Sample, organ, origin, secondary_mutation), by = "Sample")

# Define shape mapping for disease progression
progression_shapes <- c("healthy" = 16, "primary" = 17, "metastasis" = 18)

# ANOSIM test first to get statistics for annotation
organ_vec <- sample_metadata$organ[match(rownames(div_matrix_organ), sample_metadata$Sample)]

anosim_organ <- NULL
if (length(unique(organ_vec)) >= 2) {
  anosim_organ <- safe_anosim(div_matrix_organ, organ_vec, permutations = 999)

  cat("\nANOSIM test for Organ:\n")
  cat("  R statistic:", round(anosim_organ$statistic, 4), "\n")
  cat("  p-value:", anosim_organ$signif, "\n")

  write_csv(
    data.frame(
      comparison = "organ_full",
      statistic = anosim_organ$statistic,
      p_value = anosim_organ$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_organ_full.csv")
  )
}

# Format ANOSIM annotation text
anosim_text <- if (!is.null(anosim_organ)) {
  p_disp <- ifelse(anosim_organ$signif < 0.001, "p < 0.001",
                    paste0("p = ", format(anosim_organ$signif, digits = 3)))
  paste0("ANOSIM R = ", round(anosim_organ$statistic, 3), ", ", p_disp)
} else {
  ""
}

p_organ <- ggplot(mds_coords_organ, aes(x = MDS1, y = MDS2,
                                         color = secondary_mutation,
                                         shape = origin,
                                         label = Sample)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3) +
  geom_point(size = 4, alpha = 0.8) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20, show.legend = FALSE) +
  scale_color_manual(values = smut_colors, name = "Secondary Mutation") +
  scale_shape_manual(values = progression_shapes, name = "Disease Progression") +
  guides(
    color = guide_legend(override.aes = list(shape = 15, size = 4)),
    shape = guide_legend(override.aes = list(size = 4))
  ) +
  annotate("text", x = Inf, y = -Inf, label = anosim_text,
           hjust = 1.1, vjust = -0.5, size = 4, fontface = "italic") +
  theme_classic(base_size = 14) +
  coord_fixed() +
  labs(title = "MDS - Organ/Metastasis Site (All Cells)",
       subtitle = "Color = Secondary Mutation, Shape = Disease Progression",
       x = "MDS1", y = "MDS2") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

print(p_organ)
ggsave(file.path(output_dir, "mds_organ_full.png"), p_organ, width = 12, height = 9, dpi = 300, bg = "white")
```

### Divergence Heatmap - Organ

```{r heatmap_organ, fig.width=12, fig.height=11}
# Order samples by organ
sample_order_organ <- sample_metadata %>%
  dplyr::filter(Sample %in% rownames(div_matrix_organ)) %>%
  arrange(organ, origin) %>%
  pull(Sample)

div_matrix_organ_ordered <- div_matrix_organ[sample_order_organ, sample_order_organ]

anno_data_organ <- sample_metadata[match(sample_order_organ, sample_metadata$Sample), ]

col_anno_organ <- HeatmapAnnotation(
  Organ = anno_data_organ$organ,
  Origin = anno_data_organ$origin,
  col = list(Organ = organ_colors_all,
             Origin = origin_colors),
  show_legend = c(Organ = TRUE, Origin = TRUE),
  annotation_name_side = "left"
)

ht_organ <- Heatmap(
  div_matrix_organ_ordered,
  name = "KL Divergence",
  col = colorRamp2(c(0, median(div_matrix_organ), max(div_matrix_organ)),
                   c("white", "#FFC0CB", "#8B0000")),
  top_annotation = col_anno_organ,
  show_row_names = TRUE,
  show_column_names = FALSE,
  column_title = "Organ/Metastasis Site Comparison",
  row_names_gp = gpar(fontsize = 8),
  row_title = "Samples",
  cluster_rows = FALSE,
  cluster_columns = FALSE
)

png(file.path(output_dir, "heatmap_organ_full.png"),
    width = 12, height = 11, units = "in", res = 300)
draw(ht_organ)
dev.off()

draw(ht_organ)
```

### Organ - Malignant Cells Only

```{r gloscope_organ_malignant}
cat("Running GloScope analysis for Organ (Malignant Cells Only)...\n")

# Filter to malignant cells from samples with organ data
cells_organ_malignant <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_by_organ, Malignant == "Malignant") %>%
  pull(cell_id)

sample_labels_organ_malignant <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_by_organ, Malignant == "Malignant") %>%
  pull(Sample)

# Check sample representation
sample_cell_counts_organ_mal <- table(sample_labels_organ_malignant)
valid_samples_organ_mal <- names(sample_cell_counts_organ_mal)[sample_cell_counts_organ_mal >= 50]

cat("  - Total malignant cells:", length(cells_organ_malignant), "\n")
cat("  - Samples with >=50 malignant cells:", length(valid_samples_organ_mal), "\n")

if (length(valid_samples_organ_mal) >= 4) {
  # Filter to valid samples
  cells_organ_mal_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_organ_mal) %>%
    pull(cell_id)

  embedding_organ_mal <- embedding_matrix[cells_organ_mal_filtered, ]
  sample_labels_organ_mal_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_organ_mal) %>%
    pull(Sample)

  # Run GloScope
  div_matrix_organ_mal <- run_gloscope_cached(
    embedding_subset = embedding_organ_mal,
    sample_labels = sample_labels_organ_mal_filtered,
    cache_file = "divergence_organ_malignant.rds",
    analysis_name = "Organ (Malignant Cells)"
  )

  # ANOSIM test
  organ_vec_mal <- sample_metadata$organ[match(rownames(div_matrix_organ_mal), sample_metadata$Sample)]

  anosim_organ_mal <- safe_anosim(div_matrix_organ_mal, organ_vec_mal, permutations = 999)

  cat("\nANOSIM test for Organ (malignant cells):\n")
  cat("  R statistic:", round(anosim_organ_mal$statistic, 4), "\n")
  cat("  p-value:", anosim_organ_mal$signif, "\n")

  write_csv(
    tibble(
      comparison = "organ_malignant_full",
      statistic = anosim_organ_mal$statistic,
      p_value = anosim_organ_mal$signif
    ),
    file.path(output_dir, "anosim_organ_malignant_full.csv")
  )
} else {
  cat("Insufficient samples with adequate malignant cell representation\n")
}
```

### Organ - Melanocytic Cells Only

```{r gloscope_organ_melanocytic}
cat("Running GloScope analysis for Organ (Melanocytic Cells Only)...\n")

# Filter to melanocytic cells (malignant + healthy melanocytes)
cells_organ_melanocytic <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_by_organ,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(cell_id)

sample_labels_organ_melanocytic <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_by_organ,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(Sample)

# Check sample representation
sample_cell_counts_organ_mel <- table(sample_labels_organ_melanocytic)
valid_samples_organ_mel <- names(sample_cell_counts_organ_mel)[sample_cell_counts_organ_mel >= 50]

cat("  - Total melanocytic cells:", length(cells_organ_melanocytic), "\n")
cat("  - Samples with >=50 melanocytic cells:", length(valid_samples_organ_mel), "\n")

if (length(valid_samples_organ_mel) >= 4) {
  cells_organ_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_organ_mel) %>%
    pull(cell_id)

  embedding_organ_mel <- embedding_matrix[cells_organ_mel_filtered, ]
  sample_labels_organ_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_organ_mel) %>%
    pull(Sample)

  div_matrix_organ_mel <- run_gloscope_cached(
    embedding_subset = embedding_organ_mel,
    sample_labels = sample_labels_organ_mel_filtered,
    cache_file = "divergence_organ_melanocytic.rds",
    analysis_name = "Organ (Melanocytic Cells)"
  )

  # ANOSIM test
  organ_vec_mel <- sample_metadata$organ[match(rownames(div_matrix_organ_mel), sample_metadata$Sample)]

  anosim_organ_mel <- safe_anosim(div_matrix_organ_mel, organ_vec_mel, permutations = 999)

  cat("\nANOSIM test for Organ (melanocytic cells):\n")
  cat("  R statistic:", round(anosim_organ_mel$statistic, 4), "\n")
  cat("  p-value:", anosim_organ_mel$signif, "\n")

  write_csv(
    tibble(
      comparison = "organ_melanocytic_full",
      statistic = anosim_organ_mel$statistic,
      p_value = anosim_organ_mel$signif
    ),
    file.path(output_dir, "anosim_organ_melanocytic_full.csv")
  )
} else {
  cat("Insufficient samples with adequate melanocytic cell representation\n")
}
```

### Age

**Comparison:** Age groups

```{r gloscope_age}
cat("Running GloScope analysis for Age...\n")

# Filter to samples with defined age groups
samples_with_age <- sample_metadata %>%
  dplyr::filter(!is.na(Age_group) & Age_group != "Unknown") %>%
  pull(Sample)

cells_age <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_age) %>% pull(cell_id)
embedding_age <- embedding_matrix[cells_age, ]
sample_labels_age <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_age) %>% pull(Sample)

# Run GloScope
cat(sprintf("  - Cells: %d\n  - Samples: %d\n", nrow(embedding_age), length(unique(sample_labels_age))))

div_matrix_age <- run_gloscope_cached(
  embedding_subset = embedding_age,
  sample_labels = sample_labels_age,
  cache_file = "divergence_age_full_cohort.rds",
  analysis_name = "Age (All Cells)"
)

cat("Divergence matrix:", nrow(div_matrix_age), ncol(div_matrix_age), "\n")
```

```{r mds_age, fig.width=10, fig.height=8}
# MDS visualization
mds_age <- limma::plotMDS(div_matrix_age, plot = FALSE)
mds_coords_age <- data.frame(
  Sample = rownames(div_matrix_age),
  MDS1 = mds_age$x,
  MDS2 = mds_age$y
) %>%
  left_join(sample_metadata %>% dplyr::select(Sample, Age_group), by = "Sample")

# Age colors - use ageGroup_col from config (already defined in setup)

p_age <- ggplot(mds_coords_age, aes(x = MDS1, y = MDS2, color = Age_group, label = Sample)) +
  geom_point(size = 4.5, alpha = 0.7) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
  scale_color_manual(values = age_colors) +
  coord_fixed() +
  theme_minimal(base_size = 14) +
  labs(title = "MDS - Age Group (All Cells)", x = "MDS1", y = "MDS2", color = "Age Group")

print(p_age)
ggsave(file.path(output_dir, "mds_age_full.png"), p_age, width = 10, height = 8, dpi = 300, bg = "white")

# ANOSIM test
age_vec <- sample_metadata$Age_group[match(rownames(div_matrix_age), sample_metadata$Sample)]

anosim_age <- safe_anosim(div_matrix_age, age_vec, permutations = 999)

cat("\nANOSIM test for Age:\n")
cat("  R statistic:", round(anosim_age$statistic, 4), "\n")
cat("  p-value:", anosim_age$signif, "\n")

write_csv(
  tibble(
    comparison = "age_full",
    statistic = anosim_age$statistic,
    p_value = anosim_age$signif
  ),
  file.path(output_dir, "anosim_age_full.csv")
)
```

### Divergence Heatmap - Age

```{r heatmap_age, fig.width=12, fig.height=11}
# Prepare annotation — align by matrix rownames (not filter, to preserve dimension)
anno_data <- data.frame(Sample = rownames(div_matrix_age), stringsAsFactors = FALSE) %>%
  left_join(sample_metadata, by = "Sample") %>%
  dplyr::mutate(
    Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
  )

# Create column annotation
col_anno <- HeatmapAnnotation(
  Age_group = anno_data$Age_group,
  col = list(Age_group = age_colors),
  show_legend = c(Age_group = TRUE),
  annotation_name_gp = gpar(fontsize = 10),
  simple_anno_size = unit(0.4, "cm")
)

# Create heatmap
ht_age <- Heatmap(
  div_matrix_age,
  name = "KL Divergence",
  col = colorRamp2(
    c(min(div_matrix_age), median(div_matrix_age), max(div_matrix_age)),
    c("blue", "white", "red")
  ),
  top_annotation = col_anno,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  heatmap_legend_param = list(
    title = "KL Divergence",
    legend_height = unit(4, "cm")
  )
)

png(file.path(output_dir, "heatmap_age_full.png"),
    width = 2400, height = 2200, res = 300)
draw(ht_age)
dev.off()

draw(ht_age)
```

### MDS Feature Contributions - Age

```{r mds_top_genes_age}
cat("Identifying top genes driving MDS coordinates for Age...\n")

# Get MDS coordinates
mds_coords_age_mat <- data.frame(
  dim1 = mds_coords_age$MDS1,
  dim2 = mds_coords_age$MDS2,
  row.names = mds_coords_age$Sample
)

# Get sample-averaged embeddings
samples_age <- unique(sample_labels_age)
avg_pca_age <- matrix(0, nrow = length(samples_age), ncol = ncol(embedding_age),
                      dimnames = list(samples_age, colnames(embedding_age)))

for (samp in samples_age) {
  cells_in_sample <- which(sample_labels_age == samp)
  avg_pca_age[samp, ] <- colMeans(embedding_age[cells_in_sample, , drop = FALSE])
}

# Align sample order (intersect to handle missing samples)
common_age <- intersect(rownames(avg_pca_age), rownames(mds_coords_age_mat))
avg_pca_age <- avg_pca_age[common_age, ]
mds_coords_age_mat <- mds_coords_age_mat[common_age, ]

# Correlate each PC with MDS1 and MDS2
cor_mds1_age <- cor(avg_pca_age, mds_coords_age_mat[, 1])
cor_mds2_age <- cor(avg_pca_age, mds_coords_age_mat[, 2])

# Get PCA loadings
pca_loadings <- um_atlas@reductions$pca@feature.loadings

# Compute gene contributions
gene_contrib_mds1_age <- abs(pca_loadings %*% cor_mds1_age)
gene_contrib_mds2_age <- abs(pca_loadings %*% cor_mds2_age)

# Get top 30 genes for each axis
# Gene names are already in the rownames of gene_contrib matrices
contrib_vec_mds1_age <- as.vector(gene_contrib_mds1_age)
names(contrib_vec_mds1_age) <- rownames(gene_contrib_mds1_age)
top_genes_mds1_age <- names(sort(contrib_vec_mds1_age, decreasing = TRUE)[1:30])

contrib_vec_mds2_age <- as.vector(gene_contrib_mds2_age)
names(contrib_vec_mds2_age) <- rownames(gene_contrib_mds2_age)
top_genes_mds2_age <- names(sort(contrib_vec_mds2_age, decreasing = TRUE)[1:30])

cat("\nTop 30 genes driving MDS1 (Age):\n")
print(top_genes_mds1_age)

cat("\nTop 30 genes driving MDS2 (Age):\n")
print(top_genes_mds2_age)

# Save to CSV
write_csv(
  data.frame(
    Gene = top_genes_mds1_age,
    Contribution = contrib_vec_mds1_age[top_genes_mds1_age]
  ),
  file.path(output_dir, "top_genes_MDS1_age.csv")
)

write_csv(
  data.frame(
    Gene = top_genes_mds2_age,
    Contribution = contrib_vec_mds2_age[top_genes_mds2_age]
  ),
  file.path(output_dir, "top_genes_MDS2_age.csv")
)
```

---

## Secondary Mutation Analysis {.tabset}

### Secondary Mutation - All Cells

**All cells from samples with known secondary mutations (BAP1, SF3B1, EIF1AX)**

```{r gloscope_secondary_mutation_all}
cat("Running GloScope analysis for secondary mutations (All Cells)...\n")

# Filter to samples with defined secondary mutations (excluding 'none' and 'unknown')
samples_with_smut <- sample_metadata %>%
  dplyr::filter(secondary_mutation %in% c("BAP1", "SF3B1", "EIF1AX")) %>%
  pull(Sample)

cat("Found", length(samples_with_smut), "samples with known secondary mutations\n")

cells_smut <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_smut) %>% pull(cell_id)
embedding_smut <- embedding_matrix[cells_smut, ]
sample_labels_smut <- cell_metadata %>% dplyr::filter(Sample %in% samples_with_smut) %>% pull(Sample)

# Calculate or load divergence
knn_div_smut <- run_gloscope_cached(
  embedding_subset = embedding_smut,
  sample_labels = sample_labels_smut,
  cache_file = "divergence_secondary_mutation_full.rds",
  analysis_name = "Secondary Mutations (All Cells)"
)

# MDS visualization
samples_in_div_smut <- rownames(knn_div_smut)
sample_metadata_smut <- data.frame(Sample = samples_in_div_smut) %>%
  left_join(sample_metadata %>% dplyr::filter(Sample %in% samples_with_smut), by = "Sample")

mds_smut <- plotMDS(
  dist_mat = knn_div_smut,
  metadata_df = sample_metadata_smut,
  sample_id = "Sample",
  group_id = "secondary_mutation",
  k = 2
)

# ANOSIM test first to get p-value for plot
anosim_smut <- safe_anosim(
  as.dist(knn_div_smut),
  grouping = sample_metadata_smut$secondary_mutation,
  permutations = 999
)

cat("\nANOSIM test for secondary mutations (All Cells):\n")
cat("  R statistic:", round(anosim_smut$statistic, 4), "\n")
cat("  p-value:", anosim_smut$signif, "\n")

p_display_smut <- ifelse(anosim_smut$signif < 0.001, "p < 0.001",
                         paste0("p = ", format(anosim_smut$signif, digits = 3)))

p_smut_mds <- mds_smut$plot +
  scale_color_manual(values = smut_colors) +
  coord_fixed() +
  labs(
    title = "GloScope MDS: Secondary Mutations (All Cells)",
    subtitle = paste0("n = ", nrow(sample_metadata_smut), " samples | ANOSIM: R = ",
                     round(anosim_smut$statistic, 3), ", ", p_display_smut),
    color = "Secondary Mutation"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

print(p_smut_mds)

ggsave(
  file.path(output_dir, "MDS_secondary_mutation_full.png"),
  p_smut_mds,
  width = 10, height = 8, dpi = 300, bg = "white"
)

write_csv(
  tibble(
    comparison = "secondary_mutation_full",
    statistic = anosim_smut$statistic,
    p_value = anosim_smut$signif
  ),
  file.path(output_dir, "anosim_secondary_mutation_full.csv")
)
```

---

## Primary Tumor Focus {.tabset}

### Secondary Mutations (BAP1/SF3B1/EIF1AX)

**Primary uveal melanoma stratified by secondary driver mutations**

```{r gloscope_secondary_mutation}
cat("Running GloScope analysis for secondary mutations in pUM...\n")

# Include ALL primary samples (do not filter out "other"/"unknown" secondary mutations)
samples_pum <- sample_metadata %>%
  dplyr::filter(origin == "primary") %>%
  pull(Sample)

if (length(samples_pum) >= 6) {
  cells_pum <- cell_metadata %>% dplyr::filter(Sample %in% samples_pum) %>% pull(cell_id)
  embedding_pum <- embedding_matrix[cells_pum, ]
  sample_labels_pum <- cell_metadata %>% dplyr::filter(Sample %in% samples_pum) %>% pull(Sample)

  # Calculate or load divergence
  knn_div_pum <- run_gloscope_cached(
    embedding_subset = embedding_pum,
    sample_labels = sample_labels_pum,
    cache_file = "divergence_secondary_mutation_pUM_full.rds",
    analysis_name = "Secondary Mutations (pUM)"
  )

  # MDS visualization
  samples_in_div_pum <- rownames(knn_div_pum)
  sample_metadata_pum <- data.frame(Sample = samples_in_div_pum) %>%
    left_join(sample_metadata %>% dplyr::filter(Sample %in% samples_pum), by = "Sample")

  mds_pum <- plotMDS(
    dist_mat = knn_div_pum,
    metadata_df = sample_metadata_pum,
    sample_id = "Sample",
    group_id = "secondary_mutation",
    k = 2
  )

  # ANOSIM test first to get p-value for plot
  anosim_pum <- safe_anosim(
    as.dist(knn_div_pum),
    grouping = sample_metadata_pum$secondary_mutation,
    permutations = 999
  )

  cat("\nANOSIM test for secondary mutations:\n")
  cat("  R statistic:", round(anosim_pum$statistic, 4), "\n")
  cat("  p-value:", anosim_pum$signif, "\n")

  p_display_pum <- ifelse(anosim_pum$signif < 0.001, "p < 0.001",
                          paste0("p = ", format(anosim_pum$signif, digits = 3)))

  p_pum_mds <- mds_pum$plot +
    geom_point(size = 4) +
    scale_color_manual(values = smut_colors) +
    coord_fixed() +
    labs(
      title = "GloScope MDS: Secondary Mutations in Primary UM (Full Cohort)",
      subtitle = paste0("n = ", nrow(sample_metadata_pum), " samples | ANOSIM: R = ",
                       round(anosim_pum$statistic, 3), ", ", p_display_pum),
      color = "Secondary Mutation"
    ) +
    theme_minimal(base_size = 12)

  print(p_pum_mds)

  ggsave(
    file.path(output_dir, "MDS_secondary_mutation_pUM_full.png"),
    p_pum_mds,
    width = 10, height = 8, dpi = 300, bg = "white"
  )

  write_csv(
    data.frame(
      comparison = "secondary_mutation_pUM_full",
      statistic = anosim_pum$statistic,
      p_value = anosim_pum$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_secondary_mutation_pUM_full.csv")
  )

  # Divergence Heatmap
  sample_order_pum <- sample_metadata_pum %>% arrange(secondary_mutation) %>% pull(Sample)
  knn_div_ordered_pum <- knn_div_pum[sample_order_pum, sample_order_pum]

  anno_data_pum <- sample_metadata_pum[match(sample_order_pum, sample_metadata_pum$Sample), ]

  # Replace NA values with "Unknown" for annotations
  anno_data_pum_clean <- anno_data_pum %>%
    dplyr::mutate(
      Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
      Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
    )

  # Add colors for Unknown categories
  gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
  age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

  col_anno_pum <- HeatmapAnnotation(
    Patient = anno_data_pum_clean$Patient_nr,
    Gender = anno_data_pum_clean$Gender,
    Age_group = anno_data_pum_clean$Age_group,
    `Primary Mutation` = anno_data_pum_clean$primary_mutation,
    `Secondary Mutation` = anno_data_pum_clean$secondary_mutation,
    col = list(
      Gender = gender_colors_full,
      Age_group = age_colors_full,
      `Primary Mutation` = pmut_colors,
      `Secondary Mutation` = smut_colors
    ),
    annotation_name_gp = gpar(fontsize = 9),
    show_legend = c(Patient = FALSE, Gender = TRUE, Age_group = TRUE,
                    `Primary Mutation` = TRUE, `Secondary Mutation` = TRUE)
  )

  ht_pum <- Heatmap(
    knn_div_ordered_pum,
    name = "KL Divergence",
    col = colorRamp2(c(0, median(knn_div_pum), max(knn_div_pum)),
                     c("white", "orange", "red")),
    top_annotation = col_anno_pum,
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_title = "GloScope Divergence: Secondary Mutations in Primary UM (Full Cohort)",
    column_title_gp = gpar(fontsize = 14, fontface = "bold"),
    heatmap_legend_param = list(
      title = "KL Divergence",
      legend_height = unit(4, "cm")
    )
  )

  png(file.path(output_dir, "heatmap_secondary_mutation_pUM_full.png"),
      width = 2400, height = 2200, res = 300)
  draw(ht_pum)
  dev.off()

  draw(ht_pum)

  # Save heatmap object and data for recreation
  qs2::qs_save(list(ht = ht_pum, div_matrix = knn_div_ordered_pum, anno_data = anno_data_pum_clean),
               file.path(output_dir, "gloscope_heatmap_primary_tumors_objects.qs2"), nthreads = parallel::detectCores())
}
```

### Anatomical Location (Choroid/Iris/Ciliary Body)

**Primary uveal melanoma stratified by tumor location**

```{r gloscope_location}
cat("Running GloScope analysis for pUM anatomical location...\n")

samples_location <- sample_metadata %>%
  dplyr::filter(
    origin == "primary",
    LocationPrimaryTumor %in% c("choroid", "iris", "ciliary body")
  ) %>%
  pull(Sample)

if (length(samples_location) >= 6) {
  cells_location <- cell_metadata %>% dplyr::filter(Sample %in% samples_location) %>% pull(cell_id)
  embedding_location <- embedding_matrix[cells_location, ]
  sample_labels_location <- cell_metadata %>% dplyr::filter(Sample %in% samples_location) %>% pull(Sample)

  # Calculate or load divergence
  knn_div_location <- run_gloscope_cached(
    embedding_subset = embedding_location,
    sample_labels = sample_labels_location,
    cache_file = "divergence_location_pUM_full.rds",
    analysis_name = "Anatomical Location (pUM)"
  )

  # MDS visualization
  samples_in_div_location <- rownames(knn_div_location)
  sample_metadata_location <- data.frame(Sample = samples_in_div_location) %>%
    left_join(sample_metadata %>% dplyr::filter(Sample %in% samples_location), by = "Sample")

  mds_location <- plotMDS(
    dist_mat = knn_div_location,
    metadata_df = sample_metadata_location,
    sample_id = "Sample",
    group_id = "LocationPrimaryTumor",
    k = 2
  )

  # ANOSIM test first to get p-value for plot
  anosim_location <- safe_anosim(
    as.dist(knn_div_location),
    grouping = sample_metadata_location$LocationPrimaryTumor,
    permutations = 999
  )

  cat("\nANOSIM test for anatomical location:\n")
  cat("  R statistic:", round(anosim_location$statistic, 4), "\n")
  cat("  p-value:", anosim_location$signif, "\n")

  p_display_loc <- ifelse(anosim_location$signif < 0.001, "p < 0.001",
                          paste0("p = ", format(anosim_location$signif, digits = 3)))

  p_location_mds <- mds_location$plot +
    geom_point(size = 4) +
    scale_color_manual(values = primary_location_colors) +
    coord_fixed() +
    labs(
      title = "GloScope MDS: Anatomical Location in Primary UM (Full Cohort)",
      subtitle = paste0("n = ", nrow(sample_metadata_location), " samples | ANOSIM: R = ",
                       round(anosim_location$statistic, 3), ", ", p_display_loc),
      color = "Location"
    ) +
    theme_minimal(base_size = 12)

  print(p_location_mds)

  ggsave(
    file.path(output_dir, "MDS_location_pUM_full.png"),
    p_location_mds,
    width = 10, height = 8, dpi = 300, bg = "white"
  )

  write_csv(
    data.frame(
      comparison = "anatomical_location_pUM_full",
      statistic = anosim_location$statistic,
      p_value = anosim_location$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_location_pUM_full.csv")
  )

  # Divergence Heatmap
  sample_order_location <- sample_metadata_location %>% arrange(LocationPrimaryTumor) %>% pull(Sample)
  knn_div_ordered_location <- knn_div_location[sample_order_location, sample_order_location]

  anno_data_location <- sample_metadata_location[match(sample_order_location, sample_metadata_location$Sample), ]

  # Replace NA values with "Unknown" for annotations - explicit factor handling
  anno_data_location_clean <- anno_data_location %>%
    dplyr::mutate(
      Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
      Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
    )

  # Add colors for Unknown categories
  gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
  age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

  col_anno_location <- HeatmapAnnotation(
    Patient = anno_data_location_clean$Patient_nr,
    Gender = anno_data_location_clean$Gender,
    Age_group = anno_data_location_clean$Age_group,
    Location = anno_data_location_clean$LocationPrimaryTumor,
    `Primary Mutation` = anno_data_location_clean$primary_mutation,
    `Secondary Mutation` = anno_data_location_clean$secondary_mutation,
    col = list(
      Gender = gender_colors_full,
      Age_group = age_colors_full,
      Location = location_colors,
      `Primary Mutation` = pmut_colors,
      `Secondary Mutation` = smut_colors
    ),
    annotation_name_gp = gpar(fontsize = 9),
    show_legend = c(Patient = FALSE, Gender = TRUE, Age_group = TRUE,
                    Location = TRUE, `Primary Mutation` = TRUE, `Secondary Mutation` = TRUE)
  )

  ht_location <- Heatmap(
    knn_div_ordered_location,
    name = "KL Divergence",
    col = colorRamp2(c(0, median(knn_div_location), max(knn_div_location)),
                     c("white", "orange", "red")),
    top_annotation = col_anno_location,
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_title = "GloScope Divergence: Anatomical Location in Primary UM (Full Cohort)",
    column_title_gp = gpar(fontsize = 14, fontface = "bold"),
    heatmap_legend_param = list(
      title = "KL Divergence",
      legend_height = unit(4, "cm")
    )
  )

  png(file.path(output_dir, "heatmap_location_pUM_full.png"),
      width = 2400, height = 2200, res = 300)
  draw(ht_location)
  dev.off()

  draw(ht_location)
}
```

### Anatomical Location - Malignant Cells Only

```{r gloscope_location_malignant}
cat("Running GloScope analysis for Anatomical Location (Malignant Cells Only)...\n")

# Filter to malignant cells from primary tumors with location data
cells_location_mal <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_location, Malignant == "Malignant") %>%
  pull(cell_id)

sample_labels_location_mal <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_location, Malignant == "Malignant") %>%
  pull(Sample)

# Check sample representation
sample_cell_counts_loc_mal <- table(sample_labels_location_mal)
valid_samples_loc_mal <- names(sample_cell_counts_loc_mal)[sample_cell_counts_loc_mal >= 50]

cat("  - Total malignant cells from pUM:", length(cells_location_mal), "\n")
cat("  - Samples with >=50 malignant cells:", length(valid_samples_loc_mal), "\n")

if (length(valid_samples_loc_mal) >= 4) {
  cells_loc_mal_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_loc_mal) %>%
    pull(cell_id)

  embedding_loc_mal <- embedding_matrix[cells_loc_mal_filtered, ]
  sample_labels_loc_mal_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_loc_mal) %>%
    pull(Sample)

  div_matrix_loc_mal <- run_gloscope_cached(
    embedding_subset = embedding_loc_mal,
    sample_labels = sample_labels_loc_mal_filtered,
    cache_file = "divergence_location_malignant.rds",
    analysis_name = "Anatomical Location (Malignant Cells)"
  )

  # ANOSIM test
  loc_vec_mal <- sample_metadata$LocationPrimaryTumor[match(rownames(div_matrix_loc_mal), sample_metadata$Sample)]

  anosim_loc_mal <- safe_anosim(div_matrix_loc_mal, loc_vec_mal, permutations = 999)

  cat("\nANOSIM test for Location (malignant cells):\n")
  cat("  R statistic:", round(anosim_loc_mal$statistic, 4), "\n")
  cat("  p-value:", anosim_loc_mal$signif, "\n")

  write_csv(
    tibble(
      comparison = "location_malignant_full",
      statistic = anosim_loc_mal$statistic,
      p_value = anosim_loc_mal$signif
    ),
    file.path(output_dir, "anosim_location_malignant_full.csv")
  )
} else {
  cat("Insufficient samples with adequate malignant cell representation\n")
}
```

### Anatomical Location - Melanocytic Cells Only

```{r gloscope_location_melanocytic}
cat("Running GloScope analysis for Anatomical Location (Melanocytic Cells Only)...\n")

# Filter to melanocytic cells from primary tumors
cells_location_mel <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_location,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(cell_id)

sample_labels_location_mel <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_location,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(Sample)

# Check sample representation
sample_cell_counts_loc_mel <- table(sample_labels_location_mel)
valid_samples_loc_mel <- names(sample_cell_counts_loc_mel)[sample_cell_counts_loc_mel >= 50]

cat("  - Total melanocytic cells from pUM:", length(cells_location_mel), "\n")
cat("  - Samples with >=50 melanocytic cells:", length(valid_samples_loc_mel), "\n")

if (length(valid_samples_loc_mel) >= 4) {
  cells_loc_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_loc_mel) %>%
    pull(cell_id)

  embedding_loc_mel <- embedding_matrix[cells_loc_mel_filtered, ]
  sample_labels_loc_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_loc_mel) %>%
    pull(Sample)

  div_matrix_loc_mel <- run_gloscope_cached(
    embedding_subset = embedding_loc_mel,
    sample_labels = sample_labels_loc_mel_filtered,
    cache_file = "divergence_location_melanocytic.rds",
    analysis_name = "Anatomical Location (Melanocytic Cells)"
  )

  # ANOSIM test
  loc_vec_mel <- sample_metadata$LocationPrimaryTumor[match(rownames(div_matrix_loc_mel), sample_metadata$Sample)]

  anosim_loc_mel <- safe_anosim(div_matrix_loc_mel, loc_vec_mel, permutations = 999)

  cat("\nANOSIM test for Location (melanocytic cells):\n")
  cat("  R statistic:", round(anosim_loc_mel$statistic, 4), "\n")
  cat("  p-value:", anosim_loc_mel$signif, "\n")

  write_csv(
    tibble(
      comparison = "location_melanocytic_full",
      statistic = anosim_loc_mel$statistic,
      p_value = anosim_loc_mel$signif
    ),
    file.path(output_dir, "anosim_location_melanocytic_full.csv")
  )
} else {
  cat("Insufficient samples with adequate melanocytic cell representation\n")
}
```

```
## Malignant Cells Only {.tabset}

### Disease Stage (Malignant)

**Focus on tumor-intrinsic heterogeneity by analyzing malignant cells only**

```{r gloscope_malignant_origin}
cat("Running GloScope analysis for malignant cells by disease origin (FULL COHORT)...\n")

# Filter to malignant cells
cells_malignant <- cell_metadata %>%
  dplyr::filter(Malignant == "Malignant") %>%
  pull(cell_id)

embedding_malignant <- embedding_matrix[cells_malignant, ]
sample_labels_malignant <- cell_metadata %>%
  dplyr::filter(Malignant == "Malignant") %>%
  pull(Sample)

# Check sample representation (need at least 100 cells per sample for robust estimation)
sample_cell_counts <- table(sample_labels_malignant)
valid_samples <- names(sample_cell_counts)[sample_cell_counts >= 100]

cat("  - Total malignant cells:", length(cells_malignant), "\n")
cat("  - Samples with >=100 malignant cells:", length(valid_samples), "\n")

if (length(valid_samples) >= 6) {
  # Filter to valid samples
  cells_malignant_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples) %>%
    pull(cell_id)

  embedding_malignant_filtered <- embedding_matrix[cells_malignant_filtered, ]
  sample_labels_malignant_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples) %>%
    pull(Sample)

  # Calculate or load divergence
  knn_div_malignant <- run_gloscope_cached(
    embedding_subset = embedding_malignant_filtered,
    sample_labels = sample_labels_malignant_filtered,
    cache_file = "divergence_origin_malignant_full.rds",
    analysis_name = "Disease Origin (Malignant Cells)"
  )

  # MDS visualization
  samples_in_div_malignant <- rownames(knn_div_malignant)
  sample_metadata_malignant <- data.frame(Sample = samples_in_div_malignant) %>%
    left_join(sample_metadata %>% dplyr::filter(Sample %in% valid_samples), by = "Sample")

  mds_malignant <- plotMDS(
    dist_mat = knn_div_malignant,
    metadata_df = sample_metadata_malignant,
    sample_id = "Sample",
    group_id = "origin",
    k = 2
  )

  # ANOSIM test first to get p-value for plot
  anosim_malignant <- safe_anosim(
    as.dist(knn_div_malignant),
    grouping = sample_metadata_malignant$origin,
    permutations = 999
  )

  cat("\nANOSIM test for disease origin (malignant cells):\n")
  cat("  R statistic:", round(anosim_malignant$statistic, 4), "\n")
  cat("  p-value:", anosim_malignant$signif, "\n")

  # Format p-value for display
  p_display_mal <- ifelse(anosim_malignant$signif < 0.001, "p < 0.001",
                          paste0("p = ", format(anosim_malignant$signif, digits = 3)))

  p_malignant_mds <- mds_malignant$plot +
    scale_color_manual(values = origin_colors) +
    coord_fixed() +
    labs(
      title = "GloScope MDS: Disease Progression (Malignant Cells Only - Full Cohort)",
      subtitle = paste0("n = ", nrow(sample_metadata_malignant), " samples | ANOSIM: R = ",
                       round(anosim_malignant$statistic, 3), ", ", p_display_mal),
      color = "Origin"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right", plot.title = element_text(face = "bold", size = 14))

  print(p_malignant_mds)

  ggsave(
    file.path(output_dir, "MDS_origin_malignant_full.png"),
    p_malignant_mds,
    width = 10, height = 8, dpi = 300, bg = "white"
  )

  write_csv(
    data.frame(
      comparison = "disease_origin_malignant_full",
      statistic = anosim_malignant$statistic,
      p_value = anosim_malignant$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_origin_malignant_full.csv")
  )

  # Divergence Heatmap
  sample_order_malignant <- sample_metadata_malignant %>% arrange(origin) %>% pull(Sample)
  knn_div_ordered_malignant <- knn_div_malignant[sample_order_malignant, sample_order_malignant]

  anno_data_malignant <- sample_metadata_malignant[match(sample_order_malignant, sample_metadata_malignant$Sample), ]

  # Replace NA values with "Unknown" for annotations
  anno_data_malignant_clean <- anno_data_malignant %>%
    dplyr::mutate(
      Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
      Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
    )

  # Add colors for Unknown categories
  gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
  age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

  col_anno_malignant <- HeatmapAnnotation(
    Patient = anno_data_malignant_clean$Patient_nr,
    Origin = anno_data_malignant_clean$origin,
    Gender = anno_data_malignant_clean$Gender,
    Age_group = anno_data_malignant_clean$Age_group,
    `Primary Mutation` = anno_data_malignant_clean$primary_mutation,
    `Secondary Mutation` = anno_data_malignant_clean$secondary_mutation,
    col = list(
      Origin = origin_colors,
      Gender = gender_colors_full,
      Age_group = age_colors_full,
      `Primary Mutation` = pmut_colors,
      `Secondary Mutation` = smut_colors
    ),
    annotation_name_gp = gpar(fontsize = 9),
    show_legend = c(Patient = FALSE, Origin = TRUE, Gender = TRUE, Age_group = TRUE,
                    `Primary Mutation` = TRUE, `Secondary Mutation` = TRUE)
  )

  ht_malignant <- Heatmap(
    knn_div_ordered_malignant,
    name = "KL Divergence",
    col = colorRamp2(c(0, median(knn_div_malignant), max(knn_div_malignant)),
                     c("white", "orange", "red")),
    top_annotation = col_anno_malignant,
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_title = "GloScope Divergence: Disease Origin (Malignant Cells Only - Full Cohort)",
    column_title_gp = gpar(fontsize = 14, fontface = "bold"),
    heatmap_legend_param = list(
      title = "KL Divergence",
      legend_height = unit(4, "cm")
    )
  )

  png(file.path(output_dir, "heatmap_origin_malignant_full.png"),
      width = 2800, height = 2600, res = 300)
  draw(ht_malignant)
  dev.off()

  draw(ht_malignant)

  # Save heatmap object and data for recreation
  qs2::qs_save(list(ht = ht_malignant, div_matrix = knn_div_ordered_malignant, anno_data = anno_data_malignant_clean),
               file.path(output_dir, "gloscope_heatmap_malignant_cells_objects.qs2"), nthreads = parallel::detectCores())
} else {
  cat("Insufficient samples with adequate malignant cell representation\n")
}
```

### Gender (Malignant)

**Comparison:** Male vs Female (Malignant cells only)

```{r gloscope_malignant_gender}
cat("Running GloScope analysis for Gender (malignant cells only)...\n")

# Filter to malignant cells with defined gender
samples_with_gender <- sample_metadata %>%
  dplyr::filter(!is.na(Gender) & Gender %in% c("M", "F")) %>%
  pull(Sample)

cells_malignant_gender <- cell_metadata %>%
  dplyr::filter(Malignant == "Malignant", Sample %in% samples_with_gender) %>%
  pull(cell_id)

embedding_malignant_gender <- embedding_matrix[cells_malignant_gender, ]
sample_labels_malignant_gender <- cell_metadata %>%
  dplyr::filter(Malignant == "Malignant", Sample %in% samples_with_gender) %>%
  pull(Sample)

# Check sample representation
sample_cell_counts <- table(sample_labels_malignant_gender)
valid_samples <- names(sample_cell_counts)[sample_cell_counts >= 100]

cat("  - Total malignant cells:", length(cells_malignant_gender), "\n")
cat("  - Samples with >=100 malignant cells:", length(valid_samples), "\n")

if (length(valid_samples) >= 6) {
  # Filter to valid samples
  cells_malignant_gender_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples) %>%
    pull(cell_id)

  embedding_malignant_gender_filtered <- embedding_matrix[cells_malignant_gender_filtered, ]
  sample_labels_malignant_gender_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples) %>%
    pull(Sample)

  # Calculate or load divergence
  knn_div_malignant_gender <- run_gloscope_cached(
    embedding_subset = embedding_malignant_gender_filtered,
    sample_labels = sample_labels_malignant_gender_filtered,
    cache_file = "divergence_gender_malignant.rds",
    analysis_name = "Gender (Malignant Cells)"
  )

  # MDS visualization
  samples_in_div_malignant_gender <- rownames(knn_div_malignant_gender)
  sample_metadata_malignant_gender <- data.frame(Sample = samples_in_div_malignant_gender) %>%
    left_join(sample_metadata %>% dplyr::filter(Sample %in% valid_samples), by = "Sample")

  mds_malignant_gender <- plotMDS(
    dist_mat = knn_div_malignant_gender,
    metadata_df = sample_metadata_malignant_gender,
    sample_id = "Sample",
    group_id = "Gender",
    k = 2
  )

  # ANOSIM test first to get p-value for plot
  anosim_malignant_gender <- safe_anosim(
    as.dist(knn_div_malignant_gender),
    grouping = sample_metadata_malignant_gender$Gender,
    permutations = 999
  )

  cat("\nANOSIM test for Gender (malignant cells):\n")
  cat("  R statistic:", round(anosim_malignant_gender$statistic, 4), "\n")
  cat("  p-value:", anosim_malignant_gender$signif, "\n")

  p_display_gender <- ifelse(anosim_malignant_gender$signif < 0.001, "p < 0.001",
                             paste0("p = ", format(anosim_malignant_gender$signif, digits = 3)))

  p_malignant_gender_mds <- mds_malignant_gender$plot +
    scale_color_manual(values = gender_colors) +
    coord_fixed() +
    labs(
      title = "GloScope MDS: Gender (Malignant Cells Only)",
      subtitle = paste0("n = ", nrow(sample_metadata_malignant_gender), " samples | ANOSIM: R = ",
                       round(anosim_malignant_gender$statistic, 3), ", ", p_display_gender),
      color = "Gender"
    ) +
    theme_minimal(base_size = 12)

  print(p_malignant_gender_mds)

  ggsave(
    file.path(output_dir, "mds_gender_malignant.png"),
    p_malignant_gender_mds, width = 8, height = 6, dpi = 300, bg = "white"
  )

  write_csv(
    data.frame(
      comparison = "gender_malignant",
      statistic = anosim_malignant_gender$statistic,
      p_value = anosim_malignant_gender$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_gender_malignant_full.csv")
  )
} else {
  cat("Insufficient samples with adequate malignant cell representation\n")
}
```

```{r heatmap_malignant_gender, eval=length(valid_samples) >= 6}
cat("Creating divergence heatmap for Gender (malignant cells)...\n")

# Prepare data for heatmap
sample_order_malignant_gender <- sample_metadata_malignant_gender %>% arrange(Gender) %>% pull(Sample)
knn_div_ordered_malignant_gender <- knn_div_malignant_gender[sample_order_malignant_gender, sample_order_malignant_gender]

anno_data_malignant_gender <- sample_metadata_malignant_gender[match(sample_order_malignant_gender, sample_metadata_malignant_gender$Sample), ]

# Replace NA values with "Unknown" for annotations
anno_data_malignant_gender_clean <- anno_data_malignant_gender %>%
  dplyr::mutate(
    Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
    Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
  )

# Add colors for Unknown categories
gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

col_anno_malignant_gender <- HeatmapAnnotation(
  Patient = anno_data_malignant_gender_clean$Patient_nr,
  Gender = anno_data_malignant_gender_clean$Gender,
  Age_group = anno_data_malignant_gender_clean$Age_group,
  Origin = anno_data_malignant_gender_clean$origin,
  col = list(
    Gender = gender_colors_full,
    Age_group = age_colors_full,
    Origin = origin_colors
  ),
  annotation_name_gp = gpar(fontsize = 9),
  show_legend = c(Patient = FALSE, Gender = TRUE, Age_group = TRUE, Origin = TRUE)
)

ht_malignant_gender <- Heatmap(
  knn_div_ordered_malignant_gender,
  name = "KL Divergence",
  col = colorRamp2(c(0, median(knn_div_malignant_gender), max(knn_div_malignant_gender)),
                   c("white", "orange", "red")),
  top_annotation = col_anno_malignant_gender,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "Gender Comparison (Malignant Cells Only)",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  heatmap_legend_param = list(
    title = "KL Divergence",
    legend_height = unit(3, "cm")
  )
)

png(file.path(output_dir, "heatmap_gender_malignant.png"),
    width = 2800, height = 2600, res = 300)
draw(ht_malignant_gender)
dev.off()

draw(ht_malignant_gender)
```

```{r mds_top_genes_malignant_gender, eval=length(valid_samples) >= 6}
cat("Identifying top genes driving MDS coordinates for Gender (malignant cells)...\n")

# Get MDS coordinates
mds_coords_malignant_gender <- mds_malignant_gender$df[, c("dim1", "dim2")]
rownames(mds_coords_malignant_gender) <- mds_malignant_gender$df$Sample

# Get sample-averaged embeddings
samples_malignant_gender <- unique(sample_labels_malignant_gender_filtered)
avg_pca_malignant_gender <- matrix(0, nrow = length(samples_malignant_gender), ncol = ncol(embedding_malignant_gender_filtered),
                  dimnames = list(samples_malignant_gender, colnames(embedding_malignant_gender_filtered)))

for (samp in samples_malignant_gender) {
  cells_in_sample <- which(sample_labels_malignant_gender_filtered == samp)
  avg_pca_malignant_gender[samp, ] <- colMeans(embedding_malignant_gender_filtered[cells_in_sample, , drop = FALSE])
}

# Align sample order (intersect to handle missing samples)
common_mal_gender <- intersect(rownames(avg_pca_malignant_gender), rownames(mds_coords_malignant_gender))
avg_pca_malignant_gender <- avg_pca_malignant_gender[common_mal_gender, ]
mds_coords_malignant_gender <- mds_coords_malignant_gender[common_mal_gender, ]

# Correlate each PC with MDS1 and MDS2
cor_mds1_malignant_gender <- cor(avg_pca_malignant_gender, mds_coords_malignant_gender[, 1])
cor_mds2_malignant_gender <- cor(avg_pca_malignant_gender, mds_coords_malignant_gender[, 2])

# Get PCA loadings
pca_loadings <- um_atlas@reductions$pca@feature.loadings

# Compute gene contributions
gene_contrib_mds1_malignant_gender <- abs(pca_loadings %*% cor_mds1_malignant_gender)
gene_contrib_mds2_malignant_gender <- abs(pca_loadings %*% cor_mds2_malignant_gender)

# Get top 30 genes for each axis
# Gene names are already in the rownames of gene_contrib matrices
contrib_vec_mds1_malignant_gender <- as.vector(gene_contrib_mds1_malignant_gender)
names(contrib_vec_mds1_malignant_gender) <- rownames(gene_contrib_mds1_malignant_gender)
top_genes_mds1_malignant_gender <- names(sort(contrib_vec_mds1_malignant_gender, decreasing = TRUE)[1:30])

contrib_vec_mds2_malignant_gender <- as.vector(gene_contrib_mds2_malignant_gender)
names(contrib_vec_mds2_malignant_gender) <- rownames(gene_contrib_mds2_malignant_gender)
top_genes_mds2_malignant_gender <- names(sort(contrib_vec_mds2_malignant_gender, decreasing = TRUE)[1:30])

cat("\nTop 30 genes driving MDS1 (Gender - Malignant):\n")
print(top_genes_mds1_malignant_gender)

cat("\nTop 30 genes driving MDS2 (Gender - Malignant):\n")
print(top_genes_mds2_malignant_gender)

# Save to CSV
write_csv(
  data.frame(
    Gene = top_genes_mds1_malignant_gender,
    Contribution = contrib_vec_mds1_malignant_gender[top_genes_mds1_malignant_gender]
  ),
  file.path(output_dir, "top_genes_MDS1_gender_malignant.csv")
)

write_csv(
  data.frame(
    Gene = top_genes_mds2_malignant_gender,
    Contribution = contrib_vec_mds2_malignant_gender[top_genes_mds2_malignant_gender]
  ),
  file.path(output_dir, "top_genes_MDS2_gender_malignant.csv")
)
```

### Gender - Melanocytic Cells Only

```{r gloscope_gender_melanocytic}
cat("Running GloScope analysis for Gender (Melanocytic Cells Only)...\n")

# Filter to melanocytic cells with gender data
cells_gender_melanocytic <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_with_gender,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(cell_id)

sample_labels_gender_melanocytic <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_with_gender,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(Sample)

sample_cell_counts_gender_mel <- table(sample_labels_gender_melanocytic)
valid_samples_gender_mel <- names(sample_cell_counts_gender_mel)[sample_cell_counts_gender_mel >= 50]

cat("  - Total melanocytic cells:", length(cells_gender_melanocytic), "\n")
cat("  - Samples with >=50 melanocytic cells:", length(valid_samples_gender_mel), "\n")

if (length(valid_samples_gender_mel) >= 4) {
  cells_gender_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_gender_mel) %>%
    pull(cell_id)

  embedding_gender_mel <- embedding_matrix[cells_gender_mel_filtered, ]
  sample_labels_gender_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_gender_mel) %>%
    pull(Sample)

  div_matrix_gender_mel <- run_gloscope_cached(
    embedding_subset = embedding_gender_mel,
    sample_labels = sample_labels_gender_mel_filtered,
    cache_file = "divergence_gender_melanocytic.rds",
    analysis_name = "Gender (Melanocytic Cells)"
  )

  # ANOSIM test
  gender_vec_mel <- sample_metadata$Gender[match(rownames(div_matrix_gender_mel), sample_metadata$Sample)]

  anosim_gender_mel <- safe_anosim(div_matrix_gender_mel, gender_vec_mel, permutations = 999)

  cat("\nANOSIM test for Gender (melanocytic cells):\n")
  cat("  R statistic:", round(anosim_gender_mel$statistic, 4), "\n")
  cat("  p-value:", anosim_gender_mel$signif, "\n")

  write_csv(
    tibble(
      comparison = "gender_melanocytic_full",
      statistic = anosim_gender_mel$statistic,
      p_value = anosim_gender_mel$signif
    ),
    file.path(output_dir, "anosim_gender_melanocytic_full.csv")
  )
} else {
  cat("Insufficient samples with adequate melanocytic cell representation\n")
}
```

### Gender - Primary Tumors Only

```{r gloscope_gender_pUM}
cat("Running GloScope analysis for Gender (Primary Tumors Only)...\n")

# Filter to primary tumor samples with gender data
samples_pUM_gender <- sample_metadata %>%
  dplyr::filter(origin == "primary", !is.na(Gender), Gender %in% c("M", "F")) %>%
  pull(Sample)

cells_gender_pUM <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_pUM_gender) %>%
  pull(cell_id)

sample_labels_gender_pUM <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_pUM_gender) %>%
  pull(Sample)

cat("  - Total cells from pUM:", length(cells_gender_pUM), "\n")
cat("  - pUM samples with gender:", length(unique(sample_labels_gender_pUM)), "\n")

if (length(unique(sample_labels_gender_pUM)) >= 4) {
  embedding_gender_pUM <- embedding_matrix[cells_gender_pUM, ]

  div_matrix_gender_pUM <- run_gloscope_cached(
    embedding_subset = embedding_gender_pUM,
    sample_labels = sample_labels_gender_pUM,
    cache_file = "divergence_gender_pUM.rds",
    analysis_name = "Gender (Primary Tumors)"
  )

  # ANOSIM test
  gender_vec_pUM <- sample_metadata$Gender[match(rownames(div_matrix_gender_pUM), sample_metadata$Sample)]

  anosim_gender_pUM <- safe_anosim(div_matrix_gender_pUM, gender_vec_pUM, permutations = 999)

  cat("\nANOSIM test for Gender (primary tumors):\n")
  cat("  R statistic:", round(anosim_gender_pUM$statistic, 4), "\n")
  cat("  p-value:", anosim_gender_pUM$signif, "\n")

  write_csv(
    tibble(
      comparison = "gender_pUM_full",
      statistic = anosim_gender_pUM$statistic,
      p_value = anosim_gender_pUM$signif
    ),
    file.path(output_dir, "anosim_gender_pUM_full.csv")
  )
} else {
  cat("Insufficient pUM samples with gender data\n")
}
```

---

### Age (Malignant)

**Comparison:** Age groups (Young <40, Middle 40-60, Old 60+) (Malignant cells only)

```{r gloscope_malignant_age}
cat("Running GloScope analysis for Age (malignant cells only)...\n")

# Filter to malignant cells with defined age
samples_with_age <- sample_metadata %>%
  dplyr::filter(Age_group != "Unknown") %>%
  pull(Sample)

cells_malignant_age <- cell_metadata %>%
  dplyr::filter(Malignant == "Malignant", Sample %in% samples_with_age) %>%
  pull(cell_id)

embedding_malignant_age <- embedding_matrix[cells_malignant_age, ]
sample_labels_malignant_age <- cell_metadata %>%
  dplyr::filter(Malignant == "Malignant", Sample %in% samples_with_age) %>%
  pull(Sample)

# Check sample representation
sample_cell_counts <- table(sample_labels_malignant_age)
valid_samples_age <- names(sample_cell_counts)[sample_cell_counts >= 100]

cat("  - Total malignant cells:", length(cells_malignant_age), "\n")
cat("  - Samples with >=100 malignant cells:", length(valid_samples_age), "\n")

if (length(valid_samples_age) >= 6) {
  # Filter to valid samples
  cells_malignant_age_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_age) %>%
    pull(cell_id)

  embedding_malignant_age_filtered <- embedding_matrix[cells_malignant_age_filtered, ]
  sample_labels_malignant_age_filtered <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_age) %>%
    pull(Sample)

  # Calculate or load divergence
  knn_div_malignant_age <- run_gloscope_cached(
    embedding_subset = embedding_malignant_age_filtered,
    sample_labels = sample_labels_malignant_age_filtered,
    cache_file = "divergence_age_malignant.rds",
    analysis_name = "Age (Malignant Cells)"
  )

  # MDS visualization
  samples_in_div_malignant_age <- rownames(knn_div_malignant_age)
  sample_metadata_malignant_age <- data.frame(Sample = samples_in_div_malignant_age) %>%
    left_join(sample_metadata %>% dplyr::filter(Sample %in% valid_samples_age), by = "Sample")

  mds_malignant_age <- plotMDS(
    dist_mat = knn_div_malignant_age,
    metadata_df = sample_metadata_malignant_age,
    sample_id = "Sample",
    group_id = "Age_group",
    k = 2
  )

  # ANOSIM test first to get p-value for plot
  anosim_malignant_age <- safe_anosim(
    as.dist(knn_div_malignant_age),
    grouping = sample_metadata_malignant_age$Age_group,
    permutations = 999
  )

  cat("\nANOSIM test for Age (malignant cells):\n")
  cat("  R statistic:", round(anosim_malignant_age$statistic, 4), "\n")
  cat("  p-value:", anosim_malignant_age$signif, "\n")

  p_display_age <- ifelse(anosim_malignant_age$signif < 0.001, "p < 0.001",
                          paste0("p = ", format(anosim_malignant_age$signif, digits = 3)))

  p_malignant_age_mds <- mds_malignant_age$plot +
    scale_color_manual(values = age_colors) +
    coord_fixed() +
    labs(
      title = "GloScope MDS: Age Groups (Malignant Cells Only)",
      subtitle = paste0("n = ", nrow(sample_metadata_malignant_age), " samples | ANOSIM: R = ",
                       round(anosim_malignant_age$statistic, 3), ", ", p_display_age),
      color = "Age Group"
    ) +
    theme_minimal(base_size = 12)

  print(p_malignant_age_mds)

  ggsave(
    file.path(output_dir, "mds_age_malignant.png"),
    p_malignant_age_mds, width = 8, height = 6, dpi = 300, bg = "white"
  )

  write_csv(
    data.frame(
      comparison = "age_malignant",
      statistic = anosim_malignant_age$statistic,
      p_value = anosim_malignant_age$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_age_malignant_full.csv")
  )
} else {
  cat("Insufficient samples with adequate malignant cell representation\n")
}
```

```{r heatmap_malignant_age, eval=length(valid_samples_age) >= 6}
cat("Creating divergence heatmap for Age (malignant cells)...\n")

# Prepare data for heatmap
sample_order_malignant_age <- sample_metadata_malignant_age %>% arrange(Age_group) %>% pull(Sample)
knn_div_ordered_malignant_age <- knn_div_malignant_age[sample_order_malignant_age, sample_order_malignant_age]

anno_data_malignant_age <- sample_metadata_malignant_age[match(sample_order_malignant_age, sample_metadata_malignant_age$Sample), ]

# Replace NA values with "Unknown" for annotations
anno_data_malignant_age_clean <- anno_data_malignant_age %>%
  dplyr::mutate(
    Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
    Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
  )

# Add colors for Unknown categories
gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

col_anno_malignant_age <- HeatmapAnnotation(
  Patient = anno_data_malignant_age_clean$Patient_nr,
  Gender = anno_data_malignant_age_clean$Gender,
  Age_group = anno_data_malignant_age_clean$Age_group,
  Origin = anno_data_malignant_age_clean$origin,
  col = list(
    Gender = gender_colors_full,
    Age_group = age_colors_full,
    Origin = origin_colors
  ),
  annotation_name_gp = gpar(fontsize = 9),
  show_legend = c(Patient = FALSE, Gender = TRUE, Age_group = TRUE, Origin = TRUE)
)

ht_malignant_age <- Heatmap(
  knn_div_ordered_malignant_age,
  name = "KL Divergence",
  col = colorRamp2(c(0, median(knn_div_malignant_age), max(knn_div_malignant_age)),
                   c("white", "orange", "red")),
  top_annotation = col_anno_malignant_age,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "Age Group Comparison (Malignant Cells Only)",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  heatmap_legend_param = list(
    title = "KL Divergence",
    legend_height = unit(3, "cm")
  )
)

png(file.path(output_dir, "heatmap_age_malignant.png"),
    width = 2800, height = 2600, res = 300)
draw(ht_malignant_age)
dev.off()

draw(ht_malignant_age)
```

```{r mds_top_genes_malignant_age, eval=length(valid_samples_age) >= 6}
cat("Identifying top genes driving MDS coordinates for Age (malignant cells)...\n")

# Get MDS coordinates
mds_coords_malignant_age <- mds_malignant_age$df[, c("dim1", "dim2")]
rownames(mds_coords_malignant_age) <- mds_malignant_age$df$Sample

# Get sample-averaged embeddings
samples_malignant_age <- unique(sample_labels_malignant_age_filtered)
avg_pca_malignant_age <- matrix(0, nrow = length(samples_malignant_age), ncol = ncol(embedding_malignant_age_filtered),
                  dimnames = list(samples_malignant_age, colnames(embedding_malignant_age_filtered)))

for (samp in samples_malignant_age) {
  cells_in_sample <- which(sample_labels_malignant_age_filtered == samp)
  avg_pca_malignant_age[samp, ] <- colMeans(embedding_malignant_age_filtered[cells_in_sample, , drop = FALSE])
}

# Align sample order (intersect to handle missing samples)
common_mal_age <- intersect(rownames(avg_pca_malignant_age), rownames(mds_coords_malignant_age))
avg_pca_malignant_age <- avg_pca_malignant_age[common_mal_age, ]
mds_coords_malignant_age <- mds_coords_malignant_age[common_mal_age, ]

# Correlate each PC with MDS1 and MDS2
cor_mds1_malignant_age <- cor(avg_pca_malignant_age, mds_coords_malignant_age[, 1])
cor_mds2_malignant_age <- cor(avg_pca_malignant_age, mds_coords_malignant_age[, 2])

# Get PCA loadings
pca_loadings <- um_atlas@reductions$pca@feature.loadings

# Compute gene contributions
gene_contrib_mds1_malignant_age <- abs(pca_loadings %*% cor_mds1_malignant_age)
gene_contrib_mds2_malignant_age <- abs(pca_loadings %*% cor_mds2_malignant_age)

# Get top 30 genes for each axis
# Gene names are already in the rownames of gene_contrib matrices
contrib_vec_mds1_malignant_age <- as.vector(gene_contrib_mds1_malignant_age)
names(contrib_vec_mds1_malignant_age) <- rownames(gene_contrib_mds1_malignant_age)
top_genes_mds1_malignant_age <- names(sort(contrib_vec_mds1_malignant_age, decreasing = TRUE)[1:30])

contrib_vec_mds2_malignant_age <- as.vector(gene_contrib_mds2_malignant_age)
names(contrib_vec_mds2_malignant_age) <- rownames(gene_contrib_mds2_malignant_age)
top_genes_mds2_malignant_age <- names(sort(contrib_vec_mds2_malignant_age, decreasing = TRUE)[1:30])

cat("\nTop 30 genes driving MDS1 (Age - Malignant):\n")
print(top_genes_mds1_malignant_age)

cat("\nTop 30 genes driving MDS2 (Age - Malignant):\n")
print(top_genes_mds2_malignant_age)

# Save to CSV
write_csv(
  data.frame(
    Gene = top_genes_mds1_malignant_age,
    Contribution = contrib_vec_mds1_malignant_age[top_genes_mds1_malignant_age]
  ),
  file.path(output_dir, "top_genes_MDS1_age_malignant.csv")
)

write_csv(
  data.frame(
    Gene = top_genes_mds2_malignant_age,
    Contribution = contrib_vec_mds2_malignant_age[top_genes_mds2_malignant_age]
  ),
  file.path(output_dir, "top_genes_MDS2_age_malignant.csv")
)
```

### Age - Melanocytic Cells Only

```{r gloscope_age_melanocytic}
cat("Running GloScope analysis for Age (Melanocytic Cells Only)...\n")

# Filter to melanocytic cells with age data
cells_age_melanocytic <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_with_age,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(cell_id)

sample_labels_age_melanocytic <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_with_age,
                (Malignant == "Malignant" | majority_celltype == "Melanocytes")) %>%
  pull(Sample)

sample_cell_counts_age_mel <- table(sample_labels_age_melanocytic)
valid_samples_age_mel <- names(sample_cell_counts_age_mel)[sample_cell_counts_age_mel >= 50]

cat("  - Total melanocytic cells:", length(cells_age_melanocytic), "\n")
cat("  - Samples with >=50 melanocytic cells:", length(valid_samples_age_mel), "\n")

if (length(valid_samples_age_mel) >= 4) {
  cells_age_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_age_mel) %>%
    pull(cell_id)

  embedding_age_mel <- embedding_matrix[cells_age_mel_filtered, ]
  sample_labels_age_mel_filtered <- cell_metadata %>%
    dplyr::filter((Malignant == "Malignant" | majority_celltype == "Melanocytes"),
                  Sample %in% valid_samples_age_mel) %>%
    pull(Sample)

  div_matrix_age_mel <- run_gloscope_cached(
    embedding_subset = embedding_age_mel,
    sample_labels = sample_labels_age_mel_filtered,
    cache_file = "divergence_age_melanocytic.rds",
    analysis_name = "Age (Melanocytic Cells)"
  )

  # ANOSIM test
  age_vec_mel <- sample_metadata$Age_group[match(rownames(div_matrix_age_mel), sample_metadata$Sample)]

  anosim_age_mel <- safe_anosim(div_matrix_age_mel, age_vec_mel, permutations = 999)

  cat("\nANOSIM test for Age (melanocytic cells):\n")
  cat("  R statistic:", round(anosim_age_mel$statistic, 4), "\n")
  cat("  p-value:", anosim_age_mel$signif, "\n")

  write_csv(
    tibble(
      comparison = "age_melanocytic_full",
      statistic = anosim_age_mel$statistic,
      p_value = anosim_age_mel$signif
    ),
    file.path(output_dir, "anosim_age_melanocytic_full.csv")
  )
} else {
  cat("Insufficient samples with adequate melanocytic cell representation\n")
}
```

### Age - Primary Tumors Only

```{r gloscope_age_pUM}
cat("Running GloScope analysis for Age (Primary Tumors Only)...\n")

# Filter to primary tumor samples with age data
samples_pUM_age <- sample_metadata %>%
  dplyr::filter(origin == "primary", Age_group != "Unknown") %>%
  pull(Sample)

cells_age_pUM <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_pUM_age) %>%
  pull(cell_id)

sample_labels_age_pUM <- cell_metadata %>%
  dplyr::filter(Sample %in% samples_pUM_age) %>%
  pull(Sample)

cat("  - Total cells from pUM:", length(cells_age_pUM), "\n")
cat("  - pUM samples with age:", length(unique(sample_labels_age_pUM)), "\n")

if (length(unique(sample_labels_age_pUM)) >= 4) {
  embedding_age_pUM <- embedding_matrix[cells_age_pUM, ]

  div_matrix_age_pUM <- run_gloscope_cached(
    embedding_subset = embedding_age_pUM,
    sample_labels = sample_labels_age_pUM,
    cache_file = "divergence_age_pUM.rds",
    analysis_name = "Age (Primary Tumors)"
  )

  # ANOSIM test
  age_vec_pUM <- sample_metadata$Age_group[match(rownames(div_matrix_age_pUM), sample_metadata$Sample)]

  anosim_age_pUM <- safe_anosim(div_matrix_age_pUM, age_vec_pUM, permutations = 999)

  cat("\nANOSIM test for Age (primary tumors):\n")
  cat("  R statistic:", round(anosim_age_pUM$statistic, 4), "\n")
  cat("  p-value:", anosim_age_pUM$signif, "\n")

  write_csv(
    tibble(
      comparison = "age_pUM_full",
      statistic = anosim_age_pUM$statistic,
      p_value = anosim_age_pUM$signif
    ),
    file.path(output_dir, "anosim_age_pUM_full.csv")
  )
} else {
  cat("Insufficient pUM samples with age data\n")
}
```

### Melanocytic Cells (Healthy + Malignant)

**Include healthy melanocytes as negative control for comparison with malignant melanoma cells**

```{r gloscope_melanocytic_origin}
cat("Running GloScope analysis for melanocytic cells (healthy + malignant) by disease origin...\n")

# Filter to melanocytic cells: Melanocytes (healthy) + Melanoma cells (malignant)
cells_melanocytic <- cell_metadata %>%
  dplyr::filter(majority_celltype %in% c("Melanocytes", "Melanoma cells")) %>%
  pull(cell_id)

embedding_melanocytic <- embedding_matrix[cells_melanocytic, ]
sample_labels_melanocytic <- cell_metadata %>%
  dplyr::filter(majority_celltype %in% c("Melanocytes", "Melanoma cells")) %>%
  pull(Sample)

# Check sample representation (need at least 100 cells per sample)
sample_cell_counts_melanocytic <- table(sample_labels_melanocytic)
valid_samples_melanocytic <- names(sample_cell_counts_melanocytic)[sample_cell_counts_melanocytic >= 100]

cat("  - Total melanocytic cells:", length(cells_melanocytic), "\n")
cat("  - Samples with >=100 melanocytic cells:", length(valid_samples_melanocytic), "\n")

if (length(valid_samples_melanocytic) >= 6) {
  # Filter to valid samples
  cells_melanocytic_filtered <- cell_metadata %>%
    dplyr::filter(majority_celltype %in% c("Melanocytes", "Melanoma cells"),
                  Sample %in% valid_samples_melanocytic) %>%
    pull(cell_id)

  embedding_melanocytic_filtered <- embedding_matrix[cells_melanocytic_filtered, ]
  sample_labels_melanocytic_filtered <- cell_metadata %>%
    dplyr::filter(majority_celltype %in% c("Melanocytes", "Melanoma cells"),
                  Sample %in% valid_samples_melanocytic) %>%
    pull(Sample)

  # Calculate or load divergence
  knn_div_melanocytic <- run_gloscope_cached(
    embedding_subset = embedding_melanocytic_filtered,
    sample_labels = sample_labels_melanocytic_filtered,
    cache_file = "divergence_origin_melanocytic_full.rds",
    analysis_name = "Disease Origin (Melanocytic Cells: Healthy + Malignant)"
  )

  # MDS visualization
  samples_in_div_melanocytic <- rownames(knn_div_melanocytic)
  sample_metadata_melanocytic <- data.frame(Sample = samples_in_div_melanocytic) %>%
    left_join(sample_metadata %>% dplyr::filter(Sample %in% valid_samples_melanocytic), by = "Sample")

  mds_melanocytic <- plotMDS(
    dist_mat = knn_div_melanocytic,
    metadata_df = sample_metadata_melanocytic,
    sample_id = "Sample",
    group_id = "origin",
    k = 2
  )

  # ANOSIM test first to get p-value for plot
  anosim_melanocytic <- safe_anosim(
    as.dist(knn_div_melanocytic),
    grouping = sample_metadata_melanocytic$origin,
    permutations = 999
  )

  cat("\nANOSIM test for disease origin (melanocytic cells with healthy control):\n")
  cat("  R statistic:", round(anosim_melanocytic$statistic, 4), "\n")
  cat("  p-value:", anosim_melanocytic$signif, "\n")

  p_display_melanocytic <- ifelse(anosim_melanocytic$signif < 0.001, "p < 0.001",
                                  paste0("p = ", format(anosim_melanocytic$signif, digits = 3)))

  p_melanocytic_mds <- mds_melanocytic$plot +
    scale_color_manual(values = origin_colors) +
    coord_fixed() +
    labs(
      title = "GloScope MDS: Melanocytic Cells (Healthy + Malignant - Full Cohort)",
      subtitle = paste0("n = ", nrow(sample_metadata_melanocytic), " samples | ANOSIM: R = ",
                       round(anosim_melanocytic$statistic, 3), ", ", p_display_melanocytic),
      color = "Origin"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right", plot.title = element_text(face = "bold", size = 14))

  print(p_melanocytic_mds)

  ggsave(
    file.path(output_dir, "MDS_origin_melanocytic_full.png"),
    p_melanocytic_mds,
    width = 10, height = 8, dpi = 300, bg = "white"
  )

  write_csv(
    data.frame(
      comparison = "disease_origin_melanocytic_full",
      statistic = anosim_melanocytic$statistic,
      p_value = anosim_melanocytic$signif,
      permutations = 999
    ),
    file.path(output_dir, "anosim_origin_melanocytic_full.csv")
  )

  # Divergence Heatmap
  sample_order_melanocytic <- sample_metadata_melanocytic %>% arrange(origin) %>% pull(Sample)
  knn_div_ordered_melanocytic <- knn_div_melanocytic[sample_order_melanocytic, sample_order_melanocytic]

  anno_data_melanocytic <- sample_metadata_melanocytic[match(sample_order_melanocytic, sample_metadata_melanocytic$Sample), ]

  # Replace NA values with "Unknown" for annotations
  anno_data_melanocytic_clean <- anno_data_melanocytic %>%
    dplyr::mutate(
      Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
      Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
    )

  # Add colors for Unknown categories
  gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
  age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

  col_anno_melanocytic <- HeatmapAnnotation(
    Patient = anno_data_melanocytic_clean$Patient_nr,
    Origin = anno_data_melanocytic_clean$origin,
    Gender = anno_data_melanocytic_clean$Gender,
    Age_group = anno_data_melanocytic_clean$Age_group,
    `Primary Mutation` = anno_data_melanocytic_clean$primary_mutation,
    `Secondary Mutation` = anno_data_melanocytic_clean$secondary_mutation,
    col = list(
      Origin = origin_colors,
      Gender = gender_colors_full,
      Age_group = age_colors_full,
      `Primary Mutation` = pmut_colors,
      `Secondary Mutation` = smut_colors
    ),
    annotation_name_gp = gpar(fontsize = 9),
    show_legend = c(Patient = FALSE, Origin = TRUE, Gender = TRUE, Age_group = TRUE,
                    `Primary Mutation` = TRUE, `Secondary Mutation` = TRUE)
  )

  ht_melanocytic <- Heatmap(
    knn_div_ordered_melanocytic,
    name = "KL Divergence",
    col = colorRamp2(c(0, median(knn_div_melanocytic), max(knn_div_melanocytic)),
                     c("white", "orange", "red")),
    top_annotation = col_anno_melanocytic,
    show_row_names = FALSE,
    show_column_names = FALSE,
    column_title = "GloScope: Melanocytic Cells (Healthy + Malignant) by Origin",
    column_title_gp = gpar(fontsize = 14, fontface = "bold"),
    heatmap_legend_param = list(
      title = "KL Divergence",
      legend_height = unit(4, "cm")
    )
  )

  png(file.path(output_dir, "heatmap_origin_melanocytic_full.png"),
      width = 2800, height = 2600, res = 300)
  draw(ht_melanocytic)
  dev.off()

  draw(ht_melanocytic)

  # Save heatmap object and data for recreation
  qs2::qs_save(list(ht = ht_melanocytic, div_matrix = knn_div_ordered_melanocytic, anno_data = anno_data_melanocytic_clean),
               file.path(output_dir, "gloscope_heatmap_melanocytic_cells_objects.qs2"), nthreads = parallel::detectCores())

  # Summary statistics
  cat("\n=== COMPARISON SUMMARY ===\n")
  cat("Healthy melanocytes provide baseline for comparison with malignant cells\n")
  cat("  - Healthy samples:", sum(sample_metadata_melanocytic$origin == "healthy", na.rm = TRUE), "\n")
  cat("  - Primary UM samples:", sum(sample_metadata_melanocytic$origin == "primary", na.rm = TRUE), "\n")
  cat("  - Metastatic UM samples:", sum(sample_metadata_melanocytic$origin == "metastasis", na.rm = TRUE), "\n")
} else {
  cat("Insufficient samples with adequate melanocytic cell representation\n")
}
```

### Mutation Stratification (Malignant)

**Secondary mutations in malignant cells from primary tumors**

```{r gloscope_malignant_secondary}
cat("Running GloScope analysis for malignant cells by secondary mutations...\n")

samples_malignant_pum <- sample_metadata %>%
  dplyr::filter(
    origin == "primary",
    secondary_mutation %in% c("BAP1", "SF3B1", "EIF1AX")
  ) %>%
  pull(Sample)

if (length(samples_malignant_pum) >= 6) {
  cells_malignant_pum <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% samples_malignant_pum) %>%
    pull(cell_id)

  # Check sample representation
  sample_labels_malignant_pum_check <- cell_metadata %>%
    dplyr::filter(Malignant == "Malignant", Sample %in% samples_malignant_pum) %>%
    pull(Sample)

  sample_counts_pum <- table(sample_labels_malignant_pum_check)
  valid_samples_pum <- names(sample_counts_pum)[sample_counts_pum >= 100]

  cat("  - Samples with >=100 malignant cells:", length(valid_samples_pum), "\n")

  if (length(valid_samples_pum) >= 6) {
    cells_malignant_pum_filtered <- cell_metadata %>%
      dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_pum) %>%
      pull(cell_id)

    embedding_malignant_pum <- embedding_matrix[cells_malignant_pum_filtered, ]
    sample_labels_malignant_pum <- cell_metadata %>%
      dplyr::filter(Malignant == "Malignant", Sample %in% valid_samples_pum) %>%
      pull(Sample)

    # Calculate or load divergence
    knn_div_malignant_pum <- run_gloscope_cached(
      embedding_subset = embedding_malignant_pum,
      sample_labels = sample_labels_malignant_pum,
      cache_file = "divergence_secondary_mutation_malignant_full.rds",
      analysis_name = "Secondary Mutations (Malignant Cells)"
    )

    # MDS visualization
    samples_in_div_malignant_pum <- rownames(knn_div_malignant_pum)
    sample_metadata_malignant_pum <- data.frame(Sample = samples_in_div_malignant_pum) %>%
      left_join(sample_metadata %>% dplyr::filter(Sample %in% valid_samples_pum), by = "Sample")

    mds_malignant_pum <- plotMDS(
      dist_mat = knn_div_malignant_pum,
      metadata_df = sample_metadata_malignant_pum,
      sample_id = "Sample",
      group_id = "secondary_mutation",
      k = 2
    )

    # ANOSIM test first to get p-value for plot
    anosim_malignant_pum <- safe_anosim(
      as.dist(knn_div_malignant_pum),
      grouping = sample_metadata_malignant_pum$secondary_mutation,
      permutations = 999
    )

    cat("\nANOSIM test for secondary mutations (malignant cells):\n")
    cat("  R statistic:", round(anosim_malignant_pum$statistic, 4), "\n")
    cat("  p-value:", anosim_malignant_pum$signif, "\n")

    p_display_mal_pum <- ifelse(anosim_malignant_pum$signif < 0.001, "p < 0.001",
                                paste0("p = ", format(anosim_malignant_pum$signif, digits = 3)))

    p_malignant_pum_mds <- mds_malignant_pum$plot +
      scale_color_manual(values = smut_colors) +
      coord_fixed() +
      labs(
        title = "GloScope MDS: Secondary Mutations (Malignant Cells, Primary UM - Full Cohort)",
        subtitle = paste0("n = ", nrow(sample_metadata_malignant_pum), " samples | ANOSIM: R = ",
                         round(anosim_malignant_pum$statistic, 3), ", ", p_display_mal_pum),
        color = "Secondary Mutation"
      ) +
      theme_minimal(base_size = 12)

    print(p_malignant_pum_mds)

    ggsave(
      file.path(output_dir, "MDS_secondary_mutation_malignant_full.png"),
      p_malignant_pum_mds,
      width = 10, height = 8, dpi = 300, bg = "white"
    )

    write_csv(
      data.frame(
        comparison = "secondary_mutation_malignant_full",
        statistic = anosim_malignant_pum$statistic,
        p_value = anosim_malignant_pum$signif,
        permutations = 999
      ),
      file.path(output_dir, "anosim_secondary_mutation_malignant_full.csv")
    )

    # Divergence Heatmap
    sample_order_malignant_pum <- sample_metadata_malignant_pum %>% arrange(secondary_mutation) %>% pull(Sample)
    knn_div_ordered_malignant_pum <- knn_div_malignant_pum[sample_order_malignant_pum, sample_order_malignant_pum]

    anno_data_malignant_pum <- sample_metadata_malignant_pum[match(sample_order_malignant_pum, sample_metadata_malignant_pum$Sample), ]

    # Replace NA values with "Unknown" for annotations
    anno_data_malignant_pum_clean <- anno_data_malignant_pum %>%
      dplyr::mutate(
        Gender = replace(as.character(Gender), is.na(Gender) | Gender == "NA", "Unknown"),
        Age_group = replace(as.character(Age_group), is.na(Age_group) | Age_group == "NA", "Unknown")
      )

    # Add colors for Unknown categories
    gender_colors_full <- c(gender_colors, "Unknown" = "#E1E1E1FF")
    age_colors_full <- c(age_colors, "Unknown" = "#E1E1E1FF")

    col_anno_malignant_pum <- HeatmapAnnotation(
      Patient = anno_data_malignant_pum_clean$Patient_nr,
      Gender = anno_data_malignant_pum_clean$Gender,
      Age_group = anno_data_malignant_pum_clean$Age_group,
      `Primary Mutation` = anno_data_malignant_pum_clean$primary_mutation,
      `Secondary Mutation` = anno_data_malignant_pum_clean$secondary_mutation,
      col = list(
        Gender = gender_colors_full,
        Age_group = age_colors_full,
        `Primary Mutation` = pmut_colors,
        `Secondary Mutation` = smut_colors
      ),
      annotation_name_gp = gpar(fontsize = 9),
      show_legend = c(Patient = FALSE, Gender = TRUE, Age_group = TRUE,
                      `Primary Mutation` = TRUE, `Secondary Mutation` = TRUE)
    )

    ht_malignant_pum <- Heatmap(
      knn_div_ordered_malignant_pum,
      name = "KL Divergence",
      col = colorRamp2(c(0, median(knn_div_malignant_pum), max(knn_div_malignant_pum)),
                       c("white", "orange", "red")),
      top_annotation = col_anno_malignant_pum,
      show_row_names = FALSE,
      show_column_names = FALSE,
      column_title = "GloScope Divergence: Secondary Mutations (Malignant Cells - Full Cohort)",
      column_title_gp = gpar(fontsize = 14, fontface = "bold"),
      heatmap_legend_param = list(
        title = "KL Divergence",
        legend_height = unit(4, "cm")
      )
    )

    png(file.path(output_dir, "heatmap_secondary_mutation_malignant_full.png"),
        width = 2400, height = 2200, res = 300)
    draw(ht_malignant_pum)
    dev.off()

    draw(ht_malignant_pum)
  } else {
    cat("Insufficient malignant cell representation per sample\n")
  }
} else {
  cat("Insufficient samples with secondary mutations\n")
}
```

---

## Summary Statistics {.tabset}

### Overview

**GloScope ANOSIM Results Summary**

This section summarizes all ANOSIM (Analysis of Similarity) tests performed across different cell subsets and clinical/biological variables.

**Spatial/Location Comparisons** (grouped together in heatmap):

- **Disease Origin**: Compares samples by disease stage - primary uveal melanomas (pUM, from the eye) vs metastatic tumors (from liver, brain, skin, thyroid). Tests whether transcriptional divergence differs between primary and metastatic disease.
- **Organ**: Compares metastatic samples only by their organ of origin (liver, brain, skin, thyroid). Tests whether metastases to different organs show distinct transcriptional profiles.
- **Anatomical Location**: Compares primary tumors only by their intraocular location (choroid, iris, ciliary body). Tests whether the anatomical site within the eye influences transcriptional heterogeneity.

**Clinical/Biological Comparisons:**

- **Gender**: Male vs Female comparison
- **Age**: Age group comparison (Young <40, Middle 40-60, Old 60+)
- **Primary Mutation**: GNAQ vs GNA11 vs PLCB4
- **Secondary Mutation**: Presence/type of secondary mutations (SF3B1, BAP1, EIF1AX)

**Interpretation:**

- **R statistic**: Effect size (0-1), higher values indicate greater dissimilarity between groups
- **Significance**: *** p<0.001, ** p<0.01, * p<0.05, ns = not significant

### Summary Table

```{r summary_stats}
# Collect all ANOSIM results
anosim_files <- list.files(output_dir, pattern = "^anosim.*full\\.csv$", full.names = TRUE)

cat("Found", length(anosim_files), "ANOSIM result files\n")

if (length(anosim_files) > 0) {
  anosim_summary <- anosim_files %>%
    lapply(read_csv, show_col_types = FALSE) %>%
    bind_rows() %>%
    mutate(
      R_statistic = round(statistic, 4),
      p_value_formatted = format.pval(p_value, digits = 3),
      Significance = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ "ns"
      ),
      # Categorize by analysis type (cell subset)
      Analysis_Group = case_when(
        grepl("malignant", comparison, ignore.case = TRUE) ~ "Malignant Cells",
        grepl("melanocytic", comparison, ignore.case = TRUE) ~ "Melanocytic Cells",
        grepl("pUM", comparison) ~ "Primary Tumors",
        TRUE ~ "All Cells"
      ),
      # Extract base comparison type and map to clean names
      Comparison_Base = comparison %>%
        gsub("_full$", "", .) %>%
        gsub("_malignant$", "", .) %>%
        gsub("_melanocytic$", "", .) %>%
        gsub("_pUM$", "", .),
      Comparison_Clean = case_when(
        grepl("^origin", Comparison_Base) ~ "Disease Origin",
        grepl("^organ", Comparison_Base) ~ "Organ",
        grepl("^location", Comparison_Base) ~ "Anatomical Location",
        grepl("^gender", Comparison_Base) ~ "Gender",
        grepl("^age", Comparison_Base) ~ "Age",
        grepl("^primary_mutation", Comparison_Base) ~ "Primary Mutation",
        grepl("^secondary_mutation", Comparison_Base) ~ "Secondary Mutation",
        TRUE ~ tools::toTitleCase(gsub("_", " ", Comparison_Base))
      )
    )

  # Display table
  print(knitr::kable(anosim_summary %>%
                       select(Analysis_Group, Comparison_Clean, R_statistic, p_value_formatted, Significance) %>%
                       arrange(Analysis_Group, desc(R_statistic)),
                     caption = "ANOSIM Statistical Tests - Full Cohort",
                     col.names = c("Analysis Group", "Comparison", "R Statistic", "P-value", "Significance")))

  # Also create a DT table for interactivity
  cat("\n\n#### Interactive Summary Table\n\n")
  DT::datatable(anosim_summary %>%
                  select(Analysis_Group, Comparison_Clean, R_statistic, p_value_formatted, Significance),
                options = list(pageLength = 15, dom = 'ft'),
                caption = "ANOSIM Statistical Tests - Full Cohort")

  write_csv(anosim_summary, file.path(output_dir, "anosim_summary_full_cohort.csv"))

  cat("\n\n**📊 Dataframe Export for Collaborators:**\n")
  cat("The `anosim_summary` dataframe has been exported to:\n")
  cat("`", file.path(output_dir, "anosim_summary_full_cohort.csv"), "`\n\n")
  cat("This file can be used for custom heatmap modifications without re-running the analysis.\n")
} else {
  cat("No ANOSIM results available yet.\n")
}
```

### Summary Heatmap

```{r summary_heatmap, fig.width=10, fig.height=8}
if (length(anosim_files) > 0 && exists("anosim_summary")) {
  library(ggplot2)
  library(dplyr)

  # Define row ordering: spatial comparisons first, then clinical
  row_order <- c("Disease Origin", "Organ", "Anatomical Location",
                 "Gender", "Age", "Primary Mutation", "Secondary Mutation")

  # Define column ordering
  col_order <- c("All Cells", "Malignant Cells", "Melanocytic Cells", "Primary Tumors")

  # Create heatmap data
  heatmap_data <- anosim_summary %>%
    mutate(
      # Create nice labels
      Label = paste0("R=", round(R_statistic, 2), "\n", Significance),
      # -log10(p-value) for color scale, cap at 4 (p < 0.0001)
      neg_log10_p = pmin(-log10(p_value), 4),
      # Apply fixed row ordering (spatial first)
      Comparison_Clean = factor(Comparison_Clean,
                                 levels = rev(row_order[row_order %in% unique(Comparison_Clean)])),
      Analysis_Group = factor(Analysis_Group, levels = col_order)
    )

  # Create the heatmap
  p_heatmap <- ggplot(heatmap_data, aes(x = Analysis_Group, y = Comparison_Clean)) +
    geom_tile(aes(fill = R_statistic), color = "white", linewidth = 0.5) +
    geom_text(aes(label = Label), size = 3.5, color = "white", fontface = "bold") +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "#F7F7F7",
      high = "#B2182B",
      midpoint = 0.25,
      limits = c(0, 0.6),
      name = "R Statistic",
      oob = scales::squish
    ) +
    labs(
      title = "GloScope ANOSIM Results Summary",
      subtitle = "R statistic (effect size) with significance: *** p<0.001, ** p<0.01, * p<0.05, ns = not significant",
      x = "Analysis Subset",
      y = "Comparison"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 10),
      legend.position = "right"
    )

  print(p_heatmap)

  ggsave(file.path(output_dir, "anosim_summary_heatmap.png"),
         p_heatmap, width = 10, height = 8, dpi = 300, bg = "white")

  # Also create a significance-focused heatmap
  p_sig <- ggplot(heatmap_data, aes(x = Analysis_Group, y = Comparison_Clean)) +
    geom_tile(aes(fill = neg_log10_p), color = "white", linewidth = 0.5) +
    geom_text(aes(label = paste0("R=", round(R_statistic, 2))),
              size = 3.5, color = ifelse(heatmap_data$neg_log10_p > 2, "white", "black"), fontface = "bold") +
    scale_fill_viridis_c(
      option = "magma",
      name = "-log10(p-value)",
      limits = c(0, 4),
      breaks = c(0, 1, 2, 3, 4),
      labels = c("1", "0.1", "0.01", "0.001", "<0.0001")
    ) +
    labs(
      title = "GloScope ANOSIM - Statistical Significance",
      subtitle = "Darker colors indicate stronger significance",
      x = "Analysis Subset",
      y = "Comparison"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(face = "bold"),
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.subtitle = element_text(hjust = 0.5, size = 10),
      legend.position = "right"
    )

  print(p_sig)

  ggsave(file.path(output_dir, "anosim_significance_heatmap.png"),
         p_sig, width = 10, height = 8, dpi = 300, bg = "white")
} else {
  cat("No ANOSIM results available for heatmap.\n")
}
```

---

## Session Information

```{r session_info}
sessionInfo()
```

**Analysis completed:** `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

**Full cohort (NO downsampling):** ~298,640 cells

**Parallel processing:** 24 cores via BiocParallel

**Output location:** `../GloScope_Full_Outputs/`

**Cached divergence matrices:** Reusable for future analyses
