---
title: "p31662 - ProjecTILs T Cell Functional Annotation"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Extended Data (T cell subtyping with ProjecTILs CD4/CD8 references)
```

# {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

### Analysis Goals

This analysis applies **ProjecTILs** human TIL (tumor-infiltrating lymphocyte) reference mapping to functionally annotate T cells in the Uveal Melanoma TME dataset. Key objectives:

1. **Classify T cell functional states** using Carmona lab's human CD8 and CD4 TIL reference atlases
2. **Compare with existing SubTyping labels** to assess concordance and identify refined annotations
3. **Analyze compositional changes** across disease stages (Healthy vs Primary vs Metastatic) and genetic subtypes (BAP1/SF3B1/EIF1AX mutations)
4. **Validate annotations** with canonical T cell marker expression

### Dataset Information

- **Source:** Levesque UM single-cell atlas TME subset
- **Total TME cells:** 102,897 cells across 11 cell types
- **T/NK cells:** ~39,321 cells from 52 samples
- **Existing annotations:** `SubTyping` column with T cell subtype labels
- **Species:** Human

### Methods

**ProjecTILs workflow:**

1. **scGate CD4/CD8 separation:** Use scGate gene signature-based gating to classify CD4 vs CD8 T cells and remove NK cells/doublets
2. **Separate reference mapping:** Apply CD8 and CD4 human TIL reference maps independently using ProjecTILs.classifier
3. **Functional state annotation:** Each reference provides functional states (e.g., Exhausted, Effector Memory, Regulatory, etc.)
4. **Validation:** Radar plots and marker expression confirm biological accuracy of annotations

**References:**

- [ProjecTILs GitHub](https://github.com/carmonalab/ProjecTILs)
- [Bassez Breast Cancer Case Study](https://carmonalab.github.io/ProjecTILs_CaseStudies/Bassez_BC.html)
- CD8 TIL atlas: `CD8T_human_ref_v1.rds` (246MB)
- CD4 TIL atlas: `CD4T_human_ref_v1.rds` (250MB)

## Setup {.tabset}

### Libraries

```{r packages, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, knitr.table.format = "html")

library(Seurat)
library(qs2)
library(ProjecTILs)
library(scGate)
library(tidyverse)
library(patchwork)
library(ComplexHeatmap)
library(DT)
library(ezRun)
library(viridis)
library(pals)
library(future)
library(BiocParallel)
```

### Parameters

```{r parameters}
# Parallelization setup for BiocParallel
BPPARAM <- MulticoreParam(workers = 16)
register(BPPARAM)

# Future setup
plan("sequential")
options(future.globals.maxSize = 50 * 1024^3)

# Paths
input_path <- "../Levesque_UMscAtlas/01_TME_subset_integrated.qs"
output_dir <- "Outputs/ProjecTILs"
ref_cd8_path <- "../CD8T_human_ref_v1.rds"
ref_cd4_path <- "../CD4T_human_ref_v1.rds"

# Create output directory
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Seed for reproducibility
set.seed(42)

cat("Output directory:", output_dir, "\n")
cat("Input TME object:", input_path, "\n")
```

### Load TME Data

```{r load_data}
# Load TME subset
cat("Loading TME subset object...\n")
tme <- qs::qread(input_path, nthreads = 16)

cat("\n=== TME Object Summary ===\n")
print(tme)

cat("\n=== Metadata columns available ===\n")
print(colnames(tme@meta.data)[1:50])

cat("\n=== Cell type distribution ===\n")
print(table(tme$majority_celltype))

cat("\n=== Number of samples ===\n")
print(length(unique(tme$Sample)))

cat("\n=== Disease origin distribution ===\n")
print(table(tme$origin))
```

## Data Preparation {.tabset}

### Filter to T/NK Cells

```{r filter_tcells}
# Filter to T/NK cells
cat("Filtering to T/NK cells...\n")
tcells <- subset(tme, majority_celltype == "T/NK cells")

cat("\n=== T/NK Cell Subset ===\n")
cat("Number of T/NK cells:", ncol(tcells), "\n")
cat("Number of features:", nrow(tcells), "\n")
cat("Number of samples:", length(unique(tcells$Sample)), "\n")

# Clean up
rm(tme)
gc()
```

### Check Existing SubTyping

```{r check_subtyping}
# Existing T cell annotations
cat("\n=== Existing SubTyping Distribution ===\n")
subtyping_table <- table(tcells$SubTyping)
print(subtyping_table)

# Create summary table
subtyping_df <- data.frame(
  SubType = names(subtyping_table),
  Count = as.numeric(subtyping_table),
  Percentage = round(as.numeric(subtyping_table) / ncol(tcells) * 100, 2)
) |>
  arrange(desc(Count))

datatable(subtyping_df,
          caption = "Existing T Cell SubTyping Labels",
          options = list(pageLength = 20))
```

### Recompute T Cell UMAP

Recompute UMAP specifically for T cells to improve resolution of T cell populations.

```{r recompute_umap, fig.width=10, fig.height=8}
# T cell-specific dimensionality reduction
cat("Recomputing T cell-specific UMAP...\n")
DefaultAssay(tcells) <- "RNA"

tcells <- FindVariableFeatures(tcells, nfeatures = 2000, verbose = FALSE)
tcells <- ScaleData(tcells, verbose = FALSE)
tcells <- RunPCA(tcells, npcs = 50, verbose = FALSE)
tcells <- RunUMAP(tcells, dims = 1:30, verbose = FALSE, seed.use = 42)

# Plot by existing SubTyping
p_umap_orig <- DimPlot(tcells,
                       group.by = "SubTyping",
                       label = TRUE,
                       repel = TRUE,
                       raster = FALSE) +
  coord_fixed() +
  ggtitle("T/NK Cells - Existing SubTyping Labels") +
  theme(legend.position = "right")

print(p_umap_orig)

ggsave(file.path(output_dir, "umap_original_subtyping.png"),
       p_umap_orig,
       width = 12, height = 8, dpi = 300, bg = "white")
```

### Cell Count Summary

```{r cell_summary}
# Summary by sample and origin
summary_df <- tcells@meta.data |>
  group_by(Sample, origin) |>
  summarise(n_cells = n(), .groups = "drop") |>
  arrange(desc(n_cells))

datatable(summary_df,
          caption = "T Cell Counts by Sample and Disease Origin",
          options = list(pageLength = 15))

# Summary by origin
origin_summary <- tcells@meta.data |>
  group_by(origin) |>
  summarise(
    n_cells = n(),
    n_samples = n_distinct(Sample),
    pct = round(n() / ncol(tcells) * 100, 2)
  )

print(origin_summary)
```

## CD4/CD8 Identification {.tabset}

### Identify from SubTyping

Use existing SubTyping labels to identify CD4 vs CD8 T cells.

```{r identify_cd4_cd8, fig.width=12, fig.height=10}
cat("Identifying CD4 vs CD8 from SubTyping labels...\n")

# Identify CD4, CD8, MAIT, gdT, and NK cells from SubTyping
tcells$CD_type <- case_when(
  grepl("^CD4", tcells$SubTyping) ~ "CD4",
  grepl("^CD8", tcells$SubTyping) ~ "CD8",
  grepl("MAIT", tcells$SubTyping) ~ "MAIT",
  grepl("gdT|Vg9Vd2", tcells$SubTyping) ~ "gdT",
  grepl("NK", tcells$SubTyping) ~ "NK",
  TRUE ~ "Other"
)

cat("\n=== CD4/CD8 Distribution ===\n")
print(table(tcells$CD_type))

# Plot distribution
p_cd_type <- DimPlot(tcells,
                     group.by = "CD_type",
                     label = TRUE,
                     repel = TRUE,
                     raster = FALSE) +
  coord_fixed() +
  ggtitle("T Cell Types from SubTyping")

print(p_cd_type)

ggsave(file.path(output_dir, "cd4_cd8_distribution.png"),
       p_cd_type,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Cell Type Summary

Keep all T cell types (CD4, CD8, MAIT, gdT). ProjecTILs will be applied to CD4 and CD8 only.

```{r celltype_summary}
cat("T cell type distribution:\n")
print(table(tcells$CD_type))

cat("\nTotal cells:", ncol(tcells), "\n")
cat("CD8 T cells:", sum(tcells$CD_type == "CD8"), "\n")
cat("CD4 T cells:", sum(tcells$CD_type == "CD4"), "\n")
cat("MAIT cells:", sum(tcells$CD_type == "MAIT"), "\n")
cat("gdT cells:", sum(tcells$CD_type == "gdT"), "\n")
cat("NK cells:", sum(tcells$CD_type == "NK"), "\n")
cat("Other:", sum(tcells$CD_type == "Other"), "\n")
```

## ProjecTILs CD8 Annotation {.tabset}

### Load CD8 Reference

```{r load_cd8_ref, fig.width=10, fig.height=8}
cat("Loading CD8 TIL reference...\n")
ref.cd8 <- load.reference.map(ref_cd8_path)

cat("\n=== CD8 Reference Info ===\n")
print(ref.cd8)

cat("\nAvailable functional states:\n")
print(table(ref.cd8$functional.cluster))

# Plot reference UMAP
p_ref_cd8 <- DimPlot(ref.cd8,
                     cols = ref.cd8@misc$atlas.palette,
                     label = TRUE,
                     repel = TRUE,
                     raster = FALSE) +
  coord_fixed() +
  ggtitle("CD8 TIL Reference Atlas") +
  NoLegend()

print(p_ref_cd8)

ggsave(file.path(output_dir, "reference_cd8_umap.png"),
       p_ref_cd8,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Apply CD8 ProjecTILs

```{r projectils_cd8, fig.width=12, fig.height=10}
cat("Applying ProjecTILs to CD8 T cells...\n")

# Subset to CD8 cells
tcells.cd8 <- subset(tcells, CD_type == "CD8")

cat("Number of CD8 cells:", ncol(tcells.cd8), "\n")

# Join layers for Seurat v5 compatibility
DefaultAssay(tcells.cd8) <- "RNA"
tcells.cd8 <- JoinLayers(tcells.cd8)

# Convert to Seurat v4 Assay format for ProjecTILs compatibility
tcells.cd8[["RNA"]] <- as(tcells.cd8[["RNA"]], "Assay")

# Apply ProjecTILs classifier
tcells.cd8 <- ProjecTILs.classifier(
  query = tcells.cd8,
  ref = ref.cd8,
  split.by = "Sample",
  ncores = 16
)

cat("\n=== CD8 Functional States ===\n")
print(table(tcells.cd8$functional.cluster))

# Plot CD8 functional states
p_cd8_states <- DimPlot(tcells.cd8,
                        group.by = "functional.cluster",
                        cols = ref.cd8@misc$atlas.palette,
                        label = TRUE,
                        repel = TRUE,
                        order = TRUE,
                        raster = FALSE) +
  coord_fixed() +
  ggtitle("CD8 T Cells - ProjecTILs Functional States")

print(p_cd8_states)

ggsave(file.path(output_dir, "umap_cd8_functional_states.png"),
       p_cd8_states,
       width = 12, height = 10, dpi = 300, bg = "white")
```

### CD8 Radar Plots

```{r cd8_radar, fig.width=16, fig.height=12}
# CD8 marker genes for validation
cd8_markers <- c("CD8A", "CD8B", "CD3D", "CD3E",
                 "CCR7", "TCF7", "LEF1", "IL7R",  # Naive/memory
                 "GZMA", "GZMK", "GZMB", "PRF1",  # Effector/cytotoxic
                 "PDCD1", "HAVCR2", "LAG3", "TOX", # Exhaustion
                 "CXCL13", "ENTPD1")               # Tpex

# Filter to genes present in data
cd8_markers <- cd8_markers[cd8_markers %in% rownames(tcells.cd8)]

cat("CD8 markers for radar plot:", paste(cd8_markers, collapse = ", "), "\n")

# Generate radar plots
p_radar_cd8 <- plot.states.radar(
  ref = ref.cd8,
  query = tcells.cd8,
  genes4radar = cd8_markers,
  min.cells = 20
)

print(p_radar_cd8)

ggsave(file.path(output_dir, "radar_cd8_markers.png"),
       p_radar_cd8,
       width = 16, height = 12, dpi = 300, bg = "white")
```

### CD8 Proportions

```{r cd8_proportions, fig.width=14, fig.height=8}
# Proportions of CD8 functional states
cd8_prop <- tcells.cd8@meta.data |>
  group_by(functional.cluster) |>
  summarise(n = n()) |>
  mutate(pct = round(n / sum(n) * 100, 2)) |>
  arrange(desc(n))

datatable(cd8_prop,
          caption = "CD8 Functional State Proportions")

# Barplot
p_cd8_bar <- ggplot(cd8_prop, aes(x = reorder(functional.cluster, -n), y = n, fill = functional.cluster)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = ref.cd8@misc$atlas.palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Functional State", y = "Cell Count",
       title = "CD8 T Cell Functional States")

print(p_cd8_bar)

ggsave(file.path(output_dir, "barplot_cd8_states.png"),
       p_cd8_bar,
       width = 12, height = 6, dpi = 300, bg = "white")
```

## ProjecTILs CD4 Annotation {.tabset}

### Load CD4 Reference

```{r load_cd4_ref, fig.width=10, fig.height=8}
cat("Loading CD4 TIL reference...\n")
ref.cd4 <- load.reference.map(ref_cd4_path)

cat("\n=== CD4 Reference Info ===\n")
print(ref.cd4)

cat("\nAvailable functional states:\n")
print(table(ref.cd4$functional.cluster))

# Plot reference UMAP
p_ref_cd4 <- DimPlot(ref.cd4,
                     cols = ref.cd4@misc$atlas.palette,
                     label = TRUE,
                     repel = TRUE,
                     raster = FALSE) +
  coord_fixed() +
  ggtitle("CD4 TIL Reference Atlas") +
  NoLegend()

print(p_ref_cd4)

ggsave(file.path(output_dir, "reference_cd4_umap.png"),
       p_ref_cd4,
       width = 10, height = 8, dpi = 300, bg = "white")
```

### Apply CD4 ProjecTILs

```{r projectils_cd4, fig.width=12, fig.height=10}
cat("Applying ProjecTILs to CD4 T cells...\n")

# Subset to CD4 cells
tcells.cd4 <- subset(tcells, CD_type == "CD4")

cat("Number of CD4 cells:", ncol(tcells.cd4), "\n")

# Join layers for Seurat v5 compatibility
DefaultAssay(tcells.cd4) <- "RNA"
tcells.cd4 <- JoinLayers(tcells.cd4)

# Convert to Seurat v4 Assay format for ProjecTILs compatibility
tcells.cd4[["RNA"]] <- as(tcells.cd4[["RNA"]], "Assay")

# Apply ProjecTILs classifier
tcells.cd4 <- ProjecTILs.classifier(
  query = tcells.cd4,
  ref = ref.cd4,
  split.by = "Sample",
  ncores = 16
)

cat("\n=== CD4 Functional States ===\n")
print(table(tcells.cd4$functional.cluster))

# Plot CD4 functional states
p_cd4_states <- DimPlot(tcells.cd4,
                        group.by = "functional.cluster",
                        cols = ref.cd4@misc$atlas.palette,
                        label = TRUE,
                        repel = TRUE,
                        order = TRUE,
                        raster = FALSE) +
  coord_fixed() +
  ggtitle("CD4 T Cells - ProjecTILs Functional States")

print(p_cd4_states)

ggsave(file.path(output_dir, "umap_cd4_functional_states.png"),
       p_cd4_states,
       width = 12, height = 10, dpi = 300, bg = "white")
```

### CD4 Radar Plots

```{r cd4_radar, fig.width=16, fig.height=12}
# CD4 marker genes for validation
cd4_markers <- c("CD4", "CD3D", "CD3E",
                 "CCR7", "TCF7", "LEF1", "IL7R",  # Naive/CM
                 "FOXP3", "IL2RA", "CTLA4",       # Treg
                 "CXCR5", "ICOS", "BCL6",         # Tfh
                 "IFNG", "TNF", "IL2")            # Th1

# Filter to genes present in data
cd4_markers <- cd4_markers[cd4_markers %in% rownames(tcells.cd4)]

cat("CD4 markers for radar plot:", paste(cd4_markers, collapse = ", "), "\n")

# Generate radar plots
p_radar_cd4 <- plot.states.radar(
  ref = ref.cd4,
  query = tcells.cd4,
  genes4radar = cd4_markers,
  min.cells = 20
)

print(p_radar_cd4)

ggsave(file.path(output_dir, "radar_cd4_markers.png"),
       p_radar_cd4,
       width = 16, height = 12, dpi = 300, bg = "white")
```

### CD4 Proportions

```{r cd4_proportions, fig.width=14, fig.height=8}
# Proportions of CD4 functional states
cd4_prop <- tcells.cd4@meta.data |>
  group_by(functional.cluster) |>
  summarise(n = n()) |>
  mutate(pct = round(n / sum(n) * 100, 2)) |>
  arrange(desc(n))

datatable(cd4_prop,
          caption = "CD4 Functional State Proportions")

# Barplot
p_cd4_bar <- ggplot(cd4_prop, aes(x = reorder(functional.cluster, -n), y = n, fill = functional.cluster)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = ref.cd4@misc$atlas.palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Functional State", y = "Cell Count",
       title = "CD4 T Cell Functional States")

print(p_cd4_bar)

ggsave(file.path(output_dir, "barplot_cd4_states.png"),
       p_cd4_bar,
       width = 12, height = 6, dpi = 300, bg = "white")
```

## Combined Annotations {.tabset}

### Merge CD4 + CD8

```{r merge_cd4_cd8, fig.width=14, fig.height=10}
cat("Merging CD4 and CD8 annotations back to main object...\n")

# Add functional.cluster from CD8 and CD4 back to main tcells object
tcells <- AddMetaData(tcells, tcells.cd8$functional.cluster, col.name = "functional.cluster")
tcells <- AddMetaData(tcells, tcells.cd4$functional.cluster, col.name = "functional.cluster")

# Combine color palettes
combined_colors <- c(ref.cd8@misc$atlas.palette, ref.cd4@misc$atlas.palette)

cat("\n=== Combined Functional States ===\n")
print(table(tcells$functional.cluster))

# Plot combined UMAP
p_combined <- DimPlot(tcells,
                      group.by = "functional.cluster",
                      cols = combined_colors,
                      label = TRUE,
                      repel = TRUE,
                      order = TRUE,
                      raster = FALSE) +
  coord_fixed() +
  ggtitle("All T Cells - ProjecTILs Functional States")

print(p_combined)

ggsave(file.path(output_dir, "umap_all_tcells_functional.png"),
       p_combined,
       width = 14, height = 10, dpi = 300, bg = "white")
```

### Overall Proportions

```{r overall_proportions}
# Overall functional state proportions
overall_prop <- tcells@meta.data |>
  group_by(functional.cluster, CD_type) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  mutate(
    pct = round(n / nrow(tcells@meta.data) * 100, 2)
  )

datatable(overall_prop,
          caption = "All T Cell Functional State Proportions")
```

### Summary Statistics

```{r summary_stats}
# Summary by CD4 vs CD8
cd_summary <- tcells@meta.data |>
  filter(CD_type %in% c("CD4", "CD8")) |>
  group_by(CD_type) |>
  summarise(
    n_cells = n(),
    n_states = n_distinct(functional.cluster),
    n_samples = n_distinct(Sample)
  ) |>
  mutate(
    pct = round(n_cells / sum(n_cells) * 100, 2)
  )

print(cd_summary)
```

## Comparison: SubTyping vs ProjecTILs {.tabset}

### Confusion Matrix

```{r confusion_matrix, fig.width=14, fig.height=12}
cat("Creating confusion matrix...\n")

# Confusion matrix
conf_matrix <- table(
  Old_SubTyping = tcells$SubTyping,
  New_ProjecTILs = tcells$functional.cluster
)

# Export as CSV
write.csv(conf_matrix,
          file.path(output_dir, "confusion_matrix_subtyping_vs_projectils.csv"))

cat("\n=== Confusion Matrix (first 10x10) ===\n")
print(conf_matrix[1:min(10, nrow(conf_matrix)), 1:min(10, ncol(conf_matrix))])

# Heatmap visualization
library(ComplexHeatmap)
library(circlize)

ht_conf <- Heatmap(
  log10(conf_matrix + 1),
  name = "log10(cells+1)",
  col = viridis(100),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  column_title = "Confusion Matrix: SubTyping (rows) vs ProjecTILs (columns)",
  heatmap_legend_param = list(title = "log10(cells+1)")
)

print(ht_conf)

png(file.path(output_dir, "confusion_matrix_heatmap.png"),
    width = 3500, height = 3000, res = 300)
draw(ht_conf)
dev.off()
```

### Side-by-Side UMAPs

```{r comparison_umaps, fig.width=20, fig.height=8}
# Side-by-side comparison
p_old <- DimPlot(tcells,
                 group.by = "SubTyping",
                 label = TRUE,
                 repel = TRUE,
                 raster = FALSE) +
  coord_fixed() +
  ggtitle("Old: SubTyping Labels") +
  theme(legend.position = "bottom")

p_new <- DimPlot(tcells,
                 group.by = "functional.cluster",
                 cols = combined_colors,
                 label = TRUE,
                 repel = TRUE,
                 raster = FALSE) +
  coord_fixed() +
  ggtitle("New: ProjecTILs Functional States") +
  theme(legend.position = "bottom")

p_comparison <- p_old | p_new

print(p_comparison)

ggsave(file.path(output_dir, "umap_comparison_old_vs_new.png"),
       p_comparison,
       width = 20, height = 8, dpi = 300, bg = "white")
```

### Concordance Analysis

```{r concordance}
# Calculate concordance metrics
cat("\n=== Label Concordance Analysis ===\n")

# Cells per old label
old_counts <- table(tcells$SubTyping)
cat("\nOld SubTyping labels (n > 100):\n")
print(old_counts[old_counts > 100])

# Cells per new label
new_counts <- table(tcells$functional.cluster)
cat("\nNew ProjecTILs states:\n")
print(new_counts)

# For each old label, find top new labels
cat("\n=== Top ProjecTILs states for each SubTyping label ===\n")
for (old_label in names(old_counts)[old_counts > 50]) {
  subset_cells <- tcells@meta.data |>
    filter(SubTyping == old_label)

  top_new <- subset_cells |>
    group_by(functional.cluster) |>
    summarise(n = n()) |>
    arrange(desc(n)) |>
    head(3)

  cat("\n", old_label, "(n=", old_counts[old_label], "):\n", sep = "")
  print(top_new)
}
```

### Interpretation

Based on the comparison, ProjecTILs functional annotations capture:

1. **Finer-grained functional states** - Separates naive, memory, effector, and exhausted states
2. **Tpex identification** - Progenitor exhausted CD8 T cells (TCF7+, CXCL13+) may be newly identified
3. **Regulatory subtypes** - CD4 Tregs and Tfh cells more clearly delineated
4. **Functional consistency** - States reflect functional programs rather than transcriptional clusters

## Marker Validation {.tabset}

### CD8 DotPlot

```{r cd8_dotplot, fig.width=12, fig.height=10}
# CD8 marker dotplot
cd8_cells <- subset(tcells, CD_type == "CD8" & !is.na(functional.cluster))

cat("CD8 cells with functional.cluster:", ncol(cd8_cells), "\n")

# Redefine markers for this subset
cd8_markers_dotplot <- c("CD8A", "CD8B", "CD3D", "CD3E",
                         "CCR7", "TCF7", "LEF1", "IL7R",
                         "GZMA", "GZMK", "GZMB", "PRF1",
                         "PDCD1", "HAVCR2", "LAG3", "TOX",
                         "CXCL13", "ENTPD1")
cd8_markers_dotplot <- cd8_markers_dotplot[cd8_markers_dotplot %in% rownames(cd8_cells)]

cat("Available markers:", length(cd8_markers_dotplot), "\n")

if (length(cd8_markers_dotplot) > 0 && ncol(cd8_cells) > 0) {
  p_dot_cd8 <- DotPlot(cd8_cells,
                       features = unique(cd8_markers_dotplot),
                       group.by = "functional.cluster") +
    coord_flip() +
    ggtitle("CD8 T Cell Markers by Functional State") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_dot_cd8)

  ggsave(file.path(output_dir, "dotplot_cd8_markers.png"),
         p_dot_cd8,
         width = 12, height = 10, dpi = 300, bg = "white")
} else {
  cat("Skipping CD8 dotplot - no cells or markers available\n")
}
```

### CD4 DotPlot

```{r cd4_dotplot, fig.width=12, fig.height=10}
# CD4 marker dotplot
cd4_cells <- subset(tcells, CD_type == "CD4" & !is.na(functional.cluster))

cat("CD4 cells with functional.cluster:", ncol(cd4_cells), "\n")

# Redefine markers for this subset
cd4_markers_dotplot <- c("CD4", "CD3D", "CD3E",
                         "CCR7", "TCF7", "LEF1", "IL7R",
                         "FOXP3", "IL2RA", "CTLA4",
                         "PDCD1", "HAVCR2", "LAG3",
                         "CXCL13", "ICOS", "TIGIT")
cd4_markers_dotplot <- cd4_markers_dotplot[cd4_markers_dotplot %in% rownames(cd4_cells)]

cat("Available markers:", length(cd4_markers_dotplot), "\n")

if (length(cd4_markers_dotplot) > 0 && ncol(cd4_cells) > 0) {
  p_dot_cd4 <- DotPlot(cd4_cells,
                       features = unique(cd4_markers_dotplot),
                       group.by = "functional.cluster") +
    coord_flip() +
    ggtitle("CD4 T Cell Markers by Functional State") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_dot_cd4)

  ggsave(file.path(output_dir, "dotplot_cd4_markers.png"),
         p_dot_cd4,
         width = 12, height = 10, dpi = 300, bg = "white")
} else {
  cat("Skipping CD4 dotplot - no cells or markers available\n")
}
```

### Key Marker FeaturePlots

```{r feature_plots, fig.width=18, fig.height=12}
# Key markers across all T cells
key_markers <- c("CD8A", "CD4", "GZMB", "PDCD1", "FOXP3", "TCF7", "TOX", "CXCL13", "HAVCR2")
key_markers <- key_markers[key_markers %in% rownames(tcells)]

p_features <- FeaturePlot(tcells,
                          features = key_markers,
                          ncol = 3,
                          raster = FALSE,
                          order = TRUE) &
  coord_fixed()

print(p_features)

ggsave(file.path(output_dir, "featureplot_key_markers.png"),
       p_features,
       width = 18, height = 12, dpi = 300, bg = "white")
```

## Proportions Analysis {.tabset}

### By Sample

```{r props_sample, fig.width=20, fig.height=10}
# Proportions by sample
prop_sample <- tcells@meta.data |>
  group_by(Sample, functional.cluster) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(Sample) |>
  mutate(prop = n / sum(n) * 100) |>
  ungroup()

# Export
write.csv(prop_sample,
          file.path(output_dir, "proportions_by_sample.csv"),
          row.names = FALSE)

# Stacked barplot
p_prop_sample <- ggplot(prop_sample, aes(x = Sample, y = prop, fill = functional.cluster)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = combined_colors) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
        legend.position = "right") +
  labs(y = "Proportion (%)",
       title = "T Cell Functional States by Sample",
       fill = "Functional State")

print(p_prop_sample)

ggsave(file.path(output_dir, "proportions_by_sample.png"),
       p_prop_sample,
       width = 20, height = 10, dpi = 300, bg = "white")
```

### By Origin (Disease Stage)

```{r props_origin, fig.width=12, fig.height=8}
# Proportions by origin (Healthy vs Primary vs Metastatic)
prop_origin <- tcells@meta.data |>
  filter(!is.na(origin)) |>
  group_by(origin, functional.cluster) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(origin) |>
  mutate(prop = n / sum(n) * 100) |>
  ungroup()

# Export
write.csv(prop_origin,
          file.path(output_dir, "proportions_by_origin.csv"),
          row.names = FALSE)

# Stacked barplot (proportions)
p_prop_origin <- ggplot(prop_origin, aes(x = origin, y = prop, fill = functional.cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = combined_colors) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(y = "Proportion",
       x = "Disease Origin",
       title = "T Cell Functional States by Disease Stage",
       fill = "Functional State")

print(p_prop_origin)

ggsave(file.path(output_dir, "proportions_by_origin.png"),
       p_prop_origin,
       width = 12, height = 8, dpi = 300, bg = "white")

# Summary table
origin_summary <- prop_origin |>
  select(origin, functional.cluster, n, prop) |>
  arrange(origin, desc(prop))

datatable(origin_summary,
          caption = "T Cell State Proportions by Disease Origin")
```

### By Secondary Mutation

```{r props_mutation, fig.width=12, fig.height=8}
# Proportions by secondary mutation (Primary tumors only)
tcells_primary <- subset(tcells,
                         origin == "primary" &
                         secondary_mutation %in% c("BAP1", "SF3B1", "EIF1AX"))

cat("Primary tumor cells with known mutations:", ncol(tcells_primary), "\n")

prop_mut <- tcells_primary@meta.data |>
  group_by(secondary_mutation, functional.cluster) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(secondary_mutation) |>
  mutate(prop = n / sum(n) * 100) |>
  ungroup()

# Export
write.csv(prop_mut,
          file.path(output_dir, "proportions_by_mutation.csv"),
          row.names = FALSE)

# Stacked barplot
p_prop_mut <- ggplot(prop_mut, aes(x = secondary_mutation, y = prop, fill = functional.cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = combined_colors) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(y = "Proportion",
       x = "Secondary Mutation",
       title = "T Cell Functional States by Secondary Mutation (Primary Tumors)",
       fill = "Functional State")

print(p_prop_mut)

ggsave(file.path(output_dir, "proportions_by_mutation.png"),
       p_prop_mut,
       width = 12, height = 8, dpi = 300, bg = "white")

# Summary table
mut_summary <- prop_mut |>
  select(secondary_mutation, functional.cluster, n, prop) |>
  arrange(secondary_mutation, desc(prop))

datatable(mut_summary,
          caption = "T Cell State Proportions by Secondary Mutation")
```

### Proportions by CD4/CD8 Type

```{r props_cd_type, fig.width=14, fig.height=10}
# Proportions split by CD4 vs CD8
prop_cd_origin <- tcells@meta.data |>
  filter(!is.na(origin), CD_type %in% c("CD4", "CD8")) |>
  group_by(origin, CD_type, functional.cluster) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(origin, CD_type) |>
  mutate(prop = n / sum(n) * 100) |>
  ungroup()

# Faceted barplot
p_prop_cd <- ggplot(prop_cd_origin,
                    aes(x = origin, y = prop, fill = functional.cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = combined_colors) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ CD_type, ncol = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Proportion",
       x = "Disease Origin",
       title = "T Cell Functional States by Disease Stage (CD4 vs CD8)",
       fill = "Functional State")

print(p_prop_cd)

ggsave(file.path(output_dir, "proportions_by_origin_cd_type.png"),
       p_prop_cd,
       width = 14, height = 10, dpi = 300, bg = "white")
```

## Save Annotated Object

```{r save_object}
# Save annotated T cell object for future use
output_path <- "../tme_tcells_projectils.qs2"

cat("Saving annotated T cell object...\n")
qs2::qs_save(tcells, output_path, nthreads = 16)

cat("\nSaved to:", output_path, "\n")
cat("Total cells:", ncol(tcells), "\n")
cat("CD8 cells:", sum(tcells$CD_type == "CD8"), "\n")
cat("CD4 cells:", sum(tcells$CD_type == "CD4"), "\n")
cat("MAIT cells:", sum(tcells$CD_type == "MAIT"), "\n")
cat("gdT cells:", sum(tcells$CD_type == "gdT"), "\n")
cat("NK cells:", sum(tcells$CD_type == "NK"), "\n")
cat("Functional states:", length(unique(na.omit(tcells$functional.cluster))), "\n")

# Check file size
file_info <- file.info(output_path)
cat("File size:", round(file_info$size / 1024^3, 2), "GB\n")
```

## Session Info

```{r session_info, echo=FALSE}
ezSessionInfo()
```
