---
title: "p31662 - sccomp Compositional Analysis of UM TME"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Extended Data (sccomp compositional analysis of cell type proportions)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

## Overview

This analysis performs compositional analysis of the Uveal Melanoma (UM) tumor microenvironment (TME) using **sccomp** (Mangiola et al., 2023). We examine cell type proportion changes across:

1. **Disease progression**: Healthy vs Primary vs Metastatic
2. **Paired analysis**: 7 matched healthy-primary pairs
3. **Secondary mutations**: BAP1 vs SF3B1 vs EIF1AX (primary tumors only)
4. **Tumor location**: Choroid vs Iris vs Ciliary Body (primary tumors only)

**Input**: Levesque TME subset (102,897 cells, 52 samples, 11 cell types)

```{r libraries}
library(Seurat)
library(sccomp)
library(qs)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(DT)

# Set output directories
output_dir <- "Outputs/sccomp"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

## Data Loading {.tabset}

### Load TME Object

```{r load-data}
tme <- qread("../Levesque_UMscAtlas/01_TME_subset_integrated.qs", nthreads = 16)
cat("Dimensions:", dim(tme), "\n")
cat("Samples:", length(unique(tme$Sample)), "\n")
cat("Cell types:", length(unique(tme$majority_celltype)), "\n")
```

### Filter Rare Cell Types

Remove cell types present in <10% of samples (e.g., Photoreceptor cells).

```{r filter-celltypes}
# Calculate cell type frequency across samples
ct_freq <- tme@meta.data |>
  group_by(majority_celltype) |>
  summarise(
    n_cells = n(),
    n_samples = n_distinct(Sample),
    pct_samples = round(n_samples / length(unique(tme$Sample)) * 100, 1)
  ) |>
  arrange(desc(n_cells))

# Display table
DT::datatable(ct_freq, caption = "Cell Type Distribution")

# Filter to cell types in ≥10% of samples
keep_ct <- ct_freq |> filter(pct_samples >= 10) |> pull(majority_celltype)
cat("Keeping", length(keep_ct), "cell types (≥10% of samples):\n")
print(keep_ct)

tme <- subset(tme, majority_celltype %in% keep_ct)
cat("\nFiltered dimensions:", dim(tme), "\n")
```

### Sample Metadata

```{r sample-metadata}
# Create sample-level metadata for sccomp
sample_meta <- tme@meta.data |>
  group_by(Sample) |>
  summarise(
    Patient_nr = first(Patient_nr),
    origin = first(origin),
    organ = first(organ),
    primary_mutation = first(primary_mutation),
    secondary_mutation = first(secondary_mutation),
    LocationPrimaryTumor = first(LocationPrimaryTumor),
    n_cells = n()
  ) |>
  ungroup()

DT::datatable(sample_meta, caption = "Sample-Level Metadata")
```

## Q1: Disease Progression (All Cohort) {.tabset}

### Model

Full cohort analysis: Healthy vs Primary vs Metastatic with patient random effect.

```{r q1-model}
# Set factor levels
tme$origin <- factor(tme$origin, levels = c("healthy", "primary", "metastasis"))

# Run sccomp
res_origin <- tme |>
  sccomp_estimate(
    formula_composition = ~ origin + (1 | Patient_nr),
    sample = "Sample",
    cell_group = "majority_celltype",
    bimodal_mean_variability_association = TRUE,
    cores = 8
  ) |>
  sccomp_remove_outliers(cores = 8) |>
  sccomp_test(contrasts = c(
    "originprimary",
    "originmetastasis",
    "originmetastasis - originprimary"
  ))

# Save results
saveRDS(res_origin, file.path(output_dir, "sccomp_origin_results.rds"))
```

### Results

```{r q1-results}
# Extract and display results
res_origin_df <- res_origin |>
  filter(parameter != "(Intercept)") |>
  select(cell_group = majority_celltype, parameter, c_effect, c_lower, c_upper, c_pH0) |>
  mutate(
    significant = c_pH0 < 0.05,
    effect_direction = ifelse(c_effect > 0, "Increased", "Decreased")
  ) |>
  arrange(parameter, c_pH0)

DT::datatable(res_origin_df |> mutate(across(where(is.numeric), ~round(., 3))),
              caption = "Disease Progression - Compositional Changes")
```

### Boxplot

```{r q1-boxplot, fig.width=14, fig.height=10}
p_box <- res_origin |>
  sccomp_boxplot(factor = "origin", remove_unwanted_effects = TRUE)
print(p_box)

ggsave(file.path(output_dir, "boxplot_origin.png"), p_box,
       width = 14, height = 10, dpi = 300, bg = "white")
```

### Credible Intervals

```{r q1-intervals, fig.width=10, fig.height=10}
p_int <- res_origin |>
  plot_1D_intervals()
print(p_int)

ggsave(file.path(output_dir, "intervals_origin.png"), p_int,
       width = 10, height = 10, dpi = 300, bg = "white")
```

### FC Heatmap

```{r q1-heatmap, fig.width=10, fig.height=8}
# Prepare data for heatmap
fc_df <- res_origin_df |>
  select(cell_group, parameter, c_effect, c_pH0) |>
  mutate(parameter = case_when(
    parameter == "originprimary" ~ "Primary vs Healthy",
    parameter == "originmetastasis" ~ "Met vs Healthy",
    parameter == "originmetastasis - originprimary" ~ "Met vs Primary",
    TRUE ~ parameter
  ))

# Create FC matrix
fc_matrix <- fc_df |>
  select(cell_group, parameter, c_effect) |>
  pivot_wider(names_from = parameter, values_from = c_effect) |>
  column_to_rownames("cell_group") |>
  as.matrix()

# Create significance matrix
sig_matrix <- fc_df |>
  mutate(sig = case_when(
    c_pH0 < 0.001 ~ "***",
    c_pH0 < 0.01 ~ "**",
    c_pH0 < 0.05 ~ "*",
    TRUE ~ ""
  )) |>
  select(cell_group, parameter, sig) |>
  pivot_wider(names_from = parameter, values_from = sig) |>
  column_to_rownames("cell_group") |>
  as.matrix()

# Create heatmap
ht_origin <- Heatmap(
  fc_matrix,
  name = "Log2 FC",
  col = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sig_matrix[i, j], x, y, gp = gpar(fontsize = 12, fontface = "bold"))
  },
  column_title = "Disease Progression - Cell Type Compositional Changes",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 11),
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(
    title = "Log2 FC",
    legend_height = unit(4, "cm")
  )
)

draw(ht_origin)

# Save
png(file.path(output_dir, "heatmap_origin_fc.png"), width = 2400, height = 2000, res = 300)
draw(ht_origin)
dev.off()
```

## Q2: Paired Healthy vs Primary (7 Pairs) {.tabset}

### Paired Samples

```{r q2-pairs}
# Define paired samples (patients P16-P22 with healthy + primary)
paired_patients <- c("P16", "P17", "P18", "P19", "P20", "P21", "P22")

# Filter to paired samples
tme_paired <- subset(tme, Patient_nr %in% paired_patients & origin %in% c("healthy", "primary"))

# Verify pairs
paired_summary <- tme_paired@meta.data |>
  group_by(Patient_nr, origin) |>
  summarise(
    Sample = first(Sample),
    n_cells = n(),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = origin, values_from = c(Sample, n_cells))

DT::datatable(paired_summary, caption = "7 Matched Healthy-Primary Pairs")
```
### Model

```{r q2-model}
# Run paired sccomp with patient random effect
res_paired <- tme_paired |>
  sccomp_estimate(
    formula_composition = ~ origin + (1 | Patient_nr),
    sample = "Sample",
    cell_group = "majority_celltype",
    bimodal_mean_variability_association = TRUE,
    cores = 8
  ) |>
  sccomp_remove_outliers(cores = 8) |>
  sccomp_test()

# Save results
saveRDS(res_paired, file.path(output_dir, "sccomp_paired_results.rds"))
```

### Results

```{r q2-results}
res_paired_df <- res_paired |>
  filter(parameter != "(Intercept)") |>
  select(cell_group = majority_celltype, parameter, c_effect, c_lower, c_upper, c_pH0) |>
  mutate(
    significant = c_pH0 < 0.05,
    effect_direction = ifelse(c_effect > 0, "Increased in Primary", "Decreased in Primary")
  ) |>
  arrange(c_pH0)

DT::datatable(res_paired_df |> mutate(across(where(is.numeric), ~round(., 3))),
              caption = "Paired Analysis - Primary vs Healthy")
```

### Boxplot

```{r q2-boxplot, fig.width=14, fig.height=10}
p_box2 <- res_paired |>
  sccomp_boxplot(factor = "origin", remove_unwanted_effects = TRUE)
print(p_box2)

ggsave(file.path(output_dir, "boxplot_paired.png"), p_box2,
       width = 14, height = 10, dpi = 300, bg = "white")
```

### Credible Intervals

```{r q2-intervals, fig.width=10, fig.height=10}
p_int2 <- res_paired |>
  plot_1D_intervals()
print(p_int2)

ggsave(file.path(output_dir, "intervals_paired.png"), p_int2,
       width = 10, height = 10, dpi = 300, bg = "white")
```

### FC Heatmap

```{r q2-heatmap, fig.width=8, fig.height=10}
# Create matrix from paired results
fc_matrix_paired <- res_paired_df |>
  select(cell_group, c_effect) |>
  arrange(desc(c_effect))

sig_matrix_paired <- res_paired_df |>
  mutate(sig = case_when(
    c_pH0 < 0.001 ~ "***",
    c_pH0 < 0.01 ~ "**",
    c_pH0 < 0.05 ~ "*",
    TRUE ~ ""
  )) |>
  select(cell_group, sig) |>
  arrange(match(cell_group, fc_matrix_paired$cell_group))

fc_mat <- matrix(
  fc_matrix_paired$c_effect,
  ncol = 1,
  dimnames = list(fc_matrix_paired$cell_group, "Primary vs Healthy")
)

sig_mat <- matrix(
  sig_matrix_paired$sig,
  ncol = 1,
  dimnames = list(sig_matrix_paired$cell_group, "Primary vs Healthy")
)

ht_paired <- Heatmap(
  fc_mat,
  name = "Log2 FC",
  col = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sig_mat[i, j], x, y, gp = gpar(fontsize = 12, fontface = "bold"))
  },
  column_title = "Paired Analysis (7 Pairs) - Primary vs Healthy",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_names_gp = gpar(fontsize = 10),
  cluster_rows = FALSE,
  rect_gp = gpar(col = "white", lwd = 1),
  width = unit(4, "cm"),
  heatmap_legend_param = list(
    title = "Log2 FC",
    legend_height = unit(4, "cm")
  )
)

draw(ht_paired)

png(file.path(output_dir, "heatmap_paired_fc.png"), width = 2000, height = 2400, res = 300)
draw(ht_paired)
dev.off()
```

## Q3: Secondary Mutations (pUM Only) {.tabset}

### Subset to Known Mutations

```{r q3-setup}
# Primary tumors with known secondary mutations only
tme_secondary <- subset(tme,
  origin == "primary" &
  secondary_mutation %in% c("BAP1", "SF3B1", "EIF1AX")
)

tme_secondary$secondary_mutation <- factor(
  tme_secondary$secondary_mutation,
  levels = c("BAP1", "SF3B1", "EIF1AX")
)

# Summary
sec_summary <- tme_secondary@meta.data |>
  group_by(secondary_mutation) |>
  summarise(n_samples = n_distinct(Sample), n_cells = n())

DT::datatable(sec_summary, caption = "Secondary Mutation Distribution (Primary Tumors)")
```

### Model

```{r q3-model}
# Run sccomp with secondary mutation (BAP1 as reference, default contrasts)
res_secondary <- tme_secondary |>
  sccomp_estimate(
    formula_composition = ~ secondary_mutation + primary_mutation,
    sample = "Sample",
    cell_group = "majority_celltype",
    bimodal_mean_variability_association = TRUE,
    cores = 8
  ) |>
  sccomp_remove_outliers(cores = 8) |>
  sccomp_test()

saveRDS(res_secondary, file.path(output_dir, "sccomp_secondary_mutation_results.rds"))
```

### Results

```{r q3-results}
res_secondary_df <- res_secondary |>
  filter(parameter != "(Intercept)" & !grepl("primary_mutation", parameter)) |>
  select(cell_group = majority_celltype, parameter, c_effect, c_lower, c_upper, c_pH0) |>
  mutate(
    parameter = gsub("secondary_mutation", "", parameter),
    significant = c_pH0 < 0.05
  ) |>
  arrange(parameter, c_pH0)

DT::datatable(res_secondary_df |> mutate(across(where(is.numeric), ~round(., 3))),
              caption = "Secondary Mutation Contrasts")
```

### Boxplot

```{r q3-boxplot, fig.width=14, fig.height=10}
p_box3 <- res_secondary |>
  sccomp_boxplot(factor = "secondary_mutation", remove_unwanted_effects = TRUE)
print(p_box3)

ggsave(file.path(output_dir, "boxplot_secondary.png"), p_box3,
       width = 14, height = 10, dpi = 300, bg = "white")
```

### Credible Intervals

```{r q3-intervals, fig.width=10, fig.height=10}
p_int3 <- res_secondary |>
  plot_1D_intervals()
print(p_int3)

ggsave(file.path(output_dir, "intervals_secondary.png"), p_int3,
       width = 10, height = 10, dpi = 300, bg = "white")
```

### FC Heatmap

```{r q3-heatmap, fig.width=10, fig.height=8}
fc_matrix_sec <- res_secondary_df |>
  select(cell_group, parameter, c_effect) |>
  pivot_wider(names_from = parameter, values_from = c_effect) |>
  column_to_rownames("cell_group") |>
  as.matrix()

sig_matrix_sec <- res_secondary_df |>
  mutate(sig = case_when(
    c_pH0 < 0.001 ~ "***",
    c_pH0 < 0.01 ~ "**",
    c_pH0 < 0.05 ~ "*",
    TRUE ~ ""
  )) |>
  select(cell_group, parameter, sig) |>
  pivot_wider(names_from = parameter, values_from = sig) |>
  column_to_rownames("cell_group") |>
  as.matrix()

ht_secondary <- Heatmap(
  fc_matrix_sec,
  name = "Log2 FC",
  col = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sig_matrix_sec[i, j], x, y, gp = gpar(fontsize = 12, fontface = "bold"))
  },
  column_title = "Secondary Mutations - Cell Type Compositional Changes",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 11),
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 1)
)

draw(ht_secondary)

png(file.path(output_dir, "heatmap_secondary_fc.png"), width = 2400, height = 2000, res = 300)
draw(ht_secondary)
dev.off()
```

## Q4: Tumor Location (pUM Only) {.tabset}

### Subset by Location

```{r q4-setup}
# Primary tumors with known locations
tme_location <- subset(tme,
  origin == "primary" &
  LocationPrimaryTumor %in% c("choroid", "iris", "ciliary body")
)

tme_location$LocationPrimaryTumor <- factor(
  tme_location$LocationPrimaryTumor,
  levels = c("choroid", "iris", "ciliary body")
)

# Summary
loc_summary <- tme_location@meta.data |>
  group_by(LocationPrimaryTumor) |>
  summarise(n_samples = n_distinct(Sample), n_cells = n())

DT::datatable(loc_summary, caption = "Primary Tumor Location Distribution")
```

### Model

```{r q4-model}
# Run sccomp with location showing all three locations explicitly
# Use cell means coding to get all location effects
res_location <- tme_location |>
  sccomp_estimate(
    formula_composition = ~ 0 + LocationPrimaryTumor + secondary_mutation,
    sample = "Sample",
    cell_group = "majority_celltype",
    bimodal_mean_variability_association = TRUE,
    cores = 8
  ) |>
  sccomp_remove_outliers(cores = 8) |>
  sccomp_test()

saveRDS(res_location, file.path(output_dir, "sccomp_location_results.rds"))
```

### Results

```{r q4-results}
res_location_df <- res_location |>
  filter(parameter != "(Intercept)" & !grepl("secondary_mutation", parameter)) |>
  select(cell_group = majority_celltype, parameter, c_effect, c_lower, c_upper, c_pH0) |>
  mutate(
    parameter = gsub("LocationPrimaryTumor", "", parameter),
    parameter = gsub("`", "", parameter),
    significant = c_pH0 < 0.05
  ) |>
  arrange(parameter, c_pH0)

DT::datatable(res_location_df |> mutate(across(where(is.numeric), ~round(., 3))),
              caption = "Tumor Location Contrasts (vs Choroid)")
```

### Boxplot

```{r q4-boxplot, fig.width=14, fig.height=10}
p_box4 <- res_location |>
  sccomp_boxplot(factor = "LocationPrimaryTumor", remove_unwanted_effects = TRUE)
print(p_box4)

ggsave(file.path(output_dir, "boxplot_location.png"), p_box4,
       width = 14, height = 10, dpi = 300, bg = "white")
```

### Credible Intervals

```{r q4-intervals, fig.width=10, fig.height=10}
p_int4 <- res_location |>
  plot_1D_intervals()
print(p_int4)

ggsave(file.path(output_dir, "intervals_location.png"), p_int4,
       width = 10, height = 10, dpi = 300, bg = "white")
```

### FC Heatmap

```{r q4-heatmap, fig.width=8, fig.height=8}
fc_matrix_loc <- res_location_df |>
  select(cell_group, parameter, c_effect) |>
  pivot_wider(names_from = parameter, values_from = c_effect) |>
  column_to_rownames("cell_group") |>
  as.matrix()

sig_matrix_loc <- res_location_df |>
  mutate(sig = case_when(
    c_pH0 < 0.001 ~ "***",
    c_pH0 < 0.01 ~ "**",
    c_pH0 < 0.05 ~ "*",
    TRUE ~ ""
  )) |>
  select(cell_group, parameter, sig) |>
  pivot_wider(names_from = parameter, values_from = sig) |>
  column_to_rownames("cell_group") |>
  as.matrix()

ht_location <- Heatmap(
  fc_matrix_loc,
  name = "Log2 FC",
  col = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sig_matrix_loc[i, j], x, y, gp = gpar(fontsize = 12, fontface = "bold"))
  },
  column_title = "Tumor Location - Cell Type Compositional Changes (vs Choroid)",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 11),
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 1)
)

draw(ht_location)

png(file.path(output_dir, "heatmap_location_fc.png"), width = 2000, height = 2000, res = 300)
draw(ht_location)
dev.off()
```

## Q5: SubTyping Analysis (Disease Progression) {.tabset}

### Filter SubTypes

Analyze disease progression using granular SubTyping (44 categories) instead of majority_celltype.

```{r q5-filter}
# Calculate SubTyping frequency across samples
subtype_freq <- tme@meta.data |>
  group_by(SubTyping) |>
  summarise(
    n_cells = n(),
    n_samples = n_distinct(Sample),
    pct_samples = round(n_samples / length(unique(tme$Sample)) * 100, 1)
  ) |>
  arrange(desc(n_cells))

DT::datatable(subtype_freq, caption = "SubTyping Distribution (44 Categories)")

# Filter to subtypes in ≥10% of samples
keep_subtype <- subtype_freq |> filter(pct_samples >= 10) |> pull(SubTyping)
cat("Keeping", length(keep_subtype), "subtypes (≥10% of samples):\n")
print(keep_subtype)

tme_subtype <- subset(tme, SubTyping %in% keep_subtype)
cat("\nFiltered dimensions:", dim(tme_subtype), "\n")
```

### Model

```{r q5-model}
# Set factor levels
tme_subtype$origin <- factor(tme_subtype$origin, levels = c("healthy", "primary", "metastasis"))

# Run sccomp with SubTyping
res_subtype <- tme_subtype |>
  sccomp_estimate(
    formula_composition = ~ origin + (1 | Patient_nr),
    sample = "Sample",
    cell_group = "SubTyping",
    bimodal_mean_variability_association = TRUE,
    cores = 8
  ) |>
  sccomp_remove_outliers(cores = 8) |>
  sccomp_test(contrasts = c(
    "originprimary",
    "originmetastasis",
    "originmetastasis - originprimary"
  ))

saveRDS(res_subtype, file.path(output_dir, "sccomp_subtyping_results.rds"))
```

### Results

```{r q5-results}
res_subtype_df <- res_subtype |>
  filter(parameter != "(Intercept)") |>
  select(cell_group = SubTyping, parameter, c_effect, c_lower, c_upper, c_pH0) |>
  mutate(
    significant = c_pH0 < 0.05,
    effect_direction = ifelse(c_effect > 0, "Increased", "Decreased")
  ) |>
  arrange(parameter, c_pH0)

DT::datatable(res_subtype_df |> mutate(across(where(is.numeric), ~round(., 3))),
              caption = "SubTyping - Disease Progression Compositional Changes")
```

### Boxplot

```{r q5-boxplot, fig.width=16, fig.height=14}
p_box5 <- res_subtype |>
  sccomp_boxplot(factor = "origin", remove_unwanted_effects = TRUE)
print(p_box5)

ggsave(file.path(output_dir, "boxplot_subtyping.png"), p_box5,
       width = 16, height = 14, dpi = 300, bg = "white")
```

### Credible Intervals

```{r q5-intervals, fig.width=12, fig.height=14}
p_int5 <- res_subtype |>
  plot_1D_intervals()
print(p_int5)

ggsave(file.path(output_dir, "intervals_subtyping.png"), p_int5,
       width = 12, height = 14, dpi = 300, bg = "white")
```

### FC Heatmap

```{r q5-heatmap, fig.width=10, fig.height=16}
# Prepare data for heatmap
fc_df_sub <- res_subtype_df |>
  select(cell_group, parameter, c_effect, c_pH0) |>
  mutate(parameter = case_when(
    parameter == "originprimary" ~ "Primary vs Healthy",
    parameter == "originmetastasis" ~ "Met vs Healthy",
    parameter == "originmetastasis - originprimary" ~ "Met vs Primary",
    TRUE ~ parameter
  ))

# Create FC matrix
fc_matrix_sub <- fc_df_sub |>
  select(cell_group, parameter, c_effect) |>
  pivot_wider(names_from = parameter, values_from = c_effect) |>
  column_to_rownames("cell_group") |>
  as.matrix()

# Create significance matrix
sig_matrix_sub <- fc_df_sub |>
  mutate(sig = case_when(
    c_pH0 < 0.001 ~ "***",
    c_pH0 < 0.01 ~ "**",
    c_pH0 < 0.05 ~ "*",
    TRUE ~ ""
  )) |>
  select(cell_group, parameter, sig) |>
  pivot_wider(names_from = parameter, values_from = sig) |>
  column_to_rownames("cell_group") |>
  as.matrix()

# Create heatmap
ht_subtype <- Heatmap(
  fc_matrix_sub,
  name = "Log2 FC",
  col = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sig_matrix_sub[i, j], x, y, gp = gpar(fontsize = 10, fontface = "bold"))
  },
  column_title = "SubTyping - Disease Progression Compositional Changes",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 11),
  cluster_columns = FALSE,
  rect_gp = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(
    title = "Log2 FC",
    legend_height = unit(4, "cm")
  )
)

draw(ht_subtype)

# Save
png(file.path(output_dir, "heatmap_subtyping_fc.png"), width = 2400, height = 3600, res = 300)
draw(ht_subtype)
dev.off()
```

## Combined Summary Heatmap {.tabset}

### Create Combined Matrix

Combine all analyses into a single summary heatmap showing all contrasts.

```{r combined-matrix}
# Combine all results
all_results <- bind_rows(
  res_origin_df |> mutate(Analysis = "Origin", parameter = gsub("origin", "", parameter)),
  res_paired_df |> mutate(Analysis = "Paired", parameter = "Primary vs Healthy"),
  res_secondary_df |> mutate(Analysis = "Secondary", parameter = gsub("secondary_mutation", "", parameter)),
  res_location_df |> mutate(Analysis = "Location", parameter = gsub("LocationPrimaryTumor", "", parameter))
)

# Create comparison labels
all_results <- all_results |>
  mutate(Comparison = paste(Analysis, parameter, sep = ": "))

# Export combined results
write_csv(all_results, file.path(output_dir, "sccomp_all_results.csv"))
DT::datatable(all_results |> mutate(across(where(is.numeric), ~round(., 3))),
              caption = "All sccomp Results Combined")
```

### Summary Heatmap

```{r combined-heatmap, fig.width=14, fig.height=10}
# Create FC matrix - use values_fn to handle any duplicates
fc_wide <- all_results |>
  select(cell_group, Comparison, c_effect) |>
  pivot_wider(names_from = Comparison, values_from = c_effect, values_fn = first)

cell_groups <- fc_wide$cell_group
fc_combined <- fc_wide |>
  select(-cell_group) |>
  mutate(across(everything(), as.numeric)) |>
  as.matrix()
rownames(fc_combined) <- cell_groups

# Create pH0 matrix
pval_wide <- all_results |>
  select(cell_group, Comparison, c_pH0) |>
  pivot_wider(names_from = Comparison, values_from = c_pH0, values_fn = first)

pval_combined <- pval_wide |>
  select(-cell_group) |>
  mutate(across(everything(), as.numeric)) |>
  as.matrix()
rownames(pval_combined) <- cell_groups

# Column annotation by analysis type
analysis_anno <- all_results |>
  select(Comparison, Analysis) |>
  distinct() |>
  arrange(match(Analysis, c("Origin", "Paired", "Secondary", "Location")))

col_anno <- HeatmapAnnotation(
  Analysis = analysis_anno$Analysis,
  col = list(Analysis = c(
    "Origin" = "#E41A1C",
    "Paired" = "#377EB8",
    "Secondary" = "#4DAF4A",
    "Location" = "#984EA3"
  )),
  annotation_name_gp = gpar(fontsize = 10),
  annotation_legend_param = list(title = "Analysis Type")
)

# Reorder columns by analysis
fc_combined <- fc_combined[, analysis_anno$Comparison]
pval_combined <- pval_combined[, analysis_anno$Comparison]

# Create heatmap
ht_combined <- Heatmap(
  fc_combined,
  name = "Log2 FC",
  col = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  top_annotation = col_anno,
  rect_gp = gpar(col = "white", lwd = 0.5),
  cell_fun = function(j, i, x, y, width, height, fill) {
    pval <- pval_combined[i, j]
    if (!is.na(pval) && pval < 0.05) {
      grid.points(x, y, pch = 16, size = unit(2.5, "mm"),
                  gp = gpar(col = ifelse(pval < 0.01, "black", "grey50")))
    }
  },
  column_title = "sccomp Compositional Analysis Summary - UM TME",
  column_title_gp = gpar(fontsize = 16, fontface = "bold"),
  row_names_gp = gpar(fontsize = 9),
  column_names_gp = gpar(fontsize = 8),
  column_names_rot = 45,
  cluster_columns = FALSE,
  heatmap_legend_param = list(
    title = "Log2 FC",
    legend_height = unit(4, "cm")
  )
)

draw(ht_combined, annotation_legend_side = "right")

# Save high-res
png(file.path(output_dir, "heatmap_combined_summary.png"), width = 4000, height = 3000, res = 300)
draw(ht_combined, annotation_legend_side = "right")
dev.off()
```

### Legend

- **Dot in cell**: Significant at p < 0.05 (black = p < 0.01, grey = 0.01 ≤ p < 0.05)
- **Color scale**: Blue = decreased proportion, Red = increased proportion
- **Analysis types**:
  - **Origin**: Full cohort disease progression comparisons
  - **Paired**: 7 matched healthy-primary pairs only
  - **Secondary**: BAP1/SF3B1/EIF1AX mutations in primary tumors
  - **Location**: Iris/Ciliary body vs Choroid in primary tumors

## Session Info

```{r session}
sessionInfo()
```
