---
title: "SComatic Somatic Mutation Analysis - UM Atlas Cohort"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Figure 5 (SComatic somatic mutation downstream analysis: TMB, mutation burden)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 7)

# Load required libraries
library(tidyverse)
library(ggplot2)
library(patchwork)
library(viridis)
library(kableExtra)
library(MutationalPatterns)
library(BSgenome.Hsapiens.UCSC.hg38)
ref_genome <- "BSgenome.Hsapiens.UCSC.hg38"

# Set output directory
output_dir <- "../SComatic/downstream_analysis/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Set paths
scomatic_dir <- "../SComatic"
manifest_file <- file.path(scomatic_dir, "scomatic_bam_manifest.csv")
```

#  {.tabset}


## Pipeline Summary

- **SComatic Pipeline**: Steps 1-4 complete
- **Total Cohort Size**: 44 samples (33 unique patients)
- **Samples Analyzed**: 36 samples (including 4 new Iris samples: pUM14, pUM15, pUM15_16S, pUM17_16S)
- **Excluded Samples**: 2 samples
  - **1 sample excluded**: BAM alignment processing (pUM09)
  - **1 sample excluded**: Data corruption during Step 3 (mUM08_1, brain metastasis, 99 cells)
- **Total PASS Variants**: ~1,200,000+ high-confidence somatic mutations
- **Analysis Focus**: UV damage signatures (iris vs choroid comparison)
- **Iris/High_UV Samples**: 6 samples (pUM14, pUM15, pUM15_16S, pUM17_16S, pUM19, pUM21)

## Excluded Samples Details

### Samples with processing issues (1 sample)

The following sample from the full UM Atlas cohort was excluded from SComatic analysis:

| Sample | Cells | Origin | Location | Reason |
|--------|-------|--------|----------|--------|
| pUM09 | 7,494 | Primary | Choroid | BAM alignment file not available |

### Sample with data corruption (1 sample)

| Sample | Cells | Origin | Location | Reason |
|--------|-------|--------|----------|--------|
| mUM08_1 | 99 | Metastasis | Brain (choroid origin) | Data corruption in Step 3 |

This sample had corrupted output during SComatic Step 3 (BaseCellCountsMerged). Given the very low cell count (99 cells), it was excluded rather than re-processed.

### Recently Added Iris Samples (4 samples)

The following Iris/High_UV samples were added to complete the UV signature analysis:

| Sample | Cells | Origin | Location | UV Exposure |
|--------|-------|--------|----------|-------------|
| pUM14 | 575 | Primary | Iris | High_UV |
| pUM15 | 8,685 | Primary | Iris | High_UV |
| pUM15_16S | 8,012 | Primary | Iris | High_UV |
| pUM17_16S | 1,876 | Primary | Iris | High_UV |

## Objectives

1. Calculate tumor mutational burden (TMB) per sample
2. Extract COSMIC mutational signatures
3. Compare UV damage signatures (SBS7a/b) between iris and choroid samples
4. Identify tissue-specific mutation patterns
5. Compare primary vs metastatic mutation profiles

## Data Loading

### Load Sample Metadata

```{r load_metadata}
# Load sample manifest
manifest <- read_csv(manifest_file, show_col_types = FALSE) %>%
  select(sample, n_cells, origin, organ, location) %>%
  mutate(
    # Tumor type
    tumor_type = ifelse(origin == "primary", "Primary", "Metastasis"),
    # Standardize location names - ONLY for PRIMARY eye samples
    location_clean = case_when(
      # For PRIMARY eye samples, use anatomical location
      tumor_type == "Primary" & organ == "eye" & grepl("iris", location, ignore.case = TRUE) ~ "Iris",
      tumor_type == "Primary" & organ == "eye" & grepl("choroid", location, ignore.case = TRUE) ~ "Choroid",
      tumor_type == "Primary" & organ == "eye" & grepl("ciliary", location, ignore.case = TRUE) ~ "Ciliary_body",
      tumor_type == "Primary" & organ == "eye" ~ "Eye_unspecified",
      # For ALL other cases (metastasis or non-eye), use organ name
      TRUE ~ organ
    ),
    # UV exposure classification - only for PRIMARY eye samples
    uv_exposure = case_when(
      tumor_type == "Primary" & organ == "eye" & location_clean == "Iris" ~ "High_UV",
      tumor_type == "Primary" & organ == "eye" & location_clean == "Choroid" ~ "Low_UV",
      tumor_type == "Primary" & organ == "eye" & location_clean == "Ciliary_body" ~ "Medium_UV",
      TRUE ~ "Unknown"
    )
  )

# Load age metadata from UM Atlas GloScope analysis
age_metadata_file <- "../GloScope_Full_Outputs/sample_metadata.csv"
if (file.exists(age_metadata_file)) {
  um_atlas_meta <- read_csv(age_metadata_file, show_col_types = FALSE) %>%
    # Extract short sample name (pUM23, mUM11, etc.)
    mutate(sample_short = gsub(".*-", "", Sample) %>% gsub("_.*", "", .)) %>%
    # Handle duplicates (16S samples) - take first occurrence
    distinct(sample_short, .keep_all = TRUE) %>%
    select(sample_short, Age_numeric, Age_group, Gender,
           primary_mutation, secondary_mutation, Patient_nr)

  # Merge with manifest
  manifest <- manifest %>%
    left_join(um_atlas_meta, by = c("sample" = "sample_short")) %>%
    mutate(
      Age_group = ifelse(is.na(Age_group) | Age_group == "Unknown", "Unknown", Age_group),
      Gender = ifelse(is.na(Gender), "Unknown", Gender)
    )

  cat("\n=== Age Distribution in Cohort ===\n")
  print(table(manifest$Age_group))
  cat("\n=== Gender Distribution ===\n")
  print(table(manifest$Gender))
} else {
  manifest$Age_group <- "Unknown"
  manifest$Age_numeric <- NA
  manifest$Gender <- "Unknown"
  cat("Note: Age metadata file not found. Age analysis will be limited.\n")
}

# Display sample overview by organ and location
manifest %>%
  group_by(organ, tumor_type, location_clean, uv_exposure) %>%
  summarise(
    n_samples = n(),
    total_cells = sum(n_cells),
    .groups = "drop"
  ) %>%
  arrange(organ, tumor_type) %>%
  kable(caption = "Sample Distribution by Organ, Tumor Type, and Location") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Overall summary by organ
cat("\n## Summary by Organ\n\n")
manifest %>%
  group_by(organ) %>%
  summarise(
    n_samples = n(),
    total_cells = sum(n_cells),
    primary = sum(tumor_type == "Primary"),
    metastasis = sum(tumor_type == "Metastasis"),
    .groups = "drop"
  ) %>%
  arrange(desc(n_samples)) %>%
  kable(caption = "Sample Distribution by Organ") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Load Variant Data

```{r load_variants}
# Function to load SComatic variants from all samples
load_scomatic_variants <- function(scomatic_dir, manifest) {
  all_variants <- map_dfr(manifest$sample, function(sample_name) {
    variant_file <- file.path(
      scomatic_dir,
      sample_name,
      "Step4_VariantCalling",
      paste0(sample_name, ".calling.step2.tsv")
    )

    if (!file.exists(variant_file)) {
      message("Variant file not found for sample: ", sample_name)
      return(NULL)
    }

    # Read variants
    variants <- read_tsv(variant_file, comment = "#", show_col_types = FALSE,
                        col_names = c("Chr", "Start", "End", "Ref", "Alt", "FILTER",
                                    "Qual", "Up_context", "Down_context",
                                    "Cell_types", "N_cells_mut", "N_reads_mut",
                                    "VAF", "Info"))

    variants <- variants %>%
      mutate(
        sample = sample_name,
        mutation_type = paste0(Ref, ">", Alt),
        trinuc_context = paste0(Up_context, "[", Ref, ">", Alt, "]", Down_context)
      )

    return(variants)
  })

  # Add metadata
  all_variants <- all_variants %>%
    left_join(manifest, by = "sample")

  return(all_variants)
}

# Load all variants
message("Loading variants from ", nrow(manifest), " samples...")
all_variants <- load_scomatic_variants(scomatic_dir, manifest)

# Summary statistics
variant_summary <- all_variants %>%
  group_by(sample, FILTER) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = FILTER, values_from = count, values_fill = 0)

# Display summary
variant_summary %>%
  left_join(manifest %>% select(sample, origin, organ, location_clean), by = "sample") %>%
  select(sample, origin, organ, location_clean, PASS, everything()) %>%
  arrange(desc(PASS)) %>%
  head(15) %>%
  kable(caption = "Variant Summary - Top 15 Samples by PASS Variants") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11) %>%
  scroll_box(width = "100%", height = "400px")

message("Total variants loaded: ", nrow(all_variants))
message("PASS variants: ", sum(all_variants$FILTER == "PASS"))
```

## TMB Analysis

### Calculate TMB

```{r calculate_tmb}
# Calculate TMB (mutations per megabase)
# Using human genome size: ~3,000 Mb
GENOME_SIZE_MB <- 3000

tmb_results <- all_variants %>%
  filter(FILTER == "PASS") %>%
  group_by(sample, origin, organ, location_clean, uv_exposure, tumor_type, n_cells) %>%
  summarise(
    n_mutations = n(),
    n_unique_mutations = n_distinct(Chr, Start, Ref, Alt),
    .groups = "drop"
  ) %>%
  mutate(
    tmb = n_mutations / GENOME_SIZE_MB,
    tmb_per_cell = n_mutations / n_cells
  )

# Save TMB results
write_csv(tmb_results, file.path(output_dir, "TMB_results.csv"))

# Display TMB summary
tmb_results %>%
  arrange(desc(tmb)) %>%
  select(sample, origin, organ, location_clean, uv_exposure, n_mutations, tmb, n_cells, tmb_per_cell) %>%
  head(15) %>%
  kable(
    caption = "Tumor Mutational Burden - Top 15 Samples",
    digits = 3,
    col.names = c("Sample", "Origin", "Organ", "Location", "UV Exposure",
                  "Mutations", "TMB (mut/Mb)", "Cells", "TMB/Cell")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# TMB summary by organ
cat("\n### TMB Summary by Organ\n\n")
tmb_results %>%
  group_by(organ, tumor_type) %>%
  summarise(
    n_samples = n(),
    median_tmb = median(tmb),
    mean_tmb = mean(tmb),
    sd_tmb = sd(tmb),
    min_tmb = min(tmb),
    max_tmb = max(tmb),
    .groups = "drop"
  ) %>%
  arrange(organ, tumor_type) %>%
  kable(caption = "TMB Statistics by Organ and Tumor Type", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### TMB Visualization

```{r tmb_plots, fig.width=8, fig.height=10}
# TMB by organ and tumor type
p1 <- ggplot(tmb_results, aes(x = organ, y = tmb, fill = tumor_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2),
              size = 2.5, alpha = 0.6) +
  scale_fill_manual(values = c("Primary" = "#E69F00", "Metastasis" = "#56B4E9")) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Tumor Mutational Burden by Organ and Tumor Type",
    x = "Organ",
    y = "TMB (Mutations per Mb)",
    fill = "Tumor Type"
  )

# TMB by location (for eye samples only)
eye_samples <- tmb_results %>% filter(organ == "eye")
p2 <- ggplot(eye_samples, aes(x = location_clean, y = tmb, fill = tumor_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2),
              size = 2.5, alpha = 0.6) +
  scale_fill_manual(values = c("Primary" = "#E69F00", "Metastasis" = "#56B4E9")) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "TMB by Anatomical Location (Eye Samples Only)",
    x = "Anatomical Location",
    y = "TMB (Mutations per Mb)",
    fill = "Tumor Type"
  )

# TMB by UV exposure
p3 <- ggplot(tmb_results, aes(x = uv_exposure, y = tmb, fill = uv_exposure)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2.5, alpha = 0.6) +
  scale_fill_manual(
    values = c("High_UV" = "#FF6B6B", "Medium_UV" = "#FFA500",
               "Low_UV" = "#4ECDC4", "Unknown" = "grey70")
  ) +
  theme_bw(base_size = 12) +
  labs(
    title = "TMB by UV Exposure Level",
    x = "UV Exposure",
    y = "TMB (Mutations per Mb)",
    fill = "UV Exposure"
  )

# Combine plots
p_combined <- p1 / p2 / p3
print(p_combined)

ggsave(
  file.path(output_dir, "TMB_by_organ_location_UV.png"),
  p_combined,
  width = 14,
  height = 14,
  dpi = 300,
  bg = "white"
)
```

### Statistical Tests

```{r tmb_stats}
# Compare TMB between UV exposure groups
if (sum(tmb_results$uv_exposure %in% c("High_UV", "Low_UV")) >= 6) {
  tmb_uv_test <- tmb_results %>%
    filter(uv_exposure %in% c("High_UV", "Low_UV"))

  if (nrow(tmb_uv_test) > 0) {
    wilcox_result <- wilcox.test(
      tmb ~ uv_exposure,
      data = tmb_uv_test
    )

    cat("\n=== Wilcoxon Test: High UV vs Low UV ===\n")
    cat("p-value:", format.pval(wilcox_result$p.value, digits = 3), "\n")
    cat("Significance:", ifelse(wilcox_result$p.value < 0.05, "YES", "NO"), "\n\n")

    # Summary statistics
    tmb_uv_test %>%
      group_by(uv_exposure) %>%
      summarise(
        n = n(),
        median_tmb = median(tmb),
        mean_tmb = mean(tmb),
        sd_tmb = sd(tmb),
        .groups = "drop"
      ) %>%
      kable(caption = "TMB Summary by UV Exposure", digits = 3) %>%
      kable_styling(bootstrap_options = c("striped", "hover"))
  }
}

# Compare primary vs metastasis
if (sum(tmb_results$tumor_type == "Primary") > 0 &&
    sum(tmb_results$tumor_type == "Metastasis") > 0) {
  wilcox_result_tumor <- wilcox.test(
    tmb ~ tumor_type,
    data = tmb_results
  )

  cat("\n=== Wilcoxon Test: Primary vs Metastasis ===\n")
  cat("p-value:", format.pval(wilcox_result_tumor$p.value, digits = 3), "\n")
  cat("Significance:", ifelse(wilcox_result_tumor$p.value < 0.05, "YES", "NO"), "\n")
}
```

## Mutation Signatures

### Prepare Mutation Matrix

```{r prepare_mutation_matrix}
# Filter PASS variants only and keep standard chromosomes
standard_chromosomes <- paste0("chr", c(1:22, "X", "Y", "M"))

pass_variants <- all_variants %>%
  filter(FILTER == "PASS", Chr %in% standard_chromosomes)

# Create GRanges object for MutationalPatterns
library(GenomicRanges)

# Get reference genome object first
ref_genome_obj <- get(ref_genome)

vcf_like_list <- pass_variants %>%
  split(.$sample) %>%
  map(function(df) {
    if (nrow(df) == 0) return(NULL)

    # Create GRanges with seqinfo from reference
    gr <- GRanges(
      seqnames = df$Chr,
      ranges = IRanges(start = df$Start, end = df$End),
      REF = df$Ref,
      ALT = df$Alt,
      seqinfo = seqinfo(ref_genome_obj)
    )

    return(gr)
  }) %>%
  compact()

# Get trinucleotide context and mutation matrix

mutation_matrices <- map(vcf_like_list, function(gr) {
  if (is.null(gr) || length(gr) == 0) return(NULL)

  # Get mutation matrix with trinucleotide context
  mut_mat <- mut_matrix(vcf_list = list(gr), ref_genome = ref_genome_obj)

  return(mut_mat)
})

# Combine into single matrix
mut_mat_combined <- do.call(cbind, mutation_matrices)
colnames(mut_mat_combined) <- names(mutation_matrices)

message("Mutation matrix created: ", nrow(mut_mat_combined), " mutation types × ",
        ncol(mut_mat_combined), " samples")
```

### Extract COSMIC Signatures

```{r extract_signatures, fig.width=14, fig.height=10}
# Load COSMIC v3.2 signatures (latest available in MutationalPatterns)
cosmic_signatures <- get_known_signatures(source = "COSMIC_v3.2")

# Extract signatures present in our data
fit_res <- fit_to_signatures(mut_mat_combined, cosmic_signatures)

# Get signature contributions
sig_contribution <- fit_res$contribution

# Focus on UV-related signatures (SBS7a, SBS7b, SBS7c, SBS7d)
uv_signatures <- c("SBS7a", "SBS7b", "SBS7c", "SBS7d")
uv_sig_present <- uv_signatures[uv_signatures %in% rownames(sig_contribution)]

if (length(uv_sig_present) > 0) {
  cat("\nUV-related signatures detected:\n")
  cat(paste(uv_sig_present, collapse = ", "), "\n\n")

  # Plot signature contributions
  plot_contribution(
    fit_res$contribution,
    cosmic_signatures,
    coord_flip = FALSE,
    mode = "relative"
  )

  ggsave(
    file.path(output_dir, "COSMIC_signature_contributions.png"),
    width = 14,
    height = 10,
    dpi = 300,
    bg = "white"
  )
}

# Save signature contribution data with all metadata
sig_contrib_df <- as.data.frame(sig_contribution) %>%
  rownames_to_column("Signature") %>%
  pivot_longer(-Signature, names_to = "sample", values_to = "contribution") %>%
  left_join(manifest %>% select(sample, origin, organ, location_clean, uv_exposure,
                                 Age_group, Age_numeric, Gender,
                                 primary_mutation, secondary_mutation), by = "sample")

write_csv(sig_contrib_df, file.path(output_dir, "signature_contributions.csv"))
```

### UV Signature Comparison

```{r uv_signature_comparison, fig.width=12, fig.height=6}
if (length(uv_sig_present) > 0) {
  # Filter UV signature contributions
  uv_sig_data <- sig_contrib_df %>%
    filter(Signature %in% uv_sig_present) %>%
    filter(!is.na(uv_exposure))

  # Plot UV signatures by exposure level
  p_uv <- ggplot(uv_sig_data, aes(x = uv_exposure, y = contribution, fill = Signature)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Signature, scales = "free_y") +
    scale_fill_manual(values = c(
      "SBS7a" = "#FF6B6B",
      "SBS7b" = "#FFA07A",
      "SBS7c" = "#FFD700",
      "SBS7d" = "#FF8C00"
    )) +
    theme_bw(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = "UV Damage Signatures by UV Exposure Level",
      subtitle = "SBS7 signatures are characteristic of UV-induced DNA damage",
      x = "UV Exposure",
      y = "Signature Contribution (mutations)",
      fill = "UV Signature"
    )

  print(p_uv)

  ggsave(
    file.path(output_dir, "UV_signatures_by_exposure.png"),
    p_uv,
    width = 12,
    height = 8,
    dpi = 300,
    bg = "white"
  )

  # Statistical test for UV signatures
  for (sig in uv_sig_present) {
    uv_test_data <- uv_sig_data %>%
      filter(Signature == sig, uv_exposure %in% c("High_UV", "Low_UV"))

    if (nrow(uv_test_data) >= 4) {
      test_result <- wilcox.test(contribution ~ uv_exposure, data = uv_test_data)

      cat("\n=== ", sig, ": High UV vs Low UV ===\n")
      cat("p-value:", format.pval(test_result$p.value, digits = 3), "\n")
      cat("Significance:", ifelse(test_result$p.value < 0.05, "YES", "NO"), "\n")
    }
  }
}
```

## Mutation Spectrum

### 96-Channel Mutation Spectrum

```{r mutation_spectrum, fig.width=14, fig.height=10}
# Plot mutation spectrum
plot_96_profile(mut_mat_combined, condensed = TRUE)

ggsave(
  file.path(output_dir, "mutation_spectrum_96.png"),
  width = 14,
  height = 10,
  dpi = 300,
  bg = "white"
)
```

### Spectrum by Sample Group

```{r spectrum_by_group, fig.width=14, fig.height=12}
# Group samples by UV exposure
samples_high_uv <- manifest %>%
  filter(uv_exposure == "High_UV") %>%
  pull(sample)

samples_low_uv <- manifest %>%
  filter(uv_exposure == "Low_UV") %>%
  pull(sample)

# Create grouped spectra
if (length(samples_high_uv) > 0 && length(samples_low_uv) > 0) {
  # Average spectra per group
  mut_mat_high_uv <- mut_mat_combined[, colnames(mut_mat_combined) %in% samples_high_uv, drop = FALSE]
  mut_mat_low_uv <- mut_mat_combined[, colnames(mut_mat_combined) %in% samples_low_uv, drop = FALSE]

  if (ncol(mut_mat_high_uv) > 0) {
    mut_mat_high_avg <- rowMeans(mut_mat_high_uv)
  }

  if (ncol(mut_mat_low_uv) > 0) {
    mut_mat_low_avg <- rowMeans(mut_mat_low_uv)
  }

  if (exists("mut_mat_high_avg") && exists("mut_mat_low_avg")) {
    mut_mat_grouped <- cbind(
      "High_UV_Iris" = mut_mat_high_avg,
      "Low_UV_Choroid" = mut_mat_low_avg
    )

    plot_96_profile(mut_mat_grouped, condensed = TRUE)

    ggsave(
      file.path(output_dir, "mutation_spectrum_UV_comparison.png"),
      width = 14,
      height = 8,
      dpi = 300,
      bg = "white"
    )
  }
}
```

### Cosine Similarity

```{r cosine_similarity, fig.width=10, fig.height=10}
# Calculate cosine similarity between samples
cos_sim_samples <- cos_sim_matrix(mut_mat_combined, mut_mat_combined)

# Heatmap
library(pheatmap)

pheatmap(
  cos_sim_samples,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = viridis(100),
  main = "Cosine Similarity Between Sample Mutation Profiles",
  fontsize = 8
)

ggsave(
  file.path(output_dir, "cosine_similarity_samples.png"),
  width = 12,
  height = 12,
  dpi = 300,
  bg = "white"
)
```

## Summary Statistics

### Overall Summary

```{r summary_stats}
# Overall statistics
summary_stats <- list(
  total_samples = nrow(manifest),
  total_variants = nrow(all_variants),
  pass_variants = sum(all_variants$FILTER == "PASS"),
  median_tmb = median(tmb_results$tmb),
  mean_tmb = mean(tmb_results$tmb)
)

cat("\n=== SComatic Analysis Summary ===\n")
cat("Total Samples:", summary_stats$total_samples, "\n")
cat("Total Variants:", format(summary_stats$total_variants, big.mark = ","), "\n")
cat("PASS Variants:", format(summary_stats$pass_variants, big.mark = ","), "\n")
cat("Median TMB:", round(summary_stats$median_tmb, 2), "mut/Mb\n")
cat("Mean TMB:", round(summary_stats$mean_tmb, 2), "mut/Mb\n\n")

# Sample distribution
cat("=== Sample Distribution by Organ ===\n")
manifest %>%
  group_by(organ, tumor_type) %>%
  summarise(n = n(), total_cells = sum(n_cells), .groups = "drop") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\n=== Sample Distribution by Location (Eye Samples) ===\n")
manifest %>%
  filter(organ == "eye") %>%
  group_by(tumor_type, location_clean) %>%
  summarise(n = n(), .groups = "drop") %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Top Mutated Samples

```{r top_samples}
tmb_results %>%
  arrange(desc(n_mutations)) %>%
  select(sample, tumor_type, organ, location_clean, uv_exposure, n_mutations, tmb) %>%
  head(20) %>%
  kable(
    caption = "Top 20 Samples by Mutation Count",
    digits = 2,
    col.names = c("Sample", "Type", "Organ", "Location", "UV", "Mutations", "TMB")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Missing Samples Documentation {.tabset}

### Samples Missing from SComatic Analysis

The following samples from the UM Atlas could not be processed through SComatic due to missing CRAM alignment files from the o37319_CellRangerCount run:

```{r missing_samples}
# Document missing samples - especially important for iris (UV analysis)
missing_samples <- data.frame(
  Sample = c("pUM14", "pUM15", "pUM15_16S", "pUM17_16S"),
  Cells = c(575, 8685, 8012, 1876),
  Location = c("Iris", "Iris", "Iris", "Iris"),
  UV_Exposure = c("High_UV", "High_UV", "High_UV", "High_UV"),
  Reason = c(
    "CRAM file removed from o37319",
    "CRAM file removed from o37319",
    "CRAM file removed from o37319",
    "Step 4 processing incomplete"
  ),
  Impact = c(
    "Missing from UV signature analysis",
    "Missing from UV signature analysis",
    "Missing from UV signature analysis",
    "Missing from UV signature analysis"
  ),
  stringsAsFactors = FALSE
)

missing_samples %>%
  kable(
    caption = "Missing Iris Samples - Critical for UV Damage Signature Analysis",
    col.names = c("Sample", "Cells", "Location", "UV Exposure", "Reason", "Impact")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(1:4, background = "#fff3cd")

cat("\n=== Impact Summary ===\n")
cat("Total missing iris cells:", sum(missing_samples$Cells), "\n")
cat("Available iris samples: pUM19, pUM21 only\n")
cat("UV analysis limited to ~30% of available iris cells\n")
```

### Available vs Missing Sample Summary

```{r sample_summary}
# Summary of iris samples
iris_summary <- data.frame(
  Status = c("Analyzed", "Missing"),
  n_Samples = c(2, 4),
  Sample_IDs = c("pUM19, pUM21", "pUM14, pUM15, pUM15_16S, pUM17_16S"),
  Total_Cells = c(9336, 19148),
  stringsAsFactors = FALSE
)

iris_summary %>%
  kable(
    caption = "Iris Sample Availability for UV Signature Analysis",
    col.names = c("Status", "# Samples", "Sample IDs", "Total Cells")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

cat("\nNote: CRAM files for missing samples were removed from o37319_CellRangerCount run.\n")
cat("Recovery would require re-running CellRanger alignment from FASTQ files.\n")
```

## Other COSMIC Signatures Analysis {.tabset}

Beyond UV damage (SBS7), analyze other biologically relevant COSMIC signatures.

### Top Non-UV Signatures

```{r other_signatures_analysis, fig.width=12, fig.height=6}
# Define signature categories of interest
sig_categories <- list(
  "Aging" = c("SBS1", "SBS5", "SBS40"),
  "APOBEC" = c("SBS2", "SBS13"),
  "DNA_Repair_Deficiency" = c("SBS3", "SBS6", "SBS15", "SBS26"),
  "Reactive_Oxygen" = c("SBS18", "SBS36"),
  "UV_Damage" = c("SBS7a", "SBS7b", "SBS7c", "SBS7d")
)

# COSMIC signature descriptions for plot titles
sig_descriptions <- c(
  "SBS1" = "5-methylcytosine deamination (aging/clock-like)",
  "SBS2" = "APOBEC cytidine deaminase",
  "SBS3" = "Homologous recombination deficiency (BRCA1/2)",
  "SBS4" = "Tobacco smoking",
  "SBS5" = "Clock-like (unknown etiology, aging)",
  "SBS6" = "Defective DNA mismatch repair",
  "SBS7a" = "UV light exposure",
  "SBS7b" = "UV light exposure",
  "SBS7c" = "UV light exposure",
  "SBS7d" = "UV light exposure",
  "SBS8" = "Unknown (late replication errors)",
  "SBS9" = "Polymerase eta somatic hypermutation",
  "SBS10a" = "Polymerase epsilon exonuclease domain",
  "SBS10b" = "Polymerase epsilon exonuclease domain",
  "SBS11" = "Temozolomide treatment",
  "SBS12" = "Unknown",
  "SBS13" = "APOBEC cytidine deaminase",
  "SBS14" = "Concurrent POLE mutation and MMR deficiency",
  "SBS15" = "Defective DNA mismatch repair",
  "SBS16" = "Unknown",
  "SBS17a" = "Unknown (5-FU treatment possible)",
  "SBS17b" = "Unknown (5-FU treatment possible)",
  "SBS18" = "Reactive oxygen species damage",
  "SBS19" = "Unknown",
  "SBS20" = "POLD1 exonuclease domain mutations + MMR deficiency",
  "SBS21" = "Defective DNA mismatch repair",
  "SBS22" = "Aristolochic acid exposure",
  "SBS23" = "Unknown",
  "SBS24" = "Aflatoxin exposure",
  "SBS25" = "Chemotherapy treatment",
  "SBS26" = "Defective DNA mismatch repair",
  "SBS28" = "Unknown",
  "SBS29" = "Tobacco chewing",
  "SBS30" = "Defective DNA base excision repair (NTHL1)",
  "SBS31" = "Platinum chemotherapy",
  "SBS32" = "Azathioprine treatment",
  "SBS33" = "Unknown",
  "SBS34" = "Unknown",
  "SBS35" = "Platinum chemotherapy",
  "SBS36" = "Defective DNA base excision repair (MUTYH)",
  "SBS37" = "Unknown",
  "SBS38" = "Indirect UV exposure",
  "SBS39" = "Unknown",
  "SBS40" = "Clock-like (unknown etiology, aging)",
  "SBS41" = "Unknown",
  "SBS42" = "Occupational haloalkane exposure",
  "SBS44" = "Defective DNA mismatch repair"
)

# Function to get signature description
get_sig_desc <- function(sig) {
  if (sig %in% names(sig_descriptions)) {
    return(paste0(sig, " - ", sig_descriptions[sig]))
  }
  return(sig)
}

# Calculate mean contribution for each signature across all samples
sig_means <- sig_contrib_df %>%
  group_by(Signature) %>%
  summarise(
    mean_contribution = mean(contribution, na.rm = TRUE),
    sd_contribution = sd(contribution, na.rm = TRUE),
    n_samples_positive = sum(contribution > 0),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_contribution))

cat("=== Top 20 COSMIC Signatures by Mean Contribution ===\n\n")
print(head(sig_means, 20))

# Bar plot of top signatures
p_top_sigs <- sig_means %>%
  head(15) %>%
  ggplot(aes(x = reorder(Signature, mean_contribution), y = mean_contribution, fill = mean_contribution)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_contribution - sd_contribution,
                    ymax = mean_contribution + sd_contribution),
                width = 0.3) +
  scale_fill_viridis_c(option = "plasma") +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    title = "Top 15 COSMIC Signatures in UM Cohort",
    subtitle = "Mean contribution across all samples",
    x = "COSMIC Signature",
    y = "Mean Mutation Contribution",
    fill = "Contribution"
  )

print(p_top_sigs)

ggsave(
  file.path(output_dir, "top_15_cosmic_signatures.png"),
  p_top_sigs, width = 10, height = 8, dpi = 300, bg = "white"
)
```

### Aging Signatures (SBS1, SBS5)

```{r aging_signatures, fig.width=12, fig.height=8}
aging_sigs <- c("SBS1", "SBS5", "SBS40")
aging_present <- intersect(aging_sigs, unique(sig_contrib_df$Signature))

if (length(aging_present) > 0) {
  aging_data <- sig_contrib_df %>%
    filter(Signature %in% aging_present)

  # 1. By tumor type
  p_aging_type <- ggplot(aging_data, aes(x = origin, y = contribution, fill = origin)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Signature, scales = "free_y") +
    scale_fill_manual(values = c("primary" = "#E41A1C", "metastasis" = "#377EB8")) +
    theme_classic(base_size = 12) +
    labs(
      title = "Aging-Related Signatures by Tumor Origin",
      subtitle = "SBS1/SBS5 = clock-like signatures associated with cell division",
      x = "Origin",
      y = "Signature Contribution",
      fill = "Origin"
    )
  print(p_aging_type)

  # 2. By AGE CLASS - KEY NEW ANALYSIS
  age_colors <- c(
    "Young (<40)" = "#2E91E5",
    "Middle (40-60)" = "#FFCE1F",
    "Old (60+)" = "#E15759",
    "Unknown" = "#CCCCCC"
  )

  # Filter to samples with known age
  aging_data_with_age <- aging_data %>%
    filter(Age_group != "Unknown")

  if (nrow(aging_data_with_age) > 0) {
    p_aging_by_age <- ggplot(aging_data_with_age,
                              aes(x = Age_group, y = contribution, fill = Age_group)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.2, alpha = 0.4, size = 1.5) +
      facet_wrap(~Signature, scales = "free_y") +
      scale_fill_manual(values = age_colors) +
      scale_x_discrete(limits = c("Young (<40)", "Middle (40-60)", "Old (60+)")) +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(
        title = "Aging Signatures by Patient Age Class",
        subtitle = "SBS1/SBS5/SBS40 should increase with patient age if clock-like",
        x = "Age Group",
        y = "Signature Contribution",
        fill = "Age Group"
      )
    print(p_aging_by_age)

    # Statistical test for aging signatures vs age
    cat("\n=== Statistical Analysis: Aging Signatures vs Age ===\n")
    for (sig in aging_present) {
      sig_data <- aging_data_with_age %>% filter(Signature == sig)
      if (length(unique(sig_data$Age_group)) >= 2 && sum(!is.na(sig_data$Age_numeric)) > 5) {
        # Correlation with numeric age
        cor_test <- cor.test(sig_data$contribution, sig_data$Age_numeric,
                            method = "spearman", use = "complete.obs")
        cat("\n", sig, ":\n")
        cat("  Spearman correlation with age: rho =", round(cor_test$estimate, 3),
            ", p =", format.pval(cor_test$p.value, digits = 3), "\n")

        # Kruskal-Wallis test for age groups
        kw_test <- kruskal.test(contribution ~ Age_group, data = sig_data)
        cat("  Kruskal-Wallis across age groups: p =", format.pval(kw_test$p.value, digits = 3), "\n")
      }
    }

    ggsave(
      file.path(output_dir, "aging_signatures_by_age_class.png"),
      p_aging_by_age, width = 12, height = 8, dpi = 300, bg = "white"
    )
  } else {
    cat("Note: No samples with age data available for age-class analysis\n")
  }

  # 3. By organ for metastases
  p_aging_organ <- aging_data %>%
    filter(origin == "metastasis") %>%
    ggplot(aes(x = organ, y = contribution, fill = organ)) +
    geom_boxplot(alpha = 0.7) +
    facet_wrap(~Signature, scales = "free_y") +
    theme_classic(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = "Aging Signatures in Metastases by Organ",
      x = "Organ",
      y = "Signature Contribution",
      fill = "Organ"
    )
  print(p_aging_organ)

  ggsave(
    file.path(output_dir, "aging_signatures_analysis.png"),
    p_aging_type, width = 12, height = 6, dpi = 300, bg = "white"
  )
}
```

### APOBEC Signatures (SBS2, SBS13)

```{r apobec_signatures, fig.width=10, fig.height=5}
apobec_sigs <- c("SBS2", "SBS13")
apobec_present <- intersect(apobec_sigs, unique(sig_contrib_df$Signature))

if (length(apobec_present) > 0) {
  apobec_data <- sig_contrib_df %>%
    filter(Signature %in% apobec_present)

  p_apobec <- ggplot(apobec_data, aes(x = origin, y = contribution, fill = Signature)) +
    geom_boxplot(alpha = 0.7) +
    scale_fill_manual(values = c("SBS2" = "#FF6B6B", "SBS13" = "#4ECDC4")) +
    theme_classic(base_size = 14) +
    labs(
      title = "APOBEC Mutagenesis Signatures",
      subtitle = "SBS2/SBS13 = APOBEC cytidine deaminase activity (common in metastatic tumors)",
      x = "Tumor Origin",
      y = "Signature Contribution",
      fill = "Signature"
    )
  print(p_apobec)

  # Statistical comparison
  for (sig in apobec_present) {
    test_data <- apobec_data %>%
      filter(Signature == sig)

    if (length(unique(test_data$origin)) >= 2) {
      test <- wilcox.test(contribution ~ origin, data = test_data)
      cat("\n", sig, "- Primary vs Metastasis:\n")
      cat("p-value:", format.pval(test$p.value, digits = 3), "\n")
    }
  }

  ggsave(
    file.path(output_dir, "apobec_signatures_analysis.png"),
    p_apobec, width = 10, height = 6, dpi = 300, bg = "white"
  )
}
```

### Top Non-UV Signatures by All Conditions

Comprehensive analysis of top non-UV signatures across all metadata conditions.

```{r top_signatures_all_conditions, fig.width=14, fig.height=10}
# Identify top 10 non-UV signatures
uv_signatures <- c("SBS7a", "SBS7b", "SBS7c", "SBS7d")
top_non_uv <- sig_means %>%
  filter(!Signature %in% uv_signatures) %>%
  head(10) %>%
  pull(Signature)

cat("Top 10 Non-UV Signatures:", paste(top_non_uv, collapse = ", "), "\n\n")

# Create comprehensive comparison plots
for (sig in top_non_uv) {
  sig_data <- sig_contrib_df %>%
    filter(Signature == sig)

  plots <- list()

  # 1. By Origin
  plots$origin <- ggplot(sig_data, aes(x = origin, y = contribution, fill = origin)) +
    geom_boxplot(alpha = 0.7, outlier.shape = 21) +
    geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
    scale_fill_manual(values = c("primary" = "#E41A1C", "metastasis" = "#377EB8")) +
    theme_classic(base_size = 10) +
    labs(title = "By Origin", x = "", y = "Contribution") +
    theme(legend.position = "none")

  # 2. By Organ
  plots$organ <- ggplot(sig_data, aes(x = organ, y = contribution, fill = organ)) +
    geom_boxplot(alpha = 0.7, outlier.shape = 21) +
    geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
    theme_classic(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(title = "By Organ", x = "", y = "Contribution")

  # 3. By Age Group (if available)
  sig_data_age <- sig_data %>% filter(Age_group != "Unknown")
  if (nrow(sig_data_age) > 0) {
    age_colors <- c("Young (<40)" = "#2E91E5", "Middle (40-60)" = "#FFCE1F",
                    "Old (60+)" = "#E15759")
    plots$age <- ggplot(sig_data_age, aes(x = Age_group, y = contribution, fill = Age_group)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
      scale_fill_manual(values = age_colors) +
      scale_x_discrete(limits = c("Young (<40)", "Middle (40-60)", "Old (60+)")) +
      theme_classic(base_size = 10) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none") +
      labs(title = "By Age Group", x = "", y = "Contribution")
  } else {
    plots$age <- ggplot() + theme_void() + labs(title = "By Age Group (No data)")
  }

  # 4. By Gender (if available)
  sig_data_gender <- sig_data %>% filter(Gender != "Unknown" & !is.na(Gender))
  if (nrow(sig_data_gender) > 0) {
    plots$gender <- ggplot(sig_data_gender, aes(x = Gender, y = contribution, fill = Gender)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
      scale_fill_manual(values = c("M" = "#4E79A7", "F" = "#E15759")) +
      theme_classic(base_size = 10) +
      theme(legend.position = "none") +
      labs(title = "By Gender", x = "", y = "Contribution")
  } else {
    plots$gender <- ggplot() + theme_void() + labs(title = "By Gender (No data)")
  }

  # 5. By UV Exposure (for primary eye samples)
  sig_data_uv <- sig_data %>% filter(uv_exposure != "Unknown")
  if (nrow(sig_data_uv) > 0) {
    plots$uv <- ggplot(sig_data_uv, aes(x = uv_exposure, y = contribution, fill = uv_exposure)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
      scale_fill_manual(values = c("High_UV" = "#FF6B6B", "Low_UV" = "#4ECDC4", "Medium_UV" = "#FFD93D")) +
      theme_classic(base_size = 10) +
      theme(legend.position = "none") +
      labs(title = "By UV Exposure", x = "", y = "Contribution")
  } else {
    plots$uv <- ggplot() + theme_void() + labs(title = "By UV Exposure (No data)")
  }

  # 6. By Primary Mutation
  sig_data_pmut <- sig_data %>% filter(!is.na(primary_mutation) & primary_mutation != "none")
  if (nrow(sig_data_pmut) > 0) {
    plots$pmut <- ggplot(sig_data_pmut, aes(x = primary_mutation, y = contribution, fill = primary_mutation)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
      theme_classic(base_size = 10) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none") +
      labs(title = "By Primary Mutation", x = "", y = "Contribution")
  } else {
    plots$pmut <- ggplot() + theme_void() + labs(title = "By Primary Mutation (No data)")
  }

  # 7. By Secondary Mutation
  sig_data_smut <- sig_data %>% filter(!is.na(secondary_mutation) & secondary_mutation != "none" & secondary_mutation != "")
  if (nrow(sig_data_smut) > 0) {
    plots$smut <- ggplot(sig_data_smut, aes(x = secondary_mutation, y = contribution, fill = secondary_mutation)) +
      geom_boxplot(alpha = 0.7, outlier.shape = 21) +
      geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
      theme_classic(base_size = 10) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none") +
      labs(title = "By Secondary Mutation", x = "", y = "Contribution")
  } else {
    plots$smut <- ggplot() + theme_void() + labs(title = "By Secondary Mutation (No data)")
  }

  # Combine plots with signature description in title
  sig_title <- get_sig_desc(sig)
  combined <- wrap_plots(plots, ncol = 3) +
    plot_annotation(
      title = sig_title,
      subtitle = "Distribution across all metadata categories"
    )

  print(combined)
}

# Save summary figure for top 5 signatures
cat("\nSaving comprehensive comparison figure...\n")
```

### Signature Summary by Sample Group

```{r signature_heatmap_extended, fig.width=14, fig.height=12}
# Create signature heatmap for top 20 signatures
top_20_sigs <- sig_means$Signature[1:min(20, nrow(sig_means))]

sig_matrix <- sig_contrib_df %>%
  filter(Signature %in% top_20_sigs) %>%
  pivot_wider(id_cols = sample, names_from = Signature, values_from = contribution) %>%
  column_to_rownames("sample")

# Add sample annotations - include ALL metadata
sample_annot <- manifest %>%
  select(sample, origin, organ, uv_exposure, Age_group, Gender) %>%
  filter(sample %in% rownames(sig_matrix)) %>%
  mutate(
    Age_group = factor(Age_group, levels = c("Young (<40)", "Middle (40-60)", "Old (60+)", "Unknown")),
    Gender = factor(Gender, levels = c("M", "F", "Unknown"))
  ) %>%
  column_to_rownames("sample")

# Define annotation colors
annot_colors <- list(
  origin = c("primary" = "#E41A1C", "metastasis" = "#377EB8"),
  organ = c("eye" = "#E8ECFB", "liver" = "#DDA0DD", "brain" = "#FDB462",
            "skin" = "#B3DE69", "thyroid" = "#FCCDE5"),
  uv_exposure = c("High_UV" = "#FF6B6B", "Low_UV" = "#4ECDC4",
                   "Medium_UV" = "#FFD93D", "Unknown" = "#CCCCCC"),
  Age_group = c("Young (<40)" = "#2E91E5", "Middle (40-60)" = "#FFCE1F",
                "Old (60+)" = "#E15759", "Unknown" = "#CCCCCC"),
  Gender = c("M" = "#4E79A7", "F" = "#E15759", "Unknown" = "#CCCCCC")
)

# Create heatmap with pheatmap
library(pheatmap)

pheatmap(
  t(sig_matrix),
  annotation_col = sample_annot,
  annotation_colors = annot_colors,
  scale = "row",
  clustering_method = "ward.D2",
  show_colnames = TRUE,
  fontsize = 8,
  main = "Top 20 COSMIC Signatures Across UM Cohort\n(with Age, Gender, Origin, Organ, UV Exposure)",
  filename = file.path(output_dir, "heatmap_top20_signatures.png"),
  width = 14,
  height = 12
)

# Also display inline
pheatmap(
  t(sig_matrix),
  annotation_col = sample_annot,
  annotation_colors = annot_colors,
  scale = "row",
  clustering_method = "ward.D2",
  show_colnames = TRUE,
  fontsize = 8,
  main = "Top 20 COSMIC Signatures Across UM Cohort\n(with Age, Gender, Origin, Organ, UV Exposure)"
)

# Print sample summary by metadata
cat("\n=== Sample Summary in Heatmap ===\n")
cat("Total samples:", nrow(sig_matrix), "\n")
cat("\nBy Age Group:\n")
print(table(sample_annot$Age_group))
cat("\nBy Gender:\n")
print(table(sample_annot$Gender))
cat("\nBy Origin:\n")
print(table(sample_annot$origin))
```

### Biological Interpretation

```{r signature_interpretation}
cat("=== COSMIC Signature Interpretation for Uveal Melanoma ===\n\n")

cat("1. UV DAMAGE SIGNATURES (SBS7a/b/c/d)\n")
cat("   - Expected to be elevated in IRIS samples (UV-exposed)\n")
cat("   - Should be low/absent in CHOROID samples (UV-protected)\n")
cat("   - Critical for understanding UM etiology differences by location\n\n")

cat("2. AGING SIGNATURES (SBS1, SBS5)\n")
cat("   - Clock-like signatures accumulating with age\n")
cat("   - SBS1: deamination of 5-methylcytosine\n")
cat("   - SBS5: unknown etiology, associated with age\n")
cat("   - Expected in all samples, may correlate with patient age\n\n")

cat("3. APOBEC SIGNATURES (SBS2, SBS13)\n")
cat("   - APOBEC cytidine deaminases\n")
cat("   - Often elevated in metastatic disease\n")
cat("   - Associated with immune response and viral infection\n\n")

cat("4. DNA REPAIR DEFICIENCY (SBS3, SBS6, SBS15)\n")
cat("   - SBS3: BRCA1/2 deficiency (HR defect)\n")
cat("   - SBS6: MMR deficiency\n")
cat("   - May indicate therapeutic vulnerability to PARP inhibitors\n\n")

# Export signature summary
write_csv(sig_means, file.path(output_dir, "signature_summary_ranked.csv"))
cat("Signature summary exported to: signature_summary_ranked.csv\n")
```

### Session Info

```{r session_info}
sessionInfo()
```
