---
title: "UM Atlas - scSLIDE Sample-Level Trajectory Analysis (Malignant Cells)"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Extended Data (scSLIDE malignant cell trajectory inference)
```

**scSLIDE (sample-level Integration using Density Estimation) analysis** of uveal melanoma progression using malignant cells only. This analysis complements the existing GloScope analysis by using supervised learning to model disease trajectories.

# {.tabset}

## Overview

Started on `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`

### What is scSLIDE?

**scSLIDE** performs sample-level analysis of multi-sample scRNA-seq data by:

1. **Semi-supervised dimensional reduction**: Combines cell-type embedding with disease-oriented PLS (Partial Least Squares) regression
2. **Sample-level representation**: Each sample is represented as a probability distribution of cellular states
3. **Trajectory inference**: Uses diffusion maps to model continuous disease progression
4. **Supervised learning**: Leverages known phenotypes (disease status, mutations, organs) to capture relevant variation

### Analysis Scope

This report includes **three complementary trajectory analyses** on **malignant cells only** (~150-200K cells):

1. **Disease Progression**: healthy → primary → metastasis trajectory
2. **Mutation-Stratified**: BAP1 vs SF3B1 vs EIF1AX trajectories
3. **Organ-Specific Metastasis**: liver vs brain vs skin vs thyroid

Each analysis uses the same malignant cell subset but different grouping variables for supervision.

### Comparison: GloScope vs scSLIDE

| Feature | GloScope | scSLIDE |
|---------|----------|---------|
| **Approach** | Unsupervised (PCA + KL divergence) | Semi-supervised (PCA + PLS + WNN) |
| **Supervision** | None | Uses phenotype labels |
| **Visualization** | MDS scatter plots | Diffusion maps |
| **Strength** | Captures all variation | Captures disease-relevant variation |
| **Gene discovery** | Indirect (PCA loadings) | Direct (trajectory DEGs) |
| **Trajectory** | Not supported | ✅ Continuous trajectories |

**Both methods are complementary**: GloScope answers "Are samples different?", scSLIDE answers "Why are they different?".

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 8)

# Load libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(scSLIDE)
  library(qs)
  library(qs2)
  library(tidyverse)
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(BiocParallel)
  library(pls)
  library(ComplexHeatmap)
  library(circlize)
  library(vegan)
  library(DT)
})

# Patch scSLIDE::FindmmNN for Seurat 5.4+ compatibility
# Seurat 5.4.0.9000 removed fix.wnn.weights from FindModalityWeights
patched_FindmmNN <- function(object, sketch.assay = NULL, reduction.list, dims.list,
                              k.nn = 20, knn.range = 200, l2.norm = TRUE,
                              fix.wnn.weights = NULL,
                              weighted.nn.name = "weighted.nn", verbose = TRUE, ...) {
  FindModalityWeights <- getFromNamespace("FindModalityWeights", "Seurat")
  MultiModalNN <- getFromNamespace("MultiModalNN", "Seurat")
  if (k.nn <= 2) stop("k.nn must be a integer larger than 1.")
  wnn.weight <- FindModalityWeights(
    object = object, reduction.list = reduction.list,
    dims.list = dims.list, snn.far.nn = FALSE, l2.norm = l2.norm,
    k.nn = ifelse(20 >= k.nn, 20, k.nn), ...
  )
  if (is.null(sketch.assay) || !sketch.assay %in% names(object@assays)) {
    stop("sketch.assay is not correctly defined. please check.")
  }
  sketch_cells <- Cells(object[[sketch.assay]])
  if (knn.range > length(sketch_cells)) {
    knn.range <- length(sketch_cells)
    warning("knn.range exceeds landmarks. Resetting to ", length(sketch_cells))
  }
  sketch_obj <- subset(object, cells = sketch_cells)
  weighted.nn <- MultiModalNN(
    object = sketch_obj, query = object,
    k.nn = ifelse(20 >= k.nn, 20, k.nn), modality.weight = wnn.weight,
    knn.range = knn.range, verbose = TRUE, l2.norm = l2.norm, ...
  )
  object[[weighted.nn.name]] <- weighted.nn
  return(object)
}
assignInNamespace("FindmmNN", patched_FindmmNN, ns = "scSLIDE")

# Load Victoria's official color palettes
source("colors.R")

# Set up parallel processing (16 cores)
bpparam <- MulticoreParam(workers = 16, progressbar = TRUE)
register(bpparam)

# Create output directory
output_dir <- "../scSLIDE_Outputs/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Set random seed for reproducibility
set.seed(42)

cat("Parallel processing: 16 cores\n")
cat("Output directory:", output_dir, "\n")
cat("scSLIDE version:", as.character(packageVersion("scSLIDE")), "\n")
```

---

## Data Loading {.tabset}

### Load Full Dataset

```{r load_full_data}
cat("Loading Victoria's UM Atlas (Dec 2025)...\n")
um_atlas <- qs::qread(
  '../13_final_Dataset_reclustered.qs',
  nthreads = 16
)

cat("Full dataset:\n")
cat("  - Cells:", ncol(um_atlas), "\n")
cat("  - Features:", nrow(um_atlas), "\n")
cat("  - Samples:", length(unique(um_atlas$Sample)), "\n")
cat("  - Patients:", length(unique(um_atlas$Patient_nr)), "\n")

# Check available metadata
cat("\nAvailable metadata columns:\n")
print(head(colnames(um_atlas@meta.data)))
```

### Filter Malignant Cells

```{r filter_malignant}
cat("Filtering malignant cells only...\n")

# Filter to malignant cells (note: capitalized "Malignant")
um_malignant <- subset(um_atlas, subset = Malignant == "Malignant")

cat("\nMalignant subset:\n")
cat("  - Cells:", ncol(um_malignant), "\n")
cat("  - Samples:", length(unique(um_malignant$Sample)), "\n")
cat("  - Patients:", length(unique(um_malignant$Patient_nr)), "\n")

# Distribution by origin
origin_counts <- table(um_malignant$origin)
cat("\nCells by disease status:\n")
print(origin_counts)

# Samples by origin
sample_meta <- um_malignant@meta.data %>%
  group_by(Sample, origin, organ, primary_mutation, secondary_mutation) %>%
  summarise(n_cells = n(), .groups = "drop") %>%
  arrange(origin, organ)

cat("\nSamples per group:\n")
cat("  - Healthy:", sum(sample_meta$origin == "healthy"), "samples\n")
cat("  - Primary:", sum(sample_meta$origin == "primary"), "samples\n")
cat("  - Metastasis:", sum(sample_meta$origin == "metastasis"), "samples\n")

# Save malignant subset for future use
cat("\nSaving malignant subset...\n")
qs2::qs_save(
  um_malignant,
  file.path("..",
            "scSLIDE_malignant_subset.qs2"),
  nthreads = 16
)
cat("✓ Saved to: scSLIDE_malignant_subset.qs2\n")
```

### Sample Metadata Summary

```{r sample_metadata_table}
# Create sample-level summary table
sample_summary <- um_malignant@meta.data %>%
  group_by(Sample) %>%
  summarise(
    Patient = first(Patient_nr),
    Origin = first(origin),
    Organ = first(organ),
    Primary_Mutation = first(primary_mutation),
    Secondary_Mutation = first(secondary_mutation),
    Location = first(LocationPrimaryTumor),
    Gender = first(Gender),
    Age = first(Age_atDiagnosisY),
    N_Cells = n(),
    .groups = "drop"
  ) %>%
  arrange(Origin, Organ, Sample)

# Display interactive table
datatable(
  sample_summary,
  caption = "Sample-level metadata (malignant cells only)",
  options = list(
    pageLength = 15,
    scrollX = TRUE,
    columnDefs = list(list(className = 'dt-center', targets = "_all"))
  ),
  rownames = FALSE
)

cat("\nTotal samples with malignant cells:", nrow(sample_summary), "\n")
```

### Check Required Reductions

```{r check_reductions}
cat("Available dimensionality reductions:\n")
print(names(um_malignant@reductions))

# Ensure PCA exists (required for scSLIDE)
if (!"pca" %in% names(um_malignant@reductions)) {
  cat("\nComputing PCA (required for scSLIDE)...\n")
  um_malignant <- FindVariableFeatures(um_malignant, nfeatures = 3000)
  um_malignant <- ScaleData(um_malignant)
  um_malignant <- RunPCA(um_malignant, npcs = 50, verbose = FALSE)
  cat("✓ PCA computed (50 PCs)\n")
} else {
  cat("\n✓ PCA reduction already present\n")
  cat("  - PCs available:", ncol(Embeddings(um_malignant, "pca")), "\n")
}
```

---

## Disease Progression Trajectory {.tabset}

### scSLIDE Analysis

```{r scslide_origin, eval=TRUE}
cat("=== scSLIDE: Disease Progression Trajectory ===\n")
cat("Phenotype variable: origin (healthy → primary → metastasis)\n\n")

# Check for cached results
cache_origin <- file.path(output_dir, "origin_sample_object.rds")

if (file.exists(cache_origin)) {
  cat("Loading cached sample object...\n")
  sample_obj_origin <- readRDS(cache_origin)
  cat("  - Samples:", ncol(sample_obj_origin), "\n")
  cat("  - Features:", nrow(sample_obj_origin), "\n")

  # Check if DiffMap reduction exists
  if (!"DiffMap" %in% names(sample_obj_origin@reductions)) {
    cat("Computing diffusion map on cached object...\n")
    DefaultAssay(sample_obj_origin) <- "LMC"
    sample_obj_origin <- ScaleData(sample_obj_origin, features = rownames(sample_obj_origin), verbose = FALSE)
    sample_obj_origin <- RunDiffusionMap(
      object = sample_obj_origin, assay = "LMC", layer = "scale.data",
      features = rownames(sample_obj_origin), ncomp = 10
    )
    saveRDS(sample_obj_origin, cache_origin)
  }
  cat("Cached results loaded.\n")

} else {
  cat("Running PrepareSampleObject()...\n")
  cat("  - Landmark cells: 2000, PLS components: 15\n")
  cat("  - Phenotype: origin (disease status)\n\n")

  um_malignant_origin <- PrepareSampleObject(
    object = um_malignant,
    assay = "RNA",
    add.hvg = TRUE,
    group.by.CorTest = "majority_celltype",
    Y = "origin",
    sketch.training = TRUE,
    group.by.Sketch = "Sample",
    ncells.per.group = 1000,
    ncells.landmark = 2000,
    ncomp = 15,
    pls.function = "cppls",
    k.nn = 5,
    name.reduction.1 = "pca",
    dims.reduction.1 = 1:50,
    fix.wnn.weights = c(0.5, 0.5),
    verbose = TRUE
  )
  cat("\nPrepareSampleObject complete\n")

  cat("\nGenerating sample-level density matrix...\n")
  sample_obj_origin <- GenerateSampleObject(
    object = um_malignant_origin,
    sketch.assay = "LANDMARK",
    nn.name = "weighted.nn",
    group.by = "Sample",
    rename.group.by = "majority_celltype",
    k.nn = 5,
    normalization.method = "ChiSquared",
    add.meta.data = TRUE,
    return.seurat = TRUE,
    verbose = TRUE
  )
  cat("Sample object created:", ncol(sample_obj_origin), "samples,", nrow(sample_obj_origin), "features\n")

  # Scale + diffusion map
  DefaultAssay(sample_obj_origin) <- "LMC"
  sample_obj_origin <- ScaleData(sample_obj_origin, features = rownames(sample_obj_origin), verbose = FALSE)
  sample_obj_origin <- RunDiffusionMap(
    object = sample_obj_origin, assay = "LMC", layer = "scale.data",
    features = rownames(sample_obj_origin), ncomp = 10
  )
  cat("Diffusion map complete\n")

  # Cache
  saveRDS(sample_obj_origin, cache_origin)
  rm(um_malignant_origin); gc()
}

# Extract diffusion coordinates + metadata from sample object directly
# (column names differ from Sample metadata column, so we use metadata directly)
dc_coords_origin <- Embeddings(sample_obj_origin, reduction = "DiffMap")
sample_meta_origin <- sample_obj_origin@meta.data
dc_df_origin <- cbind(
  as.data.frame(dc_coords_origin),
  data.frame(
    Sample = sample_meta_origin$Sample,
    Patient = sample_meta_origin$Patient_nr,
    Origin = sample_meta_origin$origin,
    Organ = sample_meta_origin$organ,
    Primary_Mutation = sample_meta_origin$primary_mutation,
    Secondary_Mutation = sample_meta_origin$secondary_mutation,
    Location = sample_meta_origin$LocationPrimaryTumor,
    Gender = sample_meta_origin$Gender,
    stringsAsFactors = FALSE
  )
)

write.csv(dc_df_origin, file.path(output_dir, "origin_diffusion_coordinates.csv"),
          row.names = FALSE)

cat("\nDiffusion coordinates extracted for", nrow(dc_df_origin), "samples\n")
cat("Samples with Origin metadata:", sum(!is.na(dc_df_origin$Origin)), "\n")
```

### Diffusion Map Visualization

```{r plot_origin_diffusion}
# Prepare origin factor with correct order
dc_df_origin <- dc_df_origin %>%
  mutate(Origin = factor(Origin, levels = c("healthy", "primary", "metastasis")))

# Plot DC1 vs DC2 colored by origin
p1 <- ggplot(dc_df_origin, aes(x = DC_1, y = DC_2, color = Origin)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = origin_colors) +
  labs(
    title = "scSLIDE Disease Progression Trajectory",
    subtitle = "Diffusion map of malignant samples (colored by disease status)",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

print(p1)
ggsave(file.path(output_dir, "origin_DC1_DC2_by_origin.png"),
       p1, width = 10, height = 7, dpi = 300, bg = "white")

# Plot colored by organ
p2 <- ggplot(dc_df_origin, aes(x = DC_1, y = DC_2, color = Organ)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = organ_colors) +
  labs(
    title = "scSLIDE Disease Progression Trajectory",
    subtitle = "Diffusion map colored by organ",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

print(p2)
ggsave(file.path(output_dir, "origin_DC1_DC2_by_organ.png"),
       p2, width = 10, height = 7, dpi = 300, bg = "white")

# Plot colored by secondary mutation
p3 <- ggplot(dc_df_origin, aes(x = DC_1, y = DC_2, color = Secondary_Mutation)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = second_mutations_col) +
  labs(
    title = "scSLIDE Disease Progression Trajectory",
    subtitle = "Diffusion map colored by secondary mutation",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

print(p3)
ggsave(file.path(output_dir, "origin_DC1_DC2_by_secondary_mutation.png"),
       p3, width = 10, height = 7, dpi = 300, bg = "white")
```

### Statistical Testing

```{r origin_statistics}
cat("=== Statistical Testing: Disease Progression ===\n\n")

# Use dc_df_origin which has both coordinates and metadata aligned
# Filter to samples with valid metadata
dc_df_origin_valid <- dc_df_origin %>% dplyr::filter(!is.na(Origin))
cat("Samples with valid Origin metadata:", nrow(dc_df_origin_valid), "/", nrow(dc_df_origin), "\n")

# Extract coordinate columns from the joined dataframe (ensures alignment)
dc_cols <- grep("^DC_", colnames(dc_df_origin_valid), value = TRUE)
ncomp_test <- min(5, length(dc_cols))
cat("Using", ncomp_test, "diffusion components:", paste(dc_cols[1:ncomp_test], collapse = ", "), "\n")

# Ensure plain data.frame (not tibble) and proper alignment
stats_data <- as.data.frame(dc_df_origin_valid)
stats_coords <- as.matrix(stats_data[, dc_cols[1:ncomp_test]])
stats_group <- stats_data$Origin

cat("Diagnostic: nrow coords =", nrow(stats_coords), ", length group =", length(stats_group), "\n")
cat("Groups:", paste(table(stats_group), collapse = ", "), "\n")

# Compute distance matrix
dist_mat_origin <- dist(stats_coords, method = "euclidean")
cat("Distance matrix size:", attr(dist_mat_origin, "Size"), "\n")

# PERMANOVA test on diffusion coordinates
cat("\nPERMANOVA test (adonis2):\n")
cat("  - Test: Do disease groups differ in trajectory space?\n\n")

perm_origin <- tryCatch({
  adonis2(dist_mat_origin ~ stats_group, permutations = 999)
}, error = function(e) {
  cat("adonis2 error:", conditionMessage(e), "\n")
  NULL
})

if (!is.null(perm_origin)) print(perm_origin)

# ANOSIM test
cat("\nANOSIM test:\n")
anosim_origin <- tryCatch({
  anosim(dist_mat_origin, stats_group, permutations = 999)
}, error = function(e) {
  cat("ANOSIM error:", conditionMessage(e), "\n")
  NULL
})

if (!is.null(anosim_origin)) {
  cat("  - R statistic:", round(anosim_origin$statistic, 3), "\n")
  cat("  - p-value:", format.pval(anosim_origin$signif, digits = 3), "\n")
}

# Save statistics
stats_df <- data.frame(
  Test = c("PERMANOVA", "ANOSIM"),
  R_statistic = c(
    if (!is.null(perm_origin)) perm_origin$R2[1] else NA,
    if (!is.null(anosim_origin)) anosim_origin$statistic else NA
  ),
  p_value = c(
    if (!is.null(perm_origin)) perm_origin$`Pr(>F)`[1] else NA,
    if (!is.null(anosim_origin)) anosim_origin$signif else NA
  )
)

write.csv(stats_df, file.path(output_dir, "origin_statistics.csv"),
          row.names = FALSE)

cat("\n✓ Statistics saved to origin_statistics.csv\n")
```

---

## Mutation-Stratified Trajectories {.tabset}

### scSLIDE Analysis (Mutations)

```{r scslide_mutation, eval=TRUE}
cat("=== scSLIDE: Mutation-Stratified Trajectory ===\n")
cat("Grouping variable: secondary_mutation (BAP1/SF3B1/EIF1AX)\n\n")

# Filter samples with known secondary mutations
samples_with_mutations <- sample_summary %>%
  dplyr::filter(Secondary_Mutation %in% c("BAP1", "SF3B1", "EIF1AX")) %>%
  pull(Sample)

cat("Samples with secondary mutations:", length(samples_with_mutations), "\n")
cat("  - BAP1:", sum(sample_summary$Secondary_Mutation == "BAP1", na.rm = TRUE), "\n")
cat("  - SF3B1:", sum(sample_summary$Secondary_Mutation == "SF3B1", na.rm = TRUE), "\n")
cat("  - EIF1AX:", sum(sample_summary$Secondary_Mutation == "EIF1AX", na.rm = TRUE), "\n\n")

cache_mut <- file.path(output_dir, "mutation_sample_object.rds")

if (file.exists(cache_mut)) {
  cat("Loading cached mutation sample object...\n")
  sample_obj_mut <- readRDS(cache_mut)

  if (!"DiffMap" %in% names(sample_obj_mut@reductions)) {
    DefaultAssay(sample_obj_mut) <- "LMC"
    sample_obj_mut <- ScaleData(sample_obj_mut, features = rownames(sample_obj_mut), verbose = FALSE)
    sample_obj_mut <- RunDiffusionMap(
      object = sample_obj_mut, assay = "LMC", layer = "scale.data",
      features = rownames(sample_obj_mut), ncomp = 10
    )
    saveRDS(sample_obj_mut, cache_mut)
  }
  cat("Cached results loaded.\n")

} else {
  um_malignant_mut <- subset(um_malignant, subset = Sample %in% samples_with_mutations)
  cat("Cells in mutation subset:", ncol(um_malignant_mut), "\n\n")

  cat("Running PrepareSampleObject()...\n")
  um_malignant_mut <- PrepareSampleObject(
    object = um_malignant_mut,
    assay = "RNA", add.hvg = TRUE,
    group.by.CorTest = "majority_celltype",
    Y = "secondary_mutation",
    sketch.training = TRUE, group.by.Sketch = "Sample",
    ncells.per.group = 1000, ncells.landmark = 2000,
    ncomp = 15, pls.function = "cppls", k.nn = 5,
    name.reduction.1 = "pca", dims.reduction.1 = 1:50,
    fix.wnn.weights = c(0.5, 0.5), verbose = TRUE
  )
  cat("PrepareSampleObject complete\n")

  sample_obj_mut <- GenerateSampleObject(
    object = um_malignant_mut,
    sketch.assay = "LANDMARK", nn.name = "weighted.nn",
    group.by = "Sample", rename.group.by = "majority_celltype",
    k.nn = 5, normalization.method = "ChiSquared",
    add.meta.data = TRUE, return.seurat = TRUE, verbose = TRUE
  )

  DefaultAssay(sample_obj_mut) <- "LMC"
  sample_obj_mut <- ScaleData(sample_obj_mut, features = rownames(sample_obj_mut), verbose = FALSE)
  sample_obj_mut <- RunDiffusionMap(
    object = sample_obj_mut, assay = "LMC", layer = "scale.data",
    features = rownames(sample_obj_mut), ncomp = 10
  )

  saveRDS(sample_obj_mut, cache_mut)
  rm(um_malignant_mut); gc()
}

# Extract coordinates + metadata from sample object directly
dc_coords_mut <- Embeddings(sample_obj_mut, reduction = "DiffMap")
sample_meta_mut <- sample_obj_mut@meta.data
dc_df_mut <- cbind(
  as.data.frame(dc_coords_mut),
  data.frame(
    Sample = sample_meta_mut$Sample,
    Patient = sample_meta_mut$Patient_nr,
    Origin = sample_meta_mut$origin,
    Organ = sample_meta_mut$organ,
    Primary_Mutation = sample_meta_mut$primary_mutation,
    Secondary_Mutation = sample_meta_mut$secondary_mutation,
    stringsAsFactors = FALSE
  )
)

write.csv(dc_df_mut, file.path(output_dir, "mutation_diffusion_coordinates.csv"),
          row.names = FALSE)

cat("Diffusion coordinates extracted for", nrow(dc_df_mut), "samples\n")
cat("Samples with mutation metadata:", sum(!is.na(dc_df_mut$Secondary_Mutation)), "\n")
```

### Mutation Trajectories

```{r plot_mutation_diffusion}
# Plot DC1 vs DC2 colored by secondary mutation
p_mut <- ggplot(dc_df_mut, aes(x = DC_1, y = DC_2, color = Secondary_Mutation)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = second_mutations_col) +
  labs(
    title = "scSLIDE Mutation-Stratified Trajectory",
    subtitle = "Diffusion map of malignant samples (BAP1/SF3B1/EIF1AX)",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2",
    color = "Secondary Mutation"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

print(p_mut)
ggsave(file.path(output_dir, "mutation_DC1_DC2_by_mutation.png"),
       p_mut, width = 10, height = 7, dpi = 300, bg = "white")

# Also color by origin to see disease progression within mutations
p_mut_origin <- ggplot(dc_df_mut, aes(x = DC_1, y = DC_2, color = Origin)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = origin_colors) +
  facet_wrap(~ Secondary_Mutation, nrow = 1) +
  labs(
    title = "scSLIDE Mutation Trajectories",
    subtitle = "Disease progression within each mutation type",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

print(p_mut_origin)
ggsave(file.path(output_dir, "mutation_DC1_DC2_faceted_by_mutation.png"),
       p_mut_origin, width = 14, height = 5, dpi = 300, bg = "white")
```

### Statistical Testing (Mutations)

```{r mutation_statistics}
cat("=== Statistical Testing: Mutation-Stratified ===\n\n")

dc_df_mut_valid <- dc_df_mut %>% dplyr::filter(!is.na(Secondary_Mutation))
cat("Samples with valid mutation metadata:", nrow(dc_df_mut_valid), "/", nrow(dc_df_mut), "\n")

dc_cols_mut <- grep("^DC_", colnames(dc_df_mut_valid), value = TRUE)
ncomp_test_mut <- min(5, length(dc_cols_mut))

# Compute distance matrix
stats_data_mut <- as.data.frame(dc_df_mut_valid)
dist_mat_mut <- dist(stats_data_mut[, dc_cols_mut[1:ncomp_test_mut]], method = "euclidean")
stats_group_mut <- stats_data_mut$Secondary_Mutation

# PERMANOVA
perm_mut <- tryCatch({
  adonis2(dist_mat_mut ~ stats_group_mut, permutations = 999)
}, error = function(e) { cat("adonis2 error:", conditionMessage(e), "\n"); NULL })

if (!is.null(perm_mut)) print(perm_mut)

# ANOSIM
anosim_mut <- tryCatch({
  anosim(dist_mat_mut, stats_group_mut, permutations = 999)
}, error = function(e) { cat("ANOSIM error:", conditionMessage(e), "\n"); NULL })

if (!is.null(anosim_mut)) {
  cat("\nANOSIM test:\n")
  cat("  - R statistic:", round(anosim_mut$statistic, 3), "\n")
  cat("  - p-value:", format.pval(anosim_mut$signif, digits = 3), "\n")
}

# Save statistics
stats_mut <- data.frame(
  Test = c("PERMANOVA", "ANOSIM"),
  R_statistic = c(
    if (!is.null(perm_mut)) perm_mut$R2[1] else NA,
    if (!is.null(anosim_mut)) anosim_mut$statistic else NA
  ),
  p_value = c(
    if (!is.null(perm_mut)) perm_mut$`Pr(>F)`[1] else NA,
    if (!is.null(anosim_mut)) anosim_mut$signif else NA
  )
)

write.csv(stats_mut, file.path(output_dir, "mutation_statistics.csv"),
          row.names = FALSE)

cat("\n✓ Statistics saved\n")
```

---

## Organ-Specific Metastasis {.tabset}

### scSLIDE Analysis (Organs)

```{r scslide_organ, eval=TRUE}
cat("=== scSLIDE: Organ-Specific Metastasis Trajectory ===\n")
cat("Grouping variable: organ (liver/brain/skin/thyroid)\n\n")

# Filter to metastatic samples only
samples_mets <- sample_summary %>%
  dplyr::filter(Origin == "metastasis") %>%
  pull(Sample)

cat("Metastatic samples:", length(samples_mets), "\n")

organ_counts <- sample_summary %>%
  dplyr::filter(Sample %in% samples_mets) %>%
  count(Organ)
print(organ_counts)

cache_organ <- file.path(output_dir, "organ_sample_object.rds")

if (file.exists(cache_organ)) {
  cat("\nLoading cached organ sample object...\n")
  sample_obj_organ <- readRDS(cache_organ)

  if (!"DiffMap" %in% names(sample_obj_organ@reductions)) {
    DefaultAssay(sample_obj_organ) <- "LMC"
    sample_obj_organ <- ScaleData(sample_obj_organ, features = rownames(sample_obj_organ), verbose = FALSE)
    sample_obj_organ <- RunDiffusionMap(
      object = sample_obj_organ, assay = "LMC", layer = "scale.data",
      features = rownames(sample_obj_organ), ncomp = 10
    )
    saveRDS(sample_obj_organ, cache_organ)
  }
  cat("Cached results loaded.\n")

} else {
  um_malignant_mets <- subset(um_malignant, subset = Sample %in% samples_mets)
  cat("\nCells in metastasis subset:", ncol(um_malignant_mets), "\n\n")

  cat("Running PrepareSampleObject()...\n")
  um_malignant_mets <- PrepareSampleObject(
    object = um_malignant_mets,
    assay = "RNA", add.hvg = TRUE,
    group.by.CorTest = "majority_celltype",
    Y = "organ",
    sketch.training = TRUE, group.by.Sketch = "Sample",
    ncells.per.group = 1000, ncells.landmark = 2000,
    ncomp = 15, pls.function = "cppls", k.nn = 5,
    name.reduction.1 = "pca", dims.reduction.1 = 1:50,
    fix.wnn.weights = c(0.5, 0.5), verbose = TRUE
  )
  cat("PrepareSampleObject complete\n")

  sample_obj_organ <- GenerateSampleObject(
    object = um_malignant_mets,
    sketch.assay = "LANDMARK", nn.name = "weighted.nn",
    group.by = "Sample", rename.group.by = "majority_celltype",
    k.nn = 5, normalization.method = "ChiSquared",
    add.meta.data = TRUE, return.seurat = TRUE, verbose = TRUE
  )

  DefaultAssay(sample_obj_organ) <- "LMC"
  sample_obj_organ <- ScaleData(sample_obj_organ, features = rownames(sample_obj_organ), verbose = FALSE)
  sample_obj_organ <- RunDiffusionMap(
    object = sample_obj_organ, assay = "LMC", layer = "scale.data",
    features = rownames(sample_obj_organ), ncomp = 10
  )

  saveRDS(sample_obj_organ, cache_organ)
  rm(um_malignant_mets); gc()
}

# Extract coordinates + metadata from sample object directly
dc_coords_organ <- Embeddings(sample_obj_organ, reduction = "DiffMap")
sample_meta_organ <- sample_obj_organ@meta.data
dc_df_organ <- cbind(
  as.data.frame(dc_coords_organ),
  data.frame(
    Sample = sample_meta_organ$Sample,
    Patient = sample_meta_organ$Patient_nr,
    Origin = sample_meta_organ$origin,
    Organ = sample_meta_organ$organ,
    Primary_Mutation = sample_meta_organ$primary_mutation,
    Secondary_Mutation = sample_meta_organ$secondary_mutation,
    stringsAsFactors = FALSE
  )
)

write.csv(dc_df_organ, file.path(output_dir, "organ_diffusion_coordinates.csv"),
          row.names = FALSE)

cat("Diffusion coordinates extracted for", nrow(dc_df_organ), "samples\n")
cat("Samples with organ metadata:", sum(!is.na(dc_df_organ$Organ)), "\n")
```

### Organ Trajectories

```{r plot_organ_diffusion}
# Plot DC1 vs DC2 colored by organ
p_organ <- ggplot(dc_df_organ, aes(x = DC_1, y = DC_2, color = Organ)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = organ_colors) +
  labs(
    title = "scSLIDE Organ-Specific Metastasis Trajectory",
    subtitle = "Diffusion map of metastatic samples (colored by organ)",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

print(p_organ)
ggsave(file.path(output_dir, "organ_DC1_DC2_by_organ.png"),
       p_organ, width = 10, height = 7, dpi = 300, bg = "white")

# Also color by secondary mutation
p_organ_mut <- ggplot(dc_df_organ, aes(x = DC_1, y = DC_2, color = Secondary_Mutation)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = second_mutations_col) +
  labs(
    title = "scSLIDE Organ-Specific Trajectory",
    subtitle = "Colored by secondary mutation",
    x = "Diffusion Component 1",
    y = "Diffusion Component 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

print(p_organ_mut)
ggsave(file.path(output_dir, "organ_DC1_DC2_by_mutation.png"),
       p_organ_mut, width = 10, height = 7, dpi = 300, bg = "white")
```

### Organ Divergence Analysis

```{r organ_divergence}
cat("=== Organ-Specific Divergence Analysis ===\n\n")

# Compute pairwise distances between organs (using DC columns from joined df)
dc_cols_div <- grep("^DC_", colnames(dc_df_organ), value = TRUE)
ncomp_div <- min(5, length(dc_cols_div))
organ_dist <- dist(dc_df_organ[, dc_cols_div[1:ncomp_div]], method = "euclidean")
organ_dist_mat <- as.matrix(organ_dist)

# Add sample names
rownames(organ_dist_mat) <- dc_df_organ$Sample
colnames(organ_dist_mat) <- dc_df_organ$Sample

# Create annotation for heatmap
organ_annotation <- dc_df_organ %>%
  select(Sample, Organ, Secondary_Mutation) %>%
  as.data.frame()
rownames(organ_annotation) <- organ_annotation$Sample
organ_annotation$Sample <- NULL

# Create heatmap (handle potential NA values in annotations)
ha <- HeatmapAnnotation(
  Organ = organ_annotation$Organ,
  Mutation = organ_annotation$Secondary_Mutation,
  col = list(
    Organ = organ_colors[names(organ_colors) %in% unique(na.omit(organ_annotation$Organ))],
    Mutation = second_mutations_col[names(second_mutations_col) %in% unique(na.omit(organ_annotation$Secondary_Mutation))]
  ),
  na_col = "grey80",
  show_annotation_name = TRUE,
  annotation_name_side = "left"
)

ht <- Heatmap(
  organ_dist_mat,
  name = "Euclidean\nDistance",
  col = colorRamp2(c(0, max(organ_dist_mat)/2, max(organ_dist_mat)),
                   c("white", "orange", "red")),
  top_annotation = ha,
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 8),
  column_title = "Pairwise Organ Divergence (scSLIDE)",
  heatmap_legend_param = list(title_gp = gpar(fontsize = 10))
)

print(ht)

# Save heatmap
png(file.path(output_dir, "organ_divergence_heatmap.png"),
    width = 12, height = 10, units = "in", res = 300)
draw(ht)
dev.off()
```

### Statistical Testing (Organs)

```{r organ_statistics}
cat("=== Statistical Testing: Organ-Specific ===\n\n")

dc_df_organ_valid <- dc_df_organ %>% dplyr::filter(!is.na(Organ))
cat("Samples with valid organ metadata:", nrow(dc_df_organ_valid), "/", nrow(dc_df_organ), "\n")

dc_cols_organ <- grep("^DC_", colnames(dc_df_organ_valid), value = TRUE)
ncomp_test_organ <- min(5, length(dc_cols_organ))

# Compute distance matrix
stats_data_organ <- as.data.frame(dc_df_organ_valid)
dist_mat_organ <- dist(stats_data_organ[, dc_cols_organ[1:ncomp_test_organ]], method = "euclidean")
stats_group_organ <- stats_data_organ$Organ

# PERMANOVA
perm_organ <- tryCatch({
  adonis2(dist_mat_organ ~ stats_group_organ, permutations = 999)
}, error = function(e) { cat("adonis2 error:", conditionMessage(e), "\n"); NULL })

if (!is.null(perm_organ)) print(perm_organ)

# ANOSIM
anosim_organ <- tryCatch({
  anosim(dist_mat_organ, stats_group_organ, permutations = 999)
}, error = function(e) { cat("ANOSIM error:", conditionMessage(e), "\n"); NULL })

if (!is.null(anosim_organ)) {
  cat("\nANOSIM test:\n")
  cat("  - R statistic:", round(anosim_organ$statistic, 3), "\n")
  cat("  - p-value:", format.pval(anosim_organ$signif, digits = 3), "\n")
}

# Save statistics
stats_organ <- data.frame(
  Test = c("PERMANOVA", "ANOSIM"),
  R_statistic = c(
    if (!is.null(perm_organ)) perm_organ$R2[1] else NA,
    if (!is.null(anosim_organ)) anosim_organ$statistic else NA
  ),
  p_value = c(
    if (!is.null(perm_organ)) perm_organ$`Pr(>F)`[1] else NA,
    if (!is.null(anosim_organ)) anosim_organ$signif else NA
  )
)

write.csv(stats_organ, file.path(output_dir, "organ_statistics.csv"),
          row.names = FALSE)

cat("\n✓ Statistics saved\n")
```

---

## GloScope Integration {.tabset}

### Load GloScope Results

```{r load_gloscope}
cat("=== Loading GloScope Results ===\n\n")

# Load GloScope divergence matrix and compute MDS
gloscope_cache_dir <- "../GloScope_Full_Outputs/"

# Try loading malignant-specific divergence matrix first
div_file_malignant <- file.path(gloscope_cache_dir, "divergence_origin_malignant_full.rds")
div_file_full <- file.path(gloscope_cache_dir, "divergence_origin_full_cohort.rds")

if (file.exists(div_file_malignant)) {
  cat("✓ Found GloScope divergence matrix (malignant)\n")

  # Load divergence matrix
  knn_div_gloscope <- readRDS(div_file_malignant)

  cat("  - Samples in GloScope:", nrow(knn_div_gloscope), "\n")

  # Compute MDS from divergence matrix
  library(limma)
  mds_result <- plotMDS(knn_div_gloscope, plot = FALSE)

  # Extract MDS coordinates
  mds_coords <- data.frame(
    Sample = rownames(knn_div_gloscope),
    MDS1 = mds_result$x,
    MDS2 = mds_result$y
  )

  cat("  - MDS coordinates computed\n")

  # Merge with scSLIDE coordinates
  comparison_df <- dc_df_origin %>%
    left_join(mds_coords, by = "Sample") %>%
    dplyr::filter(!is.na(MDS1))  # Keep only samples in both analyses

  cat("  - Samples in both analyses:", nrow(comparison_df), "\n\n")
  if (nrow(comparison_df) < 3) {
    cat("  Too few samples overlap. Setting comparison_df to NULL.\n")
    comparison_df <- NULL
  }

} else if (file.exists(div_file_full)) {
  cat("⚠ Using full cohort divergence matrix (includes non-malignant)\n")

  # Load full cohort divergence matrix
  knn_div_gloscope <- readRDS(div_file_full)

  # Filter to samples in scSLIDE analysis
  common_samples <- intersect(rownames(knn_div_gloscope), dc_df_origin$Sample)
  knn_div_gloscope <- knn_div_gloscope[common_samples, common_samples]

  cat("  - Samples in GloScope (filtered):", nrow(knn_div_gloscope), "\n")

  # Compute MDS
  library(limma)
  mds_result <- plotMDS(knn_div_gloscope, plot = FALSE)

  mds_coords <- data.frame(
    Sample = rownames(knn_div_gloscope),
    MDS1 = mds_result$x,
    MDS2 = mds_result$y
  )

  comparison_df <- dc_df_origin %>%
    left_join(mds_coords, by = "Sample") %>%
    dplyr::filter(!is.na(MDS1))

  cat("  - Samples in both analyses:", nrow(comparison_df), "\n\n")
  if (nrow(comparison_df) < 3) {
    cat("  Too few samples overlap. Setting comparison_df to NULL.\n")
    comparison_df <- NULL
  }

} else {
  cat("⚠ GloScope results not found. Skipping integration.\n")
  cat("  Searched for:\n")
  cat("    -", div_file_malignant, "\n")
  cat("    -", div_file_full, "\n")
  comparison_df <- NULL
}
```

### Coordinate Correlation

```{r gloscope_correlation, eval=!is.null(comparison_df)}
if (!is.null(comparison_df) && nrow(comparison_df) >= 3) {
  cat("=== Correlation: scSLIDE vs GloScope ===\n\n")
  cat("Samples in comparison:", nrow(comparison_df), "\n")

  # Check that DC and MDS columns exist and have valid data
  has_dc <- "DC_1" %in% colnames(comparison_df)
  has_mds <- "MDS1" %in% colnames(comparison_df)

  if (!has_dc || !has_mds) {
    cat("Missing required columns (DC_1 or MDS1). Skipping correlation.\n")
  } else {
    valid_pairs <- complete.cases(comparison_df[, c("DC_1", "MDS1")])
    cat("Complete pairs for DC1/MDS1:", sum(valid_pairs), "\n")
  }

  # Correlation: DC1 vs MDS1
  cor_dc1_mds1 <- tryCatch(
    cor(comparison_df$DC_1, comparison_df$MDS1,
        use = "complete.obs", method = "pearson"),
    error = function(e) { cat("Correlation error:", conditionMessage(e), "\n"); NA }
  )
  if (!is.na(cor_dc1_mds1)) cat("Correlation (DC1 vs MDS1):", round(cor_dc1_mds1, 3), "\n")

  # Correlation: DC2 vs MDS2
  cor_dc2_mds2 <- tryCatch(
    cor(comparison_df$DC_2, comparison_df$MDS2,
        use = "complete.obs", method = "pearson"),
    error = function(e) { cat("Correlation error:", conditionMessage(e), "\n"); NA }
  )
  if (!is.na(cor_dc2_mds2)) cat("Correlation (DC2 vs MDS2):", round(cor_dc2_mds2, 3), "\n\n")

  # Scatter plot: DC1 vs MDS1
  p_cor1 <- ggplot(comparison_df, aes(x = MDS1, y = DC_1, color = Origin)) +
    geom_point(size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    scale_color_manual(values = origin_colors) +
    labs(
      title = "scSLIDE vs GloScope: First Component Comparison",
      subtitle = paste0("Pearson r = ", round(cor_dc1_mds1, 3)),
      x = "GloScope MDS1 (unsupervised)",
      y = "scSLIDE DC1 (supervised)"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right")

  print(p_cor1)
  ggsave(file.path(output_dir, "comparison_DC1_vs_MDS1.png"),
         p_cor1, width = 10, height = 7, dpi = 300, bg = "white")

  # Scatter plot: DC2 vs MDS2
  p_cor2 <- ggplot(comparison_df, aes(x = MDS2, y = DC_2, color = Origin)) +
    geom_point(size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    scale_color_manual(values = origin_colors) +
    labs(
      title = "scSLIDE vs GloScope: Second Component Comparison",
      subtitle = paste0("Pearson r = ", round(cor_dc2_mds2, 3)),
      x = "GloScope MDS2 (unsupervised)",
      y = "scSLIDE DC2 (supervised)"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right")

  print(p_cor2)
  ggsave(file.path(output_dir, "comparison_DC2_vs_MDS2.png"),
         p_cor2, width = 10, height = 7, dpi = 300, bg = "white")

  # Save comparison statistics
  comparison_stats <- data.frame(
    Component = c("First (DC1 vs MDS1)", "Second (DC2 vs MDS2)"),
    Correlation = c(cor_dc1_mds1, cor_dc2_mds2),
    N_Samples = c(nrow(comparison_df), nrow(comparison_df))
  )

  write.csv(comparison_stats,
            file.path(output_dir, "gloscope_scslide_correlation.csv"),
            row.names = FALSE)

  cat("✓ Comparison statistics saved\n")
}
```

### Complementarity Analysis

```{r gloscope_complementarity, eval=!is.null(comparison_df)}
if (!is.null(comparison_df)) {
  cat("=== Complementarity: Supervised vs Unsupervised ===\n\n")

  # Identify samples that diverge between methods
  # Compute residuals from DC1 ~ MDS1 regression
  lm_model <- lm(DC_1 ~ MDS1, data = comparison_df)
  comparison_df$residual_DC1 <- residuals(lm_model)

  # Find samples with large residuals (top 5 positive and negative)
  top_pos <- comparison_df %>%
    arrange(desc(residual_DC1)) %>%
    head(5)

  top_neg <- comparison_df %>%
    arrange(residual_DC1) %>%
    head(5)

  cat("Samples with highest scSLIDE score (relative to GloScope):\n")
  print(top_pos %>% select(Sample, Origin, Organ, Secondary_Mutation,
                            MDS1, DC_1, residual_DC1))

  cat("\nSamples with lowest scSLIDE score (relative to GloScope):\n")
  print(top_neg %>% select(Sample, Origin, Organ, Secondary_Mutation,
                            MDS1, DC_1, residual_DC1))

  # Plot residuals
  p_resid <- ggplot(comparison_df, aes(x = MDS1, y = residual_DC1, color = Origin)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(size = 3, alpha = 0.8) +
    scale_color_manual(values = origin_colors) +
    labs(
      title = "Divergence Between scSLIDE and GloScope",
      subtitle = "Residuals from DC1 ~ MDS1 regression",
      x = "GloScope MDS1",
      y = "Residual (scSLIDE DC1 - predicted)"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right")

  print(p_resid)
  ggsave(file.path(output_dir, "comparison_residuals.png"),
         p_resid, width = 10, height = 7, dpi = 300, bg = "white")

  cat("\n✓ Complementarity analysis complete\n")
}
```

---

## Session Information

```{r session_info}
cat("=== Session Information ===\n\n")

cat("Analysis completed:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n\n")

cat("Key package versions:\n")
cat("  - scSLIDE:", as.character(packageVersion("scSLIDE")), "\n")
cat("  - Seurat:", as.character(packageVersion("Seurat")), "\n")
cat("  - BiocParallel:", as.character(packageVersion("BiocParallel")), "\n")
cat("  - pls:", as.character(packageVersion("pls")), "\n\n")

cat("Output directory:", output_dir, "\n")
cat("Total files created:", length(list.files(output_dir)), "\n\n")

cat("=== All Files Created ===\n")
print(list.files(output_dir))

sessionInfo()
```
