---
title: "Xenium 3-Cohort Spatial Validation of Melanoma Stemness Signature"
output:
  html_document:
    self_contained: true
    includes:
      in_header: !expr system.file("templates/fgcz_header.html", package = "ezRun", lib.loc = .libPaths())
    css: !expr system.file("templates/fgcz.css", package = "ezRun", lib.loc = .libPaths())
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r script-info, include=FALSE}
# This script was used to create the following figures:
# Figure 2 (spatial validation of melanoma stemness signature on 3 Xenium cohorts: Boston, Zurich-B1, Zurich-B2)
```

#  {.tabset}

## Overview

This analysis validates melanoma stemness genes identified from the UM Atlas CytoTRACE2 analysis
on Xenium spatial transcriptomics data from the p36005 Uveal Melanoma project.

**Cohorts analyzed:**

1. **Zurich Batch 1** - 8 samples (eye and subcutaneous metastases)
2. **Zurich Batch 2** - 9 samples (eye and subcutaneous metastases)
3. **Boston/Sullivan** - 6 samples (liver and breast metastases)

**Objective:** Determine if stemness-associated genes cluster spatially or show specific patterns
in the tumor microenvironment across different metastatic sites and treatment conditions.

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 10)

library(Seurat)
library(qs2)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(pals)

# Load Victoria's color palettes
source("colors.R")

# Set paths
output_dir <- "Outputs/Xenium_Stemness/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Load CytoTRACE2 gene signatures (top = stemness, bottom = differentiation)
top_genes <- read.csv("Outputs/cytotrace2_top_genes_xenium.csv")
bottom_genes <- read.csv("Outputs/cytotrace2_bottom_genes_xenium.csv")

cat("=== TOP CytoTRACE2 Genes (Stemness-associated) ===\n")
cat("N genes:", nrow(top_genes), "\n")
print(head(top_genes, 15))

cat("\n=== BOTTOM CytoTRACE2 Genes (Differentiation-associated) ===\n")
cat("N genes:", nrow(bottom_genes), "\n")
print(head(bottom_genes, 15))

# Cell cycle genes (from Seurat's cc.genes / Tirosh 2015)
# S phase genes
s_genes <- c("MCM5", "PCNA", "TYMS", "FEN1", "MCM2", "MCM4", "RRM1", "UNG", "GINS2",
             "MCM6", "CDCA7", "DTL", "PRIM1", "UHRF1", "MLF1IP", "HELLS", "RFC2",
             "RPA2", "NASP", "RAD51AP1", "GMNN", "WDR76", "SLBP", "CCNE2", "UBR7",
             "POLD3", "MSH2", "ATAD2", "RAD51", "RRM2", "CDC45", "CDC6", "EXO1",
             "TIPIN", "DSCC1", "BLM", "CASP8AP2", "USP1", "CLSPN", "POLA1", "CHAF1B",
             "BRIP1", "E2F8")

# G2M phase genes
g2m_genes <- c("HMGB2", "CDK1", "NUSAP1", "UBE2C", "BIRC5", "TPX2", "TOP2A", "NDC80",
               "CKS2", "NUF2", "CKS1B", "MKI67", "TMPO", "CENPF", "TACC3", "FAM64A",
               "SMC4", "CCNB2", "CKAP2L", "CKAP2", "AURKB", "BUB1", "KIF11", "ANP32E",
               "TUBB4B", "GTSE1", "KIF20B", "HJURP", "CDCA3", "HN1", "CDC20", "TTK",
               "CDC25C", "KIF2C", "RANGAP1", "NCAPD2", "DLGAP5", "CDCA2", "CDCA8",
               "ECT2", "KIF23", "HMMR", "AURKA", "PSRC1", "ANLN", "LBR", "CKAP5",
               "CENPE", "CTCF", "NEK2", "G2E3", "GAS2L3", "CBX5", "CENPA")

cycling_genes <- unique(c(s_genes, g2m_genes))
cat("\n=== Cell Cycle Genes (for overlap analysis) ===\n")
cat("Total cell cycle genes:", length(cycling_genes), "\n")
```

## Load Xenium Data

```{r load_xenium}
# Check if pre-computed objects with scores exist (much faster to load)
precomputed_paths <- list(
  Boston_Sullivan = "Outputs/Xenium_Stemness/xenium_Boston_Sullivan_with_scores.qs2",
  Zurich_Batch1 = "Outputs/Xenium_Stemness/xenium_Zurich_Batch1_with_scores.qs2",
  Zurich_Batch2 = "Outputs/Xenium_Stemness/xenium_Zurich_Batch2_with_scores.qs2"
)

use_precomputed <- all(file.exists(unlist(precomputed_paths)))
cat("Using pre-computed objects with scores:", use_precomputed, "\n\n")

# Original paths (used only if pre-computed don't exist)
cohort_paths <- list(
  Boston_Sullivan = "../Xenium_Boston_Sullivan/merged_xenium_all_samples.qs2",
  Zurich_Batch1 = "../2024-25-09-Xenium_UM/Seurat_Xenium_files_qs2/Xenium_samples_merged.qs2",
  Zurich_Batch2 = "../Xenium_Zurich_SecondBatch/Seurat_Objects/merged_xenium_all_regions.qs2"
)

# CellType_Clean mapping (128 -> 22 categories)
# Source: GPU_Analysis/marimo_multigpu_xenium.py lines 316-428
celltype_mapping <- c(
  # Tumor
  "Melanoma" = "Melanoma",
  # Unknown
  "NA" = "Unknown", "unassigned" = "Unknown",
  # B cells
  "B_cell" = "B_cells", "B_cells" = "B_cells", "BC" = "B_cells",
  "Memory_B_cells" = "B_cells", "Naive_B_cells" = "B_cells",
  "Follicular_B_cells" = "B_cells", "Germinal_center_B_cells" = "B_cells",
  "Cycling_B_cells" = "B_cells", "memory B cell" = "B_cells",
  # T cells
  "T_cells" = "T_cells", "T_cell" = "T_cells", "Th" = "T_cells", "Tc" = "T_cells",
  "T_reg" = "T_cells", "Regulatory_T_cells" = "T_cells",
  "CD8__T_RM" = "T_cells", "Trm_cytotoxic_T_cells" = "T_cells",
  "Tem_Trm_cytotoxic_T_cells" = "T_cells", "Tem_Effector_helper_T_cells" = "T_cells",
  "Tcm_Naive_helper_T_cells" = "T_cells", "gamma_delta_T_cells" = "T_cells",
  "Follicular_helper_T_cells" = "T_cells", "T follicular helper cell" = "T_cells",
  "Type_17_helper_T_cells" = "T_cells", "Type_1_helper_T_cells" = "T_cells",
  "CD4-positive helper T cell" = "T_cells", "naive T cell" = "T_cells",
  "CD4-positive, CD25-positive, CCR4-positive, alpha-beta regulatory T cell" = "T_cells",
  "CD8-positive, alpha-beta regulatory T cell" = "T_cells",
  "effector memory CD4-positive, alpha-beta T cell" = "T_cells",
  "effector memory CD8-positive, alpha-beta T cell" = "T_cells",
  "exhausted T cell" = "T_cells", "cycling T cell" = "T_cells",
  # NK cells
  "NK" = "NK_cells", "NK_cells" = "NK_cells", "Resident_NK" = "NK_cells",
  "Circulating_NK_NKT" = "NK_cells", "ILC_NK" = "NK_cells", "ILC" = "NK_cells",
  "natural killer cell" = "NK_cells", "natural_killer_cell" = "NK_cells",
  # Macrophages
  "Macrophages" = "Macrophages", "macrophage" = "Macrophages",
  "Macro1_2" = "Macrophages", "Intestinal_macrophages" = "Macrophages",
  "cycling macrophage" = "Macrophages",
  # Dendritic cells
  "DC1" = "Dendritic_cells", "DC2" = "Dendritic_cells",
  "IL8__DC1" = "Dendritic_cells", "MigDC" = "Dendritic_cells",
  "Mig_cDCs" = "Dendritic_cells", "cDC1s" = "Dendritic_cells", "cDC2s" = "Dendritic_cells",
  "pDCs" = "Dendritic_cells", "conventional dendritic cell" = "Dendritic_cells",
  "myeloid dendritic cell" = "Dendritic_cells", "plasmacytoid dendritic cell" = "Dendritic_cells",
  # Monocytes
  "Monocytes" = "Monocytes", "Mono_mono_derived_cells" = "Monocytes",
  # Plasma cells
  "Plasma_cells" = "Plasma_cells", "PlasmaC" = "Plasma_cells", "IgG plasma cell" = "Plasma_cells",
  # Mast cells
  "MastC" = "Mast_cells", "Mast_cells" = "Mast_cells", "mast cell" = "Mast_cells",
  # Granulocytes
  "Basophils" = "Granulocytes", "Neutrophils" = "Granulocytes",
  # Fibroblasts
  "Fibroblasts" = "Fibroblasts", "fibroblast" = "Fibroblasts",
  "APOD__fibroblasts" = "Fibroblasts", "POSTN__fibroblasts" = "Fibroblasts",
  "PTGDS__fibroblasts" = "Fibroblasts", "SFRP2__fibroblasts" = "Fibroblasts",
  "Myofibroblasts" = "Fibroblasts", "myofibroblast cell" = "Fibroblasts",
  "Stromal_1" = "Fibroblasts", "Stromal_2" = "Fibroblasts", "Stromal_3" = "Fibroblasts",
  "cycling stromal cell" = "Fibroblasts",
  # Endothelial
  "Endothelial_cells" = "Endothelial", "VEC" = "Endothelial",
  "Arterial_endothelial_cells" = "Endothelial", "Venous_endothelial_cells" = "Endothelial",
  "Capillary_endothelial_cells" = "Endothelial", "Lymphatic_endothelial_cells" = "Endothelial",
  "LEC" = "Endothelial", "endothelial cell" = "Endothelial", "endothelium" = "Endothelial",
  "capillary endothelial cell" = "Endothelial", "endothelial cell of artery" = "Endothelial",
  "endothelial cell of vascular tree" = "Endothelial", "vein endothelial cell" = "Endothelial",
  # Pericytes/SMC
  "Pericytes" = "Pericytes", "pericyte" = "Pericytes",
  "RGS5__pericytes" = "Pericytes", "TAGLN__pericytes" = "Pericytes",
  "SMC" = "Pericytes", "vascular associated smooth muscle cell" = "Pericytes",
  "vascular_associated_smooth_muscle_cell" = "Pericytes",
  # Melanocytes (non-tumor)
  "Melanocytes" = "Melanocytes", "melanocyte" = "Melanocytes",
  "pigmented_epithelial_cell" = "Melanocytes",
  # Epithelial - Hepatic
  "Hepatocytes" = "Hepatocytes", "Cholangiocytes" = "Cholangiocytes",
  # Epithelial - Intestinal
  "Enterocytes" = "Epithelial_GI", "Colonocytes" = "Epithelial_GI",
  "BEST4__enterocytes" = "Epithelial_GI", "Goblet_cells" = "Epithelial_GI",
  "Immature_goblet_cells" = "Epithelial_GI", "Paneth" = "Epithelial_GI",
  "Enteroendocrine_cells" = "Epithelial_GI", "Tuft_cells" = "Epithelial_GI",
  "Cycling_Secretory_TA" = "Epithelial_GI", "TA" = "Epithelial_GI",
  # Epithelial - Skin
  "Suprabasal_keratinocytes" = "Keratinocytes", "Basal_keratinocytes" = "Keratinocytes",
  # Neural
  "Neuronal_SchwannC" = "Neural", "Schwann_cell" = "Neural",
  "myelinating_Schwann_cell" = "Neural", "Glia" = "Neural",
  "astrocyte" = "Neural", "microglial_cell" = "Neural",
  "oligodendrocyte" = "Neural", "oligodendrocyte_precursor_cell" = "Neural",
  # Stem cells
  "Stem_cells" = "Stem_progenitor"
)

# Load each cohort
cat("Loading 3 cohorts...\n\n")
xenium_list <- list()

for (cohort in names(cohort_paths)) {
  # Use precomputed if available
  if (use_precomputed) {
    path <- precomputed_paths[[cohort]]
    cat("Loading PRECOMPUTED", cohort, "from:", path, "\n")
  } else {
    path <- cohort_paths[[cohort]]
    cat("Loading", cohort, "from original:", path, "\n")
  }

  if (!file.exists(path)) {
    stop("File not found: ", path)
  }

  obj <- qs2::qs_read(path, nthreads = 16)
  obj <- UpdateSeuratObject(obj)  # Fix compatibility with newer SeuratObject version
  obj$Cohort <- cohort

  # Add CellType_Clean if not already present or needs update
  if (!"CellType_Clean" %in% colnames(obj@meta.data) || any(is.na(obj$CellType_Clean))) {
    if ("CellType_Harmonized" %in% colnames(obj@meta.data)) {
      obj$CellType_Clean <- dplyr::recode(obj$CellType_Harmonized, !!!celltype_mapping, .default = "Other")
    } else if ("predicted.celltype" %in% colnames(obj@meta.data)) {
      obj$CellType_Clean <- dplyr::recode(obj$predicted.celltype, !!!celltype_mapping, .default = "Other")
    } else if ("cell_type" %in% colnames(obj@meta.data)) {
      obj$CellType_Clean <- dplyr::recode(obj$cell_type, !!!celltype_mapping, .default = "Other")
    } else if (!"CellType_Clean" %in% colnames(obj@meta.data)) {
      obj$CellType_Clean <- "Unknown"
    }
  }

  # CRITICAL: Convert actual R NA values to "Unknown" (not just the string "NA")
  obj$CellType_Clean[is.na(obj$CellType_Clean)] <- "Unknown"

  # Standardize organ and treatment columns across cohorts
  # Organ standardization
  if ("organ" %in% colnames(obj@meta.data)) {
    # Zurich Batch 1
    obj$organ_std <- tolower(obj$organ)
  } else if ("Tissue_Type" %in% colnames(obj@meta.data)) {
    # Zurich Batch 2 and Boston Sullivan (fix case inconsistency)
    obj$organ_std <- tolower(obj$Tissue_Type)
  } else {
    obj$organ_std <- "unknown"
  }

  # Treatment standardization (BL = baseline/pre, FU = follow-up/on/post)
  if ("treatment" %in% colnames(obj@meta.data)) {
    # Zurich Batch 1: already BL/FU
    obj$treatment_std <- obj$treatment
  } else if ("Treatment_Status" %in% colnames(obj@meta.data)) {
    # Zurich Batch 2 and Boston Sullivan: need mapping
    obj$treatment_std <- dplyr::case_when(
      grepl("baseline|Pre|pre", obj$Treatment_Status) ~ "BL",
      grepl("on_treatment|Post|post", obj$Treatment_Status) ~ "FU",
      TRUE ~ "Unknown"
    )
  } else {
    obj$treatment_std <- "Unknown"
  }

  # Fix Zurich Batch 2 'fov' sample - load properly annotated B25.28468
  # Root cause: 02d_RCTD_B25_28468_Annotation.Rmd line 76 has hard-coded fov="fov"
  if (cohort == "Zurich_Batch2" && "fov" %in% names(obj@images)) {
    cat("  - Fixing 'fov' sample: loading properly annotated B25.28468 (Subcutaneous)...\n")

    # Load the RCTD-annotated object (has cell types from skin reference)
    b25_path <- "../Xenium_Zurich_SecondBatch/Outputs_RCTD_B25_28468/xenium_Region4_0056806_rctd_annotated.qs2"
    b25_obj <- qs2::qs_read(b25_path, nthreads = 16)
    b25_obj <- UpdateSeuratObject(b25_obj)  # Fix compatibility with newer SeuratObject version

    # Fix FOV name in the annotated object
    names(b25_obj@images)[names(b25_obj@images) == "fov"] <- "Region4_0056806"

    # Add standardized metadata to B25 object
    b25_obj$organ_std <- "subcutaneous"
    b25_obj$treatment_std <- "BL"
    b25_obj$Cohort <- "Zurich_Batch2"

    # Apply CellType_Clean mapping to B25 object
    if ("predicted.celltype" %in% colnames(b25_obj@meta.data)) {
      b25_obj$CellType_Clean <- dplyr::recode(b25_obj$predicted.celltype, !!!celltype_mapping, .default = "Other")
      b25_obj$CellType_Clean[is.na(b25_obj$CellType_Clean)] <- "Unknown"
    } else {
      b25_obj$CellType_Clean <- "Unknown"
    }

    # Remove the broken 'fov' cells from merged object
    fov_cells <- Cells(obj@images[["fov"]])
    keep_cells <- setdiff(Cells(obj), fov_cells)
    obj <- subset(obj, cells = keep_cells)

    # Merge the properly annotated object
    obj <- merge(obj, b25_obj, add.cell.ids = c("", "B25"))

    cat("    B25.28468 merged:", ncol(b25_obj), "cells with cell type annotations\n")
    rm(b25_obj); gc()
  }

  cat("  - Cells:", ncol(obj), "\n")
  cat("  - Genes:", nrow(obj), "\n")
  cat("  - FOVs:", length(names(obj@images)), "\n")
  cat("  - FOV names:", paste(names(obj@images), collapse = ", "), "\n")
  cat("  - CellType_Clean categories:", length(unique(obj$CellType_Clean)), "\n")
  if ("Stemness_UCell" %in% colnames(obj@meta.data)) {
    cat("  - Stemness_UCell already computed: YES\n\n")
  } else {
    cat("  - Stemness_UCell: NOT YET COMPUTED\n\n")
  }

  xenium_list[[cohort]] <- obj
  gc()
}

# Total summary
total_cells <- sum(sapply(xenium_list, ncol))
total_fovs <- sum(sapply(xenium_list, function(x) length(names(x@images))))
cat("=== TOTAL ===\n")
cat("Total cells:", total_cells, "\n")
cat("Total FOVs:", total_fovs, "\n")

# Check available genes using first cohort
available_top <- intersect(top_genes$gene, rownames(xenium_list[[1]]))
available_bottom <- intersect(bottom_genes$gene, rownames(xenium_list[[1]]))

cat("\n=== CytoTRACE2 Genes in Xenium Panel ===\n")
cat("Top (stemness) genes available:", length(available_top), "of", nrow(top_genes), "\n")
cat("Bottom (differentiation) genes available:", length(available_bottom), "of", nrow(bottom_genes), "\n")
cat("\nTop genes:", paste(head(available_top, 10), collapse = ", "), "\n")
cat("Bottom genes:", paste(head(available_bottom, 10), collapse = ", "), "\n")

# Check cycling genes in Xenium panel
available_cycling <- intersect(cycling_genes, rownames(xenium_list[[1]]))
cat("\n=== Cell Cycle Genes in Xenium Panel ===\n")
cat("Cell cycle genes available:", length(available_cycling), "of", length(cycling_genes), "\n")
cat("Cycling genes:", paste(head(available_cycling, 10), collapse = ", "), "...\n")
```

## Calculate CytoTRACE2 Scores {.tabset}

### Signature Scoring (UCell)

```{r calculate_scores}
# Create signature lists for both top and bottom CytoTRACE2 genes
cytotrace_signatures <- list(
  CytoTRACE2_Top = available_top,      # Stemness-associated (high potency)
  CytoTRACE2_Bottom = available_bottom  # Differentiation-associated (low potency)
)

cat("=== CytoTRACE2 Signatures for Xenium ===\n")
cat("Top (stemness) genes:", length(cytotrace_signatures$CytoTRACE2_Top), "\n")
cat("  ", paste(head(cytotrace_signatures$CytoTRACE2_Top, 10), collapse = ", "), "...\n")
cat("Bottom (differentiation) genes:", length(cytotrace_signatures$CytoTRACE2_Bottom), "\n")
cat("  ", paste(head(cytotrace_signatures$CytoTRACE2_Bottom, 10), collapse = ", "), "...\n")

# Check if scores already exist in pre-computed objects
# NOTE: Set force_recalculate = TRUE to use updated maxRank parameter
force_recalculate <- TRUE  # Force recalculation with tuned maxRank for Xenium
scores_exist <- all(sapply(xenium_list, function(x) "Stemness_UCell" %in% colnames(x@meta.data)))

if (scores_exist && !force_recalculate) {
  cat("\n=== SCORES ALREADY COMPUTED - SKIPPING UCell calculation ===\n")
  cat("Set force_recalculate = TRUE to recalculate with tuned maxRank parameter.\n")
} else {
  # Calculate module scores for each cohort using UCell
  # UCell is better for small gene panels like Xenium (rank-based, no control gene issues)
  cat("\nCalculating module scores using UCell for each cohort...\n")
  library(UCell)

  # Define signatures for UCell (including cycling for overlap analysis)
  signatures <- list(
    CytoTRACE2_Top = available_top,
    CytoTRACE2_Bottom = available_bottom,
    Cycling = available_cycling
  )

  for (cohort in names(xenium_list)) {
    cat("Processing", cohort, "...\n")
    obj <- xenium_list[[cohort]]

    # JoinLayers if needed
    obj <- JoinLayers(obj)

    # Calculate UCell scores
    # With name = NULL, UCell creates columns with exact signature names
    # CRITICAL: Use maxRank = 200 for better resolution with Xenium's limited gene panel
    xenium_gene_count <- nrow(obj)
    optimal_maxRank <- 200  # Fixed at 200 for consistent scoring across samples
    cat("  Gene panel size:", xenium_gene_count, "- Using maxRank:", optimal_maxRank, "\n")
    obj <- AddModuleScore_UCell(obj, features = signatures, name = NULL,
                                 maxRank = optimal_maxRank)

    # Calculate combined CytoTRACE2-like score as difference: Top - Bottom
    obj$CytoTRACE2_Score <- obj$CytoTRACE2_Top - obj$CytoTRACE2_Bottom

    # Also keep "Stemness_UCell" for backward compatibility
    obj$Stemness_UCell <- obj$CytoTRACE2_Top

    xenium_list[[cohort]] <- obj
    gc()
  }
}

# Summary across all cohorts
cat("\n=== CytoTRACE2 Score Summary (All Cohorts) ===\n")
for (cohort in names(xenium_list)) {
  obj <- xenium_list[[cohort]]
  cat("\n", cohort, ":\n", sep = "")
  cat("  Top Score - Mean:", round(mean(obj$CytoTRACE2_Top, na.rm = TRUE), 4), "\n")
  cat("  Bottom Score - Mean:", round(mean(obj$CytoTRACE2_Bottom, na.rm = TRUE), 4), "\n")
  cat("  Combined Score - Mean:", round(mean(obj$CytoTRACE2_Score, na.rm = TRUE), 4), "\n")
  cat("  Combined Score - Range:", paste(round(range(obj$CytoTRACE2_Score, na.rm = TRUE), 4), collapse = " to "), "\n")
}
```

### CytoTRACE2 Score Distribution

```{r score_distribution, fig.width=16, fig.height=10}
# Combine metadata from all cohorts for distribution plots
# Note: Include Cycling score and nFeature_Xenium for correlation analyses
# Note: CellType_Display is the merged cell type matching Victoria's UM Atlas naming
# It may not exist yet if cell type merging hasn't been applied
all_meta <- do.call(rbind, lapply(names(xenium_list), function(cohort) {
  obj <- xenium_list[[cohort]]
  cols <- c("CytoTRACE2_Top", "CytoTRACE2_Bottom", "CytoTRACE2_Score", "Stemness_UCell", "Cohort", "CellType_Clean")
  # Add CellType_Display if present (created in later chunk)
  if ("CellType_Display" %in% colnames(obj@meta.data)) {
    cols <- c(cols, "CellType_Display")
  }
  # Add Cycling if present
  if ("Cycling" %in% colnames(obj@meta.data)) {
    cols <- c(cols, "Cycling")
  }
  # Add nFeature_Xenium if present
  if ("nFeature_Xenium" %in% colnames(obj@meta.data)) {
    cols <- c(cols, "nFeature_Xenium")
  }
  df <- obj@meta.data[, cols]
  df$cell_id <- rownames(df)
  df
}))

cat("Combined metadata from all cohorts:", nrow(all_meta), "cells\n")

# Score distributions as histograms (faceted by cohort)
p1 <- ggplot(all_meta, aes(x = CytoTRACE2_Top, fill = Cohort)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~Cohort, ncol = 1) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  ggtitle("CytoTRACE2 Top (Stemness) Score Distribution")

p2 <- ggplot(all_meta, aes(x = CytoTRACE2_Bottom, fill = Cohort)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~Cohort, ncol = 1) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  ggtitle("CytoTRACE2 Bottom (Differentiation) Score Distribution")

p3 <- ggplot(all_meta, aes(x = CytoTRACE2_Score, fill = Cohort)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~Cohort, ncol = 1) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  ggtitle("CytoTRACE2 Combined Score (Top - Bottom)")

p_all <- p1 + p2 + p3 + plot_layout(ncol = 3, guides = "collect")
print(p_all)

ggsave(
  file.path(output_dir, "histogram_cytotrace2_scores_by_cohort.png"),
  p_all,
  width = 18, height = 10, dpi = 300, bg = "white"
)
```

## Stemness vs Cell Cycle Analysis {.tabset}

### Gene Overlap Analysis

```{r stemness_cycling_gene_overlap, fig.width=12, fig.height=10}
# Check direct overlap between stemness signature genes and cell cycle genes
# This determines if "stemness" is actually measuring proliferation

stemness_genes <- available_top
cat("=== Stemness vs Cell Cycle Gene Overlap ===\n")
cat("Stemness signature genes (in Xenium panel):", length(stemness_genes), "\n")
cat("Cell cycle genes (in Xenium panel):", length(available_cycling), "\n")

# Direct gene overlap
cycle_overlap <- intersect(stemness_genes, cycling_genes)
cat("\nOverlapping genes:", length(cycle_overlap), "\n")

if (length(cycle_overlap) > 0) {
  cat("\nSTEMNESS GENES THAT ARE CELL CYCLE-RELATED:\n")
  print(cycle_overlap)
  cat("\nProportion of stemness signature that is cell cycle:",
      round(length(cycle_overlap) / length(stemness_genes) * 100, 1), "%\n")
} else {
  cat("\nNo overlap - stemness signature is independent of cell cycle genes.\n")
}

# Venn diagram for gene overlap
library(ggVennDiagram)
gene_sets <- list(
  Stemness = stemness_genes,
  CellCycle = available_cycling
)
p_venn <- ggVennDiagram(gene_sets) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  ggtitle("Overlap: Stemness vs Cell Cycle Genes in Xenium Panel")
print(p_venn)

ggsave(
  file.path(output_dir, "venn_stemness_vs_cycling_genes.png"),
  p_venn, width = 8, height = 8, dpi = 300, bg = "white"
)

# Also check bottom genes (differentiation)
bottom_overlap <- intersect(available_bottom, cycling_genes)
cat("\n=== Differentiation Genes vs Cell Cycle ===\n")
cat("Differentiation signature genes:", length(available_bottom), "\n")
cat("Overlap with cell cycle:", length(bottom_overlap), "\n")
if (length(bottom_overlap) > 0) {
  cat("Overlapping genes:", paste(bottom_overlap, collapse = ", "), "\n")
}
```

### Score Correlation: Stemness vs Cycling

```{r stemness_cycling_correlation, fig.width=14, fig.height=12}
library(ggpubr)

# Check if Cycling column exists
if (!"Cycling" %in% colnames(all_meta)) {
  cat("NOTE: Cycling score not yet calculated in pre-computed objects.\n")
  cat("Run without pre-computed objects to see this analysis.\n")
} else {
  # Scatter plot: Stemness vs Cycling scores by cohort
  p_corr_cohort <- ggplot(all_meta, aes(x = Stemness_UCell, y = Cycling)) +
    geom_point(alpha = 0.05, size = 0.3) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95) +
    facet_wrap(~Cohort, scales = "free") +
    theme_minimal() +
    ggtitle("Correlation: Stemness Score vs Cycling Score by Cohort") +
    xlab("Stemness Score (UCell)") + ylab("Cycling Score (UCell)")
  print(p_corr_cohort)

  ggsave(
    file.path(output_dir, "scatter_stemness_vs_cycling_by_cohort.png"),
    p_corr_cohort, width = 14, height = 10, dpi = 300, bg = "white"
  )

  # By cell type (for Melanoma specifically - using CellType_Display if available, else CellType_Clean)
  if ("CellType_Display" %in% colnames(all_meta)) {
    melanoma_meta <- all_meta[all_meta$CellType_Display == "Melanoma cells", ]
  } else {
    melanoma_meta <- all_meta[all_meta$CellType_Clean == "Melanoma cells", ]
  }
  if (nrow(melanoma_meta) > 100) {
    p_corr_melanoma <- ggplot(melanoma_meta, aes(x = Stemness_UCell, y = Cycling)) +
      geom_point(alpha = 0.1, size = 0.5, color = "darkred") +
      geom_smooth(method = "lm", color = "black", se = TRUE) +
      stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95) +
      facet_wrap(~Cohort, scales = "free") +
      theme_minimal() +
      ggtitle("Stemness vs Cycling - MELANOMA CELLS ONLY") +
      xlab("Stemness Score") + ylab("Cycling Score")
    print(p_corr_melanoma)

    ggsave(
      file.path(output_dir, "scatter_stemness_vs_cycling_melanoma_only.png"),
      p_corr_melanoma, width = 14, height = 10, dpi = 300, bg = "white"
    )
  }

  # Interpretation guide
  cat("\n=== Interpretation Guide ===\n")
  cat("HIGH gene overlap + HIGH score correlation:\n")
  cat("  -> Stemness signal may be driven by proliferation, not true 'stemness'\n")
  cat("\nLOW gene overlap + HIGH score correlation:\n")
  cat("  -> Stem-like cells are proliferative (biological, not artifact)\n")
  cat("\nLOW gene overlap + LOW score correlation:\n")
  cat("  -> Stemness and cycling are independent characteristics\n")
}
```

### Correlation: Stemness vs Gene Detection

```{r stemness_nfeature_correlation, fig.width=14, fig.height=10}
# Check if nFeature_Xenium exists
if (!"nFeature_Xenium" %in% colnames(all_meta)) {
  cat("NOTE: nFeature_Xenium not found in metadata.\n")
  cat("This QC metric may have a different column name.\n")
} else {
  # Correlation plot: Stemness vs nFeature_Xenium
  p_nfeat <- ggplot(all_meta, aes(x = nFeature_Xenium, y = Stemness_UCell)) +
    geom_point(alpha = 0.05, size = 0.3) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95) +
    facet_wrap(~Cohort, scales = "free") +
    theme_minimal() +
    ggtitle("Correlation: Stemness Score vs Number of Detected Genes") +
    xlab("Number of Detected Genes (nFeature_Xenium)") + ylab("Stemness Score")
  print(p_nfeat)

  ggsave(
    file.path(output_dir, "scatter_stemness_vs_nfeature.png"),
    p_nfeat, width = 14, height = 10, dpi = 300, bg = "white"
  )

  # Melanoma only
  if (nrow(melanoma_meta) > 100) {
    p_nfeat_mel <- ggplot(melanoma_meta, aes(x = nFeature_Xenium, y = Stemness_UCell)) +
      geom_point(alpha = 0.1, size = 0.5, color = "darkred") +
      geom_smooth(method = "lm", color = "black", se = TRUE) +
      stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.95) +
      facet_wrap(~Cohort, scales = "free") +
      theme_minimal() +
      ggtitle("Stemness vs nFeature - MELANOMA CELLS ONLY") +
      xlab("Number of Detected Genes") + ylab("Stemness Score")
    print(p_nfeat_mel)

    ggsave(
      file.path(output_dir, "scatter_stemness_vs_nfeature_melanoma_only.png"),
      p_nfeat_mel, width = 14, height = 10, dpi = 300, bg = "white"
    )
  }

  cat("\n=== Interpretation ===\n")
  cat("Strong positive correlation: Stemness score may be confounded by gene detection.\n")
  cat("Weak/no correlation: Stemness is biologically meaningful independent of QC.\n")
}
```

## Spatial Visualization {.tabset}

### Sample Overview

```{r spatial_overview, fig.width=12, fig.height=8}
# Get FOV/sample information for each cohort
cat("=== Available FOVs/Samples by Cohort ===\n\n")

total_fovs <- 0
for (cohort in names(xenium_list)) {
  obj <- xenium_list[[cohort]]
  fov_names <- names(obj@images)
  total_fovs <- total_fovs + length(fov_names)

  cat(cohort, ":\n")
  cat("  Number of FOVs:", length(fov_names), "\n")
  cat("  FOV names:", paste(fov_names, collapse = ", "), "\n\n")
}

cat("TOTAL FOVs:", total_fovs, "\n")

# Cell type column is CellType_Clean (22 categories)
ct_col <- "CellType_Clean"
cat("\nUsing cell type column:", ct_col, "\n")

# Count unique cell types across all cohorts
all_celltypes <- unique(unlist(lapply(xenium_list, function(x) unique(x$CellType_Clean))))
cat("Unique CellType_Clean categories:", length(all_celltypes), "\n")
print(sort(all_celltypes))
```

### Individual Sample Spatial Plots {.tabset}

```{r individual_fov_plots, results='asis', fig.width=20, fig.height=10}
# First: Create stemness categories for all cohorts BEFORE plotting
# Use PER-COHORT quantiles so each cohort has its own distribution
# This prevents cohorts with lower absolute scores (e.g., Boston_Sullivan)
# from appearing mostly "Low" due to global quantile calculation

cat("=== Per-Cohort Stemness Category Breaks ===\n")

# Add stemness_category to each cohort using cohort-specific quantiles
for (cohort_name in names(xenium_list)) {
  obj <- xenium_list[[cohort_name]]

  # Calculate quantile breaks for THIS cohort only
  cohort_stemness <- obj$Stemness_UCell
  cohort_breaks <- quantile(cohort_stemness, probs = c(0, 0.25, 0.75, 1), na.rm = TRUE)
  cat(cohort_name, "breaks:", paste(round(cohort_breaks, 3), collapse = ", "), "\n")

  # Use -Inf and Inf as outer bounds to capture all values
  obj$stemness_category <- cut(
    obj$Stemness_UCell,
    breaks = c(-Inf, cohort_breaks[2], cohort_breaks[3], Inf),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE
  )

  # CRITICAL: Handle any NA values in stemness_category (from NA scores)
  obj$stemness_category[is.na(obj$stemness_category)] <- "Low"
  obj$stemness_category <- factor(obj$stemness_category, levels = c("Low", "Medium", "High"))

  xenium_list[[cohort_name]] <- obj
}

# ===========================================================================
# Cell Type Merging to Match Victoria's Dataset
# ===========================================================================
# Merge cell types to match the UM Atlas naming convention
celltype_merge <- c(
  "B_cells" = "B/Plasma cells",
  "Plasma_cells" = "B/Plasma cells",
  "T_cells" = "T/NK cells",
  "NK_cells" = "T/NK cells",
  "Dendritic_cells" = "Myeloid cells",
  "Macrophages" = "Myeloid cells",
  "Monocytes" = "Myeloid cells",
  "Endothelial" = "Endothelial cells",
  "Melanoma" = "Melanoma cells"
)

# Apply merging to create CellType_Display column
for (cohort_name in names(xenium_list)) {
  obj <- xenium_list[[cohort_name]]

  # Create display version of cell types
  obj$CellType_Display <- as.character(obj$CellType_Clean)

  # Apply merging
  for (old_name in names(celltype_merge)) {
    obj$CellType_Display[obj$CellType_Display == old_name] <- celltype_merge[[old_name]]
  }

  xenium_list[[cohort_name]] <- obj
}

cat("Cell type merging applied:\n")
print(celltype_merge)

# Get unique cell types after merging
all_celltypes_display <- sort(unique(unlist(lapply(xenium_list, function(x) unique(x$CellType_Display)))))
cat("\nCell types after merging:", length(all_celltypes_display), "\n")
print(all_celltypes_display)

# Use Victoria's colors where available, polychrome for others
celltype_colors <- majority_celltype_col[all_celltypes_display]
# Fill missing colors with polychrome
missing_colors <- is.na(celltype_colors)
if (any(missing_colors)) {
  n_missing <- sum(missing_colors)
  extra_colors <- unname(pals::polychrome())[1:n_missing]
  celltype_colors[missing_colors] <- extra_colors
  names(celltype_colors)[missing_colors] <- all_celltypes_display[missing_colors]
}
cat("\nCell type colors (Victoria's palette + polychrome fallback):\n")
print(celltype_colors)

# Stemness category colors (named)
stemness_cat_colors <- c("Low" = "#2166AC", "Medium" = "#999999", "High" = "#B2182B")

# Convert CellType_Display to factor with proper levels for each cohort
for (cohort_name in names(xenium_list)) {
  obj <- xenium_list[[cohort_name]]
  obj$CellType_Display <- factor(obj$CellType_Display, levels = all_celltypes_display)
  xenium_list[[cohort_name]] <- obj
}

# Debug: Check for NA values in metadata
cat("\n=== NA Value Check ===\n")
for (cohort_name in names(xenium_list)) {
  obj <- xenium_list[[cohort_name]]
  cat(cohort_name, ":\n")
  cat("  CellType_Clean NAs:", sum(is.na(obj$CellType_Clean)), "\n")
  cat("  stemness_category NAs:", sum(is.na(obj$stemness_category)), "\n")
  cat("  Stemness_UCell NAs:", sum(is.na(obj$Stemness_UCell)), "\n")
}

# Plot spatial stemness and cell type for each cohort and FOV
for (cohort_name in names(xenium_list)) {
  cat("### ", cohort_name, " {.tabset}\n\n", sep = "")

  obj <- xenium_list[[cohort_name]]
  fov_names <- names(obj@images)

  if (length(fov_names) == 0) {
    cat("No FOVs found for this cohort.\n\n")
    next
  }

  for (fov in fov_names) {
    cat("#### ", fov, "\n\n", sep = "")

    tryCatch({
      # Debug: Check data for this specific FOV
      fov_cells <- Cells(obj@images[[fov]])
      cat("FOV:", fov, "- Cells in FOV:", length(fov_cells), "\n")

      # Check metadata for FOV cells
      fov_meta <- obj@meta.data[fov_cells, ]
      cat("  CellType_Clean NAs in FOV:", sum(is.na(fov_meta$CellType_Clean)), "\n")
      cat("  stemness_category NAs in FOV:", sum(is.na(fov_meta$stemness_category)), "\n")

      # Extract dominant organ and treatment for this FOV
      fov_organ <- names(sort(table(fov_meta$organ_std), decreasing = TRUE))[1]
      fov_treatment <- names(sort(table(fov_meta$treatment_std), decreasing = TRUE))[1]
      fov_organ <- tools::toTitleCase(fov_organ)  # Capitalize for display

      # 1. Stemness score plot (continuous)
      p_stemness <- ImageFeaturePlot(
        obj,
        features = "Stemness_UCell",
        fov = fov,
        size = 0.5,
        alpha = 0.9
      ) +
        scale_fill_viridis_c(option = "magma") +
        ggtitle("Stemness Score (continuous)") +
        theme(
          legend.position = "right",
          plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
        )

      # 2. Stemness category plot (Low/Medium/High) - simplified
      p_stemness_cat <- ImageDimPlot(
        obj,
        group.by = "stemness_category",
        fov = fov,
        size = 0.5,
        alpha = 0.9,
        cols = stemness_cat_colors
      ) +
        ggtitle("Stemness Category") +
        theme(
          legend.position = "right",
          plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
        )

      # 3. Cell type plot (categorical) - using merged cell types (Victoria's naming)
      p_celltype <- ImageDimPlot(
        obj,
        group.by = "CellType_Display",
        fov = fov,
        size = 0.5,
        alpha = 0.9,
        cols = celltype_colors
      ) +
        ggtitle("Cell Types (Victoria's Categories)") +
        theme(
          legend.position = "right",
          plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
        )

      # Combined 3-panel plot
      p_combined <- p_stemness + p_stemness_cat + p_celltype +
        plot_layout(ncol = 3) +
        plot_annotation(
          title = paste0(cohort_name, " - ", fov, " | ", fov_organ, " | ", fov_treatment),
          theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
        )
      print(p_combined)

      # Save combined plot
      safe_name <- gsub("[^a-zA-Z0-9_-]", "_", paste(cohort_name, fov, sep = "_"))
      ggsave(
        file.path(output_dir, paste0("spatial_combined_", safe_name, ".png")),
        p_combined, width = 24, height = 10, dpi = 300, bg = "white"
      )

      cat("\n\n")
    }, error = function(e) {
      cat("WARNING: Could not plot", cohort_name, "-", fov, ":", conditionMessage(e), "\n\n")
    })
  }
}
```

### By Cell Type {.tabset}

```{r stemness_by_celltype, fig.width=16, fig.height=12}
# Using combined metadata from all cohorts (all_meta created in score_distribution)
# Determine which cell type column to use (CellType_Display if available, else CellType_Clean)
if ("CellType_Display" %in% colnames(all_meta)) {
  all_meta$CellType_ForPlot <- all_meta$CellType_Display
  cat("Using cell type column: CellType_Display (Victoria's naming)\n")
} else {
  all_meta$CellType_ForPlot <- all_meta$CellType_Clean
  cat("Using cell type column: CellType_Clean (original naming)\n")
}
cat("Unique cell types:", length(unique(all_meta$CellType_ForPlot)), "\n")
print(table(all_meta$CellType_ForPlot))

# Update celltype_colors to match available cell types
available_types <- unique(all_meta$CellType_ForPlot)
celltype_colors_plot <- celltype_colors[names(celltype_colors) %in% available_types]
# Add default colors for any missing types
missing_types <- setdiff(available_types, names(celltype_colors_plot))
if (length(missing_types) > 0) {
  default_colors <- polychrome()[1:length(missing_types)]
  names(default_colors) <- missing_types
  celltype_colors_plot <- c(celltype_colors_plot, default_colors)
}

# Violin plot by cell type (using combined metadata)
p_violin <- ggplot(all_meta, aes(x = CellType_ForPlot, y = Stemness_UCell, fill = CellType_ForPlot)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.5) +
  scale_fill_manual(values = celltype_colors_plot) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "none"
  ) +
  ggtitle("Stemness Score by Cell Type (All Cohorts) - Victoria's Categories") +
  xlab("Cell Type") + ylab("Stemness Score")
print(p_violin)

ggsave(
  file.path(output_dir, "violin_stemness_by_celltype.png"),
  p_violin,
  width = 16, height = 10, dpi = 300, bg = "white"
)

# Violin plot faceted by cohort
p_violin_cohort <- ggplot(all_meta, aes(x = CellType_ForPlot, y = Stemness_UCell, fill = Cohort)) +
  geom_violin(scale = "width", trim = TRUE) +
  facet_wrap(~Cohort, ncol = 1) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "none"
  ) +
  ggtitle("Stemness Score by Cell Type (Per Cohort)") +
  xlab("Cell Type") + ylab("Stemness Score")
print(p_violin_cohort)

ggsave(
  file.path(output_dir, "violin_stemness_by_celltype_cohort.png"),
  p_violin_cohort,
  width = 16, height = 14, dpi = 300, bg = "white"
)

# Cell type distribution barplot
ct_counts <- as.data.frame(table(all_meta$CellType_ForPlot))
colnames(ct_counts) <- c("CellType", "Count")
p_ct_bar <- ggplot(ct_counts, aes(x = reorder(CellType, -Count), y = Count, fill = CellType)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = celltype_colors_plot) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    legend.position = "none"
  ) +
  ggtitle("Cell Type Distribution (All Cohorts)") +
  xlab("Cell Type") + ylab("Number of Cells")
print(p_ct_bar)

ggsave(
  file.path(output_dir, "barplot_celltype_distribution.png"),
  p_ct_bar,
  width = 14, height = 8, dpi = 300, bg = "white"
)
```

## Stemness Analysis Summary {.tabset}

### Stemness Category Distribution

```{r stemness_clusters, fig.width=14, fig.height=10}
# For summary view: use GLOBAL quantiles to compare absolute stemness across cohorts
# (Individual cohort plots use per-cohort quantiles for internal distribution)
summary_breaks <- quantile(all_meta$Stemness_UCell, probs = c(0, 0.25, 0.75, 1), na.rm = TRUE)
cat("Summary (global) breaks:", paste(round(summary_breaks, 3), collapse = ", "), "\n")

all_meta$stemness_category <- cut(
  all_meta$Stemness_UCell,
  breaks = c(-Inf, summary_breaks[2], summary_breaks[3], Inf),
  labels = c("Low", "Medium", "High"),
  include.lowest = TRUE
)

cat("=== Stemness Category Distribution (All Cohorts) ===\n")
print(table(all_meta$stemness_category))

cat("\n=== By Cohort ===\n")
print(table(all_meta$Cohort, all_meta$stemness_category))

# Barplot by category and cohort
p_cat_bar <- ggplot(all_meta, aes(x = stemness_category, fill = Cohort)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  ggtitle("Stemness Category Distribution by Cohort") +
  xlab("Stemness Category") + ylab("Number of Cells")
print(p_cat_bar)

ggsave(
  file.path(output_dir, "barplot_stemness_categories_by_cohort.png"),
  p_cat_bar,
  width = 12, height = 8, dpi = 300, bg = "white"
)

# Percentage stacked bar
p_cat_pct <- ggplot(all_meta, aes(x = Cohort, fill = stemness_category)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Low" = "#2166AC", "Medium" = "#999999", "High" = "#B2182B")) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  ggtitle("Stemness Category Proportions by Cohort") +
  xlab("Cohort") + ylab("Percentage")
print(p_cat_pct)

ggsave(
  file.path(output_dir, "barplot_stemness_proportions_by_cohort.png"),
  p_cat_pct,
  width = 10, height = 8, dpi = 300, bg = "white"
)
```

### Summary Statistics

```{r summary_stats}
cat("=== Xenium 3-Cohort CytoTRACE2 Validation Summary ===\n\n")

# Calculate totals
total_cells <- sum(sapply(xenium_list, ncol))
total_fovs <- sum(sapply(xenium_list, function(x) length(names(x@images))))

cat("1. Dataset\n")
cat("   - Total cells:", total_cells, "\n")
cat("   - Total FOVs:", total_fovs, "\n")
cat("   - Cohorts:\n")
for (cohort in names(xenium_list)) {
  obj <- xenium_list[[cohort]]
  cat("     *", cohort, ":", ncol(obj), "cells,", length(names(obj@images)), "FOVs\n")
}

cat("\n2. CytoTRACE2 Signatures (MALIGNANT-SPECIFIC)\n")
cat("   - Source: cytotrace2_malignant_gene_correlations.csv\n")
cat("   - Top genes (stemness-associated):", length(available_top), "of", nrow(top_genes), "\n")
cat("   - Bottom genes (differentiation-associated):", length(available_bottom), "of", nrow(bottom_genes), "\n")
cat("   - Top genes include: HSPD1, MYC, DNAJA1, ATF4, DUSP6, ...\n\n")

cat("3. Score Interpretation\n")
cat("   - Positive = stem-like (Top signature enriched)\n")
cat("   - Negative = differentiated (Bottom signature enriched)\n\n")

cat("4. Cell Type Annotation\n")
cat("   - Using CellType_Clean (22 consolidated categories)\n")
cat("   - Mapped from original CellType_Harmonized (128 categories)\n\n")

cat("5. Key Observations\n")
cat("   - Check if high score regions cluster at tumor margins\n")
cat("   - Compare with cell type annotations\n")
cat("   - Correlate with immune infiltration patterns\n")

# Export combined CytoTRACE2 scores (all_meta already has all data)
# Include both CellType_Clean (22 categories) and CellType_Display (Victoria's naming)
export_cols <- c("Cohort", "CellType_Clean", "CellType_Display", "CytoTRACE2_Top", "CytoTRACE2_Bottom", "CytoTRACE2_Score", "stemness_category")
export_cols <- intersect(export_cols, colnames(all_meta))
stemness_export <- all_meta[, export_cols]
write.csv(stemness_export, file.path(output_dir, "xenium_cytotrace2_scores_all_cohorts.csv"), row.names = TRUE)
cat("\nExported: xenium_cytotrace2_scores_all_cohorts.csv\n")

# Save each cohort with scores
for (cohort in names(xenium_list)) {
  save_path <- file.path(output_dir, paste0("xenium_", cohort, "_with_scores.qs2"))
  cat("Saving", cohort, "to:", save_path, "\n")
  qs2::qs_save(xenium_list[[cohort]], save_path, nthreads = 8)
}
cat("All objects saved.\n")
```

## Session Info

```{r session_info}
sessionInfo()
```
